### 2015-03-03 doc/notebook/2015-02-26_Data_acquisition.Rmd
Demonstrate how we got bacterial data

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
opts_chunk$set("dev" = "png")
opts_chunk$set(results = "hold")
opts_chunk$set(fig.show = "hold")
opts_chunk$set(warning = FALSE)
opts_chunk$set(fig.align = "center")
opts_chunk$set(cache = FALSE)

opts_chunk$set(eval = FALSE)
opts_chunk$set(echo = FALSE)
```

## Get data out of ARB

> To create SSU Ref (ARB file), all sequences below 1,200 bases for Bacteria and
> Eukarya and below 900 bases for Archaea or an alignment identity below 70 or an
> alignment quality value below 50 have been removed from SSU Parc. All sequences
> with a Pintail value < 50 or an alignment quality value < 75 have been assigned
> to color group 1 in ARB (red). All Living Tree Project or  StrainInfo
> typestrains have been assigned to color group 2 in ARB (light blue). From
> http://www.arb-silva.de/documentation/release-119/

```{r get_data, engine="bash"}
wget http://www.arb-silva.de/fileadmin/silva_databases/release_119/ARB_files/SSURef_119_SILVA_14_07_14_opt.arb.tgz
wget http://www.arb-silva.de/fileadmin/silva_databases/release_119/ARB_files/SSURef_119_SILVA_14_07_14_opt.arb.tgz.md5
tar xvzf SSURef_119_SILVA_14_07_14_opt.arb.tgz
```

Within ARB, we will exclude color group 1, chloroplasts, and mitochondria.

To get good sequences...
* Go to search window
* Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
* Hit Mark Listed, Unmark Rest button (N=1493493)
* The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure
    out which taxonomy to base the analysis on. Do the following searches...
    * ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1324558)
    * ARB_color != 1 & tax_greengenes == "*Bacteria*" (N=1123144)
    * ARB_color != 1 & tax_slv == "*Bacteria*" (N=426917)
    * ARB_color != 1 & tax_embl == "*Bacteria*" (N=1207210)
    * tax_rdp | tax_greengenes | tax_embl | tax_slv (N=1376910)

Let's stick with the bacterial sequences that have an RDP taxonomy and we will
analyze them in RDP space, but preserve the taxonomy generated by the other
systems. Nicely, the RDP taxonomy does not include mitochondrial
16S rRNA gene sequences, so all we'll need to remove are the chloroplast
sequences

* Return to Search and Query window
* Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
* Change the second search field to "tax_rdp" and set it to "*Bacteria*"
* Change the third search field to "tax_rdp" and set it to "*Chloroplast*"
* Set the two Query operators to "and" and hit "Search"
* Hit Mark Listed Unmark Rest (N=1,301,541)
* Click "Write to Fields of Listed", select the "remark" field, and enter "good bacteria"
* Click "Write"
* In main ARB window go File->Export->Export to external format
* Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
* Rename `noname.fasta` to `bacteria.fasta`

Now we need the taxonomy information.
* Go Tree -> NDS
* Click "name", "acc", "tax_rdp". The "tax_rdp" field should have 250 characters
* Unclick everything else
* Click "Close"
* Go File->Export->Export fields
* Set the file name to "bacteria.taxonomy" and Column output to "TAB separated"
* Click "SAVE"

Finally, let's save the database by doing File -> Quick Save Changes and then
quit out of ARB.

## Format taxonomy file

The next thing we need to do is to clean up the bacteria.taxonomy.nds file
to make it into a proper, mothur compatible bacteria.taxonomy file. First we
want to remove the names of any sub taxa (e.g. suborder). To do this we need
to get the RDP list of names and the taxonomic level they belong to...

```{r get_rdp_names, engine="bash"}
rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata* Trainset9_032012.pds.zip

wget -N http://sourceforge.net/projects/rdp-classifier/files/RDP_Classifier_TrainingData/RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
unzip -o RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
mv RDPClassifier_16S_trainsetNo10_rawtrainingdata/trainset10_db_taxid.txt ./
rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata*
```

Now we're read to do some R'ing to get the taxonomy file formatted properly...

```{r format_taxonomy}
tax_data <- read.table(file="bacteria.taxonomy.nds", sep="\t")

#get the names to match the fasta file
seq_names <- paste(tax_data$V2, tax_data$V1, sep=".")

taxonomy <- gsub(" ", "_", tax_data$V3)  #convert any spaces to underscores
taxonomy <- gsub("[^;]*_incertae_sedis$", "", taxonomy) #remove the unknowns
taxonomy <- gsub('\"', '', taxonomy) #remove quote marks

#now we need to pull out the taxon names for the sub taxa levels
levels <- read.table(file="trainset10_db_taxid.txt", sep="*", stringsAsFactors=F)
subs <- levels[grep("sub", levels$V5),] #get the sub taxa names
sub.names <- subs$V2

tax.split <- strsplit(taxonomy, split=";")  #split the taxonomy string by semicolon

#function to see which taxa names in a list are found in the sub taxa list
remove.subs <- function(tax.vector){
,	#return(tax.vector[which(!tax.vector %in% sub.names)])
}

#apply sub taxa finding function to all sequences
no.subs <- lapply(tax.split, remove.subs)

#merge data together by semicolon for each sequence
no.subs.str <- unlist(lapply(no.subs, paste, collapse=";"))

#Tack a semicolon on to the end of each sequence
no.subs.str <- paste0(no.subs.str, ";")

write.table(cbind(seq_names, no.subs.str), "bacteria.tax", row.names=F,
                                                col.names=F, quote=F, sep="\t")
```

## Get good sequences

Now we need to know the start/end position of the sequences so that we can
make sure the reads overlap the same alignment space.

```{r get_summary, engine="bash", results="hide", echo=TRUE}
mothur "#summary.seqs(fasta=bacteria.fasta, processors=12)"
```

                Start   End     NBases  Ambigs  Polymer NumSeqs
Minimum:        1       9693    900     0       4       1
2.5%-tile:      1004    12895   1279    0       5       32539
25%-tile:       1026    13069   1355    0       5       325386
Median:         1044    13267   1382    0       5       650771
75%-tile:       1051    13825   1447    0       6       976156
97.5%-tile:     1793    13991   1522    2       7       1269003
Maximum:        6448    15767   3281    51      38      1301541
Mean:   1096.18 13442   1399.39 0.176343        5.49738
# of Seqs:      1301541

Notes...
* Looks like an archaeal sequence slipped through
* Will want to get rid of sequences with large number of Ns in them - how large?

```{r find_screening_points}
data <- read.table(file="bacteria.summary", header=T, row.names=1)
quantile(data$start, probs=seq(0.9,1,0.01))
start <- 2000

quantile(data$end, probs=seq(0,0.1,0.01))
end <- 13000

quantile(data$nbases, probs=seq(0.0,0.1,0.01))
min_length <- 1200

quantile(data$ambigs, probs=seq(0.9,1.0,0.01))
max_ambig <- 2

trimmed <- data[data$start <= start & data$end >= end &
                    data$nbases > min_length & data$ambigs <= max_ambig,]
nrow(trimmed)
summary(trimmed$nbases)
```

Now let's run these parameters using mothur. We'll also filter the sequences to
overlap in the same alignment space, unique them, and precluster them...

```{r screen_data, engine="bash"}
mothur "#set.current(processors=8);
        screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);
        filter.seqs(vertical=T, trump=.);
        unique.seqs();
        pre.cluster(fasta=current, name=current, diffs=10);"

#need to make the taxonomy file match the fasta and names files...
cut -f 1 *ster.names > precluster.accnos
mothur "#get.seqs(taxonomy=bacteria.good.tax, accnos=precluster.accnos)"

```

Now we're ready to cluster the sequences. We'll do it by the classic approach
without any cutoffs and see what we get. Let's start by splitting things at the
phylum level and cluster from there...

```{r cluster_data, engine="bash"}
mothur "#cluster.split(fasta=bacteria.good.filter.unique.precluster.fasta, name=bacteria.good.filter.unique.precluster.names, taxonomy=bacteria.good.pick.tax, taxlevel=2, classic=T, processors=8)"
```

## Get metadata

We'd like to use some metadata from the database to characterize the changes in
the representation of sequences over time, by environment, methods, etc. The
fields housed within SILVA are available at their [website](http://www.arb-silva.de/fileadmin/arb_web_db/release_115/Fields_description/SILVA_description_of_fields_16_06_2013.htm)
and there is an [FAQ](http://www.arb-silva.de/documentation/faqs/) on their site
as well. There were a number of fields that I didn't think were relevant and
instead focused on 30 fields that I thought could help the cause. I marked these
fields in the NDS feature and extracted them to `metadata.nds`. We need to
tweak this file slightly in R to get it into a format that we can use.

```{r}
metadata_fields <- c(
    "name", #internal ARB database ID, do not change!
    "acc", #accession number
    "bio_material", #identifier for the biological material from which the nucleic acid sequenced was obtained
    "clone", #clone from which the sequence was obtained
    "clone_lib", #clone library from which the sequence was obtained
    "collected_by", #name of the person who collected the specimen
    "collection_date", #date that the sample/specimen was collected
    "country", #geographical origin of sequenced sample
    "culture_collection", #institution code and identifier for the culture from which the nucleic acid sequenced was obtained, with optional collection code
    "date", #entry creation and update date separated by ;
    "description", #description
    "embl_class", #describes the data class in EMBL, e.g. CON: Constructed, WGS: Whole Genome Shotgun, see www.ebi.ac.uk/ena/about/embl_bank_format
    "env_sample", #identifies sequences derived by direct molecular isolation from a bulk environmental DNA sample (by PCR with or without subsequent cloning of the product, DGGE, or other anonymous methods) with no reliable identification of the source organism. Indicated by ‘yes’ in the ARB files
    "full_name", #organism species
    "host", #natural (as opposed to laboratory) host to the organism from which sequenced molecule was obtained.
    "identified_by", #name of the taxonomist who identified the specimen
    "insdc", #the International Nucleotide Sequence Database Collaboration (INSDC) Project Identifier that has been assigned to the entry
    "isolate", #individual isolate from which the sequence was obtained
    "isolation_source", #describes the physical, environmental and/or local geographical source of the biological sample from which the sequence was derived
    "journal", #reference location
    "lat_lon", #geographical coordinates of the location where the specimen was collected
    "publication_doi", #cross-reference DOI number
    "pubmed_id", #cross-reference Pubmed ID
    "strain", #strain from which the sequence was obtained. (t) or [T]: typestrains, [C]: cultivated, [G]: genomes
    "sub_species", #name of sub-species of organism from which sequence was obtained
    "submit_author", #submission authors from reference location
    "submit_date", #submission date from reference location
    "title", #reference title
    "depth_slv", #depth is the vertical distance below surface, e.g. for sediment or soil samples depth is measured from sediment or soil surface, respectively. Depth can be reported as an interval for subsurface samples.
    "habitat_slv" #habitat description according to EnvO-Lite
)

metadata <- read.table(file="metadata.nds", sep="\t", stringsAsFactors=FALSE,
                            col.names=metadata_fields, fill=TRUE, na.strings="",
                            quote="")
rownames(metadata) <- paste(metadata$name, metadata$acc, sep=".")
metadata <- metadata[,-c(1,2)]
```

Now we want to read in `bacteria.bad.accnos` so that we can figure out which
sequences to cull from the metadata table.

```{r}
bad_accnos_data <- read.table(file="bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
bad_accnos <- bad_accnos_data$name.accnos
metadata_good <- metadata[-c(bad_accnos),]
write.table(metadata_good, file="bacteria.good.metadata", quote=F, sep="\t")
```

## Finally...

Let's compress everything that is relevant for the rest of the project...

```{r, engine="bash"}
tar cvzf bacteria.good.tgz bacteria.good.metadata bacteria.good.filter.unique.precluster.an.list bacteria.good.tax
```

### 2015-03-13 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Rename files

--- 2015-03-03 doc/notebook/2015-02-26_Data_acquisition.Rmd
@@
 
+```
                 Start   End     NBases  Ambigs  Polymer NumSeqs
@@
 # of Seqs:      1301541
+```
 
@@
 
-Now let's run these parameters using mothur. We'll also filter the sequences to
-overlap in the same alignment space, unique them, and precluster them...
+Now let's run these parameters using mothur.
 
@@
 mothur "#set.current(processors=8);
-        screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);
-        filter.seqs(vertical=T, trump=.);
+        set.dir(input=data/raw, output=data/process)
+        screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);"
+```
+
+## Single cell genomics data
+
+Using 16S rRNA gene sequences from the single cell genomics projects we'd like
+to see whether that type of data has had a meaningful impact on the trajectory
+of the microbial census. The LSU team scraped these 251 sequences from a
+database and have provided them to me as a fasta file. Some of them are quite
+short and none of them are aligned. Let's go ahead and align them to the SILVA
+alignment and then screen them to keep sequences that overlap with the SILVA
+sequences.
+
+```{r single_cell, engine="bash"}
+mothur "#align.seqs(fasta=single_cell.bacteria.fasta, reference=bacteria.good.fasta);
+        screen.seqs(fasta=current, start=2000, end=13000, maxambig=2, minlength=1200);
+        classify.seqs(fasta=current, reference=trainset10_082014.pds.fasta, taxonomy=trainset10_082014.pds.tax, cutoff=80, processors=8)"
+```
+
+## Pooling the data
+
+Now we want to merge the SILVA and single cell genomics 16S rRNA gene sequence
+collections. We'll also filter the sequences to overlap in the same alignment
+space, unique them, and precluster them...
+
+```{r pool_data, engine="bash"}
+cat bacteria.good.fasta single_cell.bacteria.good.align > all_bacteria.align
+cat bacteria.good.tax single_cell.bacteria.good.pds.wang.taxonomy > all_bacteria.taxonomy
+
+mothur "#filter.seqs(fasta=all_bacteria.align, vertical=T, trump=.);
         unique.seqs();
@@
 cut -f 1 *ster.names > precluster.accnos
-mothur "#get.seqs(taxonomy=bacteria.good.tax, accnos=precluster.accnos)"
-
+mothur "#get.seqs(taxonomy=all_bacteria.taxonomy, accnos=precluster.accnos)"
 ```
@@
 ```{r cluster_data, engine="bash"}
-mothur "#cluster.split(fasta=bacteria.good.filter.unique.precluster.fasta, name=bacteria.good.filter.unique.precluster.names, taxonomy=bacteria.good.pick.tax, taxlevel=2, classic=T, processors=8)"
+mothur "#cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
 ```
@@
 
-metadata <- read.table(file="metadata.nds", sep="\t", stringsAsFactors=FALSE,
+metadata <- read.table(file="data/raw/metadata.nds", sep="\t", stringsAsFactors=FALSE,
                             col.names=metadata_fields, fill=TRUE, na.strings="",
@@
 ```{r}
-bad_accnos_data <- read.table(file="bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos_data <- read.table(file="data/process/bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
 metadata_good <- metadata[-c(bad_accnos),]
-write.table(metadata_good, file="bacteria.good.metadata", quote=F, sep="\t")
+write.table(metadata_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t")
 ```
 
-## Finally...
+We'd also like to get the metadata from the single cell data. We have an `xlsx` that the LSU group pulled together that we'll read in and concatenate to the end of `bacteria.good.metadata`.
 
-Let's compress everything that is relevant for the rest of the project...
+```{r}
+if(!"openxlsx" %in% rownames(installed.packages())){
+    install.packages("openxlsx")
+}
+library("openxlsx")
 
-```{r, engine="bash"}
-tar cvzf bacteria.good.tgz bacteria.good.metadata bacteria.good.filter.unique.precluster.an.list bacteria.good.tax
+spread_sheet <- read.xlsx("data/raw/BacterialSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
+bad_accnos_data <- read.table(file="data/process/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos <- bad_accnos_data$name.accnos
+spread_sheet_good <- spread_sheet[-c(bad_accnos),]
+
+write.table(spread_sheet_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
 ```

### 2015-03-13 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Add main heading
Make precluster file name specific to bacteria

@@
 ```
+
+# Getting the bacterial sequence data and metadata
 
@@
 #need to make the taxonomy file match the fasta and names files...
-cut -f 1 *ster.names > precluster.accnos
+cut -f 1 all_bacteria.filter.unique.precluster.names > precluster.accnos
 mothur "#get.seqs(taxonomy=all_bacteria.taxonomy, accnos=precluster.accnos)"

### 2015-03-13 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Adapt for archaea

--- 2015-03-13 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
@@
 
+# Getting the archaeal sequence data and metadata
+
 ## Get data out of ARB
+
+This is redundant with what we saw for the bacterial data, but for completion...
 
@@
     out which taxonomy to base the analysis on. Do the following searches...
-    * ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1324558)
-    * ARB_color != 1 & tax_greengenes == "*Bacteria*" (N=1123144)
-    * ARB_color != 1 & tax_slv == "*Bacteria*" (N=426917)
-    * ARB_color != 1 & tax_embl == "*Bacteria*" (N=1207210)
-    * tax_rdp | tax_greengenes | tax_embl | tax_slv (N=1376910)
+    * ARB_color != 1 & tax_rdp == "*Archaea*" (N=40288)
+    * ARB_color != 1 & tax_greengenes == "*Archaea*" (N=19795)
+    * ARB_color != 1 & tax_slv == "*Archaea*" (N=17641)
+    * ARB_color != 1 & tax_embl == "*Archaea*" (N=44574)
+    * ARB_color != 1 & (tax_rdp | tax_greengenes | tax_embl | tax_slv) (N=46387 )
 
-Let's stick with the bacterial sequences that have an RDP taxonomy and we will
-analyze them in RDP space, but preserve the taxonomy generated by the other
-systems. Nicely, the RDP taxonomy does not include mitochondrial
-16S rRNA gene sequences, so all we'll need to remove are the chloroplast
-sequences
+Let's stick with the archaeal sequences that have an RDP taxonomy and we will
+analyze them in RDP space, but we can always go back to the taxonomy provided by
+the other systems if we'd like. A quick spot checking suggests that the RDP
+taxonomy information is actually richer than that of the EBML taxonomy
+information.
 
@@
 * Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
-* Change the second search field to "tax_rdp" and set it to "*Bacteria*"
-* Change the third search field to "tax_rdp" and set it to "*Chloroplast*"
-* Set the two Query operators to "and" and hit "Search"
-* Hit Mark Listed Unmark Rest (N=1,301,541)
-* Click "Write to Fields of Listed", select the "remark" field, and enter "good bacteria"
+* Change the second search field to "tax_rdp" and set it to "*Archaea*"
+* Hit "Search"
+* Hit Mark Listed Unmark Rest (N=40,288)
+* Click "Write to Fields of Listed", select the "remark" field, and enter "good archaea"
 * Click "Write"
@@
 * Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
-* Rename `noname.fasta` to `bacteria.fasta`
+* Rename `noname.fasta` to `archaea.fasta`
 
@@
 * Go File->Export->Export fields
-* Set the file name to "bacteria.taxonomy" and Column output to "TAB separated"
+* Set the file name to "archaea.taxonomy" and Column output to "TAB separated"
 * Click "SAVE"
@@
 
-The next thing we need to do is to clean up the bacteria.taxonomy.nds file
-to make it into a proper, mothur compatible bacteria.taxonomy file. First we
+The next thing we need to do is to clean up the archaea.taxonomy.nds file
+to make it into a proper, mothur compatible archaea.taxonomy file. First we
 want to remove the names of any sub taxa (e.g. suborder). To do this we need
@@
 ```{r format_taxonomy}
-tax_data <- read.table(file="bacteria.taxonomy.nds", sep="\t")
+tax_data <- read.table(file="archaea.taxonomy.nds", sep="\t")
 
@@
 remove.subs <- function(tax.vector){
-,	#return(tax.vector[which(!tax.vector %in% sub.names)])
+    return(tax.vector[which(!tax.vector %in% sub.names)])
 }
@@
 
-write.table(cbind(seq_names, no.subs.str), "bacteria.tax", row.names=F,
+write.table(cbind(seq_names, no.subs.str), "archaea.tax", row.names=F,
                                                 col.names=F, quote=F, sep="\t")
@@
 ```{r get_summary, engine="bash", results="hide", echo=TRUE}
-mothur "#summary.seqs(fasta=bacteria.fasta, processors=12)"
+mothur "#summary.seqs(fasta=archaea.fasta, processors=12)"
 ```
@@
             Start    End	NBases	Ambigs	Polymer	NumSeqs
-Minimum:        1       9693    900     0       4       1
-2.5%-tile:      1004    12895   1279    0       5       32539
-25%-tile:       1026    13069   1355    0       5       325386
-Median:         1044    13267   1382    0       5       650771
-75%-tile:       1051    13825   1447    0       6       976156
-97.5%-tile:     1793    13991   1522    2       7       1269003
-Maximum:        6448    15767   3281    51      38      1301541
-Mean:   1096.18 13442   1399.39 0.176343        5.49738
-# of Seqs:      1301541
+Minimum:	1        4058	900	    0    	4	1
+2.5%-tile:	357      4692	905	    0    	5	1008
+25%-tile:	365      4783	935	    0   	5	10073
+Median: 	379      7409	1262	0   	5	20145
+75%-tile:	456      7729	1349	0    	6	30217
+97.5%-tile:	1746	 7984	1470	4   	7	39281
+Maximum:	2321	 8711	2174	46     	32	40288
+Mean:	    513.659	 6698.2	1185.3	0.3	    5.44244
+# of Seqs:	40288
 ```
@@
 Notes...
-* Looks like an archaeal sequence slipped through
-* Will want to get rid of sequences with large number of Ns in them - how large?
+* Will want to get rid of sequences with large number of Ns in them and the
+ridiculously long homopolymrs - how large?
 
 ```{r find_screening_points}
-data <- read.table(file="bacteria.summary", header=T, row.names=1)
+data <- read.table(file="archaea.summary", header=T, row.names=1)
 quantile(data$start, probs=seq(0.9,1,0.01))
@@
 quantile(data$end, probs=seq(0,0.1,0.01))
-end <- 13000
+end <- 4058
 
 quantile(data$nbases, probs=seq(0.0,0.1,0.01))
-min_length <- 1200
+min_length <- 900
 
@@
 
+quantile(data$polymer, probs=seq(0.9,1.0,0.01))
+max_polymer <- 8
+
 trimmed <- data[data$start <= start & data$end >= end &
-                    data$nbases > min_length & data$ambigs <= max_ambig,]
+                    data$polymer <= max_polymer & data$nbases > min_length &
+                    data$ambigs <= max_ambig,]
 nrow(trimmed)
@@
 mothur "#set.current(processors=8);
-        set.dir(input=data/raw, output=data/process)
-        screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);"
+        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);"
 ```
@@
 ```{r single_cell, engine="bash"}
-mothur "#align.seqs(fasta=single_cell.bacteria.fasta, reference=bacteria.good.fasta);
-        screen.seqs(fasta=current, start=2000, end=13000, maxambig=2, minlength=1200);
+mothur "#align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
+        screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
         classify.seqs(fasta=current, reference=trainset10_082014.pds.fasta, taxonomy=trainset10_082014.pds.tax, cutoff=80, processors=8)"
@@
 ```{r pool_data, engine="bash"}
-cat bacteria.good.fasta single_cell.bacteria.good.align > all_bacteria.align
-cat bacteria.good.tax single_cell.bacteria.good.pds.wang.taxonomy > all_bacteria.taxonomy
+cat archaea.good.fasta single_cell.archaea.good.align > all_archaea.align
+cat archaea.good.tax single_cell.archaea.good.pds.wang.taxonomy > all_archaea.taxonomy
 
-mothur "#filter.seqs(fasta=all_bacteria.align, vertical=T, trump=.);
+mothur "#filter.seqs(fasta=all_archaea.align, vertical=T, trump=.);
         unique.seqs();
-        pre.cluster(fasta=current, name=current, diffs=10);"
+        pre.cluster(fasta=current, name=current, diffs=9);"
 
 #need to make the taxonomy file match the fasta and names files...
-cut -f 1 *ster.names > precluster.accnos
-mothur "#get.seqs(taxonomy=all_bacteria.taxonomy, accnos=precluster.accnos)"
+cut -f 1 all_archaea.filter.unique.precluster.names > precluster.accnos
+mothur "#get.seqs(taxonomy=all_archaea.taxonomy, accnos=precluster.accnos)"
 ```
@@
 ```{r cluster_data, engine="bash"}
-mothur "#cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
+mothur "#cluster.split(fasta=all_archaea.filter.unique.precluster.fasta, name=all_archaea.filter.unique.precluster.names, taxonomy=all_archaea.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
 ```
@@
 instead focused on 30 fields that I thought could help the cause. I marked these
-fields in the NDS feature and extracted them to `metadata.nds`. We need to
+fields in the NDS feature and extracted them to `archaea_metadata.nds`. We need to
 tweak this file slightly in R to get it into a format that we can use.
@@
 
-Now we want to read in `bacteria.bad.accnos` so that we can figure out which
+Now we want to read in `archaea.bad.accnos` so that we can figure out which
 sequences to cull from the metadata table.
@@
 ```{r}
-bad_accnos_data <- read.table(file="data/process/bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos_data <- read.table(file="data/process/archaea.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
 metadata_good <- metadata[-c(bad_accnos),]
-write.table(metadata_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t")
+write.table(metadata_good, file="data/process/all_archaea.good.metadata", quote=F, sep="\t")
 ```
 
-We'd also like to get the metadata from the single cell data. We have an `xlsx` that the LSU group pulled together that we'll read in and concatenate to the end of `bacteria.good.metadata`.
+We'd also like to get the metadata from the single cell data. We have an `xlsx` that the LSU group pulled together that we'll read in and concatenate to the end of `archaea.good.metadata`.
 
@@
 
-spread_sheet <- read.xlsx("data/raw/BacterialSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
-bad_accnos_data <- read.table(file="data/process/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+spread_sheet <- read.xlsx("data/raw/ArchaealSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
+bad_accnos_data <- read.table(file="data/process/single_cell.archaea.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
@@
 
-write.table(spread_sheet_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+write.table(spread_sheet_good, file="data/process/all_archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
 ```

### 2015-03-13 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Remove folder paths for time being

@@
 
-metadata <- read.table(file="data/raw/metadata.nds", sep="\t", stringsAsFactors=FALSE,
+metadata <- read.table(file="archaea_metadata.nds", sep="\t", stringsAsFactors=FALSE,
                             col.names=metadata_fields, fill=TRUE, na.strings="",
@@
 ```{r}
-bad_accnos_data <- read.table(file="data/process/archaea.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos_data <- read.table(file="archaea.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
 metadata_good <- metadata[-c(bad_accnos),]
-write.table(metadata_good, file="data/process/all_archaea.good.metadata", quote=F, sep="\t")
+write.table(metadata_good, file="all_archaea.good.metadata", quote=F, sep="\t")
 ```

### 2015-04-22 scratch/category_parsing.R
Initial commit

@@ -1,0 +1,81 @@
+# I copied the first column of the spreadsheet into a file called `source_categories.txt`
+# First, I went through and counted the number of times each word was used and ranked
+# them in decreasing order of abundance
+
+original_words <- tolower(scan(file="source_categories.txt", quiet=T, what=""))
+original_words_freq <- table(original_words)
+head(sort(original_words_freq, decreasing=T), n=100)
+
+# From this we can see a lot of useless words like of, from, etc. But also there are
+# categories that have useful words like soil, water, sediment, etc. Let's read the
+# categories back in but doing it one line at a time this time instead of one word at
+# a time:
+
+categories <- tolower(scan(file="source_categories.txt", quiet=T, what="", sep="\n"))
+
+# let's look at some of the categories that contain words like "soil"
+
+categories[grepl("soil", categories)]
+
+# From this we see categories like: "0-20 cm bulk soil, duennwald forest". What I would
+# propose is using the grepl R function to quickly sift through these categories that 
+# contain high frequency words.
+#
+# go through and reassign any categories that contain high frequency words to 
+# the categories categories names. here are a few that I used to get going...
+
+categories[grepl("soil", categories, ignore.case=T)] <- "soil"
+categories[grepl("rhizosphere", categories, ignore.case=T)] <- "soil"
+
+categories[grepl("sediment", categories, ignore.case=T)] <- "sediment"
+
+categories[grepl("wastewater", categories, ignore.case=T)] <- "wastewater"
+categories[grepl("sludge", categories, ignore.case=T)] <- "wastewater"
+categories[grepl("sludge", categories, ignore.case=T)] <- "wastewater"
+
+categories[grepl("sea", categories, ignore.case=T)] <- "marine"
+categories[grepl("marine", categories, ignore.case=T)] <- "marine"
+
+categories[grepl("culture", categories, ignore.case=T)] <- "laboratory"
+categories[grepl("reactor", categories, ignore.case=T)] <- "laboratory"
+categories[grepl("anaerobic", categories, ignore.case=T)] <- "laboratory"
+
+categories[grepl("lake", categories, ignore.case=T)] <- "freshwater"
+categories[grepl("river", categories, ignore.case=T)] <- "freshwater"
+
+categories[grepl("feces", categories, ignore.case=T)] <- "host-associated"
+categories[grepl("patient", categories, ignore.case=T)] <- "host-associated"
[…]

### 2015-04-22 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Update code to put mothur files into data/mothur/

@@
 mothur "#set.current(processors=8);
-        set.dir(input=data/raw, output=data/process)
+        set.dir(input=data/raw, output=data/mothur)
         screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);"
@@
 ```{r}
-bad_accnos_data <- read.table(file="data/process/bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos_data <- read.table(file="data/mothur/bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
 metadata_good <- metadata[-c(bad_accnos),]
-write.table(metadata_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t")
+write.table(metadata_good, file="data/mothur/all_bacteria.good.metadata", quote=F, sep="\t")
 ```
@@
 spread_sheet <- read.xlsx("data/raw/BacterialSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
-bad_accnos_data <- read.table(file="data/process/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos_data <- read.table(file="data/mothur/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
@@
 
-write.table(spread_sheet_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+write.table(spread_sheet_good, file="data/mothur/all_bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
 ```

### 2015-04-22 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Update code to put mothur files into data/mothur/

@@
 spread_sheet <- read.xlsx("data/raw/ArchaealSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
-bad_accnos_data <- read.table(file="data/process/single_cell.archaea.bad.accnos", col.names=c("name.accnos", "reason"))
+bad_accnos_data <- read.table(file="data/mothur/single_cell.archaea.bad.accnos", col.names=c("name.accnos", "reason"))
 bad_accnos <- bad_accnos_data$name.accnos
@@
 
-write.table(spread_sheet_good, file="data/process/all_archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+write.table(spread_sheet_good, file="data/mothur/all_archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
 ```

### 2015-05-20 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Finalize through generating final metadata file

@@
 tar xvzf SSURef_119_SILVA_14_07_14_opt.arb.tgz
+mv SSURef_119_SILVA_14_07_14_opt.arb SSURef_119_SILVA_14_07_14_opt.arb.tgz.md5 data/references
+rm SSURef_119_SILVA_14_07_14_opt.arb.tgz
 ```
@@
 * Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
-* Rename `noname.fasta` to `bacteria.fasta`
+* Rename `noname.fasta` to `data/mothur/bacteria.fasta`
 
@@
 * Go File->Export->Export fields
-* Set the file name to "bacteria.taxonomy" and Column output to "TAB separated"
+* Set the file name to "data/mothur/bacteria.taxonomy" and Column output to "TAB separated"
 * Click "SAVE"
@@
 ```{r get_rdp_names, engine="bash"}
-rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata* Trainset9_032012.pds.zip
-
 wget -N http://sourceforge.net/projects/rdp-classifier/files/RDP_Classifier_TrainingData/RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
 unzip -o RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
-mv RDPClassifier_16S_trainsetNo10_rawtrainingdata/trainset10_db_taxid.txt ./
+mv RDPClassifier_16S_trainsetNo10_rawtrainingdata/* data/references/
 rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata*
@@
 ```{r format_taxonomy}
-tax_data <- read.table(file="bacteria.taxonomy.nds", sep="\t")
+tax_data <- read.table(file="data/mothur/bacteria.taxonomy.nds", sep="\t")
 
@@
 #now we need to pull out the taxon names for the sub taxa levels
-levels <- read.table(file="trainset10_db_taxid.txt", sep="*", stringsAsFactors=F)
+levels <- read.table(file="data/references/trainset10_db_taxid.txt", sep="*", stringsAsFactors=F)
 subs <- levels[grep("sub", levels$V5),] #get the sub taxa names
@@
 
-write.table(cbind(seq_names, no.subs.str), "bacteria.tax", row.names=F,
+write.table(cbind(seq_names, no.subs.str), "data/mothur/bacteria.tax", row.names=F,
                                                 col.names=F, quote=F, sep="\t")
@@
 ```{r get_summary, engine="bash", results="hide", echo=TRUE}
-mothur "#summary.seqs(fasta=bacteria.fasta, processors=12)"
+mothur "#summary.seqs(fasta=data/mothur/bacteria.fasta, processors=12, outputdir=data/mothur/)"
 ```
@@
 ```{r screen_data, engine="bash"}
-mothur "#set.current(processors=8);
-        set.dir(input=data/raw, output=data/mothur)
-        screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);"
+mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200, outputdir=data/mothur/, processors=8);"
 ```
@@
 ```{r single_cell, engine="bash"}
-mothur "#align.seqs(fasta=single_cell.bacteria.fasta, reference=bacteria.good.fasta);
+wget -N -P data/references/ http://www.mothur.org/w/images/2/24/Trainset10_082014.pds.tgz
+tar xvzf data/references/Trainset10_082014.pds.tgz -C data/references/
+mv data/references/trainset10_082014.pds/trainset10_082014.* data/references/
+rm -rf data/references/trainset10_082014.pds
+
+mothur "#set.dir(output=data/mothur);
+        align.seqs(fasta=data/raw/single_cell.bacteria.fasta, reference=data/mothur/bacteria.good.fasta);
+        set.dir(input=data/mothur);
         screen.seqs(fasta=current, start=2000, end=13000, maxambig=2, minlength=1200);
-        classify.seqs(fasta=current, reference=trainset10_082014.pds.fasta, taxonomy=trainset10_082014.pds.tax, cutoff=80, processors=8)"
+        classify.seqs(fasta=current, reference=data/references/trainset10_082014.pds.fasta, taxonomy=data/references/trainset10_082014.pds.tax, cutoff=80, processors=1)"
 ```
@@
 ```{r pool_data, engine="bash"}
-cat bacteria.good.fasta single_cell.bacteria.good.align > all_bacteria.align
-cat bacteria.good.tax single_cell.bacteria.good.pds.wang.taxonomy > all_bacteria.taxonomy
+cat data/mothur/bacteria.good.fasta data/mothur/single_cell.bacteria.good.align > data/mothur/all_bacteria.align
+cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.good.pds.wang.taxonomy > data/mothur/all_bacteria.taxonomy
 
-mothur "#filter.seqs(fasta=all_bacteria.align, vertical=T, trump=.);
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        filter.seqs(fasta=all_bacteria.align, vertical=T, trump=., processors=8);
         unique.seqs();
@@
 cut -f 1 all_bacteria.filter.unique.precluster.names > precluster.accnos
-mothur "#get.seqs(taxonomy=all_bacteria.taxonomy, accnos=precluster.accnos)"
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        get.seqs(taxonomy=all_bacteria.taxonomy, accnos=precluster.accnos)"
 ```
@@
 ```{r cluster_data, engine="bash"}
-mothur "#cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
+mothur "#set.dir(input=data/mothur, output=data/mothur, seed=1);
+        cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
 ```
@@
 instead focused on 30 fields that I thought could help the cause. I marked these
-fields in the NDS feature and extracted them to `metadata.nds`. We need to
-tweak this file slightly in R to get it into a format that we can use.
+fields in the NDS feature and extracted them to
+`data/mothur/bacteria.metadata.nds`. We need to tweak this file slightly in R to
+get it into a format that we can use.
 
@@
 
-metadata <- read.table(file="data/raw/metadata.nds", sep="\t", stringsAsFactors=FALSE,
+metadata <- read.table(file="data/mothur/bacteria.metadata.nds", sep="\t", stringsAsFactors=FALSE,
                         col.names=metadata_fields, fill=TRUE, na.strings="",
                         quote="")
-rownames(metadata) <- paste(metadata$name, metadata$acc, sep=".")
+rownames(metadata) <- paste(metadata$acc, metadata$name, sep=".")
 metadata <- metadata[,-c(1,2)]
@@
 ```{r}
-bad_accnos_data <- read.table(file="data/mothur/bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
+if(!"openxlsx" %in% rownames(installed.packages())){
+    install.packages("openxlsx")
+}
+library("openxlsx")
+
+bad_accnos_data <- read.table(file="data/mothur/bacteria.bad.accnos",
+                    col.names=c("name.accnos", "reason"), stringsAsFactors=F)
 bad_accnos <- bad_accnos_data$name.accnos
-metadata_good <- metadata[-c(bad_accnos),]
-write.table(metadata_good, file="data/mothur/all_bacteria.good.metadata", quote=F, sep="\t")
+metadata_good <- metadata[-which(rownames(metadata) %in% bad_accnos),]
+
+categories_sheet <- read.xlsx("data/raw/FinalCategories.xlsx", sheet=1, startRow=1, colNames=TRUE)
+categories <- categories_sheet[,1]
+names(categories) <- categories_sheet[,2]
+
+categories_sheet_missing <- read.xlsx("data/raw/ExtraCategoriesBacteria.xlsx", sheet=1, startRow=1, colNames=FALSE)
+categories_missing <- categories_sheet_missing[,"X1"]
+names(categories_missing) <- categories_sheet_missing[,"X2"]
+
+categories <- c(categories, categories_missing)
+
+metadata_good$category <- toupper(categories[metadata_good$isolation_source])
+
+extra <- table(metadata_good$isolation_source[is.na(metadata_good$category)])
+
+if(length(extra) == 0){
+    write.table(extra, file="missing.txt", row.names=F, col.names=F, sep="\t")
+}
+
+write.table(metadata_good, file="data/mothur/bacteria.good.metadata", quote=F, sep="\t")
 ```
@@
 ```{r}
-if(!"openxlsx" %in% rownames(installed.packages())){
-    install.packages("openxlsx")
-}
-library("openxlsx")
+spread_sheet <- read.xlsx("data/raw/FormattedBacterialMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
+rownames(spread_sheet) <- spread_sheet$name
+spread_sheet <- spread_sheet[,-c(1,2)]
 
-spread_sheet <- read.xlsx("data/raw/BacterialSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
-bad_accnos_data <- read.table(file="data/mothur/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
-bad_accnos <- bad_accnos_data$name.accnos
-spread_sheet_good <- spread_sheet[-c(bad_accnos),]
+#need actual category from LSU group
+spread_sheet$category <- NA
 
-write.table(spread_sheet_good, file="data/mothur/all_bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+metadata_genome_id <- gsub(".* (P?SCGC \\S*)", "\\1", spread_sheet$full_name)
+metadata_genome_id["SCGC AAA003-L8"] <- "SCGC AAA003-L08"
+
+#need to fix mapping between genome and gene sequences
+fasta <- scan("data/mothur/single_cell.bacteria.fasta", what="", quiet=TRUE, sep="\n")
+headers <- fasta[grepl(">", fasta)]
+seq_numbers <- gsub(">(\\d*) .*", "\\1", headers)
+names(seq_numbers) <- gsub(".* (P?SCGC \\S*) .*", "\\1", headers)
+
+#recreate a metadata table using sequence names and spread_sheet
+full_spread_sheet <- spread_sheet[names(seq_numbers) %in% metadata_genome_id,]
+rownames(full_spread_sheet) <- seq_numbers
+
+#these sequences are missing from metadata file
+#metadata_genome_id[! metadata_genome_id %in% names(seq_numbers)]
+#probably not a horrible thing since not all genomes could give a 16S rRNA gene
+#sequence
+
+bad_accnos_data <- read.table(file="data/mothur/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"),  stringsAsFactors=FALSE)
+bad_accnos <- as.character(bad_accnos_data$name.accnos)
+
+spread_sheet_good <- spread_sheet[-which(rownames(spread_sheet) %in% bad_accnos),]
+
+write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
 ```

### 2015-05-20 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Finalize through generating final metadata file

@@
 ```{r format_taxonomy}
-tax_data <- read.table(file="archaea.taxonomy.nds", sep="\t")
+tax_data <- read.table(file="data/mothur/archaea.taxonomy.nds", sep="\t")
 
@@
 
-write.table(cbind(seq_names, no.subs.str), "archaea.tax", row.names=F,
+write.table(cbind(seq_names, no.subs.str), "data/mothur/archaea.tax", row.names=F,
                                                 col.names=F, quote=F, sep="\t")
@@
 ```{r get_summary, engine="bash", results="hide", echo=TRUE}
-mothur "#summary.seqs(fasta=archaea.fasta, processors=12)"
+mothur "#summary.seqs(fasta=data/mothur/archaea.fasta, processors=12)"
 ```
@@
 ```{r find_screening_points}
-data <- read.table(file="archaea.summary", header=T, row.names=1)
+data <- read.table(file="data/mothur/archaea.summary", header=T, row.names=1)
 quantile(data$start, probs=seq(0.9,1,0.01))
@@
 mothur "#set.current(processors=8);
-        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);"
+        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900, inputdir=data/mothur/, outputdir=data/mothur/);"
 ```
@@
 ```{r single_cell, engine="bash"}
-mothur "#align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
         screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
-        classify.seqs(fasta=current, reference=trainset10_082014.pds.fasta, taxonomy=trainset10_082014.pds.tax, cutoff=80, processors=8)"
+        classify.seqs(fasta=current, reference=data/references/trainset10_082014.pds.fasta, taxonomy=data/references/trainset10_082014.pds.tax, cutoff=80, processors=8)"
 ```
@@
 ```{r pool_data, engine="bash"}
-cat archaea.good.fasta single_cell.archaea.good.align > all_archaea.align
-cat archaea.good.tax single_cell.archaea.good.pds.wang.taxonomy > all_archaea.taxonomy
+cat data/mothur/archaea.good.fasta data/mothur/single_cell.archaea.good.align > data/mothur/all_archaea.align
+cat data/mothur/archaea.good.tax data/mothur/single_cell.archaea.good.pds.wang.taxonomy > data/mothur/all_archaea.taxonomy
 
-mothur "#filter.seqs(fasta=all_archaea.align, vertical=T, trump=.);
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        filter.seqs(fasta=all_archaea.align, vertical=T, trump=.);
         unique.seqs();
@@
 #need to make the taxonomy file match the fasta and names files...
-cut -f 1 all_archaea.filter.unique.precluster.names > precluster.accnos
-mothur "#get.seqs(taxonomy=all_archaea.taxonomy, accnos=precluster.accnos)"
+cut -f 1 data/mothur/all_archaea.filter.unique.precluster.names > data/mothur/precluster.accnos
+mothur "#set.dir(input=data/mothur/, output=data/mothur/);
+        get.seqs(taxonomy=all_archaea.taxonomy, accnos=precluster.accnos)"
 ```
@@
 ```{r cluster_data, engine="bash"}
-mothur "#cluster.split(fasta=all_archaea.filter.unique.precluster.fasta, name=all_archaea.filter.unique.precluster.names, taxonomy=all_archaea.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
+mothur "#set.dir(input=data/mothur/, output=data/mothur/);
+        cluster.split(fasta=all_archaea.filter.unique.precluster.fasta, name=all_archaea.filter.unique.precluster.names, taxonomy=all_archaea.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
 ```
@@
 
-metadata <- read.table(file="archaea_metadata.nds", sep="\t", stringsAsFactors=FALSE,
-                            col.names=metadata_fields, fill=TRUE, na.strings="",
-                            quote="")
-rownames(metadata) <- paste(metadata$name, metadata$acc, sep=".")
+metadata <- read.table(file="data/mothur/archaea_metadata.nds", sep="\t",
+                        stringsAsFactors=FALSE, col.names=metadata_fields,
+                        fill=TRUE, na.strings="", quote="")
+rownames(metadata) <- paste(metadata$acc, metadata$name, sep=".")
 metadata <- metadata[,-c(1,2)]
@@
 ```{r}
-bad_accnos_data <- read.table(file="archaea.bad.accnos", col.names=c("name.accnos", "reason"))
+if(!"openxlsx" %in% rownames(installed.packages())){
+    install.packages("openxlsx")
+}
+library("openxlsx")
+
+bad_accnos_data <- read.table(file="data/mothur/archaea.bad.accnos",
+                    col.names=c("name.accnos", "reason"), stringsAsFactors=F)
 bad_accnos <- bad_accnos_data$name.accnos
-metadata_good <- metadata[-c(bad_accnos),]
-write.table(metadata_good, file="all_archaea.good.metadata", quote=F, sep="\t")
+metadata_good <- metadata[-which(rownames(metadata) %in% bad_accnos),]
+
+categories_sheet <- read.xlsx("data/raw/FinalCategories.xlsx", sheet=1, startRow=1, colNames=TRUE)
+categories <- categories_sheet[,1]
+names(categories) <- categories_sheet[,2]
+
+categories_sheet_missing <- read.xlsx("data/raw/ExtraCategoriesArchaea.xlsx", sheet=1, startRow=1, colNames=FALSE)
+categories_missing <- categories_sheet_missing[,"X1"]
+names(categories_missing) <- categories_sheet_missing[,"X2"]
+
+categories <- c(categories, categories_missing)
+
+metadata_good$category <- toupper(categories[metadata_good$isolation_source])
+
+extra <- table(metadata_good$isolation_source[is.na(metadata_good$category)])
+
+if(length(extra) == 0){
+    write.table(extra, file="missing.txt", row.names=F, col.names=F, quote=FALSE,  sep="\t")
+}
+
+write.table(metadata_good, file="data/mothur/archaea.good.metadata", quote=F, sep="\t")
 ```
@@
 ```{r}
-if(!"openxlsx" %in% rownames(installed.packages())){
-    install.packages("openxlsx")
-}
-library("openxlsx")
+spread_sheet <- read.xlsx("data/raw/FormattedArchaealMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
+rownames(spread_sheet) <- spread_sheet$name
+spread_sheet <- spread_sheet[,-c(1,2)]
 
-spread_sheet <- read.xlsx("data/raw/ArchaealSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
-bad_accnos_data <- read.table(file="data/mothur/single_cell.archaea.bad.accnos", col.names=c("name.accnos", "reason"))
-bad_accnos <- bad_accnos_data$name.accnos
-spread_sheet_good <- spread_sheet[-c(bad_accnos),]
+#need actual category from LSU group
+spread_sheet$category <- NA
 
-write.table(spread_sheet_good, file="data/mothur/all_archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+metadata_genome_id <- gsub(".* (P?SCGC \\S*).*", "\\1", spread_sheet$full_name)
+metadata_genome_id[metadata_genome_id=="SCGC AAA011-J2"] <- "SCGC AAA011-J02"
+
+#need to fix mapping between genome and gene sequences
+fasta <- scan("data/mothur/single_cell.archaea.fasta", what="", quiet=TRUE, sep="\n")
+headers <- fasta[grepl(">", fasta)]
+seq_numbers <- gsub(">(\\d*) .*", "\\1", headers)
+parse_header <- gsub(".* (P?SCGC \\S*) .*", "\\1", headers)
+parse_header <- gsub(".*SAG-(\\S*) .*", "SCGC \\1", parse_header)
+
+names(seq_numbers) <- parse_header
+
+#recreate a metadata table using sequence names and spread_sheet
+full_spread_sheet <- spread_sheet[names(seq_numbers) %in% metadata_genome_id,]
+rownames(full_spread_sheet) <- seq_numbers
+
+bad_accnos_data <- read.table(file="data/mothur/single_cell.archaea.bad.accnos", col.names=c("name.accnos", "reason"),  stringsAsFactors=FALSE)
+bad_accnos <- as.character(bad_accnos_data$name.accnos)
+
+spread_sheet_good <- spread_sheet[-which(rownames(spread_sheet) %in% bad_accnos),]
+
+write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
 ```

### 2015-06-10 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Fix variable name to match genome and gene number

@@
 
-#need actual category from LSU group
-spread_sheet$category <- NA
-
 metadata_genome_id <- gsub(".* (P?SCGC \\S*)", "\\1", spread_sheet$full_name)
-metadata_genome_id["SCGC AAA003-L8"] <- "SCGC AAA003-L08"
+metadata_genome_id[metadata_genome_id=="SCGC AAA003-L8"] <- "SCGC AAA003-L08"
 
@@
 
-#these sequences are missing from metadata file
+#these sequences are missing from fasta file
 #metadata_genome_id[! metadata_genome_id %in% names(seq_numbers)]
@@
 
-spread_sheet_good <- spread_sheet[-which(rownames(spread_sheet) %in% bad_accnos),]
+spread_sheet_good <- full_spread_sheet[-which(rownames(full_spread_sheet) %in% bad_accnos),]
 

### 2015-06-10 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Output metadata with quotes to facillitate reading back in

@@
 
-write.table(metadata_good, file="data/mothur/bacteria.good.metadata", quote=F, sep="\t")
+write.table(metadata_good, file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t")
 ```
@@
 
-write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t", col.names=F, append=T)
 ```

### 2015-06-14 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Concatenate the OTU number and taxonomy to metadata file

@@
 ```
+
+Let's now take the big metadata file that we have and concatenate on the OTU assignment and taxonomy information:
+
+```{r}
+#here we'll read in the list file and extract the OTU assignments for the 0.03
+#cutoff
+list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")[5]
+
+#let's remove the otu label and the number of OTUs at this cutoff
+list_otus <- unlist(strsplit(list_text, "\t"))[-c(1,2)]
+n_otus <- length(list_otus)
+
+#this function will take the sequence names in an OTU, split them up and then
+#assign them the number of that OTU
+split_otu <- function(counter){
+    seq_vector <- unlist(strsplit(list_otus[counter], ","))
+    n_seqs <- length(seq_vector)
+    otu_number <- rep(counter, n_seqs)
+    names(otu_number) <- seq_vector
+    otu_number
+}
+
+#now we get the OTU number for each sequence in the dataset
+otu_assignment <- unlist(lapply(1:n_otus, split_otu))
+
+#let's get the sequence classification data and remove the confidence score data
+taxonomy_file <- read.table(file="data/mothur/all_bacteria.taxonomy", header=F, row.names=1, stringsAsFactors=T)
+taxonomy_data <- gsub("\\(\\d*\\)", "", taxonomy_file$V2)
+names(taxonomy_data) <- rownames(taxonomy_file)
+
+#let's get the metadata file and paste in the OTU and taxonomy data
+metadata <- read.table(file="data/mothur/bacteria.good.metadata", header=T, stringsAsFactors=FALSE)
+metadata$otu <- otu_assignment[as.character(rownames(metadata))]
+metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]
+
+#let's make sure all of the category names are upper cased
+metadata$category <- toupper(metadata$category)
+
+#... and we output the metadata file...
+write.table(metadata, file="data/mothur/bacteria.final.metadata", sep="\t")
+
+#... and we output a group file to use for splitting up the list file
+write.table(file="data/mothur/bacteria.category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+```
+
+One thing we'd like to look at are the rarefaction curves for each environmental category. To do this we'll run the mothur command like so..
+
+```{r, engine="bash"}
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.category.groups, label=unique-0.03-0.05-0.10-0.20);
+        rarefaction.single(shared=all_bacteria.filter.unique.precluster.an.shared)"
+```

### 2015-06-14 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Concatenate the OTU number and taxonomy to metadata file

@@
 
-write.table(metadata_good, file="data/mothur/archaea.good.metadata", quote=F, sep="\t")
+write.table(metadata_good, file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t")
 ```
@@
 spread_sheet <- spread_sheet[,-c(1,2)]
-
-#need actual category from LSU group
-spread_sheet$category <- NA
 
@@
 
-spread_sheet_good <- spread_sheet[-which(rownames(spread_sheet) %in% bad_accnos),]
+spread_sheet_good <- full_spread_sheet[-which(rownames(full_spread_sheet) %in% bad_accnos),]
+colnames(spread_sheet_good) <- gsub("\\.", "_", colnames(spread_sheet_good))
 
-write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
 ```
+
+Let's now take the big metadata file that we have and concatenate on the OTU assignment and taxonomy information:
+
+```{r}
+#here we'll read in the list file and extract the OTU assignments for the 0.03
+#cutoff
+list_text <- scan(file="data/mothur/all_archaea.filter.unique.precluster.an.list", sep="\n", what="")[5]
+
+#let's remove the otu label and the number of OTUs at this cutoff
+list_otus <- unlist(strsplit(list_text, "\t"))[-c(1,2)]
+n_otus <- length(list_otus)
+
+#this function will take the sequence names in an OTU, split them up and then
+#assign them the number of that OTU
+split_otu <- function(counter){
+    seq_vector <- unlist(strsplit(list_otus[counter], ","))
+    n_seqs <- length(seq_vector)
+    otu_number <- rep(counter, n_seqs)
+    names(otu_number) <- seq_vector
+    otu_number
+}
+
+#now we get the OTU number for each sequence in the dataset
+otu_assignment <- unlist(lapply(1:n_otus, split_otu))
+
+#let's get the sequence classification data and remove the confidence score data
+taxonomy_file <- read.table(file="data/mothur/all_archaea.taxonomy", header=F, row.names=1, stringsAsFactors=T)
+taxonomy_data <- gsub("\\(\\d*\\)", "", taxonomy_file$V2)
+names(taxonomy_data) <- rownames(taxonomy_file)
+
+#let's get the metadata file and paste in the OTU and taxonomy data
+metadata <- read.table(file="data/mothur/archaea.good.metadata", header=T, stringsAsFactors=FALSE)
+metadata$otu <- otu_assignment[as.character(rownames(metadata))]
+metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]
+
+#let's make sure all of the category names are upper cased
+metadata$category <- toupper(metadata$category)
+
+#... and we output the metadata file...
+write.table(metadata, file="data/mothur/archaea.final.metadata", sep="\t")
+
+#... and we output a group file to use for splitting up the list file
+write.table(file="data/mothur/archaea.category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+```
+
+One thing we'd like to look at are the rarefaction curves for each environmental category. To do this we'll run the mothur command like so..
+
+```{r, engine="bash"}
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.category.groups, label=unique-0.03-0.05-0.10-0.20);
+        rarefaction.single(shared=all_archaea.filter.unique.precluster.an.shared)"
+```

### 2015-07-03 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
Add groupings defined by Rene

@@
 
+#correct typos:
+metadata$category[metadata$category == 'AF'] <- 'AE'
+metadata$category[metadata$category == 'AQBF'] <- 'AQB'
+metadata$category[metadata$category == 'AQD'] <- 'AQO'
+metadata$category[metadata$category == 'AQHS'] <- 'AQH'
+metadata$category[metadata$category == 'AQS'] <- 'AQMS'
+metadata$category[metadata$category == 'AWF'] <- 'AQF'
+metadata$category[metadata$category == 'DS'] <- 'BS'
+metadata$category[metadata$category == 'S'] <- 'SO'
+metadata$category[metadata$category == 'SI'] <- 'BI'
+metadata$category[metadata$category == 'V'] <- 'BO'
+metadata$category[metadata$category == 'VD'] <- 'BD'
+metadata$category[metadata$category == 'ZOZO'] <- 'ZO'
+metadata$category[metadata$category == 'AH'] <- 'AQH'
+metadata$category[metadata$category == 'AQOS'] <- 'AQS'
+metadata$category[metadata$category == 'BS'] <- 'BD'
+
 #... and we output the metadata file...
@@
 
-#... and we output a group file to use for splitting up the list file
-write.table(file="data/mothur/bacteria.category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+#... and we output a group file to use for splitting up the list file into the individual categories...
+write.table(file="data/mothur/bacteria.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+
+#... and we output a group file to use for splitting up the list file into the pooled categories...
+#   For bacteria: AQF, AQFS, AQM, AQMS, BD, BI, BP, OT, PR, SO, ZO, and ZV and the others into one category
+#   per Rene 6/24/2015...
+pool1 <- c("AQF", "AQFS", "AQM", "AQMS", "BD", "BI", "BP", "OT", "PR", "SO", "ZO", "ZV")
+pool_categories <- rep(NA, nrow(metadata))
+pool_categories[metadata$category %in% pool1] <- "pool1"
+pool_categories[!metadata$category %in% pool1] <- "pool2"
+pool_categories[is.na(metadata$category)] <- "NA"
+write.table(file="data/mothur/bacteria.pool_category.groups", x=cbind(rownames(metadata), pool_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+
+#... and we pool the categories by environment
+env_categories <- rep(NA, nrow(metadata))
+env_categories[metadata$category == "AE"] <- "aerosol"
+env_categories[metadata$category == "OT"] <- "other"
+env_categories[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
+env_categories[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
+env_categories[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
+env_categories[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
+env_categories[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
+write.table(file="data/mothur/bacteria.env_category.groups", x=cbind(rownames(metadata), env_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
 ```
 
-One thing we'd like to look at are the rarefaction curves for each environmental category. To do this we'll run the mothur command like so..
+One thing we'd like to look at are the rarefaction curves for each set of group files and the entire dataset. To do this we'll run the mothur command like so..
 
@@
 mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.category.groups, label=unique-0.03-0.05-0.10-0.20);
-        rarefaction.single(shared=all_bacteria.filter.unique.precluster.an.shared)"
+        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.all_category.groups, label=unique-0.03-0.05-0.10-0.20);
+        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.all_categories.shared);
+        rarefaction.single(shared=all_bacteria.all_categories.shared);
+        summary.single(shared=all_bacteria.all_categories.shared, calc=nseqs-sobs-coverage)"
+
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.pool_category.groups, label=unique-0.03-0.05-0.10-0.20);
+        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.pool_category.shared);
+        rarefaction.single(shared=all_bacteria.pool_category.shared);
+        summary.single(shared=all_bacteria.pool_category.shared, calc=nseqs-sobs-coverage)"
+
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.env_category.groups, label=unique-0.03-0.05-0.10-0.20);
+        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.env_category.shared);
+        rarefaction.single(shared=all_bacteria.env_category.shared);
+        summary.single(shared=all_bacteria.env_category.shared, calc=nseqs-sobs-coverage)"
+
+mothur "#rarefaction.single(list=data/mothur/all_bacteria.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20);"
 ```

### 2015-07-03 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Add groupings defined by Rene

@@
 
+#correct typos:
+metadata$category[metadata$category == 'AQD'] <- 'AQO'
+metadata$category[metadata$category == 'AQHS'] <- 'AQH'
+metadata$category[metadata$category == 'AQS'] <- 'AQMS'
+metadata$category[metadata$category == 'VD'] <- 'BD'
+metadata$category[metadata$category == 'BM'] <- 'BI'
+metadata$category[metadata$category == 'AH'] <- 'AQH'
+
 #... and we output the metadata file...
@@
 
-#... and we output a group file to use for splitting up the list file
-write.table(file="data/mothur/archaea.category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+#... and we output a group file to use for splitting up the list file into the individual categories...
+write.table(file="data/mothur/archaea.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+
+#... and we output a group file to use for splitting up the list file into the pooled categories...
+#   For archaea: AQH, AQM, AQMS, and ZV into one. the rest into the other
+#   per Rene 6/24/2015...
+pool1 <- c("AQH", "AQM", "AQMS", "ZV")
+pool_categories <- rep(NA, nrow(metadata))
+pool_categories[metadata$category %in% pool1] <- "pool1"
+pool_categories[!metadata$category %in% pool1] <- "pool2"
+pool_categories[is.na(metadata$category)] <- "NA"
+write.table(file="data/mothur/archaea.pool_category.groups", x=cbind(rownames(metadata), pool_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
+
+#... and we pool the categories by environment
+env_categories <- rep(NA, nrow(metadata))
+env_categories[metadata$category == "AE"] <- "aerosol"
+env_categories[metadata$category == "OT"] <- "other"
+env_categories[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
+env_categories[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
+env_categories[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
+env_categories[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
+env_categories[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
+write.table(file="data/mothur/archaea.env_category.groups", x=cbind(rownames(metadata), env_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
 ```
 
-One thing we'd like to look at are the rarefaction curves for each environmental category. To do this we'll run the mothur command like so..
+One thing we'd like to look at are the rarefaction curves for each set of group files and the entire dataset. To do this we'll run the mothur command like so..
 
@@
 mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.category.groups, label=unique-0.03-0.05-0.10-0.20);
-        rarefaction.single(shared=all_archaea.filter.unique.precluster.an.shared)"
+        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.all_category.groups, label=unique-0.03-0.05-0.10-0.20);
+        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.all_categories.shared);
+        rarefaction.single(shared=all_archaea.all_categories.shared);
+        summary.single(shared=all_archaea.all_categories.shared, calc=nseqs-sobs-coverage)"
+
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.pool_category.groups, label=unique-0.03-0.05-0.10-0.20);
+        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.pool_category.shared);
+        rarefaction.single(shared=all_archaea.pool_category.shared);
+        summary.single(shared=all_archaea.pool_category.shared, calc=nseqs-sobs-coverage)"
+
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.env_category.groups, label=unique-0.03-0.05-0.10-0.20);
+        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.env_category.shared);
+        rarefaction.single(shared=all_archaea.env_category.shared);
+        summary.single(shared=all_archaea.env_category.shared, calc=nseqs-sobs-coverage)"
+
+mothur "#rarefaction.single(list=data/mothur/all_archaea.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20);"
 ```

### 2015-10-20 doc/notebook/Archaeal_data_acquisition.Rmd
Rename files

--- 2015-05-20 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
@@
 
-spread_sheet_good <- spread_sheet[-which(rownames(spread_sheet) %in% bad_accnos),]
+spread_sheet_good <- full_spread_sheet[-which(rownames(full_spread_sheet) %in% bad_accnos),]
+colnames(spread_sheet_good) <- gsub("\\.", "_", colnames(spread_sheet_good))
 
-write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
+write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
 ```

### 2015-10-20 doc/notebook/Archaeal_data_acquisition.Rmd
Fix how rendered to markdown
Fix bullet list formatting

--- 2015-07-03 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
@@
+---
+output: md_document
+---
+
 ```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
@@
 opts_chunk$set(eval = FALSE)
-opts_chunk$set(echo = FALSE)
+opts_chunk$set(echo = TRUE)
 ```
@@
 To get good sequences...
+
 * Go to search window
@@
 * Hit Mark Listed, Unmark Rest button (N=1493493)
-* The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure
-    out which taxonomy to base the analysis on. Do the following searches...
+
+The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure out which taxonomy to base the analysis on. Do the following searches...
+
 * ARB_color != 1 & tax_rdp == "*Archaea*" (N=40288)
@@
 Now we need the taxonomy information.
+
 * Go Tree -> NDS

### 2015-10-20 doc/notebook/Bacterial_data_acquisition.Rmd
Fix how rendered to markdown
Fix bullet list formatting

--- 2015-07-03 doc/notebook/2015-02-26_Bacterial_data_acquisition.Rmd
@@
 * Hit Mark Listed, Unmark Rest button (N=1493493)
-* The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure
-    out which taxonomy to base the analysis on. Do the following searches...
+
+The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure out which taxonomy to base the analysis on. Do the following searches...
+
 * ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1324558)

### 2015-10-20 Schloss_Census2_XXXX_2015.Rmd
Copy Cameron's text into Rmd

---
title: 'The status of the microbial census: an update'
author: 
output: md_document
---

Updating the status of the microbial census

Patrick D. Schloss1*, Rene Girard2, Thomas Martin2, Joshua Edwards2, J. Cameron Thrash2*

1.	Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109, U.S.A.
2.	Department of Biological Sciences, Louisiana State University, Baton Rouge, LA 70803, U.S.A.

*Correspondence: pschloss@umich.edu, thrashc@lsu.edu

## Abstract

## Introduction

The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology. At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process immensely. Furthermore, creating standards for delimiting what makes one microbe “different” from another has proved contentious at best. In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, and metagenomics.
	Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene. Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science. While reliance on this gene subjects us to biases created by primer selection*, differences in amplification strength* and fidelity*, internal features which may disrupt traditional measurements*, and potentially misleading classification due to infrequent horizontal gene transfer*, the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker. Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
	In 2004, the Status of the Microbial Census analyzed 56,215 partial rRNA gene sequences available in the Ribosomal Database Project and concluded that “either current sampling methods are not adequate to identify 107 to 109 different species or these estimates are high.” (Schloss, 2004) Since that time, the explosion of short-read high throughput sequencing technology has caused a dramatic increase in the number of sequences deposited in public databases. Currently there are over XXXXXX 16S rRNA gene sequences in SRA. Recent estimates based on this short read data place the total number of microbial species at around XXXXX.**
However, the increase in sequencing volume has come at the cost of sequence length. The commonly used iTag sequencing technology amplifies an approximately 250 bp region of the 16S rRNA gene at the V4 hypervariable region, and other schemes have used different parts of the gene. The decrease in length means a decrease in informative sites, making direct comparisons between short read sequencing and mostly or completely full-length sequences difficult. Youssef et al. (2009) demonstrated high variability between the number of OTUs that different hypervariable sites estimated compared to full-length 16S rRNA gene sequences, and also that sample location influenced this behavior. Thus, it remains unclear to what degree richness estimates from short read technology over or underestimate the numbers from full-length sequences.
Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences. In the 12 years since the collection of data for Schloss and Handelsman’s analysis, the number of available sequences has risen over twenty-fold, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study.

## Materials and Methods

Sequence data
Primary sequences were downloaded from the ARB-SILVA database on XXXX, 2015 with the requirement that sequences were ≥ XXXX bp. This resulted in XXXXXXXX unique sequences. Additional sequence data was included from single-cell genomes available on IMG, many of which were recently obtained via the GEBA-MDM effort in Rinke et al. (2014). “SCGC” was searched on the IMG database 12 Mar 2015, manually separated into bacterial and archaeal SSU genes, and associated metadata was also downloaded.

Classification of organisms based on source metadata
Organisms were classified based on seven broad “coarse” categories, and refined further to facilitate additional analyses with twenty-six more specific “fine” categories (Table 1). These were assigned based on manual curation of the metadata associated with sequences obtained from the ARB-SILVA and IMG databases. The original metadata table was edited through sorting the “Source” category alphabetically and filtering for unique terms. The resulting XXXXXX terms were then manually assigned to the coarse and fine categories in Table 1. For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse “Other” category.

OTU clustering with Mothur

Plots
Rarefaction curves were completed in R using modifications to the script available in Supplementary Information.

## Results and Discussion

Caveats
Recent data suggests that a considerable diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers. This dataset does not include sequences from metagenomic assemblies but Brown et al. (2015) have used such assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called “Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys. Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015). Thus, for many environments, the estimates within must be considered as lower bounds. 
Table 1. Habitat classifications

### 2015-10-23 doc/notebook/Bacterial_data_acquisition.Rmd
Use github flavored markdown
Add metadata for release 123

@@
 ---
-output: md_document
+output:
+  md_document:
+    variant: markdown_github
 ---
@@
 ```{r get_data, engine="bash"}
-wget http://www.arb-silva.de/fileadmin/silva_databases/release_119/ARB_files/SSURef_119_SILVA_14_07_14_opt.arb.tgz
-wget http://www.arb-silva.de/fileadmin/silva_databases/release_119/ARB_files/SSURef_119_SILVA_14_07_14_opt.arb.tgz.md5
-tar xvzf SSURef_119_SILVA_14_07_14_opt.arb.tgz
-mv SSURef_119_SILVA_14_07_14_opt.arb SSURef_119_SILVA_14_07_14_opt.arb.tgz.md5 data/references
-rm SSURef_119_SILVA_14_07_14_opt.arb.tgz
+wget http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz
+wget http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz.md5
+tar xvzf SSURef_123_SILVA_19_07_15_opt.arb.tgz         
+mv SSURef_123_SILVA_19_07_15_opt.arb SSURef_123_SILVA_19_07_15_opt.arb.tgz.md5 data/references
+rm SSURef_123_SILVA_19_07_15_opt.arb.tgz
 ```
@@
 * Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
-* Hit Mark Listed, Unmark Rest button (N=1493493)
+* Hit Mark Listed, Unmark Rest button (N=1,652,440)
 
@@
 
-* ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1324558)
-* ARB_color != 1 & tax_greengenes == "*Bacteria*" (N=1123144)
-* ARB_color != 1 & tax_slv == "*Bacteria*" (N=426917)
-* ARB_color != 1 & tax_embl == "*Bacteria*" (N=1207210)
-* tax_rdp | tax_greengenes | tax_embl | tax_slv (N=1376910)
+* ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1,384,528)
+* ARB_color != 1 & tax_greengenes == "*Bacteria*" (N=1,123,279)
+* ARB_color != 1 & tax_slv == "*Bacteria*" (N=1,526,523)
+* ARB_color != 1 & tax_embl == "*Bacteria*" (N=1,335,715)
+* ARB_color != 1 & (tax_rdp | tax_greengenes | tax_embl | tax_slv) (N=1,529,365)
 
-Let's stick with the bacterial sequences that have an RDP taxonomy and we will
-analyze them in RDP space, but preserve the taxonomy generated by the other
-systems. Nicely, the RDP taxonomy does not include mitochondrial
-16S rRNA gene sequences, so all we'll need to remove are the chloroplast
-sequences
+Let's stick with the bacterial sequences that have an Silva taxonomy and we will
+analyze them in Silva space, but preserve the taxonomy generated by the other
+systems. The Silva taxonomy includes mitochondrial and chloroplast 16S rRNA gene
+sequences, so we'll need to remove them.
 
@@
 * Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
-* Change the second search field to "tax_rdp" and set it to "*Bacteria*"
-* Change the third search field to "tax_rdp" and set it to "*Chloroplast*"
-* Set the two Query operators to "and" and hit "Search"
-* Hit Mark Listed Unmark Rest (N=1,301,541)
+* Change the second search field to "tax_slv" and set it to "*Bacteria*"
+* Click "Mark Listed, Unmark Rest"
+* Change the first search field to "tax_slv" and set it to "*Chloroplast*", set to "not equal"
+* Change the second search field to "tax_slv" and set it to "*Mitochondria*", set to "not equal"
+* Set the two Query operators to "and"
+* Click "Keep species" and "that match query"
+* Hit "Search"
+* Hit Mark Listed Unmark Rest (N=1,515,024)
 * Click "Write to Fields of Listed", select the "remark" field, and enter "good bacteria"
@@
 * Go Tree -> NDS
-* Click "name", "acc", "tax_rdp". The "tax_rdp" field should have 250 characters
+* Click "name", "acc", "tax_slv". The "tax_slv" field should have 250 characters
 * Unclick everything else
@@
 The next thing we need to do is to clean up the bacteria.taxonomy.nds file
-to make it into a proper, mothur compatible bacteria.taxonomy file. First we
-want to remove the names of any sub taxa (e.g. suborder). To do this we need
-to get the RDP list of names and the taxonomic level they belong to...
-
-```{r get_rdp_names, engine="bash"}
-wget -N http://sourceforge.net/projects/rdp-classifier/files/RDP_Classifier_TrainingData/RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
-unzip -o RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
-mv RDPClassifier_16S_trainsetNo10_rawtrainingdata/* data/references/
-rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata*
-```
-
-Now we're read to do some R'ing to get the taxonomy file formatted properly...
+to make it into a proper, mothur compatible bacteria.taxonomy file. We'll do
+some R'ing to get the taxonomy file formatted properly...
 
 ```{r format_taxonomy}
-tax_data <- read.table(file="data/mothur/bacteria.taxonomy.nds", sep="\t")
+tax_data <- read.table(file="data/mothur/bacteria.taxonomy.nds", sep="\t", stringsAsFactors=FALSE)
 
@@
 taxonomy <- gsub(" ", "_", tax_data$V3)  #convert any spaces to underscores
-taxonomy <- gsub("[^;]*_incertae_sedis$", "", taxonomy) #remove the unknowns
 taxonomy <- gsub('\"', '', taxonomy) #remove quote marks
 
-#now we need to pull out the taxon names for the sub taxa levels
-levels <- read.table(file="data/references/trainset10_db_taxid.txt", sep="*", stringsAsFactors=F)
-subs <- levels[grep("sub", levels$V5),] #get the sub taxa names
-sub.names <- subs$V2
-
-tax.split <- strsplit(taxonomy, split=";")  #split the taxonomy string by semicolon
-
-#function to see which taxa names in a list are found in the sub taxa list
-remove.subs <- function(tax.vector){
-    return(tax.vector[which(!tax.vector %in% sub.names)])
-}
-
-#apply sub taxa finding function to all sequences
-no.subs <- lapply(tax.split, remove.subs)
-
-#merge data together by semicolon for each sequence
-no.subs.str <- unlist(lapply(no.subs, paste, collapse=";"))
-
-#Tack a semicolon on to the end of each sequence
-no.subs.str <- paste0(no.subs.str, ";")
-
-write.table(cbind(seq_names, no.subs.str), "data/mothur/bacteria.tax", row.names=F,
+write.table(cbind(seq_names, taxonomy), "data/mothur/bacteria.tax", row.names=F,
                                                 col.names=F, quote=F, sep="\t")
@@
             Start	End		NBases	Ambigs	Polymer	NumSeqs
-Minimum:        1       9693    900     0       4       1
-2.5%-tile:      1004    12895   1279    0       5       32539
-25%-tile:       1026    13069   1355    0       5       325386
-Median:         1044    13267   1382    0       5       650771
-75%-tile:       1051    13825   1447    0       6       976156
-97.5%-tile:     1793    13991   1522    2       7       1269003
-Maximum:        6448    15767   3281    51      38      1301541
-Mean:   1096.18 13442   1399.39 0.176343        5.49738
-# of Seqs:      1301541
+Minimum:    1		10725	900		0		4	1
+2.5%-tile:	1003	14692	1274	0		5	37876
+25%-tile:	1011	14865	1355	0		5	378757
+Median: 	1044	15082	1387	0		5	757513
+75%-tile:	1048	15672	1455	0		6	1136269
+97.5%-tile:	1815	15771	1539	2		7	1477149
+Maximum:	7036	17738	3469	64		38	1515024
+Mean:		1104.14	15254.4	1404	0.181	5.509
+# of Seqs:	1515024
 ```
@@
 Notes...
+
 * Looks like an archaeal sequence slipped through
@@
 ```{r find_screening_points}
-data <- read.table(file="bacteria.summary", header=T, row.names=1)
+data <- read.table(file="data/mothur/bacteria.summary", header=T, row.names=1)
 quantile(data$start, probs=seq(0.9,1,0.01))
@@
 quantile(data$end, probs=seq(0,0.1,0.01))
-end <- 13000
+end <- 14853
 
@@
 ```{r screen_data, engine="bash"}
-mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200, outputdir=data/mothur/, processors=8);"
+mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=14853, maxambig=2, minlength=1200, outputdir=data/mothur/, processors=8);"
 ```
@@
 ```{r single_cell, engine="bash"}
-wget -N -P data/references/ http://www.mothur.org/w/images/2/24/Trainset10_082014.pds.tgz
-tar xvzf data/references/Trainset10_082014.pds.tgz -C data/references/
-mv data/references/trainset10_082014.pds/trainset10_082014.* data/references/
-rm -rf data/references/trainset10_082014.pds
+#wget -N -P data/references/ http://www.mothur.org/w/images/2/24/Trainset10_082014.pds.tgz
+#tar xvzf data/references/Trainset10_082014.pds.tgz -C data/references/
+#mv data/references/trainset10_082014.pds/trainset10_082014.* data/references/
+#rm -rf data/references/trainset10_082014.pds
 
 mothur "#set.dir(output=data/mothur);
-        align.seqs(fasta=data/raw/single_cell.bacteria.fasta, reference=data/mothur/bacteria.good.fasta);
+        align.seqs(fasta=data/mothur/single_cell.bacteria.fasta, reference=data/mothur/bacteria.good.fasta);
         set.dir(input=data/mothur);
         screen.seqs(fasta=current, start=2000, end=13000, maxambig=2, minlength=1200);
-        classify.seqs(fasta=current, reference=data/references/trainset10_082014.pds.fasta, taxonomy=data/references/trainset10_082014.pds.tax, cutoff=80, processors=1)"
+        classify.seqs(fasta=current, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80)"
 ```
@@
 cat data/mothur/bacteria.good.fasta data/mothur/single_cell.bacteria.good.align > data/mothur/all_bacteria.align
-cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.good.pds.wang.taxonomy > data/mothur/all_bacteria.taxonomy
+cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.good.good.wang.taxonomy > data/mothur/all_bacteria.taxonomy
 
@@
 
-metadata <- read.table(file="data/mothur/bacteria.metadata.nds", sep="\t", stringsAsFactors=FALSE,
-                        col.names=metadata_fields, fill=TRUE, na.strings="",
-                        quote="")
+metadata <- read.table(file="data/mothur/bacteria.metadata.nds", sep="\t",
+						stringsAsFactors=FALSE, col.names=metadata_fields,
+						fill=TRUE, na.strings="", quote="")
 rownames(metadata) <- paste(metadata$acc, metadata$name, sep=".")
@@
 
-if(length(extra) == 0){
+if(length(extra) != 0){
     write.table(extra, file="missing.txt", row.names=F, col.names=F, sep="\t")

### 2015-10-24 doc/notebook/Bacterial_data_acquisition.Rmd
Add path to precluster.accnos steps

@@
 #need to make the taxonomy file match the fasta and names files...
-cut -f 1 all_bacteria.filter.unique.precluster.names > precluster.accnos
+cut -f 1 data/mothur/all_bacteria.filter.unique.precluster.names > precluster.accnos
 mothur "#set.dir(input=data/mothur, output=data/mothur);

### 2015-11-09 doc/notebook/Bacterial_data_acquisition.Rmd
Update to add new data from v.123

@@
 #here we'll read in the list file and extract the OTU assignments for the 0.03
-#cutoff
-list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")[5]
+#cutoff (for some reason mothur was repeating the header line for each OTU threshold)
+list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")
+
+list_text_03 <- list_text[grepl("^0\\.03", list_text)]
 
 #let's remove the otu label and the number of OTUs at this cutoff
-list_otus <- unlist(strsplit(list_text, "\t"))[-c(1,2)]
+list_otus <- unlist(strsplit(list_text_03, "\t"))[-c(1,2)]
 n_otus <- length(list_otus)
@@
 metadata$category[metadata$category == 'BS'] <- 'BD'
+metadata$category[metadata$category == 'AQS'] <- 'AQFS'
+metadata$category[metadata$category == 'ZL'] <- 'ZV'
+metadata$category[metadata$category == 'SR'] <- 'PR'
+metadata$category[metadata$category == 'M'] <- 'AQM'
+metadata$category[metadata$category == 'AQFM'] <- 'AQFS'
+metadata$category[metadata$category == 'SV'] <- 'ZV'
 

### 2015-11-09 doc/notebook/Archaeal_data_acquisition.Rmd
Use github flavored markdown

@@
 * Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
-* Hit Mark Listed, Unmark Rest button (N=1493493)
+* Hit Mark Listed, Unmark Rest button (N=1652440)
 
@@
 
-* ARB_color != 1 & tax_rdp == "*Archaea*" (N=40288)
-* ARB_color != 1 & tax_greengenes == "*Archaea*" (N=19795)
-* ARB_color != 1 & tax_slv == "*Archaea*" (N=17641)
-* ARB_color != 1 & tax_embl == "*Archaea*" (N=44574)
-* ARB_color != 1 & (tax_rdp | tax_greengenes | tax_embl | tax_slv) (N=46387 )
+* ARB_color != 1 & tax_rdp == "*Archaea*" (N=46146)
+* ARB_color != 1 & tax_greengenes == "*Archaea*" (N=19830)
+* ARB_color != 1 & tax_slv == "*Archaea*" (N=59240)
+* ARB_color != 1 & tax_embl == "*Archaea*" (N=57863)
 
-Let's stick with the archaeal sequences that have an RDP taxonomy and we will
-analyze them in RDP space, but we can always go back to the taxonomy provided by
-the other systems if we'd like. A quick spot checking suggests that the RDP
-taxonomy information is actually richer than that of the EBML taxonomy
-information.
+Let's stick with the archaeal sequences that have an Silva taxonomy and we will
+analyze them in Silva space, but we can always go back to the taxonomy provided by
+the other systems if we'd like.
 
@@
 * Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
-* Change the second search field to "tax_rdp" and set it to "*Archaea*"
+* Change the second search field to "tax_silva" and set it to "*Archaea*"
 * Hit "Search"
-* Hit Mark Listed Unmark Rest (N=40,288)
+* Hit Mark Listed Unmark Rest (N=59,240)
 * Click "Write to Fields of Listed", select the "remark" field, and enter "good archaea"
@@
 * Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
-* Rename `noname.fasta` to `archaea.fasta`
+* Rename `noname.fasta` to `data/mothur/archaea.fasta`
 
@@
 The next thing we need to do is to clean up the archaea.taxonomy.nds file
-to make it into a proper, mothur compatible archaea.taxonomy file. First we
-want to remove the names of any sub taxa (e.g. suborder). To do this we need
-to get the RDP list of names and the taxonomic level they belong to...
-
-```{r get_rdp_names, engine="bash"}
-wget -N http://sourceforge.net/projects/rdp-classifier/files/RDP_Classifier_TrainingData/RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
-unzip -o RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
-mv RDPClassifier_16S_trainsetNo10_rawtrainingdata/* data/references/
-rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata*
-```
-
-Now we're read to do some R'ing to get the taxonomy file formatted properly...
+to make it into a proper, mothur compatible archaea.taxonomy file. Now we're
+ready to do some R'ing to get the taxonomy file formatted properly...
 
@@
 taxonomy <- gsub(" ", "_", tax_data$V3)  #convert any spaces to underscores
-taxonomy <- gsub("[^;]*_incertae_sedis$", "", taxonomy) #remove the unknowns
 taxonomy <- gsub('\"', '', taxonomy) #remove quote marks
 
-#now we need to pull out the taxon names for the sub taxa levels
-levels <- read.table(file="data/references/trainset10_db_taxid.txt", sep="*", stringsAsFactors=F)
-subs <- levels[grep("sub", levels$V5),] #get the sub taxa names
-sub.names <- subs$V2
-
-tax.split <- strsplit(taxonomy, split=";")  #split the taxonomy string by semicolon
-
-#function to see which taxa names in a list are found in the sub taxa list
-remove.subs <- function(tax.vector){
-    return(tax.vector[which(!tax.vector %in% sub.names)])
-}
-
-#apply sub taxa finding function to all sequences
-no.subs <- lapply(tax.split, remove.subs)
-
-#merge data together by semicolon for each sequence
-no.subs.str <- unlist(lapply(no.subs, paste, collapse=";"))
-
-#Tack a semicolon on to the end of each sequence
-no.subs.str <- paste0(no.subs.str, ";")
-
-write.table(cbind(seq_names, no.subs.str), "data/mothur/archaea.tax", row.names=F,
+write.table(cbind(seq_names, taxonomy), "data/mothur/archaea.tax", row.names=F,
                                                 col.names=F, quote=F, sep="\t")
@@
 			Start	End		NBases	Ambigs	Polymer	NumSeqs
-Minimum:	1        4058	900	    0    	4	1
-2.5%-tile:	357      4692	905	    0    	5	1008
-25%-tile:	365      4783	935	    0   	5	10073
-Median: 	379      7409	1262	0   	5	20145
-75%-tile:	456      7729	1349	0    	6	30217
-97.5%-tile:	1746	 7984	1470	4   	7	39281
-Maximum:	2321	 8711	2174	46     	32	40288
-Mean:	    513.659	 6698.2	1185.3	0.3	    5.44244
-# of Seqs:	40288
+Minimum:	1		6806	900		0		4		1
+2.5%-tile:	1004	7718	905		0		5		1482
+25%-tile:	1006	8794	943		0		5		14811
+Median: 	1039	11168	1308	0		5		29621
+75%-tile:	1056	11747	1398	0		6		44431
+97.5%-tile:	3020	11909	1469	3		8		57760
+Maximum:	5159	12877	2962	46		25		59240
+Mean:		1270.65	10393.5	1204.49	0.26683	5.53047
+# of Seqs:	59240
 ```
@@
 quantile(data$start, probs=seq(0.9,1,0.01))
-start <- 2000
+start <- 3000
 
 quantile(data$end, probs=seq(0,0.1,0.01))
-end <- 4058
+end <- 7800
 
@@
 mothur "#set.current(processors=8);
-        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900, inputdir=data/mothur/, outputdir=data/mothur/);"
+        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900, inputdir=data/mothur/, outputdir=data/mothur/);"
 ```
@@
         screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
-        classify.seqs(fasta=current, reference=data/references/trainset10_082014.pds.fasta, taxonomy=data/references/trainset10_082014.pds.tax, cutoff=80, processors=8)"
+        classify.seqs(fasta=current, reference=archaea.good.fasta, taxonomy=archaea.good.tax, cutoff=80, processors=8)"
 ```
@@
 
-if(length(extra) == 0){
+if(length(extra) != 0){
     write.table(extra, file="missing.txt", row.names=F, col.names=F, quote=FALSE,  sep="\t")

### 2015-11-09 code/time_course_plots.R
Add time course analysis

@@ -1,0 +1,105 @@
+bact <- read.table(file='data/mothur/bacteria.final.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
+
+arch <- read.table(file='data/mothur/archaea.final.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
+
+
+pdf(file="results/figures/sequences_deposited.pdf")
+	deposits_table <- table(bact_year_deposited)
+	bact_deposits_by_year <- as.numeric(deposits_table)
+	names(bact_deposits_by_year) <- names(deposits_table)
+	bact_deposits_by_year <- bact_deposits_by_year[!(names(bact_deposits_by_year) %in% c("2015"))]
+
+	plot(bact_deposits_by_year~as.numeric(names(bact_deposits_by_year)), log='y', type="l", col="blue", lwd=2, xlab="Year", ylab="Number of sequences per year", ylim=c(1,1e6), xlim=c(1980, 2015), axes=F)
+
+	deposits_table <- table(arch_year_deposited)
+	arch_deposits_by_year <- as.numeric(deposits_table)
+	names(arch_deposits_by_year) <- names(deposits_table)
+	arch_deposits_by_year <- arch_deposits_by_year[!(names(arch_deposits_by_year) %in% c("2015"))]
+
+	points(arch_deposits_by_year~as.numeric(names(arch_deposits_by_year)), type="l", col="red", lwd=2)
+	axis(1)
+	axis(2, at=c(1,10,1e2,1e3,1e4,1e5,1e6), labels=parse(text=paste("10^", 0:6)), las=1)
+	box()
+	legend(x=2003, y=100, legend=c("Bacteria", "Archaea"), col=c("blue", "red"), lty=1, lwd=2)
+dev.off()
+
+
+pdf(file="results/figures/otus_deposited.pdf")
+	bact_otus_table <- table(bact$otu, bact_year_deposited)
+	bact_otu_appearance <- rep("", nrow(bact_otus_table))
+	names(bact_otu_appearance) <- as.character(1:length(bact_otu_appearance))
+
[…]

### 2015-11-09 code/time_course_plots.R
Fixed x-axis on number of authors plot

@@ -1,3 +1,5 @@
+year_of_analysis <- "2015"
+
 bact <- read.table(file='data/mothur/bacteria.final.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
 
@@ -9,14 +11,14 @@ pdf(file="results/figures/sequences_deposited.pdf")
 	deposits_table <- table(bact_year_deposited)
 	bact_deposits_by_year <- as.numeric(deposits_table)
 	names(bact_deposits_by_year) <- names(deposits_table)
-	bact_deposits_by_year <- bact_deposits_by_year[!(names(bact_deposits_by_year) %in% c("2015"))]
+	bact_deposits_by_year <- bact_deposits_by_year[!(names(bact_deposits_by_year) %in% year_of_analysis)]
 
 	plot(bact_deposits_by_year~as.numeric(names(bact_deposits_by_year)), log='y', type="l", col="blue", lwd=2, xlab="Year", ylab="Number of sequences per year", ylim=c(1,1e6), xlim=c(1980, 2015), axes=F)
 
 	deposits_table <- table(arch_year_deposited)
 	arch_deposits_by_year <- as.numeric(deposits_table)
 	names(arch_deposits_by_year) <- names(deposits_table)
-	arch_deposits_by_year <- arch_deposits_by_year[!(names(arch_deposits_by_year) %in% c("2015"))]
+	arch_deposits_by_year <- arch_deposits_by_year[!(names(arch_deposits_by_year) %in% year_of_analysis)]
 
 	points(arch_deposits_by_year~as.numeric(names(arch_deposits_by_year)), type="l", col="red", lwd=2)
 	axis(1)
@@ -92,7 +94,7 @@ pdf("results/figures/fifty_authors_deposited.pdf")
 	bact_fifty <- unlist(sapply(levels(factor(bact_year_deposited)), bact_how_many))
 	bact_fifty <- bact_fifty[!(names(bact_fifty) %in% c("2015"))]
 
-	plot(bact_fifty~as.numeric(names(bact_fifty)), type="l", lwd=2, xlab="Year", ylab="Number of studies accounting for 50% of sequences deposited", col="blue")
[…]

### 2015-11-09 Schloss_Census2_XXXX_2015.Rmd
Change to github-flavored markdown
Editing and completion of Methods section

@@
 ---
-title: 'The status of the microbial census: an update'
-author: 
-output: md_document
+title: "The status of the microbial census: an update"
+output:
+  md_document:
+    variant: markdown_github
 ---
 
-Updating the status of the microbial census
+# The status of the microbial census: an update
 
-Patrick D. Schloss1*, Rene Girard2, Thomas Martin2, Joshua Edwards2, J. Cameron Thrash2*
+Patrick D. Schloss<sup>1*</sup>, Rene Girard<sup>2</sup>, Thomas Martin<sup>2</sup>, Joshua Edwards<sup>2</sup>, J. Cameron Thrash<sup>2*</sup>
 
@@
 
-*Correspondence: pschloss@umich.edu, thrashc@lsu.edu
+\* Correspondence: pschloss@umich.edu, thrashc@lsu.edu
+
+\newpage
 
 ## Abstract
+
+Abstract goes here.
 
@@
 At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process immensely.
-Furthermore, creating standards for delimiting what makes one microbe “different” from another has proved contentious at best.
+Furthermore, creating standards for delimiting what makes one microbe "different” from another has proved contentious at best.
 In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, and metagenomics.
+
 Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
 Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science.
-While reliance on this gene subjects us to biases created by primer selection*, differences in amplification strength* and fidelity*, internal features which may disrupt traditional measurements*, and potentially misleading classification due to infrequent horizontal gene transfer*, the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
+While reliance on this gene subjects us to biases created by primer selection [REFS], differences in amplification strength [REFS] and fidelity [REFS], internal features which may disrupt traditional measurements [REFS], and potentially misleading classification due to infrequent horizontal gene transfer [REFS], the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
 Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
-	In 2004, the Status of the Microbial Census analyzed 56,215 partial rRNA gene sequences available in the Ribosomal Database Project and concluded that “either current sampling methods are not adequate to identify 107 to 109 different species or these estimates are high.” (Schloss, 2004) Since that time, the explosion of short-read high throughput sequencing technology has caused a dramatic increase in the number of sequences deposited in public databases.
+
+In 2004, Schloss and Handelsman analyzed 56,215 partial rRNA gene sequences available in the Ribosomal Database Project and concluded that "either current sampling methods are not adequate to identify 10<sup>7</sup> to 10<sup>9</sup> different species or these estimates are high (Schloss, 2004)." Since that time, the explosion of short-read high throughput sequencing technology has caused a dramatic increase in the number of sequences deposited in public databases.
 Currently there are over XXXXXX 16S rRNA gene sequences in SRA.
-Recent estimates based on this short read data place the total number of microbial species at around XXXXX.**
+Recent estimates based on this short read data place the total number of microbial species at around XXXXX.
+
 However, the increase in sequencing volume has come at the cost of sequence length.
-The commonly used iTag sequencing technology amplifies an approximately 250 bp region of the 16S rRNA gene at the V4 hypervariable region, and other schemes have used different parts of the gene.
+The commonly used MiSeq-based sequencing technology amplifies the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are shorter than 500 bp.
 The decrease in length means a decrease in informative sites, making direct comparisons between short read sequencing and mostly or completely full-length sequences difficult.
@@
 Thus, it remains unclear to what degree richness estimates from short read technology over or underestimate the numbers from full-length sequences.
+
 Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences.
@@
 
+* Lack of references
+
+Issues we sought to solve...
+* How have the number of full-length sequences changed over the past N years?
+* How does this vary by environment?
+* How has the coverage of the OTUs represented by these sequences changed over the last N years?
+* How has the number of sequences per study changed over the past N years?
+* How has Figure 1 changed over the past N years?
+
 ## Materials and Methods
 
-Sequence data
-Primary sequences were downloaded from the ARB-SILVA database on XXXX, 2015 with the requirement that sequences were ≥ XXXX bp.
-This resulted in XXXXXXXX unique sequences.
-Additional sequence data was included from single-cell genomes available on IMG, many of which were recently obtained via the GEBA-MDM effort in Rinke et al.
-(2014).
-“SCGC” was searched on the IMG database 12 Mar 2015, manually separated into bacterial and archaeal SSU genes, and associated metadata was also downloaded.
+***Sequence data curation.***
+The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@REF].
+This release is based on the EMBL-EBI/ENA Release 123, which was released in March 2015.
+The SILVA curators identify potential SSU sequences using keyword searches and sequence-based search using RNAmmer (http://www.arb-silva.de/documentation/release-123/).
+The SILVA curators then screened the 7,168,241 resulting sequences based on a minimum length criteria (<300 nt), number of ambiguous base calls (>2%), length of sequence homopolymers (>2%), presence of vector contamination (>2%), low alignment quality value (<75), and likelihood of being chimeric (Pintail value < 50).
+Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nt and the archaeal reference set retained those archaeal sequences longer than 900 nt.
+The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata.
+Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], many of which were recently obtained via the GEBA-MDM effort in Rinke et al.
+[@REF].
+"SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata.
+The IMG sequences were aligned against the respective SILVA-based reference using mothur [@REF].
+The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur.
+Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates while maintaining a minimum sequence length of 1,200 and 900 nt for the bacterial and archeael datasets, respectively.
+The final datasets contained 1,412,682 bacterial and 53,619 archaeal 16S rRNA gene sequences.
+Sequences were assigned to operational taxonomic units (OTUs) with a 3% distance threshold using the average neighbor clustering algorithm.
 
-Classification of organisms based on source metadata
-Organisms were classified based on seven broad “coarse” categories, and refined further to facilitate additional analyses with twenty-six more specific “fine” categories (Table 1).
-These were assigned based on manual curation of the metadata associated with sequences obtained from the ARB-SILVA and IMG databases.
-The original metadata table was edited through sorting the “Source” category alphabetically and filtering for unique terms.
-The resulting XXXXXX terms were then manually assigned to the coarse and fine categories in Table 1.
-For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse “Other” category.
+***Metadata curation.***
+The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse” categories, and further refined to facilitate additional analyses with twenty-six more specific "fine” categories (Table 1).
+These were assigned based on manual curation of the `isolation_source` category within the ARB database associated with each of the sequences.
+There were 20,084 bacterial and 1,613 archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1.
+For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
+There were 150,378 bacterial and 2,604 archaeal sequences where an `isolation_source` term was not collected.
+Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare???].
 
-OTU clustering with Mothur
-
-Plots
-Rarefaction curves were completed in R using modifications to the script available in Supplementary Information.
+***Data analysis.***
+Our analysis made use of ARB [OS X v.6.0, @Pruesse2007], mothur mothur [v.1.37.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+Within R we utilized knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic] and openxlsx [v.
+`r packageDescription("openxlsx")$Version`; @Walker2015] packages.
+A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_XXXX_2015.
 
@@
 
-Caveats
+***Caveats.***
 Recent data suggests that a considerable diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers.
 This dataset does not include sequences from metagenomic assemblies but Brown et al.
-(2015) have used such assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called “Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
+(2015) have used such assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
 Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015).
 Thus, for many environments, the estimates within must be considered as lower bounds.
-Table 1. Habitat classifications
+
+**Table 1. Habitat classifications**

### 2015-11-09 Schloss_Census2_mBio_2015.Rmd
Integrate LaTeX stuff

--- 2015-11-09 Schloss_Census2_XXXX_2015.Rmd
@@
 ---
-title: "The status of the microbial census: an update"
+title: "**The status of the microbial census: an update**"
+bibliography: references.bib
 output:
-  md_document:
-    variant: markdown_github
+  pdf_document:
+    includes:
+      in_header: header.tex
+csl: mbio.csl
+fontsize: 11pt
+geometry: margin=1.0in
 ---
 
-# The status of the microbial census: an update
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = FALSE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("cache" = FALSE)
 
-Patrick D. Schloss<sup>1*</sup>, Rene Girard<sup>2</sup>, Thomas Martin<sup>2</sup>, Joshua Edwards<sup>2</sup>, J. Cameron Thrash<sup>2*</sup>
+inline_hook <- function(x){
+	print(x)
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
+		}
 
-1.	Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109, U.S.A.
-2.	Department of Biological Sciences, Louisiana State University, Baton Rouge, LA 70803, U.S.A.
+	} else {
+    	paste(x)      
+	}
+}
+knitr::knit_hooks$set(inline=inline_hook)
+```
+\vspace{35mm}
 
-\* Correspondence: pschloss@umich.edu, thrashc@lsu.edu
+Patrick D. Schloss^1$\dagger$^, Rene Girard^2^, Thomas Martin^2^, Joshua Edwards^2^, and J. Cameron Thrash^2$\dagger$^
+
+\vspace{40mm}
+
+$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu and thrashc@lsu.edu
+
+1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
+
+2\. Department of Biological Sciences, Louisiana State University, Baton Rouge, LA 70803
 
 \newpage
+\linenumbers
 
@@
 Abstract goes here.
+
+## Importance
+
+\newpage
 

### 2015-11-10 Schloss_Census2_mBio_2015.Rmd
Fill in Introduction

@@
 
+\newpage
+
 ## Importance
+
+Importance goes here.
 
@@
 ## Introduction
-
-The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology.
-At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process immensely.
-Furthermore, creating standards for delimiting what makes one microbe "different” from another has proved contentious at best.
+In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was deposited into NCBI's GenBank making it the first of more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978].
+GenBank accessions represent nearly one-third of all sequences deposited in the database making it the best-represented gene.
+As Sanger sequencing has given way to to-called "next generation sequencing" technologies, hundreds of millions of 16S rRNA gene sequences have been deposited into the NCBI's Sequence Read Archive.
+The expansion in sequencing throughput and increased access to sequencing technology has allowed for more environments to be sequenced at a deeper coverage resulting in the identification of novel taxa.
+The ability to obtain sequence data from microorganisms without cultivation has radically altered our perspective of their role in nearly every environment from deep ocean sediment cores (e.g.
+accession AY436526) to the International Space Station (e.g.
+accession DQ497748).
+The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@reference].
+At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process.
+Furthermore, creating standards for delimiting what makes one microbe "different” from another has been contentious [@reference].
 In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, and metagenomics.
 
-Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
-Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science.
-While reliance on this gene subjects us to biases created by primer selection [REFS], differences in amplification strength [REFS] and fidelity [REFS], internal features which may disrupt traditional measurements [REFS], and potentially misleading classification due to infrequent horizontal gene transfer [REFS], the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
-Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
+The increase in sequencing volume has come at the cost of sequence length.
+The commonly used MiSeq-based sequencing platform from Illumina is widely used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
+Perhaps most disconcerting about this development is the sense that the increased read depth is being gained using short read platforms rather than the full-length sequences.
+Because these short reads are used for classification to existing taxa, we are missing the opportunity to propose novel candidate taxa and vastly underappreciating the biodiversity of microbial life.
+We likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
 
-In 2004, Schloss and Handelsman analyzed 56,215 partial rRNA gene sequences available in the Ribosomal Database Project and concluded that "either current sampling methods are not adequate to identify 10<sup>7</sup> to 10<sup>9</sup> different species or these estimates are high (Schloss, 2004)." Since that time, the explosion of short-read high throughput sequencing technology has caused a dramatic increase in the number of sequences deposited in public databases.
-Currently there are over XXXXXX 16S rRNA gene sequences in SRA.
-Recent estimates based on this short read data place the total number of microbial species at around XXXXX.
-
-However, the increase in sequencing volume has come at the cost of sequence length.
-The commonly used MiSeq-based sequencing technology amplifies the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are shorter than 500 bp.
-The decrease in length means a decrease in informative sites, making direct comparisons between short read sequencing and mostly or completely full-length sequences difficult.
-Youssef et al.
-(2009) demonstrated high variability between the number of OTUs that different hypervariable sites estimated compared to full-length 16S rRNA gene sequences, and also that sample location influenced this behavior.
-Thus, it remains unclear to what degree richness estimates from short read technology over or underestimate the numbers from full-length sequences.
+Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998, @Curtis2002].
+That census called for a broader and deeper characterization of all environments.
+Refreshingly, this challenge was largely met.
+There have been major investments in studying the Earth's microbiome using 16S rRNA gene sequencing through initiatives such as the Human Microbiome Project [@reference], the Earth Microbiome Project [@reference], and the International Census of Marine Microorganisms.
+But most important, the original census was performed on the cusp of radical developments in sequencing technologies.
+That advancement has largely moved the bulk of sequencing throughput from large sequencing centers to individual investigators and leveraged their diversity to expand the representation of organisms and environments represented in public databases.
 
 Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences.
-In the 12 years since the collection of data for Schloss and Handelsman’s analysis, the number of available sequences has risen over twenty-fold, despite the overwhelming contemporary focus by most researchers on short-read technologies.
+In the 13 years since the collection of data for Schloss and Handelsman’s analysis, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies.
 This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study.
-
-* Lack of references
-
-Issues we sought to solve...
-* How have the number of full-length sequences changed over the past N years?
-* How does this vary by environment?
-* How has the coverage of the OTUs represented by these sequences changed over the last N years?
-* How has the number of sequences per study changed over the past N years?
-* How has Figure 1 changed over the past N years?
+Although there has been a robust growth in the number of full-length sequences deposited to GenBank since its creation in 1983, the rate of growth has stalled over the past 5 years and the deposits have been dominated by a handful of research groups studying a limited number of environments.
+As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.
 
@@
 There were 150,378 bacterial and 2,604 archaeal sequences where an `isolation_source` term was not collected.
-Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare???].
+Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].
 
@@
 `r packageDescription("openxlsx")$Version`; @Walker2015] packages.
-A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_XXXX_2015.
+A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_mBio_2015.
 
 ## Results and Discussion
+
+Issues we sought to solve...
+* How have the number of full-length sequences changed over the past N years?
+* How does this vary by environment?
+* How has the coverage of the OTUs represented by these sequences changed over the last N years?
+* How has the number of sequences per study changed over the past N years?
+* How has Figure 1 changed over the past N years?
 
@@
 
+Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
+Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science.
+While reliance on this gene subjects us to biases created by primer selection [REFS], differences in amplification strength [REFS] and fidelity [REFS], internal features which may disrupt traditional measurements [REFS], and potentially misleading classification due to infrequent horizontal gene transfer [REFS], the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
+Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
+
+The first 16S sequence was published in 1978, not deposited until 1983. A bit of an allegory for our time.
+
 **Table 1. Habitat classifications**

### 2015-11-10 Schloss_Census2_mBio_2015.Rmd
Trying to get version numbers in citation parentheses

@@
 ***Data analysis.***
-Our analysis made use of ARB [OS X v.6.0, @Pruesse2007], mothur mothur [v.1.37.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
-Within R we utilized knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic] and openxlsx [v.
-`r packageDescription("openxlsx")$Version`; @Walker2015] packages.
+Our analysis made use of ARB (OS X v.6.0) [@Pruesse2007], mothur (v.1.37.0) [@Schloss2009], and R (v.`r paste(version$major, version$minor, sep='.')`) [@language2015].
+Within R we utilized knitr (v.`r packageDescription("knitr")$Version`) [@xie2013dynamic] and openxlsx (v.
+`r packageDescription("openxlsx")$Version`) [@Walker2015] packages.
 A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_mBio_2015.
@@
 Issues we sought to solve...
+
 * How have the number of full-length sequences changed over the past N years?

### 2015-11-10 scratch/category_parsing.R
Remove parsing script that wasn't used

@@ -1,81 +1,0 @@
-# I copied the first column of the spreadsheet into a file called `source_categories.txt`
-# First, I went through and counted the number of times each word was used and ranked
-# them in decreasing order of abundance
-
-original_words <- tolower(scan(file="source_categories.txt", quiet=T, what=""))
-original_words_freq <- table(original_words)
-head(sort(original_words_freq, decreasing=T), n=100)
-
-# From this we can see a lot of useless words like of, from, etc. But also there are
-# categories that have useful words like soil, water, sediment, etc. Let's read the
-# categories back in but doing it one line at a time this time instead of one word at
-# a time:
-
-categories <- tolower(scan(file="source_categories.txt", quiet=T, what="", sep="\n"))
-
-# let's look at some of the categories that contain words like "soil"
-
-categories[grepl("soil", categories)]
-
-# From this we see categories like: "0-20 cm bulk soil, duennwald forest". What I would
-# propose is using the grepl R function to quickly sift through these categories that 
-# contain high frequency words.
-#
-# go through and reassign any categories that contain high frequency words to 
-# the categories categories names. here are a few that I used to get going...
-
-categories[grepl("soil", categories, ignore.case=T)] <- "soil"
-categories[grepl("rhizosphere", categories, ignore.case=T)] <- "soil"
-
-categories[grepl("sediment", categories, ignore.case=T)] <- "sediment"
-
-categories[grepl("wastewater", categories, ignore.case=T)] <- "wastewater"
-categories[grepl("sludge", categories, ignore.case=T)] <- "wastewater"
-categories[grepl("sludge", categories, ignore.case=T)] <- "wastewater"
-
-categories[grepl("sea", categories, ignore.case=T)] <- "marine"
-categories[grepl("marine", categories, ignore.case=T)] <- "marine"
-
-categories[grepl("culture", categories, ignore.case=T)] <- "laboratory"
-categories[grepl("reactor", categories, ignore.case=T)] <- "laboratory"
-categories[grepl("anaerobic", categories, ignore.case=T)] <- "laboratory"
-
-categories[grepl("lake", categories, ignore.case=T)] <- "freshwater"
-categories[grepl("river", categories, ignore.case=T)] <- "freshwater"
-
-categories[grepl("feces", categories, ignore.case=T)] <- "host-associated"
-categories[grepl("patient", categories, ignore.case=T)] <- "host-associated"
[…]

### 2015-11-10 Schloss_Census2_mBio_2015.Rmd
Build Table 1 and other numbers into Rmd file

@@
 knitr::knit_hooks$set(inline=inline_hook)
+
+year_of_analysis <- "2015"
+
+bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
+
+arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
 ```
@@
 Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates while maintaining a minimum sequence length of 1,200 and 900 nt for the bacterial and archeael datasets, respectively.
-The final datasets contained 1,412,682 bacterial and 53,619 archaeal 16S rRNA gene sequences.
+The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences.
 Sequences were assigned to operational taxonomic units (OTUs) with a 3% distance threshold using the average neighbor clustering algorithm.
@@
 These were assigned based on manual curation of the `isolation_source` category within the ARB database associated with each of the sequences.
-There were 20,084 bacterial and 1,613 archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1.
+There were `r length(unique(bact$isolation_source))-1` bacterial and `r length(unique(arch$isolation_source))-1` archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1.
 For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
-There were 150,378 bacterial and 2,604 archaeal sequences where an `isolation_source` term was not collected.
+There were `r sum(is.na(bact$isolation_source))` bacterial and `r sum(is.na(arch$isolation_source))` archaeal sequences where an `isolation_source` term was not collected.
 Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].
@@
 
+## Conclusions
+
+Future for PacBio in generating full-length sequences
+
 The first 16S sequence was published in 1978, not deposited until 1983. A bit of an allegory for our time.
 
+\newpage
+\singlespacing
+
 **Table 1. Habitat classifications**
+
+```{r table1}
+
+coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Zoological", "", "", "", "Other", "No isolation_source", "Total")
+
+fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Hydrothermal vent", "Ice", "Marine", "Marine sediment", "Other", "Digesters", "Food-associated", "Industrial/mining", "Other", "Pollution associated", "Other", "Root", "Surface", "Agriculture", "Desert", "Other", "Permafrost", "Arthropod", "Non-arthropod invertebrate", "Other", "Vertebrate", "", "", "")
+
+abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "BD", "BF", "BI", "BO", "BP", "PO", "PR", "PS", "SA", "SD", "SO", "SP", "ZA", "ZN", "ZO", "ZV", "OT")
+
+bact_count <- table(bact$category)[abbreviation]
+bact_count[is.na(bact_count)] <- 0
+bact_count <- c(bact_count, nrow(bact) - sum(bact_count))
+bact_count <- c(bact_count, nrow(bact))
+
+arch_count <- table(arch$category)[abbreviation]
+arch_count[is.na(arch_count)] <- 0
+arch_count <- c(arch_count, nrow(arch) - sum(arch_count))
+arch_count <- c(arch_count, nrow(arch))
+
+abbreviation <- c(abbreviation, "", "")
+composite <- cbind(coarse, fine, abbreviation, bact_count, arch_count)
+
+kable(composite, row.names=FALSE, col.names=c("Coarse", "Fine", "Abbreviation", "Bacterial", "Archaeal"), align=c('l', 'l', 'c', 'c', 'c'), format='latex')
+
+```
+
+\newpage
+\doublespacing
+
+## References

### 2015-11-10 code/time_course_plots.R

@@ -1,9 +1,9 @@
 year_of_analysis <- "2015"
 
-bact <- read.table(file='data/mothur/bacteria.final.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
 
-arch <- read.table(file='data/mothur/archaea.final.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
 
 

### 2015-11-12 code/Arc_3_percent_Uniq.R
Added R code from Rene

@@ -1,0 +1,45 @@
+data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_archaea.env_category.groups.rarefaction", header=T)
+category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
+plant_columns <- grep(category, colnames(data))
+category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
+aerosol_columns <- grep(category, colnames(data))
+category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
+aquatic_columns <- grep(category, colnames(data))
+category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
+built_columns <- grep(category, colnames(data))
+category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
+other_columns <- grep(category, colnames(data))
+category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
+soil_columns <- grep(category, colnames(data))
+category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
+zoonotic_columns <- grep(category, colnames(data))
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
+nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
+pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_3per.pdf")
+
+par(mar=c(5,5,0.5,0.5))
+plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
+     xlim=c(0,max(data$numsampled)),
+     xlab="Number of sequences sampled",
+     ylab="",
+     axes=FALSE)
+
[…]

### 2015-11-12 code/Bac_5_percent_uniq.R

@@ -1,0 +1,46 @@
+data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
+category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
+plant_columns <- grep(category, colnames(data))
+category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
+aerosol_columns <- grep(category, colnames(data))
+category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
+aquatic_columns <- grep(category, colnames(data))
+category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
+built_columns <- grep(category, colnames(data))
+category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
+other_columns <- grep(category, colnames(data))
+category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
+soil_columns <- grep(category, colnames(data))
+category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
+zoonotic_columns <- grep(category, colnames(data))
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
+nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
+pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_5per.pdf")
+
+
+par(mar=c(5,5,0.5,0.5))
+plot(NA, type="n", ylim=c(0,max(30000)),
+     xlim=c(0,max(data$numsampled)),
+     xlab="Number of sequences sampled",
+     ylab="",
+     axes=FALSE)
+
+points(data$X0.05.plant~data$numsampled, type="l", col="red", lwd=2)
[…]

### 2015-11-13 Schloss_Census2_mBio_2015.Rmd

@@
 As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.
+In this, continued technological advances, such as the reconstruction of nearly full-length sequences from metagenomics with EMIRGE (Miller 2011) and inovations in culturing (references needed: iChip, Epstein dormancy, Thrash 2015; Gifford 2014), will be vital.
 
@@
 
-* How have the number of full-length sequences changed over the past N years?
-* How does this vary by environment?
-* How has the coverage of the OTUs represented by these sequences changed over the last N years?
-* How has the number of sequences per study changed over the past N years?
+* Figure 1A. How have the number of full-length sequences changed over the past N years?
+* Figure 1B. How does this vary by environment?
+
+* Figure 2. How has the number of sequences per study changed over the past N years?
+
+* Figure 3. How has the coverage of the OTUs represented by these sequences changed over the last N years?
+
 * How has Figure 1 changed over the past N years?
+
+* In what environments do we have the best handle on extant diversity?
+  Figure 4A,B. What are the differences in sampling effort/sequence coverage at the 97% level for the different coarse habitats, and how is this different between Bacteria and Archaea?
+  Figure 5A,B. For the most sampled coarse habitats (Zoonotic for Bacteria, Aquatic for Archaea), what are the fine-scale habitats where they have been sampled most heavily?
+
+* In what habitats do we have the poorest understanding of diversity (and thus expect to yield considerable new knowledge over the years)?
+* How is this different for Bacteria vs. Archaea?
+* What are the most significant improvements since 2004, and where are we still lacking the most data?
+* For which phyla has there been substantial increases in sampled diversity since 2004?
+* Can we make any recommendations for habitats where biodiversity is under-sampled, perhaps calls for conservation in certain environments based on how poorly we understand the microbial diversity?
 
@@
 The first 16S sequence was published in 1978, not deposited until 1983. A bit of an allegory for our time.
+
+Impact of EMIRGE
+
+Renewed call for cultivation
 
@@
 
+**Figure 1. Sequences deposited by year**
+
+Pat, this is a placeholder for you
+
+**Figure 2. Sequence deposition by study**
+
+Pat, ibid.
+
+**Figure 3. Sampling by OTU cutoff**
+
+I only have an old code from Rene for this, and I don't think it reports all the sequences
+
+**Figure 4A. Bacterial sampling by environment**
+
+```{r figure4a}
+data <- read.table(file="data/mothur/all_bacteria.env_category.groups.rarefaction", header=T)
+category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
+plant_columns <- grep(category, colnames(data))
+category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
+aerosol_columns <- grep(category, colnames(data))
+category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
+aquatic_columns <- grep(category, colnames(data))
+category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
+built_columns <- grep(category, colnames(data))
+category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
+other_columns <- grep(category, colnames(data))
+category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
+soil_columns <- grep(category, colnames(data))
+category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
+zoonotic_columns <- grep(category, colnames(data))
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
+nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
+pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per.pdf")
+
+par(mar=c(5,5,0.5,0.5))
+plot(NA, type="n", ylim=c(0,max(40000)),
+     xlim=c(0,max(data$numsampled)),
+     xlab="Number of sequences sampled",
+     ylab="",
+     axes=FALSE)
+
+points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
+points(data$X0.03.aerosol~data$numsampled, type="l", col="green", lwd=2)
+points(data$X0.03.aquatic~data$numsampled, type="l", col="blue", lwd=2)
+points(data$X0.03.built~data$numsampled, type="l", col="orange", lwd=2)
+points(data$X0.03.other~data$numsampled, type="l", col="black", lwd=2)
+points(data$X0.03.soil~data$numsampled, type="l", col="purple", lwd=2)
+points(data$X0.03.zoonotic~data$numsampled, type="l", col="deeppink", lwd=2)
+
+legend(x=5e5, y=20000, lwd=2, lty=1,
+       legend=c("Other", "Plant", "Aerosol", "Aquatic", "Built", "Soil", "Zoonotic"),
+       col=c("black", "red", "green", "blue", "orange", "purple", "deeppink"))
+
+axis(1, at=seq(0,700000,1e5), label=c("0", "100,000", "200,000", "300,000", "400,000", "500,000", "600,000", "700,000"), cex.axis=0.8)
+axis(2, at=seq(0,40000,5000), label=c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "35,000", "40,000"), las=2, cex.axis=0.8)
+box()
+mtext(side=2, "Number of OTUs", line=4)
+
+dev.off()
+
+```
+
+**Figure 4B. Archaeal sampling by environment**
+
+```{r figure4b}
+
+data <- read.table(file="data/mothur/all_archaea.env_category.groups.rarefaction", header=T)
+category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
+plant_columns <- grep(category, colnames(data))
+category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
+aerosol_columns <- grep(category, colnames(data))
+category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
+aquatic_columns <- grep(category, colnames(data))
+category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
+built_columns <- grep(category, colnames(data))
+category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
+other_columns <- grep(category, colnames(data))
+category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
+soil_columns <- grep(category, colnames(data))
+category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
+zoonotic_columns <- grep(category, colnames(data))
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
+nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
+plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
+pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_3per.pdf")
+
+par(mar=c(5,5,0.5,0.5))
+plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
+     xlim=c(0,max(data$numsampled)),
+     xlab="Number of sequences sampled",
+     ylab="",
+     axes=FALSE)
+
+points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
+points(data$X0.03.aerosol~data$numsampled, type="l", col="green", lwd=2)
+points(data$X0.03.aquatic~data$numsampled, type="l", col="blue", lwd=2)
+points(data$X0.03.built~data$numsampled, type="l", col="orange", lwd=2)
+points(data$X0.03.other~data$numsampled, type="l", col="black", lwd=2)
+points(data$X0.03.soil~data$numsampled, type="l", col="purple", lwd=2)
+points(data$X0.03.zoonotic~data$numsampled, type="l", col="deeppink", lwd=2)
+
+legend(x=15000, y=1550, lwd=2, lty=1,
+       legend=c("Other", "Plant", "Aerosol", "Aquatic", "Built", "Soil", "Zoonotic"),
+       col=c("black", "red", "green", "blue", "orange", "purple", "deeppink"))
+
+axis(1, at=seq(0,35000,5000), label=c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "35,000"), cex.axis=0.8)
+axis(2, at=seq(0,5250,750), label=c("0", "750", "1500", "2250", "3,000", "3750", "4500", "5250"), las=2, cex.axis=0.8)
+box()
+mtext(side=2, "Number of OTUs", line=4)
+
+dev.off()
+```
+
 \newpage

### 2015-11-17 Schloss_Census2_mBio_2015.Rmd
Move Table 1 from paper Rmd to separate file add data

@@
 
-**Table 1. Habitat classifications**
-
-```{r table1}
-
-coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Zoological", "", "", "", "Other", "No isolation_source", "Total")
-
-fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Hydrothermal vent", "Ice", "Marine", "Marine sediment", "Other", "Digesters", "Food-associated", "Industrial/mining", "Other", "Pollution associated", "Other", "Root", "Surface", "Agriculture", "Desert", "Other", "Permafrost", "Arthropod", "Non-arthropod invertebrate", "Other", "Vertebrate", "", "", "")
-
-abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "BD", "BF", "BI", "BO", "BP", "PO", "PR", "PS", "SA", "SD", "SO", "SP", "ZA", "ZN", "ZO", "ZV", "OT")
-
-bact_count <- table(bact$category)[abbreviation]
-bact_count[is.na(bact_count)] <- 0
-bact_count <- c(bact_count, nrow(bact) - sum(bact_count))
-bact_count <- c(bact_count, nrow(bact))
-
-arch_count <- table(arch$category)[abbreviation]
-arch_count[is.na(arch_count)] <- 0
-arch_count <- c(arch_count, nrow(arch) - sum(arch_count))
-arch_count <- c(arch_count, nrow(arch))
-
-abbreviation <- c(abbreviation, "", "")
-composite <- cbind(coarse, fine, abbreviation, bact_count, arch_count)
-
-kable(composite, row.names=FALSE, col.names=c("Coarse", "Fine", "Abbreviation", "Bacterial", "Archaeal"), align=c('l', 'l', 'c', 'c', 'c'), format='latex')
-
-```
-
 **Figure 1. Sequences deposited by year**

### 2015-11-17 results/tables/build_table_1.Rmd
Move Table 1 from paper Rmd to separate file add data

---
output:
  pdf_document:
    includes:
      in_header: table_1_header.tex
tables: true
fontsize: 8pt
geometry: margin=0.8in
---

**Table 1. Habitat classifications**

```{r table1, results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)

bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)

arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)

coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Zoological", "", "", "", "Other", "No source data", "Total")

fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Hydrothermal vent", "Ice", "Marine", "Marine sediment", "Other", "Digesters", "Food-associated", "Industrial/mining", "Other", "Pollution associated", "Other", "Root", "Surface", "Agriculture", "Desert", "Other", "Permafrost", "Arthropod", "Other invertebrate", "Other", "Vertebrate", "", "", "")

abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "BD", "BF", "BI", "BP", "BO", "PR", "PS", "PO", "SA", "SD", "SP", "SO", "ZA", "ZN", "ZV", "ZO", "OT")

get_sobs <- function(x, db){
	if(x == "NA"){
		length(table(db[is.na(db$category), "otu"]))
	} else if(x == "total"){
		length(table(db[, "otu"]))
	} else {
		length(table(db[db$category==x, "otu"]))
	}
}

get_coverage <- function(x, db){
	if(x=="NA"){
		x_table <- table(db[is.na(db$category), "otu"])
	} else if(x == "total") {
		x_table <- table(db[, "otu"])
	} else {
		x_table <- table(db[db$category==x, "otu"])
	}

	singletons <- sum(x_table == 1)
	total_seqs <- sum(x_table)

	if(total_seqs != 0){
		100*(1-singletons/total_seqs)
	} else {
		NA
	}

}

bact_count <- table(bact$category)[abbreviation]
bact_count[is.na(bact_count)] <- 0
bact_count <- c(bact_count, nrow(bact) - sum(bact_count))
bact_count <- format(c(bact_count, nrow(bact)), big.mark=',')
bact_sobs <- format(sapply(c(abbreviation, "NA", "total"), get_sobs, bact), big.mark=',')
bact_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_coverage, bact), nsmall=1, digits=3)

arch_count <- table(arch$category)[abbreviation]
arch_count[is.na(arch_count)] <- 0
arch_count <- c(arch_count, nrow(arch) - sum(arch_count))
arch_count <- format(c(arch_count, nrow(arch)), big.mark=',')
arch_sobs <- format(sapply(c(abbreviation, "NA", "total"), get_sobs, arch), big.mark=',')
arch_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_coverage, arch), nsmall=1, digits=3)

#abbreviation <- c(abbreviation, "", "")
composite <- cbind(coarse, fine, bact_count, bact_sobs, bact_coverage, arch_count, arch_sobs, arch_coverage)
composite <- rbind(c("", "", "Sequences", "OTUs", "Coverage (%)", "Sequences", "OTUs", "Coverage (%)"), composite)

composite <- xtable(composite)

addtorow <- list(pos=list(0, 1),
				command=c('\\multirow{2}{*}{Coarse} & \\multirow{2}{*}{Fine} & \\multicolumn{3}{c}{Bacterial} & \\multicolumn{3}{c}{Archaeal} \\\\ \\cmidrule(r){3-5}	\\cmidrule(r){6-8}', '\\hline '))

align(composite) <- "lllcccccc"
print(composite, add.to.row=addtorow, include.rownames=FALSE, include.colnames=FALSE, hline.after = c(-1,nrow(composite)-1, nrow(composite)), comment=FALSE, size="small")
```

### 2015-11-17 code/domain_rarefaction.R
Generate plot for domain-level rarefaction

@@ -1,0 +1,53 @@
+my_col <- c(unique=rainbow(5)[1], X0.03='orange', X0.05=rainbow(5)[3], X0.10=rainbow(5)[4], X0.20=rainbow(5)[5] )
+my_pch <- c(unique=16, X0.03=16, X0.05=17, X0.10=21, X0.20=22)
+
+plot_curves <- function(rarefaction){
+	nseqs <- rarefaction$numsampled
+	rarefaction <- rarefaction[,-1]
+	rarefaction <- rarefaction[,!grepl("ci", colnames(rarefaction))]
+
+	xlim <- c(1,max(nseqs))
+	ylim <- c(1, 1.05* max(rarefaction$unique))
+
+	plot(NA, xlim=xlim, ylim=ylim, type='n', xlab="", ylab="", main="", axes=FALSE)
+	for(cutoff in colnames(rarefaction)){
+		points(rarefaction[,cutoff]~nseqs, type="l", lwd=2, col=my_col[cutoff])
+
+		spacing <- floor(seq(1,length(nseqs),length.out=10))[-1]
+		points(rarefaction[spacing,cutoff]~nseqs[spacing], col=my_col[cutoff], pch=my_pch[cutoff], bg="white", cex=1.25)
+	}
+	box()
+
+}
+
+bact <- read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T)
+arch <- read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T)
+
+pdf('results/figures/domain_rarefaction.pdf', width=7.5, height=3.5)
+par(mar=c(2,4,0.5,0.5))
+
+layout(matrix(c(3,1,2,0,4,4), nrow=2, byrow=T), widths=c(0.1,1,1), heights=c(1,0.10))
+
+cex_size <- 0.9
+
+plot_curves(bact)
+axis(1, at=seq(0,1.4e6,2e5), label=c(0, expression('2' %*% 10^5), expression(4 %*% 10^5), expression(6 %*% 10^5), expression(8 %*% 10^5), expression(1 %*% 10^6), expression(1.2 %*% 10^6), expression(1.4 %*% 10^6)), cex.axis=cex_size)
[…]

### 2015-11-18 Schloss_Census2_mBio_2015.Rmd
Add some ideas to Results section outline

@@
 ## Results and Discussion
+
+* What does the ongonig census look like?
+  - flat
+* Why might it be flat?
+  - Good's coverage
+  - Schloss's coverage
+  - % of NGS reads that map to database
+* Why aren't we doig a great job?
+  - Rate of change in full-length sequences has plateaued
+  - New sequences are submitted by a small number of investigators for a limited number of environments
+* What can we do to fix this?
+  - Characterize coverage by environments
+  - Cultured vs. uncultured
+  - Single cell genomics
 

### 2015-11-18 results/tables/build_table_1.Rmd
Fix label ordering; add OTU coverage

@@
 
-fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Hydrothermal vent", "Ice", "Marine", "Marine sediment", "Other", "Digesters", "Food-associated", "Industrial/mining", "Other", "Pollution associated", "Other", "Root", "Surface", "Agriculture", "Desert", "Other", "Permafrost", "Arthropod", "Other invertebrate", "Other", "Vertebrate", "", "", "")
+fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Hydrothermal vent", "Ice", "Marine", "Marine sediment", "Other", "Digesters", "Food-associated", "Industrial/mining", "Pollution associated", "Other", "Root", "Surface", "Other", "Agriculture", "Desert", "Permafrost", "Other", "Arthropod", "Other invertebrate", "Vertebrate", "Other", "", "", "")
 
@@
 
-get_coverage <- function(x, db){
+get_goods_coverage <- function(x, db){
 	if(x=="NA"){
@@
 
+get_schloss_coverage <- function(x, db){
+	if(x=="NA"){
+		x_table <- table(db[is.na(db$category), "otu"])
+	} else if(x == "total") {
+		x_table <- table(db[, "otu"])
+	} else {
+		x_table <- table(db[db$category==x, "otu"])
+	}
+
+	singletons <- sum(x_table == 1)
+	total_otus <- length(x_table)
+
+	if(total_otus != 0){
+		100*(1-singletons/total_otus)
+	} else {
+		NA
+	}
+
+}
+
 bact_count <- table(bact$category)[abbreviation]
@@
 bact_sobs <- format(sapply(c(abbreviation, "NA", "total"), get_sobs, bact), big.mark=',')
-bact_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_coverage, bact), nsmall=1, digits=3)
+bact_good_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_goods_coverage, bact), nsmall=1, digits=3)
+bact_schloss_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_schloss_coverage, bact), nsmall=1, digits=3)
 
@@
 arch_sobs <- format(sapply(c(abbreviation, "NA", "total"), get_sobs, arch), big.mark=',')
-arch_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_coverage, arch), nsmall=1, digits=3)
+arch_good_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_goods_coverage, arch), nsmall=1, digits=3)
+arch_schloss_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_schloss_coverage, arch), nsmall=1, digits=3)
 
 #abbreviation <- c(abbreviation, "", "")
-composite <- cbind(coarse, fine, bact_count, bact_sobs, bact_coverage, arch_count, arch_sobs, arch_coverage)
-composite <- rbind(c("", "", "Sequences", "OTUs", "Coverage (%)", "Sequences", "OTUs", "Coverage (%)"), composite)
+composite <- cbind(coarse, fine, bact_count, bact_sobs, bact_good_coverage, bact_schloss_coverage, arch_count, arch_sobs, arch_good_coverage, arch_schloss_coverage)
+composite <- rbind(c("", "", "Sequences", "OTUs", "Seq.", "OTU", "Sequences", "OTUs", "Seq.", "OTU"), composite)
 
@@
 addtorow <- list(pos=list(0, 1),
-				command=c('\\multirow{2}{*}{Coarse} & \\multirow{2}{*}{Fine} & \\multicolumn{3}{c}{Bacterial} & \\multicolumn{3}{c}{Archaeal} \\\\ \\cmidrule(r){3-5}	\\cmidrule(r){6-8}', '\\hline '))
+				command=c('\\multirow{2}{*}{Coarse} & \\multirow{2}{*}{Fine} & \\multicolumn{4}{c}{Bacterial} & \\multicolumn{4}{c}{Archaeal} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
 
-align(composite) <- "lllcccccc"
+align(composite) <- "lllcccccccc"
 print(composite, add.to.row=addtorow, include.rownames=FALSE, include.colnames=FALSE, hline.after = c(-1,nrow(composite)-1, nrow(composite)), comment=FALSE, size="small")

### 2015-11-18 results/tables/build_table_1.Rmd
Fix formatting and caption

@@
 tables: true
-fontsize: 8pt
-geometry: margin=0.8in
+geometry: margin=1.0in
 ---
 
-**Table 1. Habitat classifications**
+**Table 1.
+Status of microbial census by habitat classifications and domain.** The isolation_source field from the SILVA reference database was manually curated to assign bacterial and archaeal sequences coarse and fine scale habitat classifications.
+We calculated the number of sequences and OTUs observed and the percent coverage on a sequence or OTU basis for each classification and domain.
 
@@
 composite <- cbind(coarse, fine, bact_count, bact_sobs, bact_good_coverage, bact_schloss_coverage, arch_count, arch_sobs, arch_good_coverage, arch_schloss_coverage)
-composite <- rbind(c("", "", "Sequences", "OTUs", "Seq.", "OTU", "Sequences", "OTUs", "Seq.", "OTU"), composite)
+composite <- rbind(c("", "", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
+\\\\ Coverage}", "\\bigcell{c}{\\% OTU \\\\ Coverage}", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
+\\\\ Coverage}", "\\bigcell{c}{\\% OTU \\\\ Coverage}"), composite)
 
@@
 addtorow <- list(pos=list(0, 1),
-				command=c('\\multirow{2}{*}{Coarse} & \\multirow{2}{*}{Fine} & \\multicolumn{4}{c}{Bacterial} & \\multicolumn{4}{c}{Archaeal} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
+				command=c('\\multirow{4}{*}{Coarse} & \\multirow{4}{*}{Fine} & \\multicolumn{4}{c}{Bacterial} & \\multicolumn{4}{c}{Archaeal} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
 
 align(composite) <- "lllcccccccc"
-print(composite, add.to.row=addtorow, include.rownames=FALSE, include.colnames=FALSE, hline.after = c(-1,nrow(composite)-1, nrow(composite)), comment=FALSE, size="small")
+print(composite, add.to.row=addtorow, include.rownames=FALSE, include.colnames=FALSE, hline.after = c(-1,nrow(composite)-1, nrow(composite)), comment=FALSE, size="scriptsize")
 ```

### 2015-11-18 code/time_course_plots.R
Generate annual summary file

@@ -1,10 +1,16 @@
-year_of_analysis <- "2015"
+start_year <- 1983
+year_of_analysis <- 2015
 
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
+
+summary_table <- data.frame(matrix(rep(0, 32*8), ncol=8))
+rownames(summary_table) <- start_year:(year_of_analysis-1)
+colnames(summary_table) <- c('b_nseqs', 'b_sobs', 'b_n50', 'b_phyla',
+							'a_nseqs', 'a_sobs', 'a_n50', 'a_phyla')
 
 
 pdf(file="results/figures/sequences_deposited.pdf")
@@ -13,12 +19,16 @@ pdf(file="results/figures/sequences_deposited.pdf")
 	names(bact_deposits_by_year) <- names(deposits_table)
 	bact_deposits_by_year <- bact_deposits_by_year[!(names(bact_deposits_by_year) %in% year_of_analysis)]
 
+	summary_table[names(bact_deposits_by_year),'b_nseqs'] <- bact_deposits_by_year
+
 	plot(bact_deposits_by_year~as.numeric(names(bact_deposits_by_year)), log='y', type="l", col="blue", lwd=2, xlab="Year", ylab="Number of sequences per year", ylim=c(1,1e6), xlim=c(1980, 2015), axes=F)
 
 	deposits_table <- table(arch_year_deposited)
 	arch_deposits_by_year <- as.numeric(deposits_table)
 	names(arch_deposits_by_year) <- names(deposits_table)
 	arch_deposits_by_year <- arch_deposits_by_year[!(names(arch_deposits_by_year) %in% year_of_analysis)]
+
+	summary_table[names(arch_deposits_by_year),'a_nseqs'] <- arch_deposits_by_year
 
 	points(arch_deposits_by_year~as.numeric(names(arch_deposits_by_year)), type="l", col="red", lwd=2)
 	axis(1)
[…]

### 2015-11-18 Schloss_Census2_mBio_2015.Rmd
Start filling in Results section

--- 2015-11-17 Schloss_Census2_mBio_2015.Rmd
@@
 		} else {
-			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
+			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
 		}
@@
 
-year_of_analysis <- "2015"
-
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
 ```
@@
 The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences.
-Sequences were assigned to operational taxonomic units (OTUs) with a 3% distance threshold using the average neighbor clustering algorithm.
+Sequences were assigned to OTUs using the average neighbor clustering algorithm [@REF].
 
@@
 
-Issues we sought to solve...
+```{r rarefaction}
+get_sobs <- function(data){
+	rownames(data) <- data[,1]
+	data <- data[,-1]
+	data <- data[,!grepl('ci', colnames(data))]
+	unlist(data[nrow(data),])
+}
+b_rare <- read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T)
+b_sobs <- get_sobs(b_rare)
+b_goods <- 100*(1-sum(table(bact$otu)==1)/nrow(bact))
+b_schloss <- 100*(1-sum(table(bact$otu)==1)/b_sobs["X0.03"])
+
+a_rare <- read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T)
+a_sobs <- get_sobs(a_rare)
+a_goods <- 100*(1-sum(table(arch$otu)==1)/nrow(arch))
+a_schloss <- 100*(1-sum(table(arch$otu)==1)/a_sobs["X0.03"])
+
+```
+
+***The status of the bacterial and archaeal census.***
+To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%.
+Although it is not possible to link a specific taxonomic level (e.g.
+species, genus, family, etc.) to a specific distance threshold, we selected distances of 0, 3, 5, 10, and 20% because they are widely regarded as representing the range of genetic diversity of the 16S rRNA gene within each domain.
+By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
+After sampling `r nrow(bact)` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1).
+Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once.
+In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once.
+Paralleling the bacterial results, after sampling `r nrow(arch)` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1).
+Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once.
+In contrast, only `r a_schloss`% of the OTUs that were observed had been seen more than once.
+The results for the bacterial and archaeal censuses indicate that regardless of the domain, if we continue sampling with the current strategies we will continue to sample OTUs that have already been observed even though a large fraction of OTUs have only been sampled once.
+Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once it is likely that an even larger number of OTUs have yet to be sampled for both domains.
+
+* Why aren't we doing a great job?
+  - Rate of change in full-length sequences has plateaued
+  - New sequences are submitted by a small number of investigators for a limited number of environments
+
+```{r time_course}
+  	time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE)
+
+	get_n50_count <- function(year){
+		sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]]
+	}
+	dominant_studies <- sum(unlist(lapply(2009:2012, get_n50_count)))
+
+	get_n50_pm_ids <- function(year){
+		authors <- names(sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]])
+		dom_pm_id <- unique(bact[bact$submit_author %in% authors, c('submit_author', 'pubmed_id')])$pubmed_id
+		as.character(dom_pm_id[!is.na(dom_pm_id)])
+	}
+	dominant_titles <- unlist(lapply(2009:2012, get_n50_pm_ids))
+
+	stopifnot(dominant_titles == c('19478181', '22310478', '20668241', '22310478', '22719818', '22832344'))
+
+	require(knitcitations, quiet=TRUE)
+	citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
+
+	get_n50_categories <- function(year){
+		authors <- names(sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]])
+		dom_pm_id <- unique(bact[bact$submit_author %in% authors, c('submit_author', 'category')])$category
+		as.character(dom_pm_id[!is.na(dom_pm_id)])
+	}
+	dominant_categories <- unlist(lapply(c(2009:2012,2014), get_n50_categories))
+
+```
+
+***The dynamics of the bacterial census.***
+One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
+In fact, since 2009 the number of new bacterial sequences generated has plateaued at an average of `r mean(time_data[as.character(2009:2014),'b_nseqs'])` sequences per year (Figure 2A).
+Although this is still an impressive number of sequences, since 2008 the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(2006:2014),'b_sobs'])` new OTUs per year (Figure 2B).
+Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms, we expected that the large number of sequences were being deposited by a handful of large projects.
+Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C).
+Between 2009 and 2012, `r sum(time_data[as.character(2009:2012), 'b_nseqs'])` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])`) submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_studies` (`r 100*dominant_studies/sum(time_data[as.character(2009:2012), 'b_nseqs'])`% of all sequences).
+These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
+In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year.
+Although these deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.
+
+***The dynamics of the archaeal census.***
+The depth of sequencing being done to advance the archaeal census is `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census.
+In contrast to the plateau that was observed in bacterial sequence submissions, prior to 2012 the submission of archaeal sequences generally increased each year; however between 2012 and 2014 the number of archaeal sequences submitted decreased from `r time_data['2012', 'a_nseqs']` to `r time_data['2014', 'a_nseqs']` sequences per year.
+The number of new OTUs contributed each year has largely paralleled that of the bacterial census with a plateau starting in 2005 at `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year.
+Although the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has never been more than `r max(time_data[,'a_n50'])`, with the exception of 2012 and 2014 there has been an average of `r mean(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions responsible for more than 50% of the sequences each year since 1996.
+The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea.
+
+* What can we do to fix this?
+  - Characterize coverage by environments
+  - Cultured vs. uncultured
+  - Single cell genomics
 
@@
 
-```{r figure4a}
+```{r figure4a, eval=FALSE}
 data <- read.table(file="data/mothur/all_bacteria.env_category.groups.rarefaction", header=T)
@@
 
-```{r figure4b}
+```{r figure4b, eval=FALSE}
 

### 2015-11-18 Schloss_Census2_mBio_2015.Rmd
Fix some typos

@@
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
 
@@
 
-* Why aren't we doing a great job?
-  - Rate of change in full-length sequences has plateaued
-  - New sequences are submitted by a small number of investigators for a limited number of environments
-
 ```{r time_course}
@@
 
-	require(knitcitations, quiet=TRUE)
-	citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
+	#require(knitcitations, quiet=TRUE)
+	#citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
 
@@
 One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
-In fact, since 2009 the number of new bacterial sequences generated has plateaued at an average of `r mean(time_data[as.character(2009:2014),'b_nseqs'])` sequences per year (Figure 2A).
+In fact, since 2009 the number of new bacterial sequences generated has plateaued at an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
 Although this is still an impressive number of sequences, since 2008 the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(2006:2014),'b_sobs'])` new OTUs per year (Figure 2B).
@@
 Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C).
-Between 2009 and 2012, `r sum(time_data[as.character(2009:2012), 'b_nseqs'])` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])`) submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_studies` (`r 100*dominant_studies/sum(time_data[as.character(2009:2012), 'b_nseqs'])`% of all sequences).
+Between 2009 and 2012, `r sum(time_data[as.character(2009:2012), 'b_nseqs'])` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_studies` (`r 100*dominant_studies/sum(time_data[as.character(2009:2012), 'b_nseqs'])`% of all sequences).
 These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
@@
 Although these deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.
+
+==> The archaea census section is pretty rough still...
 

### 2015-11-19 results/tables/build_table_1.Rmd
Fix sanitizing of text
Generate output text file

@@
 require(xtable, quietly=TRUE)
+options(xtable.sanitize.text.function=identity)
 
@@
 
-fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Hydrothermal vent", "Ice", "Marine", "Marine sediment", "Other", "Digesters", "Food-associated", "Industrial/mining", "Pollution associated", "Other", "Root", "Surface", "Other", "Agriculture", "Desert", "Permafrost", "Other", "Arthropod", "Other invertebrate", "Vertebrate", "Other", "", "", "")
+fine <- c("", "Brackish", "Brackish sediment", "Freshwater", "Freshwater sediment", "Marine", "Marine sediment", "Hydrothermal vent", "Ice", "Other", "Digesters", "Food-associated", "Industrial/mining", "Pollution associated", "Other", "Root", "Surface", "Other", "Agriculture", "Desert", "Permafrost", "Other", "Vertebrate", "Arthropod", "Other invertebrate", "Other", "", "", "")
 
-abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "BD", "BF", "BI", "BP", "BO", "PR", "PS", "PO", "SA", "SD", "SP", "SO", "ZA", "ZN", "ZV", "ZO", "OT")
+abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQM", "AQMS", "AQH", "AQI", "AQO", "BD", "BF", "BI", "BP", "BO", "PR", "PS", "PO", "SA", "SD", "SP", "SO", "ZV", "ZA", "ZN",  "ZO", "OT")
 
@@
 #abbreviation <- c(abbreviation, "", "")
+
 composite <- cbind(coarse, fine, bact_count, bact_sobs, bact_good_coverage, bact_schloss_coverage, arch_count, arch_sobs, arch_good_coverage, arch_schloss_coverage)
+write.table(composite, file=paste0(PROJHOME, "/data/process/table1_data.tsv"))
+
 composite <- rbind(c("", "", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
@@
 addtorow <- list(pos=list(0, 1),
-				command=c('\\multirow{4}{*}{Coarse} & \\multirow{4}{*}{Fine} & \\multicolumn{4}{c}{Bacterial} & \\multicolumn{4}{c}{Archaeal} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
+				command=c('\\multirow{4}{*}{Coarse} & \\multirow{4}{*}{Fine} & \\multicolumn{4}{c}{Bacteria} & \\multicolumn{4}{c}{Archaea} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
 

### 2015-11-19 Schloss_Census2_mBio_2015.Rmd
Working through results section

@@
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
 ```
@@
 ***Metadata curation.***
-The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse” categories, and further refined to facilitate additional analyses with twenty-six more specific "fine” categories (Table 1).
-These were assigned based on manual curation of the `isolation_source` category within the ARB database associated with each of the sequences.
+The metadata that was contained within the SSU Ref database was used to expand our analysis beyond a basic count of sequences and the number of OTUs in each domain.
+The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table 1).
+These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences.
 There were `r length(unique(bact$isolation_source))-1` bacterial and `r length(unique(arch$isolation_source))-1` archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1.
 For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
-There were `r sum(is.na(bact$isolation_source))` bacterial and `r sum(is.na(arch$isolation_source))` archaeal sequences where an `isolation_source` term was not collected.
+There were `r sum(is.na(bact$isolation_source))` bacterial and `r sum(is.na(arch$isolation_source))` archaeal sequences where an "isolation_source" term was not collected.
+We ascertained whether a sequence came from a bacterial isolate using the "strain" and "isolate" fields within the database.
+If a sequence had a value for either of these fields it was considered an isolate.
 Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].
@@
 
-	stopifnot(dominant_titles == c('19478181', '22310478', '20668241', '22310478', '22719818', '22832344'))
-
-	#require(knitcitations, quiet=TRUE)
-	#citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
-
 	get_n50_categories <- function(year){
@@
 
+	stopifnot(dominant_titles == c('19478181', '22310478', '20668241', '22310478', '22719818', '22832344'))
+
+	#require(knitcitations, quiet=TRUE)
+	#citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
 ```
@@
 One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
-In fact, since 2009 the number of new bacterial sequences generated has plateaued at an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
-Although this is still an impressive number of sequences, since 2008 the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(2006:2014),'b_sobs'])` new OTUs per year (Figure 2B).
+In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
+Although this is still an impressive number of sequences, since `r start<-2007;start` the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(start:2014),'b_sobs'])` new OTUs per year (Figure 2B).
 Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms, we expected that the large number of sequences were being deposited by a handful of large projects.
@@
 
-==> The archaea census section is pretty rough still...
-
 ***The dynamics of the archaeal census.***
-The depth of sequencing being done to advance the archaeal census is `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census.
-In contrast to the plateau that was observed in bacterial sequence submissions, prior to 2012 the submission of archaeal sequences generally increased each year; however between 2012 and 2014 the number of archaeal sequences submitted decreased from `r time_data['2012', 'a_nseqs']` to `r time_data['2014', 'a_nseqs']` sequences per year.
-The number of new OTUs contributed each year has largely paralleled that of the bacterial census with a plateau starting in 2005 at `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year.
-Although the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has never been more than `r max(time_data[,'a_n50'])`, with the exception of 2012 and 2014 there has been an average of `r mean(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions responsible for more than 50% of the sequences each year since 1996.
+The depth of sequencing being done to advance the archaeal census is `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census (Table 1).
+The annual number of sequences submitted has largely paralleled that of the bacterial census with a plateau starting in 2009 and an average of `r mean(time_data[as.character(2009:2014),'a_nseqs'])` sequences each year since then.
+The number of new archaeal OTUs represented by these sequences began to slow in 2005 with an average of `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year.
+With the exception of 2012 and 2014, the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has varied between `r min(time_data[c(1996:2011,2013),'a_n50'])` and `r max(time_data[c(1996:2011,2013),'a_n50'])` submissions per year.
 The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea.
 
-* What can we do to fix this?
-  - Characterize coverage by environments
-  - Cultured vs. uncultured
+```{r environment}
+	strip <- function(x){	as.numeric(gsub(',', '', x))	}
+
+	table1_raw <- read.table(file="data/process/table1_data.tsv", row.names=NULL, stringsAsFactors=FALSE)
+	table1_raw <- table1_raw[-nrow(table1_raw),]
+	rownames(table1_raw) <- table1_raw$row.names
+	rownames(table1_raw)[nrow(table1_raw)] <- 'no_source_data'
+	table1_raw <- table1_raw[,-c(1,2)]
+	table1_raw[c("AE", "OT", "no_source_data"),"fine"] <- c("Aerosol", "Other", "no_source_data")
+
+	for(col in 2:ncol(table1_raw)){
+		table1_raw[,col] <- strip(table1_raw[,col])
+	}
+
+	table1_good <- table1_raw[-nrow(table1_raw),]
+
+	b_class_percent <- 100*sum(table1_good$bact_count)/sum(table1_raw$bact_count)
+	a_class_percent <- 100*sum(table1_good$arch_count)/sum(table1_raw$arch_count)
+
+	coarse_map <- c("A"=="aerosol", "AQ" = "aquatic", "B" = "built environment", "O" = "other", "P" = "plant", "S" = "soil", "Z" = "zoonotic")
+
+	coarse <- gsub("^(.)[A-Z]$", "\\1", rownames(table1_good))
+	coarse <- gsub("^(..)[A-Z]+$", "\\1", coarse)
+
+	b_agg_count <- aggregate(table1_good$bact_count, by=list(coarse_map[coarse]), sum)
+	a_agg_count <- aggregate(table1_good$arch_count, by=list(coarse_map[coarse]), sum)
+
+```
+
+***Focusing the census by environment***
+We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database.
+Across these broad categories there was wide variation in the number of sequences that have been sampled.
+Among the bacterial sequences the three best represented groups were from `r b_agg_count[1]`, `r b_agg_count[2]`, and `r b_agg_count[3]` sources and among the bacterial sequences the three best represented groups were from `r a_agg_count[1]`, `r a_agg_count[2]`, and `r a_agg_count[3]` sources.
+For both domains, soil samples were the fourth most represented category.
+The orders of these categories was surprising considering soil and aquatic environments harbor the most microbial biomass and biodiversity [@Whitman1998].
+In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range for the bacterial OTU-based coverage only varied between `r quantile(table1_good$bact_schloss_coverage, prob=c(0.25))` to `r quantile(table1_good$bact_schloss_coverage, prob=c(0.75))` categories (median coverage=`r median(table1_good$bact_schloss_coverage, na.rm=T)`%).
+The interquartile range in the OTU-based coverage by environment for the archaeal data was `r quantile(table1_good$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to `r quantile(table1_good$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(table1_good$arch_schloss_coverage, na.rm=T)`%).
+The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
+Given the relatively low number of archaeal sequences that have been deposited for many of these categories, it is possible that the coverage for many categories may not be reliable (e.g.
+aerosol and plant surfaces).
+By decomposing the census by environmental categories, it is clear that even among the best sampled environments, our ability to claim anything but a basic characterization of microbial biodiversity is limited.
+
+```{r culture}
+	b_is_cult <- (!is.na(bact$strain) | !is.na(bact$isolate)) & !grepl("\\.Unc", rownames(bact))
+	a_is_cult <- (!is.na(arch$strain) | !is.na(arch$isolate)) & !grepl("\\.Unc", rownames(arch))
+
+	b_ncult_full <- sum(b_is_cult)
+	a_ncult_full <- sum(a_is_cult)
+
+	b_fract_cult <- 100 * b_ncult_full / nrow(bact)
+	a_fract_cult <- 100 * a_ncult_full / nrow(arch)
+
+	b_ncult_2005 <- sum((!is.na(bact$strain) | !is.na(bact$isolate)) & bact_year_deposited <= 2005, na.rm=T)
+	a_ncult_2005 <- sum((!is.na(arch$strain) | !is.na(arch$isolate)) & arch_year_deposited <= 2005, na.rm=T)
+
+	b_ncult_delta <- b_ncult_full - b_ncult_2005
+
+	b_before_2005 <- sum(bact_year_deposited <= 2005, na.rm=T)
+	b_since_2005 <- sum(bact_year_deposited > 2005, na.rm=T)
+
+	b_slow_cult <- (b_since_2005/b_before_2005) / (b_ncult_delta/b_ncult_2005)
+
+	a_ncult_delta <- a_ncult_full - a_ncult_2005
+
+	a_before_2005 <- sum(arch_year_deposited <= 2005, na.rm=T)
+	a_since_2005 <- sum(arch_year_deposited > 2005, na.rm=T)
+
+	a_slow_cult <- (a_since_2005/a_before_2005) / (a_ncult_delta/a_ncult_2005)
+
+	bact_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", bact$taxonomy)
+	bact_phylum[bact_phylum=="Euryarchaeota"] <- NA
+	bact_phylum[bact_phylum=="aquifer1"] <- NA
+
+	b_is_cult[rownames(bact) == 'AJ224540.MxeCultu'] <- FALSE
+#	aquifer1 - scrap
+#	bact[bact_phylum=="Candidate_division_OP3" & !is.na(bact$isolate),] - AJ224540.MxeCultu
+
+#	bact[bact_phylum=="CKC4" & !is.na(bact$isolate),] -> 2 - 1 is a genome, 1 is an isolate from same research group
+#	bact[bact_phylum=="PAUC34f" & (!is.na(bact$isolate) | !is.na(bact$strain)),] - legit
+#	bact[bact_phylum=="TM6" & (!is.na(bact$isolate) | !is.na(bact$strain)) & !grepl("\\.Unc", rownames(bact)),] - legit
+
+	b_phylum_cult_table <- table(bact_phylum, b_is_cult)
+	b_phylum_cult_rate <- b_phylum_cult_table[,"TRUE"] / apply(b_phylum_cult_table, 1, sum)
+
+	arch_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", arch$taxonomy)
+	arch_phylum[arch_phylum=="Archaea;"] <- "Unclassified"
+	a_phylum_cult_table <- table(arch_phylum, a_is_cult)
+	a_phylum_cult_rate <- a_phylum_cult_table[,"TRUE"] / apply(a_phylum_cult_table, 1, sum)
+
+cult_otus <- unique(bact$otu[b_is_cult])			# otus with a cultured representative
+b_in_cult_otu <- bact$otu %in% cult_otus
+
+#total number of OTUs by phylum
+n_all_otus_phylum <- aggregate(bact$otu, by=list(bact_phylum), function(x){length(unique(x))})
+
+#total number of OTUs with a cultured rep by phylum
+n_cult_otus_phylum <- aggregate(bact$otu[b_in_cult_otu], by=list(bact_phylum[b_in_cult_otu]), length)
+
+#need to merge these two tables...
+
+```
+
+***The cultured census***
+In our previous analysis there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+
+* fraction of sequences from cultured / uncultured bugs (2005 vs 2014) by domain
+	- 18.9% of bacterial sequences and 7.2% of archaeal sequences are from isolates (b_fract_cult & a_fract_cult)
+	- culture rates lagging total submissions by 2.7-fold (bacteria / archaea; b_slow_cult / a_slow_cult)
+
+* fraction of sequences from cultured / uncultured bugs (2014) by phylum
+	- b_phylum_cult_rate
+	- a_phylum_cult_rate
+
+* % of OTUs with a cultured representative by phylum
+	- basically done.
+
+* do these new cultures represent new OTUs?
+
+***The census is taxonomically skewed***
+* For which phyla has there been substantial increases in sampled diversity since 2004?
+
+***New technologies to access novel biodiversity***
   - Single cell genomics
-
-* Figure 1A. How have the number of full-length sequences changed over the past N years?
-* Figure 1B. How does this vary by environment?
-
-* Figure 2. How has the number of sequences per study changed over the past N years?
-
-* Figure 3. How has the coverage of the OTUs represented by these sequences changed over the last N years?
-
-* How has Figure 1 changed over the past N years?
-
-* In what environments do we have the best handle on extant diversity?
-  Figure 4A,B. What are the differences in sampling effort/sequence coverage at the 97% level for the different coarse habitats, and how is this different between Bacteria and Archaea?
-  Figure 5A,B. For the most sampled coarse habitats (Zoonotic for Bacteria, Aquatic for Archaea), what are the fine-scale habitats where they have been sampled most heavily?
-
-* In what habitats do we have the poorest understanding of diversity (and thus expect to yield considerable new knowledge over the years)?
-* How is this different for Bacteria vs. Archaea?
-* What are the most significant improvements since 2004, and where are we still lacking the most data?
-* For which phyla has there been substantial increases in sampled diversity since 2004?
-* Can we make any recommendations for habitats where biodiversity is under-sampled, perhaps calls for conservation in certain environments based on how poorly we understand the microbial diversity?
 
@@
 Renewed call for cultivation
+
+What are the most significant improvements since 2004, and where are we still lacking the most data?
 

### 2015-11-20 code/build_phylum_category_heatmap.R
Generate new phylum-level plots

@@ -1,0 +1,69 @@
+require("RColorBrewer")
+
+bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", bact$taxonomy)
+bact_phylum[bact_phylum=="Euryarchaeota"] <- NA
+bact_phylum[bact_phylum=="aquifer1"] <- NA
+
+arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", arch$taxonomy)
+arch_phylum[arch_phylum=="Archaea;"] <- "Unclassified"
+
+bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA" #to remove
+arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD" #to remove
+
+
+categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other")
+
+
+b_phylum_cat_count <- table(bact_phylum, bact$category)
+b_phylum_cat_ra <- prop.table(b_phylum_cat_count, margin=2)
+b_phylum_cat_max_ra <- apply(b_phylum_cat_ra, 1, max)
+o <- order(b_phylum_cat_max_ra, decreasing=T)
+b_ordered_subset <- b_phylum_cat_ra[o[1:10],rev(names(categories))]
+b_average <- table(bact_phylum)[o[1:10]] / length(bact_phylum)
+
+
[…]

### 2015-11-20 code/build_phylum_effort.R

@@ -1,0 +1,69 @@
+bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
+
+bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA"
+
+
+b_phylum_year <- table(bact_year_deposited, bact_phylum)
+b_n_phylum_2015 <- ncol(b_phylum_year)
+b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
+b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
+b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
+b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
+
+b_phylum_new <- b_phylum_year[rownames(b_phylum_year) %in% as.character(2007:2014),]
+b_n_new_seqs <- sum(b_phylum_new)						#how many sequences are there since 2006
+b_n_new_seqs_phyla <- apply(b_phylum_new, 2, sum)		#how many sequences in each phyla since 2006?
+
+enough <- (b_n_new_seqs_phyla + b_n_old_seqs_phyla) > 1000
+phyla_ratio <- b_n_new_seqs_phyla/b_n_old_seqs_phyla
+seq_ratio <- b_n_new_seqs/b_n_old_seqs
+
+b_enough_ratio <- sort(phyla_ratio[enough]/seq_ratio)
+
+
+arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
+
+arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD"
+
+a_phylum_year <- table(arch_year_deposited, arch_phylum)
+a_n_phylum_2015 <- ncol(a_phylum_year)
+a_phylum_2006 <- a_phylum_year[rownames(a_phylum_year) %in% as.character(1983:2006),]
[…]

### 2015-11-20 Schloss_Census2_mBio_2015.Rmd
Add text/code for taxonomic analysis

@@
 
+bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA"
+
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
+
+arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD"
+
 ```
@@
 #	aquifer1 - scrap
-#	bact[bact_phylum=="Candidate_division_OP3" & !is.na(bact$isolate),] - AJ224540.MxeCultu
+#	bact[bact_phylum=="Candidate_division_OP3" & !is.na(bact$isolate) & !is.na(bact_phylum),] - AJ224540.MxeCultu
 
@@
 
-cult_otus <- unique(bact$otu[b_is_cult])			# otus with a cultured representative
-b_in_cult_otu <- bact$otu %in% cult_otus
+	b_cult_otus <- unique(bact$otu[b_is_cult])
+	b_in_cult_otu <- bact$otu %in% b_cult_otus
 
-#total number of OTUs by phylum
 	n_all_otus_phylum <- aggregate(bact$otu, by=list(bact_phylum), function(x){length(unique(x))})
+	n_cult_otus_phylum <- aggregate(bact$otu[b_in_cult_otu], by=list(bact_phylum[b_in_cult_otu]), function(x){length(unique(x))})
 
-#total number of OTUs with a cultured rep by phylum
-n_cult_otus_phylum <- aggregate(bact$otu[b_in_cult_otu], by=list(bact_phylum[b_in_cult_otu]), length)
+	b_otus_phylum <- data.frame(matrix(rep(0, nrow(n_all_otus_phylum)*2), ncol=2, dimnames=list(n_all_otus_phylum$Group.1, c('cultured', 'total'))))
 
-#need to merge these two tables...
+	b_otus_phylum[n_all_otus_phylum$Group.1,"total"] <- n_all_otus_phylum$x
+	b_otus_phylum[n_cult_otus_phylum$Group.1,"cultured"] <- n_cult_otus_phylum$x
 
+	a_cult_otus <- unique(arch$otu[a_is_cult])
+	a_in_cult_otu <- arch$otu %in% a_cult_otus
+
+	n_all_otus_phylum <- aggregate(arch$otu, by=list(arch_phylum), function(x){length(unique(x))})
+	n_cult_otus_phylum <- aggregate(arch$otu[a_in_cult_otu], by=list(arch_phylum[a_in_cult_otu]), function(x){length(unique(x))})
+
+	a_otus_phylum <- data.frame(matrix(rep(0, nrow(n_all_otus_phylum)*2), ncol=2, dimnames=list(n_all_otus_phylum$Group.1, c('cultured', 'total'))))
+
+	a_otus_phylum[n_all_otus_phylum$Group.1,"total"] <- n_all_otus_phylum$x
+	a_otus_phylum[n_cult_otus_phylum$Group.1,"cultured"] <- n_cult_otus_phylum$x
+
+	b_n1_cult_otus <- sum(table(bact$otu[b_in_cult_otu])==1)/length(table(bact$otu[b_in_cult_otu]))
+	a_n1_cult_otus <- sum(table(arch$otu[a_in_cult_otu])==1)/length(table(arch$otu[a_in_cult_otu]))
 ```
@@
 
-* % of OTUs with a cultured representative by phylum
-	- basically done.
+* % of OTUs with a cultured representative by phylum and in aggregate
+	- b_otus_phylum
+	- a_otus_phylum
 
-* do these new cultures represent new OTUs?
+* do these new cultures represent new OTUs by phylum and in aggregate
+	- b_n1_cult_otus
+	- a_n1_cult_otus
+
+* need a plot showing % cultured sequences and % cultured OTUs
 
 ***The census is taxonomically skewed***
+
+```{r taxon_census}
+	b_phylum_year <- table(bact_year_deposited, bact_phylum)
+	b_n_phylum_2015 <- ncol(b_phylum_year)
+	b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
+	b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
+	b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
+	b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
+
+	b_new_phyla <- colnames(b_phylum_year)[apply(b_phylum_2006, 2, sum) == 0] #what are the new phyla?
+	b_new_phyla_names <- paste(paste(b_new_phyla[1:(length(b_new_phyla)-1)], collapse=", "),
+							 b_new_phyla[length(b_new_phyla)], sep=", and ")
+
+	a_phylum_year <- table(arch_year_deposited, arch_phylum)
+	a_n_phylum_2015 <- ncol(a_phylum_year)
+	a_phylum_2006 <- a_phylum_year[rownames(a_phylum_year) %in% as.character(1983:2006),]
+	a_n_phylum_2006 <- sum(apply(a_phylum_2006, 2, sum) > 0) #how many phyla were there?
+	a_n_old_seqs <- sum(a_phylum_2006)						#how many seqs in 2006?
+	a_n_old_seqs_phyla <- apply(a_phylum_2006, 2, sum)		#how many sequences in each phyla?
+
+	a_new_phyla <- colnames(a_phylum_year)[apply(a_phylum_2006, 2, sum) == 0] #what are the new phyla?
+	a_new_phyla_names <- paste(a_new_phyla, collapse=" and ")
+	a_new_phyla_names <- gsub("_", " ", a_new_phyla_names)
+	a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
+```
+
 * For which phyla has there been substantial increases in sampled diversity since 2004?
+	- new phyla: b_new_phyla_names, a_new_phyla_names
+	- Figure phylum_effort.pdf
+
+* generate a heatmap of category by phylum relative abundance
+	- Figure phylum_category_heatmap
 
@@
   - Single cell genomics
+  - Metagenomes
 

### 2015-11-20 doc/notebook/Archaeal_data_acquisition.Rmd
Change metadat files to be *.v123.metadata
Metadata typos

@@
 metadata$category[metadata$category == 'AH'] <- 'AQH'
+metadata$category[metadata$category == 'BDBD'] <- 'BD'
 
 #... and we output the metadata file...
-write.table(metadata, file="data/mothur/archaea.final.metadata", sep="\t")
+write.table(metadata, file="data/process/archaea.v123.metadata", sep="\t")
 

### 2015-11-23 doc/notebook/Bacterial_data_acquisition.Rmd
Change metadat files to be *.v123.metadata
Fix path issue btwn pre.cluster and cluster.split

@@
 
-* ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1,384,528)
-* ARB_color != 1 & tax_greengenes == "*Bacteria*" (N=1,123,279)
-* ARB_color != 1 & tax_slv == "*Bacteria*" (N=1,526,523)
-* ARB_color != 1 & tax_embl == "*Bacteria*" (N=1,335,715)
-* ARB_color != 1 & (tax_rdp | tax_greengenes | tax_embl | tax_slv) (N=1,529,365)
+* ARB_color != 1 & tax_rdp == "Bacteria*" (N=1,376,114)
+* ARB_color != 1 & tax_greengenes == "k__Bacteria*" (N=1,119,394)
+* ARB_color != 1 & tax_slv == "Bacteria*" (N=1,516,624)
+* ARB_color != 1 & tax_embl == "Bacteria*" (N=1,331,557)
 
@@
 * Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
-* Change the second search field to "tax_slv" and set it to "*Bacteria*"
+* Change the second search field to "tax_slv" and set it to "Bacteria*"
 * Click "Mark Listed, Unmark Rest"
@@
 * Hit "Search"
-* Hit Mark Listed Unmark Rest (N=1,515,024)
-* Click "Write to Fields of Listed", select the "remark" field, and enter "good bacteria"
-* Click "Write"
+* Hit Mark Listed Unmark Rest (N=1,505,125)
 * In main ARB window go File->Export->Export to external format
@@
 * Go File->Export->Export fields
-* Set the file name to "data/mothur/bacteria.taxonomy" and Column output to "TAB separated"
+* Set the file name to "data/mothur/bacteria.taxonomy.nds" and Column output to "TAB separated"
 * Click "SAVE"
@@
 				Start   End     NBases  Ambigs  Polymer NumSeqs
-Minimum:    1		10725	900		0		4	1
-2.5%-tile:	1003	14692	1274	0		5	37876
-25%-tile:	1011	14865	1355	0		5	378757
-Median: 	1044	15082	1387	0		5	757513
-75%-tile:	1048	15672	1455	0		6	1136269
-97.5%-tile:	1815	15771	1539	2		7	1477149
-Maximum:	7036	17738	3469	64		38	1515024
-Mean:		1104.14	15254.4	1404	0.181	5.509
-# of Seqs:	1515024
+Minimum:        1       12530   1200    0       4       1
+2.5%-tile:      1003    14538   1282    0       5       37629
+25%-tile:       1011    14707   1356    0       5       376282
+Median:         1044    14926   1387    0       5       752563
+75%-tile:       1048    15514   1455    0       6       1128844
+97.5%-tile:     1795    15613   1539    2       7       1467497
+Maximum:        5300    17580   3469    64      38      1505125
+Mean:   		1099.25 15100.4 1405.16 0.180003        5.50868
+# of Seqs:      1505125
 ```
@@
 data <- read.table(file="data/mothur/bacteria.summary", header=T, row.names=1)
+
 quantile(data$start, probs=seq(0.9,1,0.01))
@@
 quantile(data$end, probs=seq(0,0.1,0.01))
-end <- 14853
+end <- 14459
 
@@
                     data$nbases > min_length & data$ambigs <= max_ambig,]
-nrow(trimmed)
+
+nrow(trimmed) #[1] 1431222
+
 summary(trimmed$nbases)
+#Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
+#1201    1359    1390    1409    1457    3469
 ```
@@
 ```{r screen_data, engine="bash"}
-mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=14853, maxambig=2, minlength=1200, outputdir=data/mothur/, processors=8);"
+mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=14459, maxambig=2, minlength=1200, outputdir=data/mothur/, processors=8);"
 ```
@@
 ```{r single_cell, engine="bash"}
-#wget -N -P data/references/ http://www.mothur.org/w/images/2/24/Trainset10_082014.pds.tgz
-#tar xvzf data/references/Trainset10_082014.pds.tgz -C data/references/
-#mv data/references/trainset10_082014.pds/trainset10_082014.* data/references/
-#rm -rf data/references/trainset10_082014.pds
-
 mothur "#set.dir(output=data/mothur);
@@
 #need to make the taxonomy file match the fasta and names files...
-cut -f 1 data/mothur/all_bacteria.filter.unique.precluster.names > precluster.accnos
+cut -f 1 data/mothur/all_bacteria.filter.unique.precluster.names > data/mothur/precluster.accnos
 mothur "#set.dir(input=data/mothur, output=data/mothur);
@@
 metadata$category[metadata$category == 'SV'] <- 'ZV'
+metadata$category[metadata$category == 'ZI'] <- 'ZA'
 
 #... and we output the metadata file...
-write.table(metadata, file="data/mothur/bacteria.final.metadata", sep="\t")
+write.table(metadata, file="data/process/bacteria.v123.metadata", sep="\t")
 

### 2015-11-23 Schloss_Census2_mBio_2015.Rmd
Enter E. coli numbering for where sequence data are located

@@
 The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur.
-Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates while maintaining a minimum sequence length of 1,200 and 900 nt for the bacterial and archeael datasets, respectively.
+Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates.
+The sequences in the resulting bacterial dataset overlapped bases 113 through 1344 of an *E.
+coli* reference sequence (V00348) and had a median length of 1,227 nt.
+The sequences in the resulting archaeal dataset overlapped positions 404 to 973 and had a median length of 580 nt.
+The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences.
 The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences.
@@
 
-***Caveats.***
-Recent data suggests that a considerable diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers.
-This dataset does not include sequences from metagenomic assemblies but Brown et al.
-(2015) have used such assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
-Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015).
-Thus, for many environments, the estimates within must be considered as lower bounds.
-
-Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
-Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science.
-While reliance on this gene subjects us to biases created by primer selection [REFS], differences in amplification strength [REFS] and fidelity [REFS], internal features which may disrupt traditional measurements [REFS], and potentially misleading classification due to infrequent horizontal gene transfer [REFS], the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
-Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
-
 ## Conclusions
@@
 What are the most significant improvements since 2004, and where are we still lacking the most data?
+
+Recent data suggests that a considerable diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers.
+This dataset does not include sequences from metagenomic assemblies but Brown et al.
+(2015) have used such assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
+Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015).
+Thus, for many environments, the estimates within must be considered as lower bounds.
+
+Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
+Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science.
+While reliance on this gene subjects us to biases created by primer selection [REFS], differences in amplification strength [REFS] and fidelity [REFS], internal features which may disrupt traditional measurements [REFS], and potentially misleading classification due to infrequent horizontal gene transfer [REFS], the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
+Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
 

### 2015-11-24 code/time_course_plots.R
Restructure code to separate data processing from plotting

@@ -1,133 +1,65 @@
-start_year <- 1983
-year_of_analysis <- 2015
-
-bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
-
-arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
-
-summary_table <- data.frame(matrix(rep(0, 32*8), ncol=8))
-rownames(summary_table) <- start_year:(year_of_analysis-1)
-colnames(summary_table) <- c('b_nseqs', 'b_sobs', 'b_n50', 'b_phyla',
-							'a_nseqs', 'a_sobs', 'a_n50', 'a_phyla')
+time_data <- read.table(file="data/process/by_year_analysis.tsv", header=T)
+year <- as.numeric(rownames(time_data))
 
 
-pdf(file="results/figures/sequences_deposited.pdf")
-	deposits_table <- table(bact_year_deposited)
-	bact_deposits_by_year <- as.numeric(deposits_table)
-	names(bact_deposits_by_year) <- names(deposits_table)
-	bact_deposits_by_year <- bact_deposits_by_year[!(names(bact_deposits_by_year) %in% year_of_analysis)]
+pdf(file="results/figures/time_course_figure.pdf", width=3.75, height=6.0)
+	layout(matrix(c(1,2,3,4), nrow=4), height=c(1,1,1,0.25))
 
-	summary_table[names(bact_deposits_by_year),'b_nseqs'] <- bact_deposits_by_year
+	#A
+	par(mar=c(0.5,5.5,0.5,0.5))
 
-	plot(bact_deposits_by_year~as.numeric(names(bact_deposits_by_year)), log='y', type="l", col="blue", lwd=2, xlab="Year", ylab="Number of sequences per year", ylim=c(1,1e6), xlim=c(1980, 2015), axes=F)
+	time_data[time_data$a_nseqs == 0, "a_nseqs"] <- 1
+	time_data[time_data$b_nseqs == 0, "b_nseqs"] <- 1
 
-	deposits_table <- table(arch_year_deposited)
-	arch_deposits_by_year <- as.numeric(deposits_table)
[…]

### 2015-11-24 code/time_course_submission_data.R

@@ -1,0 +1,94 @@
+first_year <- 1983
+final_year  <- 2015
+
+get_year <- function(date_vector){
+	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date_vector)
+}
+
+
+bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact_year_deposited <- get_year(bact$date)
+
+arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch_year_deposited <- get_year(arch$date)
+
+summary_table <- data.frame(matrix(rep(0, (final_year-first_year+1)*8), ncol=8))
+rownames(summary_table) <- first_year:final_year
+colnames(summary_table) <- c('b_nseqs', 'b_sobs', 'b_n50', 'b_n50_doi',
+							'a_nseqs', 'a_sobs', 'a_n50', 'a_n50_doi')
+
+
+get_annual_seq_deposits <- function(year_vector){
+	deposits_by_year <- as.numeric(table(year_vector))
+	names(deposits_by_year) <- levels(factor(year_vector))
+	deposits_by_year
+}
+
+bact_data <- get_annual_seq_deposits(bact_year_deposited)
+summary_table[names(bact_data),'b_nseqs'] <- bact_data
+
+arch_data <- get_annual_seq_deposits(arch_year_deposited)
+summary_table[names(arch_data),'a_nseqs'] <- arch_data
+
+
+
+get_annual_otu_deposits <- function(year_deposited, otus){
+
+	#make a table to indicate the years each OTU appeared
+	otus_table <- table(otus, year_deposited)
+
+	#make a vector to indicate the first year each OTU appeared
+	otu_appearance <- rep("", nrow(otus_table))
+	names(otu_appearance) <- as.character(1:length(otu_appearance))
+
+	#get the first year that each OTU appeared
+	otu_appearance <- colnames(otus_table)[apply(otus_table != 0, 1, function(x){ifelse(sum(x) == 0, NA, which.max(x))})]
+
+	#count the number of new OTUs in each year
+	new_otus_table <- table(otu_appearance)
+	new_otus_by_year <- as.numeric(new_otus_table)
+	names(new_otus_by_year) <- names(new_otus_table)
+	new_otus_by_year
+}
+
[…]

### 2015-11-24 Schloss_Census2_mBio_2015.Rmd
Make archaeal numbering against reference of Sulfolobus
Update code through the archaeal census paragraph

@@
 
-bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+#bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
 
-bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA"
+#arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+#arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
+```
 
-arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
-
-arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD"
-
-```
 \vspace{35mm}
@@
 Furthermore, creating standards for delimiting what makes one microbe "different” from another has been contentious [@reference].
-In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, and metagenomics.
+In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, single cell technologies, and metagenomics.
 
@@
 coli* reference sequence (V00348) and had a median length of 1,227 nt.
-The sequences in the resulting archaeal dataset overlapped positions 404 to 973 and had a median length of 580 nt.
+The sequences in the resulting archaeal dataset overlapped positions 362 to 937 of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of 580 nt.
 The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences.
@@
 There were `r sum(is.na(bact$isolation_source))` bacterial and `r sum(is.na(arch$isolation_source))` archaeal sequences where an "isolation_source" term was not collected.
-We ascertained whether a sequence came from a bacterial isolate using the "strain" and "isolate" fields within the database.
-If a sequence had a value for either of these fields it was considered an isolate.
+We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms.
 Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].
@@
 ```{r rarefaction}
+# take these data out of the rarefaciton files so that we can have the unique,
+# 0.03, 0.05, 0.10, and 0.20 OTU definitions for both the bacterial and archaeal
+# data
+
 get_sobs <- function(data){
@@
 }
-b_rare <- read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T)
-b_sobs <- get_sobs(b_rare)
-b_goods <- 100*(1-sum(table(bact$otu)==1)/nrow(bact))
-b_schloss <- 100*(1-sum(table(bact$otu)==1)/b_sobs["X0.03"])
 
-a_rare <- read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T)
-a_sobs <- get_sobs(a_rare)
-a_goods <- 100*(1-sum(table(arch$otu)==1)/nrow(arch))
-a_schloss <- 100*(1-sum(table(arch$otu)==1)/a_sobs["X0.03"])
+b_sobs <- get_sobs(read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T))
+a_sobs <- get_sobs(read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T))
 
+category_summary <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
+b_goods <- category_summary["total", "bact_good_coverage"]
+b_schloss <- category_summary["total", "bact_schloss_coverage"]
+
+a_goods <- category_summary["total", "arch_good_coverage"]
+a_schloss <- category_summary["total", "arch_schloss_coverage"]
 ```
@@
 ```{r time_course}
-  	time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE)
+time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE, stringsAsFactors=FALSE)
 
-	get_n50_count <- function(year){
-		sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]]
-	}
-	dominant_studies <- sum(unlist(lapply(2009:2012, get_n50_count)))
+all_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_nseqs'])
+dominant_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_n50_nseqs'])
 
-	get_n50_pm_ids <- function(year){
-		authors <- names(sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]])
-		dom_pm_id <- unique(bact[bact$submit_author %in% authors, c('submit_author', 'pubmed_id')])$pubmed_id
-		as.character(dom_pm_id[!is.na(dom_pm_id)])
-	}
-	dominant_titles <- unlist(lapply(2009:2012, get_n50_pm_ids))
-
-	get_n50_categories <- function(year){
-		authors <- names(sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]])
-		dom_pm_id <- unique(bact[bact$submit_author %in% authors, c('submit_author', 'category')])$category
-		as.character(dom_pm_id[!is.na(dom_pm_id)])
-	}
-	dominant_categories <- unlist(lapply(c(2009:2012,2014), get_n50_categories))
-
-	stopifnot(dominant_titles == c('19478181', '22310478', '20668241', '22310478', '22719818', '22832344'))
+# what were the papers that these sequences came from?
+dominant_pmids_2009_2012 <- unique(unlist(strsplit(time_data[as.character(2009:2012), "b_n50_doi"], split=',')))
 
@@
 #citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
+
+stopifnot(dominant_pmids_2009_2012 == c("19478181", "22310478", "20668241", "22719818", "22832344"))
+#if this stops then will need to update stuff below
 ```
 
-***The dynamics of the bacterial census.***
+***The status of the bacterial census.***
 One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
@@
 Although this is still an impressive number of sequences, since `r start<-2007;start` the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(start:2014),'b_sobs'])` new OTUs per year (Figure 2B).
-Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms, we expected that the large number of sequences were being deposited by a handful of large projects.
+Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects.
 Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C).
-Between 2009 and 2012, `r sum(time_data[as.character(2009:2012), 'b_nseqs'])` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_studies` (`r 100*dominant_studies/sum(time_data[as.character(2009:2012), 'b_nseqs'])`% of all sequences).
+Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences).
 These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
@@
 
-***The dynamics of the archaeal census.***
-The depth of sequencing being done to advance the archaeal census is `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census (Table 1).
+***The status of the archaeal census.***
+The depth of sequencing being done to advance the archaeal census has been `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census (Table 1).
 The annual number of sequences submitted has largely paralleled that of the bacterial census with a plateau starting in 2009 and an average of `r mean(time_data[as.character(2009:2014),'a_nseqs'])` sequences each year since then.
 The number of new archaeal OTUs represented by these sequences began to slow in 2005 with an average of `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year.
-With the exception of 2012 and 2014, the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has varied between `r min(time_data[c(1996:2011,2013),'a_n50'])` and `r max(time_data[c(1996:2011,2013),'a_n50'])` submissions per year.
+With the exception of 2012 and 2014, the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has varied between `r min(time_data[as.character(c(1996:2011,2013)),'a_n50'])` and `r max(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions per year.
 The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea.
+
+```{r taxon_census}
+	b_phylum_year <- table(bact_year_deposited, bact_phylum)
+	b_phylum_n <- apply(b_phylum_year, 2, sum)
+	b_n_phylum_2015 <- ncol(b_phylum_year)
+	b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
+	b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
+	b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
+	b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
+
+	b_new_phyla <- colnames(b_phylum_year)[apply(b_phylum_2006, 2, sum) == 0] #what are the new phyla?
+	b_new_phyla_names <- paste(paste(b_new_phyla[1:(length(b_new_phyla)-1)], collapse=", "),
+							 b_new_phyla[length(b_new_phyla)], sep=", and ")
+
+	a_phylum_year <- table(arch_year_deposited, arch_phylum)
+	a_phylum_n <- apply(a_phylum_year, 2, sum)
+	a_n_phylum_2015 <- ncol(a_phylum_year)
+	a_phylum_2006 <- a_phylum_year[rownames(a_phylum_year) %in% as.character(1983:2006),]
+	a_n_phylum_2006 <- sum(apply(a_phylum_2006, 2, sum) > 0) #how many phyla were there?
+	a_n_old_seqs <- sum(a_phylum_2006)						#how many seqs in 2006?
+	a_n_old_seqs_phyla <- apply(a_phylum_2006, 2, sum)		#how many sequences in each phyla?
+
+	a_new_phyla <- colnames(a_phylum_year)[apply(a_phylum_2006, 2, sum) == 0] #what are the new phyla?
+	a_new_phyla_names <- paste(a_new_phyla, collapse=" and ")
+	a_new_phyla_names <- gsub("_", " ", a_new_phyla_names)
+	a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
+```
+
+***The ability to sample microbial life is taxonomically skewed*** <-- Sucky title
+The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum_n[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes")])/sum(b_phylum_n)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum_n[c("Euryarchaeota",  "Thaumarchaeota")])/sum(a_phylum_n)`% of the archaeal sequences, yet we sought to understand how the representation individual phyla has changed relative to the state of the census in 2006.
+We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published and ushered in a radical change in how microbial communities are studied [@Sogin2006].
+Since then, the number of phyla within each domain has continued to increase.
+In 2006 there were `r b_n_phylum_2006` bacterial and `r a_n_phylum_2006` phyla.
+Since then there have been `r b_new_phyla` new  bacterial (`r b_new_phyla_names`) and `r a_new_phyla` new archaeal candidate phyla (`r a_new_phyla_names`).
+Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure phylum_effort.pdf).
+Among the bacterial phyla with at least 1,000 sequences, Kazan−3B−09 (`r something`) and Atribacteria (`r something`) were sequenced 4-fold more often while Deinococcus−Thermus (`r something`) and Tenericutes (`r something`) were sequenced 4-fold less often than would have been expected since 2006.
+Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r something`-fold more often and the Crenarchaeota were sequenced `r something`-fold less often than expected.
+Overall, these results suggest <something...>
 
@@
 The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
-Given the relatively low number of archaeal sequences that have been deposited for many of these categories, it is possible that the coverage for many categories may not be reliable (e.g.
-aerosol and plant surfaces).
-By decomposing the census by environmental categories, it is clear that even among the best sampled environments, our ability to claim anything but a basic characterization of microbial biodiversity is limited.
+Across all categories, the bacterial and archaeal sequencing data represented a limited number of phyla (Figure category_phylum_heatmap.pdf).
+Regardless, there were clear phylum-level signatures that differentiated the various categories .
+The bacterial sequences were assigned to `r XX` environmental categories where `r XX` were dominated by Proteobacteria, `r XX` by Firmicutes, `r XX` by Bacteroidetes, and `r XX` by Actinobacteria.
+The archaeal sequences were assigned to `r XX ` categories where `r XX` were dominated by Euryarchaeota and `r XX` by Thaumarchaeota.
+Within each of the bacterial and archaeal phyla, there was considerable variation in the relative abundance of each across the categories confirming that taxonomic signatures exist to differentiate different environments even at a broad taxonomic level.
 
@@
 
-	b_ncult_2005 <- sum((!is.na(bact$strain) | !is.na(bact$isolate)) & bact_year_deposited <= 2005, na.rm=T)
-	a_ncult_2005 <- sum((!is.na(arch$strain) | !is.na(arch$isolate)) & arch_year_deposited <= 2005, na.rm=T)
+	b_ncult_2005 <- sum(b_is_cult & bact_year_deposited <= 2005, na.rm=T)
+	a_ncult_2005 <- sum(a_is_cult & arch_year_deposited <= 2005, na.rm=T)
 
@@
 
-#	bact[bact_phylum=="CKC4" & !is.na(bact$isolate),] -> 2 - 1 is a genome, 1 is an isolate from same research group
+#	bact[bact_phylum=="CKC4" & !is.na(bact$strain) & !is.na(bact_phylum),] -> 2 - 1 is a genome, 1 is an isolate from same research group
 #	bact[bact_phylum=="PAUC34f" & (!is.na(bact$isolate) | !is.na(bact$strain)),] - legit
@@
 
-	b_n1_cult_otus <- sum(table(bact$otu[b_in_cult_otu])==1)/length(table(bact$otu[b_in_cult_otu]))
-	a_n1_cult_otus <- sum(table(arch$otu[a_in_cult_otu])==1)/length(table(arch$otu[a_in_cult_otu]))
+	#not sure what this is actually getting...
+	b_n1_cult_otus <- 100*sum(table(bact$otu[b_in_cult_otu])==1)/length(table(bact$otu[b_in_cult_otu]))
+	a_n1_cult_otus <- 100*sum(table(arch$otu[a_in_cult_otu])==1)/length(table(arch$otu[a_in_cult_otu]))
 ```
@@
 ***The cultured census***
-In our previous analysis there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
+Overall, `r b_fract_cult`% of bacterial sequences and `r a_fract_cult`% of archaeal sequences have come from isolated organisms.
+Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_slow_cult` and `r a_slow_cult`-fold for bacteria and archaea, respectively.
+Among the `r length(b_phylum_cult_rate)` bacterial phyla, `r sum(b_phylum_cult_rate == 0)` have no cultured representatives and `r sum(b_phylum_cult_rate == 0)` of the `r length(a_phylum_cult_rate)` archaeal phyla have cultured representatives.
+Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_phylum_cult_rate[b_phylum_cult_rate!=0])*100`% for the bacterial phyla and `r median(a_phylum_cult_rate[a_phylum_cult_rate!=0])*100`% for the archaeal phyla (Figure XXXXXX).
+When we asked what percentage of OTUs had at least one cultured respresentative we found that `r 100*sum(b_otus_phylum$cultured)/sum(b_otus_phylum$total)`% of the `r sum(b_otus_phylum$total)` bacterial OTUs and `r 100*sum(a_otus_phylum$cultured)/sum(a_otus_phylum$total)`% of the `r sum(a_otus_phylum$total)` archeael OTUs had at least one cultured representative.
+Of those OTUs that had a sequence from a culture, `r b_n1_cult_otus`
+and `r a_n1_cult_otus`% of them only contained sequences from cultures. These values suggest that bacterial cultivation efforts are biased towards
 
-* fraction of sequences from cultured / uncultured bugs (2005 vs 2014) by domain
-	- 18.9% of bacterial sequences and 7.2% of archaeal sequences are from isolates (b_fract_cult & a_fract_cult)
-	- culture rates lagging total submissions by 2.7-fold (bacteria / archaea; b_slow_cult / a_slow_cult)
-
+* need a plot showing % cultured sequences and % cultured OTUs
 	* fraction of sequences from cultured / uncultured bugs (2014) by phylum
@@
 		- a_phylum_cult_rate
-
 	* % of OTUs with a cultured representative by phylum and in aggregate
@@
 
-* do these new cultures represent new OTUs by phylum and in aggregate
-	- b_n1_cult_otus
-	- a_n1_cult_otus
+```{r single_cell}
 
-* need a plot showing % cultured sequences and % cultured OTUs
+grepl("^[^\\.]*$", rownames(bact))
 
-***The census is taxonomically skewed***
+dim(bact[bact$publication_doi=="10.1038/nature12352" & !is.na(bact$publication_doi),])
+#[1] 95 31
 
-```{r taxon_census}
-	b_phylum_year <- table(bact_year_deposited, bact_phylum)
-	b_n_phylum_2015 <- ncol(b_phylum_year)
-	b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
-	b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
-	b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
-	b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
+```
+***New technologies to access novel biodiversity***
+Assembly of metagenomic and single cell shotgun sequence data offers the hope of identifying large fragments of genomic data from as yet uncultured organisms.
+To test the ability of single cell technologies to identify
 
-	b_new_phyla <- colnames(b_phylum_year)[apply(b_phylum_2006, 2, sum) == 0] #what are the new phyla?
-	b_new_phyla_names <- paste(paste(b_new_phyla[1:(length(b_new_phyla)-1)], collapse=", "),
-							 b_new_phyla[length(b_new_phyla)], sep=", and ")
-
-	a_phylum_year <- table(arch_year_deposited, arch_phylum)
-	a_n_phylum_2015 <- ncol(a_phylum_year)
-	a_phylum_2006 <- a_phylum_year[rownames(a_phylum_year) %in% as.character(1983:2006),]
-	a_n_phylum_2006 <- sum(apply(a_phylum_2006, 2, sum) > 0) #how many phyla were there?
-	a_n_old_seqs <- sum(a_phylum_2006)						#how many seqs in 2006?
-	a_n_old_seqs_phyla <- apply(a_phylum_2006, 2, sum)		#how many sequences in each phyla?
-
-	a_new_phyla <- colnames(a_phylum_year)[apply(a_phylum_2006, 2, sum) == 0] #what are the new phyla?
-	a_new_phyla_names <- paste(a_new_phyla, collapse=" and ")
-	a_new_phyla_names <- gsub("_", " ", a_new_phyla_names)
-	a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
-```
-
-* For which phyla has there been substantial increases in sampled diversity since 2004?
-	- new phyla: b_new_phyla_names, a_new_phyla_names
-	- Figure phylum_effort.pdf
-
-* generate a heatmap of category by phylum relative abundance
-	- Figure phylum_category_heatmap
-
-***New technologies to access novel biodiversity***
   - Single cell genomics
@@
 Impact of EMIRGE
+- http://www.genomebiology.com/2011/12/5/R44
 
@@
 Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
+
+Given the relatively low number of archaeal sequences that have been deposited for many of these categories, it is possible that the coverage for many categories may not be reliable (e.g.
+aerosol and plant surfaces).
+By decomposing the census by environmental categories, it is clear that even among the best sampled environments, our ability to claim anything but a basic characterization of microbial biodiversity is limited.
 

### 2015-11-24 code/time_course_plots.R

@@ -1,6 +1,10 @@
 time_data <- read.table(file="data/process/by_year_analysis.tsv", header=T)
 year <- as.numeric(rownames(time_data))
 
+exclude_year <- 2015
+
+time_data <- time_data[year != exclude_year,]
+year <- year[year != exclude_year]
 
 pdf(file="results/figures/time_course_figure.pdf", width=3.75, height=6.0)
 	layout(matrix(c(1,2,3,4), nrow=4), height=c(1,1,1,0.25))

### 2015-11-24 code/time_course_submission_data.R

@@ -12,10 +12,10 @@ bact_year_deposited <- get_year(bact$date)
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 arch_year_deposited <- get_year(arch$date)
 
-summary_table <- data.frame(matrix(rep(0, (final_year-first_year+1)*8), ncol=8))
+summary_table <- data.frame(matrix(rep(0, (final_year-first_year+1)*10), ncol=10))
 rownames(summary_table) <- first_year:final_year
-colnames(summary_table) <- c('b_nseqs', 'b_sobs', 'b_n50', 'b_n50_doi',
-							'a_nseqs', 'a_sobs', 'a_n50', 'a_n50_doi')
+colnames(summary_table) <- c('b_nseqs', 'b_sobs', 'b_n50', 'b_n50_doi', 'b_n50_nseqs',
+							'a_nseqs', 'a_sobs', 'a_n50', 'a_n50_doi', 'a_n50_nseqs')
 
 
 get_annual_seq_deposits <- function(year_vector){
@@ -58,21 +58,21 @@ arch_data <- get_annual_otu_deposits(arch_year_deposited, arch$otu)
 summary_table[names(arch_data),'a_sobs'] <- arch_data
 
 
-how_many <- function(year, year_vector, db, fraction=0.50){
+get_n50_data <- function(year, year_vector, db, fraction=0.50){
 
 	year_db <- db[year_vector==as.character(year) & !is.na(year_vector),]
 	dummy_variable <- paste(year_db$submit_author, year_db$pubmed_id, sep='_')
-	ranked_order <- cumsum(sort(table(dummy_variable), decreasing=T))/length(dummy_variable)
+	cumsum_count <- cumsum(sort(table(dummy_variable), decreasing=T))
+	cumsum_fraction <- cumsum_count/length(dummy_variable)
 
-	count <- unname(which.max(ranked_order>=fraction))
+	count <- unname(which.max(cumsum_fraction>=fraction))
 
 	if(length(count) == 0){
-		count <- 0
-		names(count) <- "NA"
+		list(count=0, names="NA", nseqs=0)
 	} else {
-		names(count) <- paste(names(ranked_order)[1:count], collapse='|')
+		list(count=count, names=paste(names(cumsum_count)[1:count], collapse='|'), nseqs=unname(cumsum_count[count]))
 	}
-	count
+
[…]

### 2015-11-25 Schloss_Census2_mBio_2015.Rmd
Fix r code in M&M section

@@
 
+```{r overview}
+category_summary <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
+
+b_nseqs <- category_summary["total", "bact_count"]
+b_goods <- category_summary["total", "bact_good_coverage"]
+b_schloss <- category_summary["total", "bact_schloss_coverage"]
+
+a_nseqs <- category_summary["total", "arch_count"]
+a_goods <- category_summary["total", "arch_good_coverage"]
+a_schloss <- category_summary["total", "arch_schloss_coverage"]
+```
+
 ***Sequence data curation.***
@@
 The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences.
-The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences.
+The final datasets contained `r b_nseqs` bacterial and `r a_nseqs` archaeal 16S rRNA gene sequences.
 Sequences were assigned to OTUs using the average neighbor clustering algorithm [@REF].
@@
 The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table 1).
-These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences.
-There were `r length(unique(bact$isolation_source))-1` bacterial and `r length(unique(arch$isolation_source))-1` archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1.
+These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences (Table 1).
 For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
-There were `r sum(is.na(bact$isolation_source))` bacterial and `r sum(is.na(arch$isolation_source))` archaeal sequences where an "isolation_source" term was not collected.
+There were `r category_summary["no_source", "bact_count"]` bacterial and `r category_summary["no_source", "arch_count"]` archaeal sequences where an "isolation_source" term was not collected.
 We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms.
-Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].
+Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environemtnal categorizations are available at [@figshare].
 
@@
 a_sobs <- get_sobs(read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T))
-
-category_summary <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
-b_goods <- category_summary["total", "bact_good_coverage"]
-b_schloss <- category_summary["total", "bact_schloss_coverage"]
-
-a_goods <- category_summary["total", "arch_good_coverage"]
-a_schloss <- category_summary["total", "arch_schloss_coverage"]
 ```
@@
 By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
-After sampling `r nrow(bact)` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1).
+After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1).
 Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once.
@@
 
+##### here #######
 ```{r taxon_census}

### 2015-11-25 code/build_phylum_effort.R
Separate out phylum counting

@@ -1,44 +1,9 @@
-bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
+count_threshold <- 1000
 
-bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA"
+bacteria <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
+archaea <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
 
-
-b_phylum_year <- table(bact_year_deposited, bact_phylum)
-b_n_phylum_2015 <- ncol(b_phylum_year)
-b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
-b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
-b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
-b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
-
-b_phylum_new <- b_phylum_year[rownames(b_phylum_year) %in% as.character(2007:2014),]
-b_n_new_seqs <- sum(b_phylum_new)						#how many sequences are there since 2006
-b_n_new_seqs_phyla <- apply(b_phylum_new, 2, sum)		#how many sequences in each phyla since 2006?
-
-enough <- (b_n_new_seqs_phyla + b_n_old_seqs_phyla) > 1000
-phyla_ratio <- b_n_new_seqs_phyla/b_n_old_seqs_phyla
-seq_ratio <- b_n_new_seqs/b_n_old_seqs
-
-b_enough_ratio <- sort(phyla_ratio[enough]/seq_ratio)
-
-
-arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
-
-arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD"
-
-a_phylum_year <- table(arch_year_deposited, arch_phylum)
-a_n_phylum_2015 <- ncol(a_phylum_year)
[…]

### 2015-11-25 code/get_phylum_counts.R

@@ -1,0 +1,34 @@
+threshold <- 2006
+
+get_year <- function(date){
+	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date)
+}
+
+get_phylum <- function(taxonomy){
+	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
+	phylum[grepl(';', phylum)] <- 'Unclassified'
+	phylum
+}
+
+summarize_phylum_data <- function(db){
+	year_deposited <- get_year(db$date)
+	phylum <- get_phylum(db$taxonomy)
+
+	years <- as.numeric(levels(factor(year_deposited)))
+	pre <- years <= threshold
+	post <- years > threshold
+
+	phylum_year <- table(year_deposited, phylum)
+	pre <- apply(phylum_year[pre,], 2, sum)
+	post <- apply(phylum_year[post,], 2, sum)
+	all <- apply(phylum_year, 2, sum)
+	cbind(pre, post, all)
+}
+
+generate_table <- function(domain){
+	metadata_file <- paste0('data/process/', domain, '.v123.metadata')
+	count_file <- paste0('data/process/', domain, '.phyla.counts.tsv')
+
+	input <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
+	write.table(summarize_phylum_data(input), file=count_file, quote=FALSE, sep='\t')
+}

### 2015-11-25 code/get_phylum_counts.R
Add ratio of ratios column

@@ -1,4 +1,4 @@
-threshold <- 2006
+
 
 get_year <- function(date){
 	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date)
@@ -10,25 +10,32 @@ get_phylum <- function(taxonomy){
 	phylum
 }
 
-summarize_phylum_data <- function(db){
+get_ratio_of_ratios <- function(pre, post){
+	phyla_ratio <- post/pre
+	seq_ratio <- sum(post)/sum(pre)
+	phyla_ratio / seq_ratio
+}
+
+summarize_phylum_data <- function(db, threshold){
 	year_deposited <- get_year(db$date)
 	phylum <- get_phylum(db$taxonomy)
 
 	years <- as.numeric(levels(factor(year_deposited)))
-	pre <- years <= threshold
-	post <- years > threshold
+	pre_years <- years <= threshold
+	post_years <- years > threshold
 
 	phylum_year <- table(year_deposited, phylum)
-	pre <- apply(phylum_year[pre,], 2, sum)
-	post <- apply(phylum_year[post,], 2, sum)
+	pre <- apply(phylum_year[pre_years,], 2, sum)
+	post <- apply(phylum_year[post_years,], 2, sum)
 	all <- apply(phylum_year, 2, sum)
-	cbind(pre, post, all)
+	ratio_of_ratios <- get_ratio_of_ratios(pre, post)
+	cbind(pre, post, all, ratio_of_ratios)
 }
 
-generate_table <- function(domain){
+generate_table <- function(domain, threhold=2006){
 	metadata_file <- paste0('data/process/', domain, '.v123.metadata')
 	count_file <- paste0('data/process/', domain, '.phyla.counts.tsv')
 
 	input <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	write.table(summarize_phylum_data(input), file=count_file, quote=FALSE, sep='\t')
+	write.table(summarize_phylum_data(input, threshold), file=count_file, quote=FALSE, sep='\t')
 }

### 2015-11-25 code/get_phylum_counts.R
Fix typo in threshold variable name

@@ -32,7 +32,7 @@ summarize_phylum_data <- function(db, threshold){
 	cbind(pre, post, all, ratio_of_ratios)
 }
 
-generate_table <- function(domain, threhold=2006){
+generate_table <- function(domain, threshold=2006){
 	metadata_file <- paste0('data/process/', domain, '.v123.metadata')
 	count_file <- paste0('data/process/', domain, '.phyla.counts.tsv')
 

### 2015-11-25 code/build_phylum_effort_plot.R
Generate phylum effort plot from separated code

@@ -3,32 +3,39 @@ count_threshold <- 1000
 bacteria <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
 archaea <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
 
-enough <- (a_n_new_seqs_phyla + a_n_old_seqs_phyla) > count_threshold
-phyla_ratio <- a_n_new_seqs_phyla/a_n_old_seqs_phyla
-seq_ratio <- a_n_new_seqs/a_n_old_seqs
+enough <- bacteria$all >= count_threshold
+bact_ratios <- log2(bacteria$ratio_of_ratios[enough])
+names(bact_ratios) <- gsub("_\\(SAR406_clade\\)", "", rownames(bacteria)[enough])
 
-a_enough_ratio <- sort(phyla_ratio[enough]/seq_ratio)
-names(a_enough_ratio) <- gsub("Miscellaneous_Crenarchaeotic_Group", "Misc. Crenarchaeota", names(a_enough_ratio))
+bact_ratios <- sort(bact_ratios)
+n_bact_ratios <- length(bact_ratios)
 
-n_b_enough <- length(b_enough_ratio)
-n_a_enough <- length(a_enough_ratio)
-n_enough <- n_b_enough+n_a_enough
+
+enough <- archaea$all >= count_threshold
+arch_ratios <- log2(archaea$ratio_of_ratios[enough])
+names(arch_ratios) <- gsub("Miscellaneous_Crenarchaeotic_Group", "Misc. Crenarchaeota", rownames(archaea)[enough])
+
+arch_ratios <- sort(arch_ratios)
+n_arch_ratios <- length(arch_ratios)
 
 pdf(file="results/figures/phylum_effort.pdf", width=3.75, height=5)
-	par(mar=c(5,10, 0.5, 0.5), cex=0.7)
+	par(mar=c(5,10, 1.0, 0.5), cex=0.7)
 
-	plot("", ylim=c(1,n_enough+1), xlim=c(0, max(c(a_enough_ratio, b_enough_ratio))), type="n", xlab="", ylab="", axes=F)
+	plot("", ylim=c(1,n_bact_ratios+n_arch_ratios+2), xlim=c(min(c(bact_ratios, arch_ratios)), ceiling(max(c(bact_ratios, arch_ratios)))), type="n", xlab="", ylab="", axes=F)
 
[…]

### 2015-11-25 Schloss_Census2_mBio_2015.Rmd
Update code for change in phylum-level representation

@@
 
-##### here #######
 ```{r taxon_census}
-	b_phylum_year <- table(bact_year_deposited, bact_phylum)
-	b_phylum_n <- apply(b_phylum_year, 2, sum)
-	b_n_phylum_2015 <- ncol(b_phylum_year)
-	b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
-	b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
-	b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
-	b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
+oxford_comma <- function(ids){
+	if(length(ids)==0){
+		"ERROR"
+	} else if(length(ids) == 1){
+		ids
+	} else if(length(ids) == 2){
+		paste(ids, collapse=" and ")
+	} else {
+		paste(paste(ids[1:(length(ids)-1)], collapse=", "), ids[length(ids)], sep=", and ")
+	}
+}
 
-	b_new_phyla <- colnames(b_phylum_year)[apply(b_phylum_2006, 2, sum) == 0] #what are the new phyla?
-	b_new_phyla_names <- paste(paste(b_new_phyla[1:(length(b_new_phyla)-1)], collapse=", "),
-							 b_new_phyla[length(b_new_phyla)], sep=", and ")
+b_phylum <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
+b_phylum <- b_phylum[!rownames(b_phylum) %in% 'aquifer1',]
+rownames(b_phylum) <- gsub("Candidate_division_", "", rownames(b_phylum))
+rownames(b_phylum) <- gsub("_", " ", rownames(b_phylum))
 
-	a_phylum_year <- table(arch_year_deposited, arch_phylum)
-	a_phylum_n <- apply(a_phylum_year, 2, sum)
-	a_n_phylum_2015 <- ncol(a_phylum_year)
-	a_phylum_2006 <- a_phylum_year[rownames(a_phylum_year) %in% as.character(1983:2006),]
-	a_n_phylum_2006 <- sum(apply(a_phylum_2006, 2, sum) > 0) #how many phyla were there?
-	a_n_old_seqs <- sum(a_phylum_2006)						#how many seqs in 2006?
-	a_n_old_seqs_phyla <- apply(a_phylum_2006, 2, sum)		#how many sequences in each phyla?
+b_new_phyla <- rownames(b_phylum)[b_phylum$pre == 0]
+b_new_phyla_names <- oxford_comma(b_new_phyla_names)
 
-	a_new_phyla <- colnames(a_phylum_year)[apply(a_phylum_2006, 2, sum) == 0] #what are the new phyla?
-	a_new_phyla_names <- paste(a_new_phyla, collapse=" and ")
+a_phylum <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
+a_new_phyla <- rownames(a_phylum)[a_phylum$pre == 0]
+
+a_new_phyla_names <- oxford_comma(a_new_phyla_names)
+a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
 a_new_phyla_names <- gsub("_", " ", a_new_phyla_names)
-	a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
 ```
 
-***The ability to sample microbial life is taxonomically skewed*** <-- Sucky title
-The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum_n[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes")])/sum(b_phylum_n)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum_n[c("Euryarchaeota",  "Thaumarchaeota")])/sum(a_phylum_n)`% of the archaeal sequences, yet we sought to understand how the representation individual phyla has changed relative to the state of the census in 2006.
+***The ability to sample microbial life is taxonomically skewed (meh.)***
+The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences.
+We sought to understand how the representation of individual phyla has changed relative to the state of the census in 2006.
 We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published and ushered in a radical change in how microbial communities are studied [@Sogin2006].
-Since then, the number of phyla within each domain has continued to increase.
-In 2006 there were `r b_n_phylum_2006` bacterial and `r a_n_phylum_2006` phyla.
+In 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
 Since then there have been `r b_new_phyla` new  bacterial (`r b_new_phyla_names`) and `r a_new_phyla` new archaeal candidate phyla (`r a_new_phyla_names`).
 Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure phylum_effort.pdf).
-Among the bacterial phyla with at least 1,000 sequences, Kazan−3B−09 (`r something`) and Atribacteria (`r something`) were sequenced 4-fold more often while Deinococcus−Thermus (`r something`) and Tenericutes (`r something`) were sequenced 4-fold less often than would have been expected since 2006.
-Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r something`-fold more often and the Crenarchaeota were sequenced `r something`-fold less often than expected.
-Overall, these results suggest <something...>
+Among the bacterial phyla with at least 1,000 sequences, `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios >= 4.0 & b_phylum$all >=1000,])))` were sequenced 4-fold more often while `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios < 0.5 & b_phylum$all >=1000,])))` were sequenced 2-fold less often than would have been expected since 2006.
+Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r a_phylum["Thaumarchaeota", "ratio_of_ratios"]`-fold more often and the Crenarchaeota were sequenced `r 1/a_phylum["Crenarchaeota", "ratio_of_ratios"]`-fold less often than expected.
+Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006.
 

### 2015-11-25 code/build_phylum_category_heatmap.R
Separate out data curation and heatmapping for phylum/category

@@ -1,69 +1,54 @@
 require("RColorBrewer")
 
-bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", bact$taxonomy)
-bact_phylum[bact_phylum=="Euryarchaeota"] <- NA
-bact_phylum[bact_phylum=="aquifer1"] <- NA
+categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other", total="Total")
 
-arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", arch$taxonomy)
-arch_phylum[arch_phylum=="Archaea;"] <- "Unclassified"
+rev_categories <- length(categories):1
 
-bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA" #to remove
-arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD" #to remove
+data <- read.table(file="data/process/phylum_category_counts.tsv", header=T)
 
+bacteria <- data[data$domain=="bacteria",-1]
+bacteria <- bacteria[order(bacteria$total, decreasing=T),]
+bacteria_top <- as.matrix(bacteria[1:10,rev_categories])
 
[…]

### 2015-11-25 code/get_phylum_category_counts.R
Separate out data curation and heatmapping for phylum/category

@@ -1,0 +1,36 @@
+get_phylum <- function(taxonomy){
+	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
+	phylum[grepl(';', phylum)] <- "Unclassified"
+	phylum
+}
+
+
+categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other")
+
+bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA" #to remove
+
+bact_phylum <- get_phylum(bact$taxonomy)
+bact_phylum[bact_phylum=="Euryarchaeota"] <- NA #to be removed
+bact_phylum[bact_phylum=="aquifer1"] <- NA
+
+b_phylum_cat_count <- table(bact_phylum, bact$category)
+b_phylum_cat_ra <- prop.table(b_phylum_cat_count, margin=2)
+b_phylum_cat_ra <- cbind(b_phylum_cat_ra, total=table(bact_phylum)/length(bact_phylum))
+rownames(b_phylum_cat_ra) <- gsub("_?\\(.*\\)", "", rownames(b_phylum_cat_ra) )
+bacterial_data <- cbind(domain=rep("bacteria", nrow(b_phylum_cat_ra)), b_phylum_cat_ra)
+
+
+arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD" #to remove
+
+arch_phylum <- get_phylum(arch$taxonomy)
+
+a_phylum_cat_count <- table(arch_phylum, factor(arch$category, levels=names(categories)))
[…]

### 2015-11-25 Schloss_Census2_mBio_2015.Rmd
Use phylum_category_counts.tsv data for phylum category analysis

@@
 
-	table1_raw <- read.table(file="data/process/table1_data.tsv", row.names=NULL, stringsAsFactors=FALSE)
-	table1_raw <- table1_raw[-nrow(table1_raw),]
-	rownames(table1_raw) <- table1_raw$row.names
-	rownames(table1_raw)[nrow(table1_raw)] <- 'no_source_data'
-	table1_raw <- table1_raw[,-c(1,2)]
-	table1_raw[c("AE", "OT", "no_source_data"),"fine"] <- c("Aerosol", "Other", "no_source_data")
+coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
 
-	for(col in 2:ncol(table1_raw)){
-		table1_raw[,col] <- strip(table1_raw[,col])
-	}
+time <- grepl('_pre', colnames(coverage_by_category_and_time)) | grepl('_post', colnames(coverage_by_category_and_time))
+coverage_by_category <- coverage_by_category_and_time[,!time]
 
-	table1_good <- table1_raw[-nrow(table1_raw),]
+totals <- unlist(coverage_by_category[nrow(coverage_by_category),])
+coverage_by_category <- coverage_by_category[-nrow(coverage_by_category),]
 
-	b_class_percent <- 100*sum(table1_good$bact_count)/sum(table1_raw$bact_count)
-	a_class_percent <- 100*sum(table1_good$arch_count)/sum(table1_raw$arch_count)
+b_class_percent <- 100*sum(coverage_by_category[-nrow(coverage_by_category), "bact_count"])/sum(totals["bact_count"])
+
+a_class_percent <- 100*sum(coverage_by_category[-nrow(coverage_by_category), "arch_count"])/sum(totals["arch_count"])
 
@@
 
-	coarse <- gsub("^(.)[A-Z]$", "\\1", rownames(table1_good))
+coarse <- gsub("^(.)[A-Z]$", "\\1", rownames(coverage_by_category))
 coarse <- gsub("^(..)[A-Z]+$", "\\1", coarse)
 
-	b_agg_count <- aggregate(table1_good$bact_count, by=list(coarse_map[coarse]), sum)
-	a_agg_count <- aggregate(table1_good$arch_count, by=list(coarse_map[coarse]), sum)
+b_agg_count <- aggregate(coverage_by_category$bact_count, by=list(coarse_map[coarse]), sum)
+b_text <- paste0(b_agg_count$Group.1, ' (N=', format(b_agg_count$x, big.mark=',', trim=TRUE), ')')[order(b_agg_count$x, decreasing=T)]
+
+a_agg_count <- aggregate(coverage_by_category$arch_count, by=list(coarse_map[coarse]), sum)
+a_text <- paste0(a_agg_count$Group.1, ' (N=', format(a_agg_count$x, big.mark=',', trim=TRUE), ')')[order(a_agg_count$x, decreasing=T)]
+
+phylum_category <- read.table(file="data/process/phylum_category_counts.tsv")
+bacteria <- phylum_category[phylum_category$domain == "bacteria", -1]
+
+dominant_bacteria <- sort(table(apply(bacteria, 2, which.max)), decreasing=T)
+names(dominant_bacteria) <- rownames(bacteria)[as.numeric(names(dominant_bacteria))]
+dominant_bacteria_list <- oxford_comma(paste0(names(dominant_bacteria), ' (N=', dominant_bacteria, ')'))
+
+archaea <- phylum_category[phylum_category$domain == "archaea", -1]
+dominant_archaea <- sort(table(apply(archaea, 2, function(x){z<-which.max(x);if(length(z)==0){z<-NA};z})), decreasing=T)
+names(dominant_archaea) <- rownames(archaea)[as.numeric(names(dominant_archaea))]
+dominant_archaea_list <- oxford_comma(paste0(names(dominant_archaea), ' (N=', dominant_archaea, ')'))
 
@@
 Across these broad categories there was wide variation in the number of sequences that have been sampled.
-Among the bacterial sequences the three best represented groups were from `r b_agg_count[1]`, `r b_agg_count[2]`, and `r b_agg_count[3]` sources and among the bacterial sequences the three best represented groups were from `r a_agg_count[1]`, `r a_agg_count[2]`, and `r a_agg_count[3]` sources.
-For both domains, soil samples were the fourth most represented category.
+Among the bacterial sequences the three best represented groups were from `r oxford_comma(b_text[1:3])` sources and among the archaeal sequences the three best represented groups were from `r oxford_comma(a_text[1:3])` sources.
+For both domains, soil samples were the fourth most represented category (bacteria: `r b_agg_count[b_agg_count$Group.1=="soil", "x"]`; archaea: `r a_agg_count[a_agg_count$Group.1=="soil", "x"]`).
 The orders of these categories was surprising considering soil and aquatic environments harbor the most microbial biomass and biodiversity [@Whitman1998].
-In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range for the bacterial OTU-based coverage only varied between `r quantile(table1_good$bact_schloss_coverage, prob=c(0.25))` to `r quantile(table1_good$bact_schloss_coverage, prob=c(0.75))` categories (median coverage=`r median(table1_good$bact_schloss_coverage, na.rm=T)`%).
-The interquartile range in the OTU-based coverage by environment for the archaeal data was `r quantile(table1_good$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to `r quantile(table1_good$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(table1_good$arch_schloss_coverage, na.rm=T)`%).
+In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range across the fine-level categories for the bacterial OTU-based coverage only varied between `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.25))` to `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.75))` (median coverage=`r median(coverage_by_category$bact_schloss_coverage, na.rm=T)`%).
+The interquartile range in the OTU-based coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to`r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
 The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
 Across all categories, the bacterial and archaeal sequencing data represented a limited number of phyla (Figure category_phylum_heatmap.pdf).
-Regardless, there were clear phylum-level signatures that differentiated the various categories .
-The bacterial sequences were assigned to `r XX` environmental categories where `r XX` were dominated by Proteobacteria, `r XX` by Firmicutes, `r XX` by Bacteroidetes, and `r XX` by Actinobacteria.
-The archaeal sequences were assigned to `r XX ` categories where `r XX` were dominated by Euryarchaeota and `r XX` by Thaumarchaeota.
+Among the bacterial data, the fine-scale categories were dominated by `r dominant_bacteria_list` and among the archaeal data, they were dominated by `r dominant_archaea_list`.
+Regardless, there were clear phylum-level signatures that differentiated the various categories.
 Within each of the bacterial and archaeal phyla, there was considerable variation in the relative abundance of each across the categories confirming that taxonomic signatures exist to differentiate different environments even at a broad taxonomic level.

### 2015-11-26 code/get_culture_by_time_counts.R
Generate phylum by culture and time data

@@ -1,0 +1,54 @@
+get_phylum <- function(taxonomy){
+	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
+	phylum[grepl(';', phylum)] <- "Unclassified"
+	phylum <- gsub("_?\\(.*\\)", "", phylum)
+	phylum
+}
+
+get_year <- function(date){
+	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date)
+}
+
+is_cultured <- function(db){
+	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db))
+}
+
+#bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+#rownames(bact)[rownames(bact) == 'AJ224540.MxeCultu'] <- 'AJ224540.UncMxeCultu'
+
+
+get_cultured_data <- function(cultured, phyla, otu){
+	nseqs <- table(phyla, cultured)
+	nseqs <- cbind(cult_nseqs=nseqs[,"TRUE"], total_nseqs=apply(nseqs, 1, sum))
+
+	sobs <- apply(table(phyla, otu) > 0, 1, sum)
+	sobs_cultured <- apply(table(phyla[cultured], otu[cultured]) > 0, 1, sum)
+	sobs <- cbind(cult_sobs=sobs_cultured, total_sobs=sobs)
+
+	cbind(nseqs, sobs)
+}
+
+get_domain_data <- function(db, threshold=2006){
+
+	phyla <- factor(get_phylum(db$taxonomy))
+	cultured <- is_cultured(db)
+
+	pre <- get_year(db$date) < threshold
+
+	all <- get_cultured_data(cultured, phyla, db$otu)
+	pre <- get_cultured_data(cultured[pre], phyla[pre], db$otu[pre])
+	post <- get_cultured_data(cultured[!pre], phyla[!pre], db$otu[!pre])
+	pool <- cbind(pre, post, all)
+	colnames(pool) <- paste0(c(rep("pre_", 3), rep("post_", 3), rep("all_", 3)), colnames(pool))
+	pool
+}
+
+
+run_domain <- function(domain){
+	metadata_file <- paste0('data/process/', domain, '.v123.metadata')
+	output_file <- paste0('data/process/', domain, '.cultured_by_time_counts.tsv')
+
[…]

### 2015-11-30 Schloss_Census2_mBio_2015.Rmd
Incorporate culturability analysis

@@
 ```{r culture}
-	b_is_cult <- (!is.na(bact$strain) | !is.na(bact$isolate)) & !grepl("\\.Unc", rownames(bact))
-	a_is_cult <- (!is.na(arch$strain) | !is.na(arch$isolate)) & !grepl("\\.Unc", rownames(arch))
+b_cult_time <- read.table(file='data/process/bacteria.cultured_by_time_counts.tsv', header=T)
+a_cult_time <- read.table(file='data/process/archaea.cultured_by_time_counts.tsv', header=T)
 
-	b_ncult_full <- sum(b_is_cult)
-	a_ncult_full <- sum(a_is_cult)
+b_percent_cult_nseqs_all <- 100*sum(b_cult_time[,"all_cult_nseqs"])/sum(b_cult_time[,"all_total_nseqs"])
+a_percent_cult_nseqs_all <- 100*sum(a_cult_time[,"all_cult_nseqs"])/sum(a_cult_time[,"all_total_nseqs"])
 
-	b_fract_cult <- 100 * b_ncult_full / nrow(bact)
-	a_fract_cult <- 100 * a_ncult_full / nrow(arch)
+b_shift_cult_nseqs_all <- 1/((sum(b_cult_time[,"post_cult_nseqs"])/sum(b_cult_time[,"pre_cult_nseqs"])) / (sum(b_cult_time[,"post_total_nseqs"])/sum(b_cult_time[,"pre_total_nseqs"])))
 
-	b_ncult_2005 <- sum(b_is_cult & bact_year_deposited <= 2005, na.rm=T)
-	a_ncult_2005 <- sum(a_is_cult & arch_year_deposited <= 2005, na.rm=T)
+a_shift_cult_nseqs_all <- 1/((sum(a_cult_time[,"post_cult_nseqs"])/sum(a_cult_time[,"pre_cult_nseqs"])) / (sum(a_cult_time[,"post_total_nseqs"])/sum(a_cult_time[,"pre_total_nseqs"])))
 
-	b_ncult_delta <- b_ncult_full - b_ncult_2005
+b_rate_cult_nseqs_phyulm <- b_cult_time[, 'all_cult_nseqs'] / b_cult_time[, 'all_total_nseqs']
+a_rate_cult_nseqs_phyulm <- a_cult_time[, 'all_cult_nseqs'] / a_cult_time[, 'all_total_nseqs']
 
-	b_before_2005 <- sum(bact_year_deposited <= 2005, na.rm=T)
-	b_since_2005 <- sum(bact_year_deposited > 2005, na.rm=T)
-
-	b_slow_cult <- (b_since_2005/b_before_2005) / (b_ncult_delta/b_ncult_2005)
-
-	a_ncult_delta <- a_ncult_full - a_ncult_2005
-
-	a_before_2005 <- sum(arch_year_deposited <= 2005, na.rm=T)
-	a_since_2005 <- sum(arch_year_deposited > 2005, na.rm=T)
-
-	a_slow_cult <- (a_since_2005/a_before_2005) / (a_ncult_delta/a_ncult_2005)
-
-	bact_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", bact$taxonomy)
-	bact_phylum[bact_phylum=="Euryarchaeota"] <- NA
-	bact_phylum[bact_phylum=="aquifer1"] <- NA
-
-	b_is_cult[rownames(bact) == 'AJ224540.MxeCultu'] <- FALSE
-#	aquifer1 - scrap
-#	bact[bact_phylum=="Candidate_division_OP3" & !is.na(bact$isolate) & !is.na(bact_phylum),] - AJ224540.MxeCultu
-
-#	bact[bact_phylum=="CKC4" & !is.na(bact$strain) & !is.na(bact_phylum),] -> 2 - 1 is a genome, 1 is an isolate from same research group
-#	bact[bact_phylum=="PAUC34f" & (!is.na(bact$isolate) | !is.na(bact$strain)),] - legit
-#	bact[bact_phylum=="TM6" & (!is.na(bact$isolate) | !is.na(bact$strain)) & !grepl("\\.Unc", rownames(bact)),] - legit
-
-	b_phylum_cult_table <- table(bact_phylum, b_is_cult)
-	b_phylum_cult_rate <- b_phylum_cult_table[,"TRUE"] / apply(b_phylum_cult_table, 1, sum)
-
-	arch_phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", arch$taxonomy)
-	arch_phylum[arch_phylum=="Archaea;"] <- "Unclassified"
-	a_phylum_cult_table <- table(arch_phylum, a_is_cult)
-	a_phylum_cult_rate <- a_phylum_cult_table[,"TRUE"] / apply(a_phylum_cult_table, 1, sum)
-
-	b_cult_otus <- unique(bact$otu[b_is_cult])
-	b_in_cult_otu <- bact$otu %in% b_cult_otus
-
-	n_all_otus_phylum <- aggregate(bact$otu, by=list(bact_phylum), function(x){length(unique(x))})
-	n_cult_otus_phylum <- aggregate(bact$otu[b_in_cult_otu], by=list(bact_phylum[b_in_cult_otu]), function(x){length(unique(x))})
-
-	b_otus_phylum <- data.frame(matrix(rep(0, nrow(n_all_otus_phylum)*2), ncol=2, dimnames=list(n_all_otus_phylum$Group.1, c('cultured', 'total'))))
-
-	b_otus_phylum[n_all_otus_phylum$Group.1,"total"] <- n_all_otus_phylum$x
-	b_otus_phylum[n_cult_otus_phylum$Group.1,"cultured"] <- n_cult_otus_phylum$x
-
-	a_cult_otus <- unique(arch$otu[a_is_cult])
-	a_in_cult_otu <- arch$otu %in% a_cult_otus
-
-	n_all_otus_phylum <- aggregate(arch$otu, by=list(arch_phylum), function(x){length(unique(x))})
-	n_cult_otus_phylum <- aggregate(arch$otu[a_in_cult_otu], by=list(arch_phylum[a_in_cult_otu]), function(x){length(unique(x))})
-
-	a_otus_phylum <- data.frame(matrix(rep(0, nrow(n_all_otus_phylum)*2), ncol=2, dimnames=list(n_all_otus_phylum$Group.1, c('cultured', 'total'))))
-
-	a_otus_phylum[n_all_otus_phylum$Group.1,"total"] <- n_all_otus_phylum$x
-	a_otus_phylum[n_cult_otus_phylum$Group.1,"cultured"] <- n_cult_otus_phylum$x
-
-	#not sure what this is actually getting...
-	b_n1_cult_otus <- 100*sum(table(bact$otu[b_in_cult_otu])==1)/length(table(bact$otu[b_in_cult_otu]))
-	a_n1_cult_otus <- 100*sum(table(arch$otu[a_in_cult_otu])==1)/length(table(arch$otu[a_in_cult_otu]))
+b_percent_cult_otu_all <- 100*sum(b_cult_time[,'all_cult_sobs'])/sum(b_cult_time[,'all_total_sobs'])
+a_percent_cult_otu_all <- 100*sum(a_cult_time[,'all_cult_sobs'])/sum(a_cult_time[,'all_total_sobs'])
 ```
@@
 To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
-Overall, `r b_fract_cult`% of bacterial sequences and `r a_fract_cult`% of archaeal sequences have come from isolated organisms.
-Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_slow_cult` and `r a_slow_cult`-fold for bacteria and archaea, respectively.
-Among the `r length(b_phylum_cult_rate)` bacterial phyla, `r sum(b_phylum_cult_rate == 0)` have no cultured representatives and `r sum(b_phylum_cult_rate == 0)` of the `r length(a_phylum_cult_rate)` archaeal phyla have cultured representatives.
-Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_phylum_cult_rate[b_phylum_cult_rate!=0])*100`% for the bacterial phyla and `r median(a_phylum_cult_rate[a_phylum_cult_rate!=0])*100`% for the archaeal phyla (Figure XXXXXX).
-When we asked what percentage of OTUs had at least one cultured respresentative we found that `r 100*sum(b_otus_phylum$cultured)/sum(b_otus_phylum$total)`% of the `r sum(b_otus_phylum$total)` bacterial OTUs and `r 100*sum(a_otus_phylum$cultured)/sum(a_otus_phylum$total)`% of the `r sum(a_otus_phylum$total)` archeael OTUs had at least one cultured representative.
-Of those OTUs that had a sequence from a culture, `r b_n1_cult_otus`
-and `r a_n1_cult_otus`% of them only contained sequences from cultures. These values suggest that bacterial cultivation efforts are biased towards
+Overall, `r b_percent_cult_nseqs_all`% of bacterial sequences and `r a_percent_cult_nseqs_all`% of archaeal sequences have come from isolated organisms.
+Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_shift_cult_nseqs_all` and `r a_shift_cult_nseqs_all`-fold for bacteria and archaea, respectively.
+Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r nrow(a_cult_time)` of the `r sum(a_cult_time$all_cult_nseqs == 0)` archaeal phyla have cultured representatives.
+This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
+Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure phylum_effort.pdf).
+So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
+Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
+We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative.
+Considering that the percentage of sequences from bacterial cultures is greater than the percentage of OTUs from bacterial cultures, we suspect that bacterial cultivation efforts are largely resampling the same diversity relative to culture-independent approaches.
+In contrast, the relationship is reversed among the archaea indicating that archaeal cultivation efforts continue to sample greater biodiversity than culture-independent approaches.
 
-* need a plot showing % cultured sequences and % cultured OTUs
-	* fraction of sequences from cultured / uncultured bugs (2014) by phylum
-		- b_phylum_cult_rate
-		- a_phylum_cult_rate
-	* % of OTUs with a cultured representative by phylum and in aggregate
-		- b_otus_phylum
-		- a_otus_phylum
+###### here #######
 

### 2015-11-30 code/build_culture_effort_plot.R

@@ -1,0 +1,57 @@
+
+b_cult_time <- read.table(file='data/process/bacteria.cultured_by_time_counts.tsv', header=T)
+b_cult_time <- b_cult_time[b_cult_time$all_total_nseqs >= 1000,]
+rownames(b_cult_time) <- gsub("Candidate_division_", "", rownames(b_cult_time))
+
+b_phyla <- rownames(b_cult_time)
+b_cult_time[,"nseqs_precent"] <- 100*b_cult_time[,"all_cult_nseqs"]/b_cult_time[,"all_total_nseqs"]
+b_cult_time[,"sobs_precent"] <- 100*b_cult_time[,"all_cult_sobs"]/b_cult_time[,"all_total_sobs"]
+b_is_cultured <- b_cult_time$sobs_precent > 0
+
+b_cultured <- b_cult_time[b_is_cultured,colnames(b_cult_time) %in% c("nseqs_precent", "sobs_precent")]
+b_order <- order(b_cultured$sobs_precent, decreasing=T)
+b_cultured <- b_cultured[b_order,]
+b_cultured <- b_cultured[rownames(b_cultured) != "Euryarchaeota",] #needs to be removed
+b_n_taxa <- nrow(b_cultured)
+
+
+
+a_cult_time <- read.table(file='data/process/archaea.cultured_by_time_counts.tsv', header=T)
+a_cult_time <- a_cult_time[a_cult_time$all_total_nseqs >= 1000,]
+
+a_phyla <- rownames(a_cult_time)
+a_cult_time[,"nseqs_precent"] <- 100*a_cult_time[,"all_cult_nseqs"]/a_cult_time[,"all_total_nseqs"]
+a_cult_time[,"sobs_precent"] <- 100*a_cult_time[,"all_cult_sobs"]/a_cult_time[,"all_total_sobs"]
+a_is_cultured <- a_cult_time$sobs_precent > 0
+
+a_cultured <- a_cult_time[a_is_cultured,colnames(a_cult_time) %in% c("nseqs_precent", "sobs_precent")]
+a_order <- order(a_cultured$sobs_precent, decreasing=T)
+a_cultured <- a_cultured[a_order,]
+a_n_taxa <- nrow(a_cultured)
+
+total_lines <- b_n_taxa + a_n_taxa + 2
+
[…]

### 2015-11-30 code/build_phylum_effort_plot.R

@@ -21,7 +21,7 @@ n_arch_ratios <- length(arch_ratios)
 pdf(file="results/figures/phylum_effort.pdf", width=3.75, height=5)
 	par(mar=c(5,10, 1.0, 0.5), cex=0.7)
 
-	plot("", ylim=c(1,n_bact_ratios+n_arch_ratios+2), xlim=c(min(c(bact_ratios, arch_ratios)), ceiling(max(c(bact_ratios, arch_ratios)))), type="n", xlab="", ylab="", axes=F)
+	plot("", ylim=c(1,n_bact_ratios+n_arch_ratios+2), xlim=c(-3,4), type="n", xlab="", ylab="", axes=F)
 
 	abline(v=0, col="gray")
 	abline(h=(n_arch_ratios+3):(n_bact_ratios+n_arch_ratios+2), lty=3, col="gray")
@@ -31,11 +31,11 @@ pdf(file="results/figures/phylum_effort.pdf", width=3.75, height=5)
 	points(x=arch_ratios, y=1:n_arch_ratios)
 	box()
 
-	axis(1, at=-2:4, label=2^(-2:4))
+	axis(1, at=-3:4, label=-3:4)
 	axis(2, at=c((n_arch_ratios+3):(n_bact_ratios+n_arch_ratios+2), 1:n_arch_ratios), label=c(names(bact_ratios), names(arch_ratios)), las=1, tick=F)
 
 	mtext("Bacteria", side=2, line=5, at=n_bact_ratios+n_arch_ratios+3.2, las=1, font=2, cex=0.8)
 	mtext("Archaea", side=2, line=5, at=n_arch_ratios+1.2, las=1, font=2, cex=0.8)
 
-	title(xlab="Fold difference in sequencing\neffort before and after 2006", line=3.5)
+	title(xlab="Fold difference in sequencing effort\nbefore and after 2006 (log2 scale)", line=3.5)
 dev.off()

### 2015-11-30 code/get_culture_by_category_counts.R

@@ -5,10 +5,6 @@ get_phylum <- function(taxonomy){
 	phylum
 }
 
-get_year <- function(date){
-	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date)
-}
-
 is_cultured <- function(db){
 	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db))
 }
@@ -35,7 +31,7 @@ get_domain_data <- function(db, threshold=2006){
 	pre <- get_cultured_data(cultured[pre], phyla[pre], db$otu[pre])
 	post <- get_cultured_data(cultured[!pre], phyla[!pre], db$otu[!pre])
 	pool <- cbind(pre, post, all)
-	colnames(pool) <- paste0(c(rep("pre_", 4), rep("post_", 4), rep("all_", 4)), colnames(pool))
+	colnames(pool) <- paste0(c(rep("pre_", 3), rep("post_", 3), rep("all_", 3)), colnames(pool))
 	pool
 }
 

### 2015-11-30 Schloss_Census2_mBio_2015.Rmd
Polish code to make it render to pdf

@@
 
-#bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)
-
-#arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-#arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
 ```
@@
 In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once.
-Paralleling the bacterial results, after sampling `r nrow(arch)` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1).
+Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1).
 Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once.
@@
 Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C).
-Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences).
+Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(dominant_pmids_2009_2012)` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences).
 These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
@@
 b_new_phyla <- rownames(b_phylum)[b_phylum$pre == 0]
-b_new_phyla_names <- oxford_comma(b_new_phyla_names)
+b_new_phyla_names <- oxford_comma(b_new_phyla)
 
@@
 
-a_new_phyla_names <- oxford_comma(a_new_phyla_names)
+a_new_phyla_names <- oxford_comma(a_new_phyla)
 a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
@@
 In 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
-Since then there have been `r b_new_phyla` new  bacterial (`r b_new_phyla_names`) and `r a_new_phyla` new archaeal candidate phyla (`r a_new_phyla_names`).
+Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`).
 Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure phylum_effort.pdf).
@@
 
-```{r single_cell}
+```{r single_cell, eval=FALSE}
 

### 2015-12-01 Schloss_Census2_mBio_2015.Rmd
Generate Venn diagram data for OTU overlap between methods

@@
 
-###### here #######
-
 ```{r single_cell, eval=FALSE}
 
-grepl("^[^\\.]*$", rownames(bact))
+```
 
-dim(bact[bact$publication_doi=="10.1038/nature12352" & !is.na(bact$publication_doi),])
-#[1] 95 31
-
-```
 ***New technologies to access novel biodiversity***

### 2015-12-01 code/build_otu_overlap_by_method_venn.R

@@ -1,0 +1,47 @@
+require("VennDiagram")
+
+get_category_names <- function(counts){
+		c(paste0("Cultured (N=", format(counts[1], big.mark=','), ")"),
+			paste0("PCR (N=", format(counts[2], big.mark=','), ")"),
+			paste0("Single cell (N=", format(counts[3], big.mark=','), ")"))
+}
+
+count_data <- read.table(file='data/process/otu_overlap_by_method.tsv', header=T)
+
+
+pdf(file="results/figures/venn_otu_by_method.pdf")
+
+layout(matrix(c(1,2), nrow=2), height=c(1,1), width=1)
+
+bact_cats <- get_category_names(count_data[c("cult_all", "pcr_all", "sc_all"), "bact_counts"])
+
+draw.triple.venn(area1=count_data["cult_all","bact_counts"],
+				area2=count_data["pcr_all","bact_counts"],
+				area3=count_data["sc_all","bact_counts"],
+				n12=count_data["cult-pcr","bact_counts"],
+				n23=count_data["pcr-sc","bact_counts"],
+				n13=count_data["cult-sc","bact_counts"],
+				n123=count_data["cult-pcr-sc","bact_counts"],,
+				category=bact_cats,
+				fontfamily = rep("sans", 7),
+				cat.fontfamily = rep("sans", 3),
+				euler.d = FALSE, scaled=FALSE
+				)
+
+arch_cats <- get_category_names(count_data[c("cult_all", "pcr_all", "sc_all"), "arch_counts"])
+
+plot.new()
+draw.triple.venn(area1=count_data["cult_all","arch_counts"],
+				area2=count_data["pcr_all","arch_counts"],
+				area3=count_data["sc_all","arch_counts"],
+				n12=count_data["cult-pcr","arch_counts"],
+				n23=count_data["pcr-sc","arch_counts"],
+				n13=count_data["cult-sc","arch_counts"],
+				n123=count_data["cult-pcr-sc","arch_counts"],,
+				category=arch_cats,
+				fontfamily = rep("sans", 7),
+				cat.fontfamily = rep("sans", 3),
+				euler.d = FALSE, scaled=FALSE
+				)
+
+dev.off()

### 2015-12-01 code/get_otu_overlap_by_method.R

@@ -1,0 +1,58 @@
+is_cultured <- function(db){
+	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db))
+}
+
+is_pcr <- function(db){
+	!((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db))
+}
+
+is_single_cell <- function(db){
+	grepl("^[^.]*$", rownames(db)) | db$publication_doi == "10.1038/nature12352"
+}
+
+
+get_data <- function(db){
+	cultured_otus <- unique(db[is_cultured(db), "otu"])
+	pcr_otus <- unique(db[is_pcr(db), "otu"])
+	sc_otus <- unique(db[is_single_cell(db), "otu"])
+
+
+	total_otus <- max(db$otu)	#Total OTUs
+
+	n_cult_otus <- length(cultured_otus) #Cultured OTUs	1
+	n_pcr_otus <- length(pcr_otus) #PCR OTUs				2
+	n_sc_otus <- length(sc_otus) #Single cell OTUs		3
+
+	n_cult_only <- sum(!cultured_otus %in% union(pcr_otus, sc_otus))
+	n_pcr_only <- sum(!pcr_otus %in% union(cultured_otus, sc_otus))
+	n_sc_only <- sum(!sc_otus %in% union(pcr_otus, cultured_otus))
+
+	n_cult_pcr_otus <- sum(cultured_otus %in% pcr_otus)	#cult-pcr		12
+	n_cult_sc_otus <- sum(cultured_otus %in% sc_otus)	#cult-single	13
+	n_pcr_sc_otus <- sum(pcr_otus %in% sc_otus)			#pcr-single		23
+	n_three_otus <- length(intersect(intersect(cultured_otus, sc_otus), pcr_otus)) #cult-pcr-single
+
+	c("cult_all"=n_cult_otus,
+	"pcr_all"=n_pcr_otus,
+	"sc_all"=n_sc_otus,
+	"cult_only"=n_cult_only,
+	"pcr_only"=n_pcr_only,
+	"sc_only"=n_sc_only,
+	"cult-pcr"=n_cult_pcr_otus,
+	"cult-sc"=n_cult_sc_otus,
+	"pcr-sc"=n_pcr_sc_otus,
+	"cult-pcr-sc"=n_three_otus,
+	"total"=total_otus)
+}
+
+
+
+
+
[…]

### 2015-12-01 code/get_otu_overlap_by_method.R
Add sequence counts for each method

@@ -12,9 +12,14 @@ is_single_cell <- function(db){
 
 
 get_data <- function(db){
-	cultured_otus <- unique(db[is_cultured(db), "otu"])
-	pcr_otus <- unique(db[is_pcr(db), "otu"])
-	sc_otus <- unique(db[is_single_cell(db), "otu"])
+	cultured <- is_cultured(db)
+	cultured_otus <- unique(db[cultured, "otu"])
+
+	pcrd <- is_pcr(db)
+	pcr_otus <- unique(db[pcrd, "otu"])
+
+	sc <- is_single_cell(db)
+	sc_otus <- unique(db[sc, "otu"])
 
 
 	total_otus <- max(db$otu)	#Total OTUs
@@ -32,7 +37,10 @@ get_data <- function(db){
 	n_pcr_sc_otus <- sum(pcr_otus %in% sc_otus)			#pcr-single		23
 	n_three_otus <- length(intersect(intersect(cultured_otus, sc_otus), pcr_otus)) #cult-pcr-single
 
-	c("cult_all"=n_cult_otus,
+	c("n_cult"=sum(cultured),
+	"n_pcr"=sum(pcrd),
+	"n_sc"=sum(sc),
+	"cult_all"=n_cult_otus,
 	"pcr_all"=n_pcr_otus,
 	"sc_all"=n_sc_otus,
 	"cult_only"=n_cult_only,

### 2015-12-01 Schloss_Census2_mBio_2015.Rmd
Fix count of sequences to remove NAs

@@
 Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
-We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative.
+We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure venn_otu_by_method.pdf).
 Considering that the percentage of sequences from bacterial cultures is greater than the percentage of OTUs from bacterial cultures, we suspect that bacterial cultivation efforts are largely resampling the same diversity relative to culture-independent approaches.
@@
 
-```{r single_cell, eval=FALSE}
+```{r single_cell}
+overlap_counts <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
+
+b_per_unique_sc <- 100*overlap_counts["sc_only", "bact_counts"]/overlap_counts["sc_all", "bact_counts"]
+
+a_per_unique_sc <- 100*overlap_counts["sc_only", "arch_counts"]/overlap_counts["sc_all", "arch_counts"]
+
+b_per_shared <- 100*overlap_counts["cult-pcr-sc", "bact_counts"]/overlap_counts["sc_all","bact_counts"]
+
+b_expand_genome <- 100*(overlap_counts["sc_only", "bact_counts"]+overlap_counts["pcr-sc","bact_counts"]-overlap_counts["cult-pcr-sc","bact_counts"])/overlap_counts["sc_all", "bact_counts"]
+
+a_expand_genome <- 100*(overlap_counts["sc_only", "arch_counts"]+overlap_counts["pcr-sc","arch_counts"]-overlap_counts["cult-pcr-sc","arch_counts"])/overlap_counts["sc_all", "arch_counts"]
 
@@
 Assembly of metagenomic and single cell shotgun sequence data offers the hope of identifying large fragments of genomic data from as yet uncultured organisms.
-To test the ability of single cell technologies to identify
-
-  - Single cell genomics
-  - Metagenomes
+To test the ability of single cell technologies to expand our knowledge of microbial diversity beyond that of the 16S rRNA gene and pure cultures, we compared the overlap of OTUs found by the three methods (Figure venn_otu_by_method.pdf).
+Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts['n_sc', 'bact_counts']` bacterial and `r overlap_counts['n_sc', 'arch_counts']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts["sc_all", "bact_counts"]` and `r overlap_counts["sc_all", "arch_counts"]`.
+Interestingly, only `r b_per_unique_sc` and `r a_per_unique_sc`% of the bacterial and archaeal OTUs, respectively, that the single-cell 16S rRNA gene sequences belonged to had not been observed by cultivation or PCR-based efforts.
+For both domains, the fractions of the single-cell 16S rRNA gene sequences that were recovered by cultivation and PCR were similar.
+Furthermore, among the bacteria `r b_per_shared`% of the single-cell OTUs were previously observed by both cultivation and PCR-based methods.
+Although the majority of single-cell genome projects have originated from a single study [@REF], `r b_expand_genome` and `r a_expand_genome` of the bacterial and archaeal OTUs, respectively, were from previously uncultured organisms.
+This represents an encouraging avenue to expanding our knowledge of bacterial diversity beyond the 16S rRNA gene.
 

### 2015-12-01 code/get_otu_overlap_by_method.R

@@ -37,9 +37,9 @@ get_data <- function(db){
 	n_pcr_sc_otus <- sum(pcr_otus %in% sc_otus)			#pcr-single		23
 	n_three_otus <- length(intersect(intersect(cultured_otus, sc_otus), pcr_otus)) #cult-pcr-single
 
-	c("n_cult"=sum(cultured),
-	"n_pcr"=sum(pcrd),
-	"n_sc"=sum(sc),
+	c("n_cult"=sum(cultured, na.rm=T),
+	"n_pcr"=sum(pcrd, na.rm=T),
+	"n_sc"=sum(sc, na.rm=T),
 	"cult_all"=n_cult_otus,
 	"pcr_all"=n_pcr_otus,
 	"sc_all"=n_sc_otus,

### 2015-12-01 Schloss_Census2_mBio_2015.Rmd
Fix typo

@@
 To test the ability of single cell technologies to expand our knowledge of microbial diversity beyond that of the 16S rRNA gene and pure cultures, we compared the overlap of OTUs found by the three methods (Figure venn_otu_by_method.pdf).
-Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts['n_sc', 'bact_counts']` bacterial and `r overlap_counts['n_sc', 'arch_counts']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts["sc_all", "bact_counts"]` and `r overlap_counts["sc_all", "arch_counts"]`.
+Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts['n_sc', 'bact_counts']` bacterial and `r overlap_counts['n_sc', 'arch_counts']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts["sc_all", "bact_counts"]` and `r overlap_counts["sc_all", "arch_counts"]` bacterial and archaeal OTUs, respectively.
 Interestingly, only `r b_per_unique_sc` and `r a_per_unique_sc`% of the bacterial and archaeal OTUs, respectively, that the single-cell 16S rRNA gene sequences belonged to had not been observed by cultivation or PCR-based efforts.

### 2015-12-01 code/build_otu_overlap_by_method_venn.R
Fix venn diagram; Closes #11

@@ -1,4 +1,42 @@
-require("VennDiagram")
+require("plotrix")
+
+pds_venn <- function(A, B, C, AB, AC, BC, ABC, categories,
+	cir.lwd = 2
+){
+
+	par(mar=c(1,2,3,2), cex=0.7)
+	plot(NA, xlim=c(0,1), ylim=c(0,1), axes=F, xlab="", ylab="")
+
+	radius <- 0.325
+	draw.circle(radius, 1-radius, radius, lwd=cir.lwd)	#A
+	draw.circle(1-radius, 1-radius, radius, lwd=cir.lwd)	#B
+	draw.circle(0.5, radius, radius, lwd=cir.lwd)	#C
+
+	n_ABC <- ABC
+	n_A <- A - AB - AC + n_ABC
+	n_B <- B - AB - BC + n_ABC
+	n_C <- C - AC - BC + n_ABC
+
+	n_AB <- AB - n_ABC
+	n_AC <- AC - n_ABC
+	n_BC <- BC - n_ABC
+
+	text(x=0.2, y=0.7, format(n_A, big.mark=','))
+	text(x=0.8, y=0.7, format(n_B, big.mark=','))
+	text(x=0.5, y=0.2, format(n_C, big.mark=','))
+
+	text(x=0.3, y=0.43, format(n_AC, big.mark=','))
+	text(x=0.7, y=0.43, format(n_BC, big.mark=','))
+	text(x=0.5, y=0.75, format(n_AB, big.mark=','))
+
+	text(x=0.5, y=0.56, format(n_ABC, big.mark=','), pos=1)
+
+	text(x=-0.025, y=1.025, categories[1], font=2, adj=0, xpd=T)
+	text(x=1.025, y=1.025, categories[2], font=2, adj=1, xpd=T)
+	text(x=0.5, y=0, categories[3], font=2, pos=1, xpd=T)
+
+}
+
 
 get_category_names <- function(counts){
 		c(paste0("Cultured (N=", format(counts[1], big.mark=','), ")"),
@@ -6,42 +44,37 @@ get_category_names <- function(counts){
 			paste0("Single cell (N=", format(counts[3], big.mark=','), ")"))
 }
 
+
 count_data <- read.table(file='data/process/otu_overlap_by_method.tsv', header=T)
 
 
-pdf(file="results/figures/venn_otu_by_method.pdf")
-
[…]

### 2015-12-01 code/build_phylum_category_heatmap.R
Fix order of categories

@@ -2,7 +2,7 @@ require("RColorBrewer")
 
 categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other", total="Total")
 
-rev_categories <- length(categories):1
+rev_categories <- rev(names(categories))
 
 data <- read.table(file="data/process/phylum_category_counts.tsv", header=T)
 

### 2015-12-02 code/get_emigre_fasta.sh
Scrape journals and lab websites for EMIGRE data

@@ -1,0 +1,71 @@
+# EMIRGE: reconstruction of full-length ribosomal genes from microbial community short read sequencing data.
+# Miller CS, Baker BJ, Thomas BC, Singer SW, Banfield JF.
+# Genome Biol. 2011;12(5):R44. doi: 10.1186/gb-2011-12-5-r44. Epub 2011 May 19.
+# PMID: 21595876
+# http://www.genomebiology.com/2011/12/5/R44
+
+wget http://www.genomebiology.com/content/supplementary/gb-2011-12-5-r44-s3.fasta -O data/raw/miller_2011_emigre.fasta
+
+
+
+# Short-read assembly of full-length 16S amplicons reveals bacterial diversity in subsurface sediments.
+# Miller CS, Handley KM, Wrighton KC, Frischkorn KR, Thomas BC, Banfield JF.
+# PLoS One. 2013;8(2):e56018. doi: 10.1371/journal.pone.0056018. Epub 2013 Feb 6.
+# PMID: 23405248
+
+wget http://banfieldlab.berkeley.edu/EMIRGE/data/paper2/EMIRGE_reconstructed_16S.fasta.gz -O data/raw/miller_2013_emigre.fasta.gz
+gunzip data/raw/miller_2013_emigre.fasta.gz
+
+
+
+# Fermentation, hydrogen, and sulfur metabolism in multiple uncultivated bacterial phyla.
+# Wrighton KC, Thomas BC, Sharon I, Miller CS, Castelle CJ, VerBerkmoes NC, Wilkins MJ, Hettich RL, Lipton MS, Williams KH, Long PE, Banfield JF.
+# Science. 2012 Sep 28;337(6102):1661-5. Erratum in: Science. 2012 Nov 9;338(6108):742.
+# PMID: 23019650
+
+wget http://www.sciencemag.org/content/suppl/2012/09/26/337.6102.1661.DC1/1224041_SupplmentalDownload.zip
+unzip 1224041_SupplmentalDownload.zip FASTA_files/16S_EMIRGE_CloneSilva108db_p0.005.fasta
+mv FASTA_files/16S_EMIRGE_CloneSilva108db_p0.005.fasta data/raw/wrighton_2012_emigre.fasta
+rm -rf FASTA_files 1224041_SupplmentalDownload.zip
+
+
+
[…]

### 2015-12-03 Schloss_Census2_mBio_2015.Rmd

@@
 
-**Figure 1. Sequences deposited by year**
+**Figure results/figures/domain_rarefaction.pdf.
+Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** The number of bacterial and archaeal OTUs observed among the longest sequences in the SILVA database continues to grow at a rate too slow to ever reach estimates of 10^6^ to 10^11^ bacterial species.
 
-Pat, this is a placeholder for you
+**Figure results/figures/time_course_figure.pdf.
+Progression of the microbial census since the first full-length 16S rRNA gene sequence was deposited into GenBank in 1983.*** The number of bacterial and archaeal 16S rRNA gene sequences deposited (A) and the new OTUs they represent (B) has increased exponentially until the last several years when the rate of change has plateaued.
+For both bacterial and archaeal sequences, the number of studies that are responsible for depositing more than 50% of the seuqences each year has been relatively small (C).
 
-**Figure 2. Sequence deposition by study**
+**Figure results/figures/category_phylum_heatmap.pdf.
+Heatmap depicting the relative abundance of the most common bacterial and archaeal phyla across different environments.** Each environmental category exhibited a phylum-level signature although the bacterial census was dominated by sequences from the Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes and the archaeal census was dominated by sequences from the Euryarchaeota and Thaumarchaeota.
 
-Pat, ibid.
+**Figure results/figures/phylum_effort.pdf.
+Relative rate of sequence deposition for each bacterial and archaeal phylum before and after 2006 relative to the sequencing of all bacteria.** The figure shows the relative rates for those phyla with at least 1,000 sequences and the x-axis is on a log2 scale.
 
-**Figure 3. Sampling by OTU cutoff**
+**Figure results/figures/phylum_cultured.pdf.
+The rate that sequences and OTUs are generated from bacterial and archaeal cultures relative to all sequences and OTUs.** The phyla shown have at least 1,000 total sequences.
 
-I only have an old code from Rene for this, and I don't think it reports all the sequences
-
-**Figure 4A. Bacterial sampling by environment**
+**Figure results/figures/venn_otu_by_method.pdf. The number of OTUs that were found by either cultivation, PCR, single-cell genomics or multiple methods for bacterial and archaeal sequences.**
 

### 2015-12-03 results/tables/build_table_1.Rmd

@@
 
-bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+data <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
+int_cols <- grepl("count", colnames(data)) | grepl("sobs", colnames(data))
+float_cols <- !int_cols
 
-arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
+data[,int_cols] <- format(data[,int_cols], big.mark=',')
+data[,float_cols] <- format(data[,float_cols], big.mark=',', nsmall=1, digits=1)
+
+data_current <- data[,!(grepl("pre", colnames(data)) | grepl("post", colnames(data)))]
 
@@
 
-abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQM", "AQMS", "AQH", "AQI", "AQO", "BD", "BF", "BI", "BP", "BO", "PR", "PS", "PO", "SA", "SD", "SP", "SO", "ZV", "ZA", "ZN",  "ZO", "OT")
+composite <- cbind(coarse, fine, data_current)
 
-get_sobs <- function(x, db){
-	if(x == "NA"){
-		length(table(db[is.na(db$category), "otu"]))
-	} else if(x == "total"){
-		length(table(db[, "otu"]))
-	} else {
-		length(table(db[db$category==x, "otu"]))
-	}
-}
+header <- c("", "", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
+\\\\ Coverage}", "\\bigcell{c}{\\% OTU \\\\ Coverage}", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
+\\\\ Coverage}", "\\bigcell{c}{\\% OTU \\\\ Coverage}")
 
-get_goods_coverage <- function(x, db){
-	if(x=="NA"){
-		x_table <- table(db[is.na(db$category), "otu"])
-	} else if(x == "total") {
-		x_table <- table(db[, "otu"])
-	} else {
-		x_table <- table(db[db$category==x, "otu"])
-	}
-
-	singletons <- sum(x_table == 1)
-	total_seqs <- sum(x_table)
-
-	if(total_seqs != 0){
-		100*(1-singletons/total_seqs)
-	} else {
-		NA
-	}
-
-}
-
-get_schloss_coverage <- function(x, db){
-	if(x=="NA"){
-		x_table <- table(db[is.na(db$category), "otu"])
-	} else if(x == "total") {
-		x_table <- table(db[, "otu"])
-	} else {
-		x_table <- table(db[db$category==x, "otu"])
-	}
-
-	singletons <- sum(x_table == 1)
-	total_otus <- length(x_table)
-
-	if(total_otus != 0){
-		100*(1-singletons/total_otus)
-	} else {
-		NA
-	}
-
-}
-
-bact_count <- table(bact$category)[abbreviation]
-bact_count[is.na(bact_count)] <- 0
-bact_count <- c(bact_count, nrow(bact) - sum(bact_count))
-bact_count <- format(c(bact_count, nrow(bact)), big.mark=',')
-bact_sobs <- format(sapply(c(abbreviation, "NA", "total"), get_sobs, bact), big.mark=',')
-bact_good_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_goods_coverage, bact), nsmall=1, digits=3)
-bact_schloss_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_schloss_coverage, bact), nsmall=1, digits=3)
-
-arch_count <- table(arch$category)[abbreviation]
-arch_count[is.na(arch_count)] <- 0
-arch_count <- c(arch_count, nrow(arch) - sum(arch_count))
-arch_count <- format(c(arch_count, nrow(arch)), big.mark=',')
-arch_sobs <- format(sapply(c(abbreviation, "NA", "total"), get_sobs, arch), big.mark=',')
-arch_good_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_goods_coverage, arch), nsmall=1, digits=3)
-arch_schloss_coverage <- format(sapply(c(abbreviation, "NA", "total"), get_schloss_coverage, arch), nsmall=1, digits=3)
-
-#abbreviation <- c(abbreviation, "", "")
-
-composite <- cbind(coarse, fine, bact_count, bact_sobs, bact_good_coverage, bact_schloss_coverage, arch_count, arch_sobs, arch_good_coverage, arch_schloss_coverage)
-write.table(composite, file=paste0(PROJHOME, "/data/process/table1_data.tsv"))
-
-composite <- rbind(c("", "", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
-\\\\ Coverage}", "\\bigcell{c}{\\% OTU \\\\ Coverage}", "\\bigcell{c}{Sequences \\\\ (N)}", "\\bigcell{c}{OTUs \\\\ (N)}", "\\bigcell{c}{\\% Seq.
-\\\\ Coverage}", "\\bigcell{c}{\\% OTU \\\\ Coverage}"), composite)
+composite <- rbind(header, composite)
 

### 2015-12-03 code/get_emigre_fasta.sh
Fix backslash exiting problem

@@ -37,7 +37,7 @@ rm -rf FASTA_files 1224041_SupplmentalDownload.zip
 
 wget "http://journals.plos.org/plosone/article/asset?unique&id=info:doi/10.1371/journal.pone.0057819.s003" -O data/raw/wilkins_2013.docx
 pandoc -s data/raw/wilkins_2013.docx -t markdown -o data/raw/wilkins_2013.md
-R -e "write(gsub('[\\\\\]', '', scan(file='data/raw/wilkins_2013.md', what='', sep='\\\n', quiet=T))[-1], 'data/raw/wilkins_2013_emigre.fasta')"
+R -e "write(gsub('[\\\\]', '', scan(file='data/raw/wilkins_2013.md', what='', sep='\\n', quiet=T))[-1], 'data/raw/wilkins_2013_emigre.fasta')"
 rm data/raw/wilkins_2013.docx data/raw/wilkins_2013.md
 
 

### 2015-12-03 code/get_emigre_fasta.sh
Fix docx to fasta conversion

@@ -36,7 +36,7 @@ rm -rf FASTA_files 1224041_SupplmentalDownload.zip
 # PMID: 23472107
 
 wget "http://journals.plos.org/plosone/article/asset?unique&id=info:doi/10.1371/journal.pone.0057819.s003" -O data/raw/wilkins_2013.docx
-pandoc -s data/raw/wilkins_2013.docx -t markdown -o data/raw/wilkins_2013.md
+pandoc -s data/raw/wilkins_2013.docx -t asciidoc -o data/raw/wilkins_2013.md
 R -e "write(gsub('[\\\\]', '', scan(file='data/raw/wilkins_2013.md', what='', sep='\\n', quiet=T))[-1], 'data/raw/wilkins_2013_emigre.fasta')"
 rm data/raw/wilkins_2013.docx data/raw/wilkins_2013.md
 

### 2015-12-03 code/get_emirge_fasta.sh
Fix misspellings of emirge

@@ -4,7 +4,7 @@
 # PMID: 21595876
 # http://www.genomebiology.com/2011/12/5/R44
 
-wget http://www.genomebiology.com/content/supplementary/gb-2011-12-5-r44-s3.fasta -O data/raw/miller_2011_emigre.fasta
+wget http://www.genomebiology.com/content/supplementary/gb-2011-12-5-r44-s3.fasta -O data/raw/miller_2011_emirge.fasta
 
 
 
@@ -13,8 +13,8 @@ wget http://www.genomebiology.com/content/supplementary/gb-2011-12-5-r44-s3.fast
 # PLoS One. 2013;8(2):e56018. doi: 10.1371/journal.pone.0056018. Epub 2013 Feb 6.
 # PMID: 23405248
 
-wget http://banfieldlab.berkeley.edu/EMIRGE/data/paper2/EMIRGE_reconstructed_16S.fasta.gz -O data/raw/miller_2013_emigre.fasta.gz
-gunzip data/raw/miller_2013_emigre.fasta.gz
+wget http://banfieldlab.berkeley.edu/EMIRGE/data/paper2/EMIRGE_reconstructed_16S.fasta.gz -O data/raw/miller_2013_emirge.fasta.gz
+gunzip data/raw/miller_2013_emirge.fasta.gz
 
 
 
@@ -25,7 +25,7 @@ gunzip data/raw/miller_2013_emigre.fasta.gz
 
 wget http://www.sciencemag.org/content/suppl/2012/09/26/337.6102.1661.DC1/1224041_SupplmentalDownload.zip
 unzip 1224041_SupplmentalDownload.zip FASTA_files/16S_EMIRGE_CloneSilva108db_p0.005.fasta
-mv FASTA_files/16S_EMIRGE_CloneSilva108db_p0.005.fasta data/raw/wrighton_2012_emigre.fasta
+mv FASTA_files/16S_EMIRGE_CloneSilva108db_p0.005.fasta data/raw/wrighton_2012_emirge.fasta
 rm -rf FASTA_files 1224041_SupplmentalDownload.zip
 
 
@@ -37,7 +37,7 @@ rm -rf FASTA_files 1224041_SupplmentalDownload.zip
 
 wget "http://journals.plos.org/plosone/article/asset?unique&id=info:doi/10.1371/journal.pone.0057819.s003" -O data/raw/wilkins_2013.docx
[…]

### 2015-12-03 code/get_emirge_metadata.sh

@@ -4,7 +4,7 @@
 # PMID: 21595876
 # http://www.genomebiology.com/2011/12/5/R44
 
-R -e "source('code/build_metadata.R'); build_metadata('data/raw/miller_2011_emigre.fasta', pub_date='2011-05-19', publication_doi='10.1186/gb-2011-12-5-r44', category='BI')"
+R -e "source('code/build_metadata.R'); build_metadata('data/raw/miller_2011_emirge.fasta', pub_date='2011-05-19', publication_doi='10.1186/gb-2011-12-5-r44', category='BI')"
 
 
 
@@ -14,7 +14,7 @@ R -e "source('code/build_metadata.R'); build_metadata('data/raw/miller_2011_emig
 # PLoS One. 2013;8(2):e56018. doi: 10.1371/journal.pone.0056018. Epub 2013 Feb 6.
 # PMID: 23405248
 
-R -e "source('code/build_metadata.R'); build_metadata('data/raw/miller_2013_emigre.fasta', pub_date='2013-02-06', publication_doi='10.1371/journal.pone.0056018', category='AQF')"
+R -e "source('code/build_metadata.R'); build_metadata('data/raw/miller_2013_emirge.fasta', pub_date='2013-02-06', publication_doi='10.1371/journal.pone.0056018', category='AQF')"
 
 
 
@@ -23,7 +23,7 @@ R -e "source('code/build_metadata.R'); build_metadata('data/raw/miller_2013_emig
 # Science. 2012 Sep 28;337(6102):1661-5. Erratum in: Science. 2012 Nov 9;338(6108):742.
 # PMID: 23019650
 
-R -e "source('code/build_metadata.R'); build_metadata('data/raw/wrighton_2012_emigre.fasta', pub_date='2012-11-09', publication_doi='10.1126/science.1224041', category='AQF')"
+R -e "source('code/build_metadata.R'); build_metadata('data/raw/wrighton_2012_emirge.fasta', pub_date='2012-11-09', publication_doi='10.1126/science.1224041', category='AQF')"
 
 
 
@@ -32,7 +32,7 @@ R -e "source('code/build_metadata.R'); build_metadata('data/raw/wrighton_2012_em
 # PLoS One. 2013;8(3):e57819. doi: 10.1371/journal.pone.0057819. Epub 2013 Mar 5.
 # PMID: 23472107
 
[…]

### 2015-12-03 code/get_emirge_fasta.sh
Use RDP taxnomy to classify sequences to the proper domain

@@ -67,5 +67,23 @@ mv "Supplementary Material BSC_EMIRGE_outputs/BSC_Metagenome_SSUrRNA_EMIRGEoutpu
 rm -rf "Supplementary Material BSC_EMIRGE_outputs" DataSheet1.ZIP
 
 
-
+# Pool the fasta data
 cat data/raw/*_*_emirge.fasta > data/raw/emirge.fasta
+
+
+# Now we want to classify the sequences to split the sequences by domain...
+mothur "#classify.seqs(fasta=data/raw/emirge.fasta, reference=data/references/trainset10_082014.pds.fasta, taxonomy=data/references/trainset10_082014.pds.tax, cutoff=80, processors=8)"
+
+
+# Pull out the bacteria...
+mothur "#get.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Bacteria)"
+
+mv data/raw/emirge.pds.wang.pick.taxonomy data/mothur/emirge.bacteria.taxonomy
+mv data/raw/emirge.pick.fasta data/mothur/emirge.bacteria.fasta
+
+
+# Pull out the archaea...
+mothur "#get.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Archaea)"
+
+mv data/raw/emirge.pds.wang.pick.taxonomy data/mothur/emirge.archaea.taxonomy
+mv data/raw/emirge.pick.fasta data/mothur/emirge.archaea.fasta

### 2015-12-03 code/get_emirge_fasta.sh
Remove the separated taxonomy files

@@ -78,12 +78,12 @@ mothur "#classify.seqs(fasta=data/raw/emirge.fasta, reference=data/references/tr
 # Pull out the bacteria...
 mothur "#get.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Bacteria)"
 
-mv data/raw/emirge.pds.wang.pick.taxonomy data/mothur/emirge.bacteria.taxonomy
+rm data/raw/emirge.pds.wang.pick.taxonomy
 mv data/raw/emirge.pick.fasta data/mothur/emirge.bacteria.fasta
 
 
 # Pull out the archaea...
 mothur "#get.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Archaea)"
 
-mv data/raw/emirge.pds.wang.pick.taxonomy data/mothur/emirge.archaea.taxonomy
+rm data/raw/emirge.pds.wang.pick.taxonomy
 mv data/raw/emirge.pick.fasta data/mothur/emirge.archaea.fasta

### 2015-12-03 doc/notebook/Bacterial_data_acquisition.Rmd
Add steps to process extra EMIRGE data

@@
         classify.seqs(fasta=current, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80)"
+```
+
+## EMIRGE sequence data
+
+EMIRGE is a method of either assembling 16S rRNA gene sequences from metagenomic
+data or from sheared amplicons. This is a method of generating full-length
+sequences using short read technologies. We've identified a handful of studies
+that have used EMIRGE to generate these data and that actually made their data
+publicly available. Some of these are already in the SILVA database, but many
+are only available in supplementary materials or on lab websites. To generate
+the pooled fasta and metadata files we run the following. Note that because
+we're scraping data out of some supplements these scripts assume that you have
+access to the journals...
+
+```{r get_emirge, engine="bash"}
+bash code/get_emirge_fasta.sh
+bash code/get_emirge_metadata.sh
+```
+
+This will bring in 81,616 sequences: 81608 from bacteria, 7 from archaea, and 1 from a eukaryote.
+We'll ignore the eukaryote.
+The `code/get_emirge_fasta.sh` script split the sequences by domain and put the fasta sequences into `data/mothur/` for further processing.
+We need to process these sequences similar to what we did with the single cell data.
+
+```{r process_emirge, engine="bash"}
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        align.seqs(fasta=data/mothur/emirge.bacteria.fasta, reference=data/mothur/bacteria.good.fasta);
+        screen.seqs(fasta=current, start=2000, end=14459, maxambig=2, minlength=1200);
+        classify.seqs(fasta=data/raw/EMIRGE.fasta, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80, processors=8)"
 ```
@@
 
+#need to change the name of this sequence to make it clear that it was not cultured
+rownames(metadata)[rownames(metadata) == 'AJ224540.MxeCultu'] <- 'AJ224540.UncMxeCultu'
+
 #... and we output the metadata file...

### 2015-12-03 code/get_emirge_fasta.sh
Remove Mitochondria and Chloroplast sequences as well

@@ -76,14 +76,14 @@ mothur "#classify.seqs(fasta=data/raw/emirge.fasta, reference=data/references/tr
 
 
 # Pull out the bacteria...
-mothur "#get.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Bacteria)"
+mothur "#remove.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Archaea-Eukaryota-Chloroplast-Mitochondria)"
 
 rm data/raw/emirge.pds.wang.pick.taxonomy
 mv data/raw/emirge.pick.fasta data/mothur/emirge.bacteria.fasta
 
 
 # Pull out the archaea...
-mothur "#get.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Archaea)"
+mothur "#remove.lineage(fasta=data/raw/emirge.fasta, taxonomy=data/raw/emirge.pds.wang.taxonomy, taxon=Bacteria-Eukaryota-Chloroplast-Mitochondria)"
 
 rm data/raw/emirge.pds.wang.pick.taxonomy
 mv data/raw/emirge.pick.fasta data/mothur/emirge.archaea.fasta

### 2015-12-04 doc/notebook/Bacterial_data_acquisition.Rmd
Add EMIRGE code to work flow

@@
         set.dir(input=data/mothur);
-        screen.seqs(fasta=current, start=2000, end=13000, maxambig=2, minlength=1200);
+        screen.seqs(fasta=current, start=2000, end=14459, maxambig=2, minlength=1200);
         classify.seqs(fasta=current, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80)"
 ```
+
+Once we run everything through, we see that there were 199 sequences that met our `screen.seqs` criteria.
 
@@
 
-This will bring in 81,616 sequences: 81608 from bacteria, 7 from archaea, and 1 from a eukaryote.
+This will bring in 81,616 sequences: 81,608 from bacteria, 7 from archaea, and 1 from a eukaryote.
 We'll ignore the eukaryote.
@@
 mothur "#set.dir(input=data/mothur, output=data/mothur);
-        align.seqs(fasta=data/mothur/emirge.bacteria.fasta, reference=data/mothur/bacteria.good.fasta);
+        align.seqs(fasta=data/mothur/emirge.bacteria.fasta, reference=data/references/bacteria.seed.fasta);
         screen.seqs(fasta=current, start=2000, end=14459, maxambig=2, minlength=1200);
-        classify.seqs(fasta=data/raw/EMIRGE.fasta, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80, processors=8)"
+        classify.seqs(fasta=current, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80, processors=8)"
 ```
+
+Once we run everything through, we see that there were 58,975 sequences that met our `screen.seqs` criteria.
 
@@
 collections. We'll also filter the sequences to overlap in the same alignment
-space, unique them, and precluster them...
+space, `unique.seqs` them, and `pre.cluster` them...
 
 ```{r pool_data, engine="bash"}
-cat data/mothur/bacteria.good.fasta data/mothur/single_cell.bacteria.good.align > data/mothur/all_bacteria.align
-cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.good.good.wang.taxonomy > data/mothur/all_bacteria.taxonomy
+cat data/mothur/bacteria.good.fasta data/mothur/single_cell.bacteria.good.align data/mothur/emirge.bacteria.good.align > data/mothur/all_bacteria.align
+cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.good.good.wang.taxonomy data/mothur/emirge.bacteria.good.good.wang.taxonomy > data/mothur/all_bacteria.taxonomy
 

### 2015-12-04 code/get_silva_seed.sh

@@ -1,0 +1,16 @@
+wget http://mothur.org/w/images/1/15/Silva.seed_v123.tgz -O data/references/Silva.seed_v123.tgz
+tar xvzf data/references/Silva.seed_v123.tgz silva.seed_v123.tax
+mv silva.seed_v123.tax data/references/
+rm data/references/Silva.seed_v123.tgz
+
+grep "Bacteria;" data/references/silva.seed_v123.tax | cut -f 1 > data/references/silva.seed_v123.bacteria.accnos
+
+mothur "#get.seqs(fasta=data/mothur/bacteria.fasta, accnos=data/references/silva.seed_v123.bacteria.accnos, outputdir=data/references)"
+
+mv data/references/bacteria.pick.fasta data/references/bacteria.seed.fasta
+
+#grep "Archaea;" data/references/silva.seed_v123.tax | cut -f 1 > data/references/silva.seed_v123.archaea.accnos
+#
+#mothur "#get.seqs(fasta=data/mothur/archaea.fasta, accnos=data/references/silva.seed_v123.archaea.accnos, outputdir=data/references)"
+#
+#align the archaeal sequences against data/mothur/archaea.fasta

### 2015-12-04 doc/notebook/Archaeal_data_acquisition.Rmd
Add EMIGRE code to work flow

@@
         align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
-        screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
+        screen.seqs(fasta=current, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900);
         classify.seqs(fasta=current, reference=archaea.good.fasta, taxonomy=archaea.good.tax, cutoff=80, processors=8)"
 ```
+
+Once we run everything through, we see that there were 44 sequences that met our `screen.seqs` criteria.
+
+## EMIRGE sequence data
+
+EMIRGE is a method of either assembling 16S rRNA gene sequences from metagenomic
+data or from sheared amplicons. This is a method of generating full-length
+sequences using short read technologies. We've identified a handful of studies
+that have used EMIRGE to generate these data and that actually made their data
+publicly available. Some of these are already in the SILVA database, but many
+are only available in supplementary materials or on lab websites. To generate
+the pooled fasta and metadata files we run the following. Note that because
+we're scraping data out of some supplements these scripts assume that you have
+access to the journals...
+
+```{r get_emirge, engine="bash"}
+bash code/get_emirge_fasta.sh
+bash code/get_emirge_metadata.sh
+```
+
+This will bring in 81,616 sequences: 81,608 from bacteria, 7 from archaea, and 1 from a eukaryote.
+We'll ignore the eukaryote.
+The `code/get_emirge_fasta.sh` script split the sequences by domain and put the fasta sequences into `data/mothur/` for further processing.
+We need to process these sequences similar to what we did with the single cell data.
+Because we're worried about the archaea for this notebook and there are so few, we'll process them slightly differently from the bacterial sequences by aligning them to `archaea.good.fasta`.
+
+```{r process_emirge, engine="bash"}
+mothur "#set.dir(input=data/mothur, output=data/mothur);
+        align.seqs(fasta=data/mothur/emirge.archaea.fasta, reference=data/mothur/archaea.good.fasta);
+        screen.seqs(fasta=current, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900);
+        classify.seqs(fasta=current, reference=data/mothur/archaea.good.fasta, taxonomy=data/mothur/archaea.good.tax, cutoff=80, processors=8)"
+```
+
+Once we run everything through, we see that there were 6 sequences that met our `screen.seqs` criteria.
 
@@
 ```{r pool_data, engine="bash"}
-cat data/mothur/archaea.good.fasta data/mothur/single_cell.archaea.good.align > data/mothur/all_archaea.align
-cat data/mothur/archaea.good.tax data/mothur/single_cell.archaea.good.pds.wang.taxonomy > data/mothur/all_archaea.taxonomy
+cat data/mothur/archaea.good.fasta data/mothur/single_cell.archaea.good.align data/mothur/emirge.archaea.align > data/mothur/all_archaea.align
+cat data/mothur/archaea.good.tax data/mothur/single_cell.archaea.good.good.wang.taxonomy data/mothur/emirge.archaea.good.good.wang.taxonomy > data/mothur/all_archaea.taxonomy
 

### 2015-12-04 doc/notebook/Archaeal_data_acquisition.Rmd
Fix screen.seqs settings and clean up taxonomy

@@
 ```{r screen_data, engine="bash"}
-mothur "#set.current(processors=8);
-        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900, inputdir=data/mothur/, outputdir=data/mothur/);"
+mothur "#screen.seqs(fasta=data/mothur/archaea.fasta, taxonomy=data/mothur/archaea.tax, start=3000, end=7800, maxambig=2, maxhomop=8, processors=8);"
 ```
@@
         align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
-        screen.seqs(fasta=current, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900);
-        classify.seqs(fasta=current, reference=archaea.good.fasta, taxonomy=archaea.good.tax, cutoff=80, processors=8)"
+        screen.seqs(fasta=current, start=3000, end=7800, maxambig=2, maxhomop=8);
+        classify.seqs(fasta=current, reference=archaea.good.fasta, taxonomy=archaea.good.tax, cutoff=80)"
 ```
@@
 mothur "#set.dir(input=data/mothur, output=data/mothur);
-        align.seqs(fasta=data/mothur/emirge.archaea.fasta, reference=data/mothur/archaea.good.fasta);
-        screen.seqs(fasta=current, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900);
-        classify.seqs(fasta=current, reference=data/mothur/archaea.good.fasta, taxonomy=data/mothur/archaea.good.tax, cutoff=80, processors=8)"
+        align.seqs(fasta=emirge.archaea.fasta, reference=archaea.good.fasta);
+        screen.seqs(fasta=current, start=3000, end=7800, maxambig=2, maxhomop=8);
+        classify.seqs(fasta=current, reference=archaea.good.fasta, taxonomy=archaea.good.tax, cutoff=80)"
 ```
@@
 
+Before we can press on with clustering, we need to clean up the taxonomy files that we generated for the EMIRGE and single cell data a bit. We'll do this with R
+
+```{r tax_clean}
+tax_clean <- function(tax_file, out_file){
+	tax <- scan(file=tax_file, sep='\n', what="", quiet=T)
+	tax_clean <- gsub("\\(\\d*\\);", ";", tax)
+	tax_clean <- gsub("unclassified;", "", tax_clean)
+	write(tax_clean, file=out_file)
+}
+
+tax_clean('data/mothur/single_cell.archaea.good.good.wang.taxonomy', 'data/mothur/single_cell.archaea.taxonomy')
+
+tax_clean('data/mothur/emirge.archaea.good.good.wang.taxonomy', 'data/mothur/emirge.archaea.taxonomy')
+```
+
 Now we want to merge the SILVA and single cell genomics 16S rRNA gene sequence
@@
 ```{r pool_data, engine="bash"}
-cat data/mothur/archaea.good.fasta data/mothur/single_cell.archaea.good.align data/mothur/emirge.archaea.align > data/mothur/all_archaea.align
-cat data/mothur/archaea.good.tax data/mothur/single_cell.archaea.good.good.wang.taxonomy data/mothur/emirge.archaea.good.good.wang.taxonomy > data/mothur/all_archaea.taxonomy
+cat data/mothur/archaea.good.fasta data/mothur/single_cell.archaea.good.align data/mothur/emirge.archaea.good.align > data/mothur/all_archaea.align
+cat data/mothur/archaea.good.tax data/mothur/single_cell.archaea.taxonomy data/mothur/emirge.archaea.taxonomy > data/mothur/all_archaea.taxonomy
 
@@
         unique.seqs();
-        pre.cluster(fasta=current, name=current, diffs=9);"
+        pre.cluster(fasta=current, name=current, diffs=5);"
 
@@
 ```
+
+XXXXXXXXXXXXXXXX
 

### 2015-12-04 doc/notebook/Bacterial_data_acquisition.Rmd
Fix screen.seqs settings and clean up taxonomy

@@
 ```{r screen_data, engine="bash"}
-mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=14459, maxambig=2, minlength=1200, outputdir=data/mothur/, processors=8);"
+mothur "#screen.seqs(fasta=data/mothur/bacteria.fasta, taxonomy=data/mothur/bacteria.tax, start=2000, end=14459, maxambig=2, maxhomop=8, outputdir=data/mothur/, processors=8);"
 ```
@@
         set.dir(input=data/mothur);
-        screen.seqs(fasta=current, start=2000, end=14459, maxambig=2, minlength=1200);
+        screen.seqs(fasta=current, start=2000, end=14459, maxambig=2);
         classify.seqs(fasta=current, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80)"
@@
         align.seqs(fasta=data/mothur/emirge.bacteria.fasta, reference=data/references/bacteria.seed.fasta);
-        screen.seqs(fasta=current, start=2000, end=14459, maxambig=2, minlength=1200);
+        screen.seqs(fasta=current, start=2000, end=14459, maxambig=2);
         classify.seqs(fasta=current, reference=data/mothur/bacteria.good.fasta, taxonomy=data/mothur/bacteria.good.tax, cutoff=80, processors=8)"
@@
 
+XXXXXXXXXXXXXXXX
+
 ## Pooling the data
+
+Before we can press on with clustering, we need to clean up the taxonomy files that we generated for the EMIRGE and single cell data a bit. We'll do this with R
+
+```{r tax_clean}
+tax_clean <- function(tax_file, out_file){
+	tax <- scan(file=tax_file, sep='\n', what="", quiet=T)
+	tax_clean <- gsub("\\(\\d*\\);", ";", tax)
+	tax_clean <- gsub("unclassified;", "", tax_clean)
+	write(tax_clean, file=out_file)
+}
+
+tax_clean('data/mothur/single_cell.bacteria.good.good.wang.taxonomy', 'data/mothur/single_cell.bacteria.taxonomy')
+
+tax_clean('data/mothur/emirge.bacteria.good.good.wang.taxonomy', 'data/mothur/emirge.bacteria.taxonomy')
+```
 
@@
 cat data/mothur/bacteria.good.fasta data/mothur/single_cell.bacteria.good.align data/mothur/emirge.bacteria.good.align > data/mothur/all_bacteria.align
-cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.good.good.wang.taxonomy data/mothur/emirge.bacteria.good.good.wang.taxonomy > data/mothur/all_bacteria.taxonomy
+cat data/mothur/bacteria.good.tax data/mothur/single_cell.bacteria.taxonomy data/mothur/emirge.bacteria.taxonomy > data/mothur/all_bacteria.taxonomy
 

### 2015-12-04 code/build_domain_rarefaction_plot.R
Add category-based rarefaction curves

@@ -1,7 +1,7 @@
-my_col <- c(unique=rainbow(5)[1], X0.03='orange', X0.05=rainbow(5)[3], X0.10=rainbow(5)[4], X0.20=rainbow(5)[5] )
-my_pch <- c(unique=16, X0.03=16, X0.05=17, X0.10=21, X0.20=22)
+my_domain_col <- c(unique=rainbow(5)[1], X0.03='orange', X0.05=rainbow(5)[3], X0.10=rainbow(5)[4], X0.20=rainbow(5)[5] )
+my_domain_pch <- c(unique=16, X0.03=16, X0.05=17, X0.10=21, X0.20=22)
 
-plot_curves <- function(rarefaction){
+plot_domain_curves <- function(rarefaction){
 	nseqs <- rarefaction$numsampled
 	rarefaction <- rarefaction[,-1]
 	rarefaction <- rarefaction[,!grepl("ci", colnames(rarefaction))]
@@ -11,37 +11,89 @@ plot_curves <- function(rarefaction){
 
 	plot(NA, xlim=xlim, ylim=ylim, type='n', xlab="", ylab="", main="", axes=FALSE)
 	for(cutoff in colnames(rarefaction)){
-		points(rarefaction[,cutoff]~nseqs, type="l", lwd=2, col=my_col[cutoff])
+		points(rarefaction[,cutoff]~nseqs, type="l", lwd=2, col=my_domain_col[cutoff])
 
 		spacing <- floor(seq(1,length(nseqs),length.out=10))[-1]
-		points(rarefaction[spacing,cutoff]~nseqs[spacing], col=my_col[cutoff], pch=my_pch[cutoff], bg="white", cex=1.25)
+		points(rarefaction[spacing,cutoff]~nseqs[spacing], col=my_domain_col[cutoff], pch=my_domain_pch[cutoff], bg="white", cex=1.25)
 	}
 	box()
 
 }
 
-bact <- read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T)
-arch <- read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T)
+my_categories <- c(aerosol="Aerosol", aquatic="Aquatic", built="Built", plant="Plant", soil="Soil", zoonotic="Zoonotic", other="Other")
 
-pdf('results/figures/domain_rarefaction.pdf', width=7.5, height=3.5)
[…]

### 2015-12-04 doc/notebook/Archaeal_data_acquisition.Rmd
Add emirge metadata; fix output of rarefaction to data/process

@@
 
-XXXXXXXXXXXXXXXX
-
 ## Get metadata
@@
 
+We'd also like to get the metadata from the EMIRGE data.
+We have an metadata file was created above with `code/get_emirge.metadata.sh` that we'll read in and concatenate to the end of `archaea.good.metadata`.
+
+```{r emirge_metadata}
+emirge_metadata <- read.table(file="data/raw/emirge.metadata", row.names=1)
+archaea_sequences <- read.table(file="data/mothur/emirge.archaea.taxonomy")[1]
+archaea_emirge_metadata <- emirge_metadata[archaea_sequences$V1,]
+
+write.table(archaea_emirge_metadata, file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
+```
+
 Let's now take the big metadata file that we have and concatenate on the OTU assignment and taxonomy information:
@@
         system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.all_categories.shared);
-        rarefaction.single(shared=all_archaea.all_categories.shared);
-        summary.single(shared=all_archaea.all_categories.shared, calc=nseqs-sobs-coverage)"
+        rarefaction.single(shared=all_archaea.all_categories.shared, outputdir=data/process)"
 
@@
         system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.pool_category.shared);
-        rarefaction.single(shared=all_archaea.pool_category.shared);
-        summary.single(shared=all_archaea.pool_category.shared, calc=nseqs-sobs-coverage)"
+        rarefaction.single(shared=all_archaea.pool_category.shared, outputdir=data/process)"
 
@@
         system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.env_category.shared);
-        rarefaction.single(shared=all_archaea.env_category.shared);
-        summary.single(shared=all_archaea.env_category.shared, calc=nseqs-sobs-coverage)"
+        rarefaction.single(shared=all_archaea.env_category.shared, outputdir=data/process)"
 
-mothur "#rarefaction.single(list=data/mothur/all_archaea.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20);"
+mothur "#rarefaction.single(list=data/mothur/all_archaea.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20, outputdir=data/process);"
+
+rm data/mothur/*rabund data/mothur/mothur*logfile mothur*logfile
 ```

### 2015-12-04 doc/notebook/Archaeal_data_acquisition.Rmd
Fix problem from reading in strings as factors

@@
 emirge_metadata <- read.table(file="data/raw/emirge.metadata", row.names=1)
-archaea_sequences <- read.table(file="data/mothur/emirge.archaea.taxonomy")[1]
+archaea_sequences <- read.table(file="data/mothur/emirge.archaea.taxonomy", stringsAsFactors=FALSE)[1]
 archaea_emirge_metadata <- emirge_metadata[archaea_sequences$V1,]

### 2015-12-04 results/tables/build_table_1.Rmd
Moved some stuff around

@@
 data <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
-int_cols <- grepl("count", colnames(data)) | grepl("sobs", colnames(data))
-float_cols <- !int_cols
-
-data[,int_cols] <- format(data[,int_cols], big.mark=',')
-data[,float_cols] <- format(data[,float_cols], big.mark=',', nsmall=1, digits=1)
-
 data_current <- data[,!(grepl("pre", colnames(data)) | grepl("post", colnames(data)))]
@@
 composite <- cbind(coarse, fine, data_current)
+
+int_cols <- grepl("count", colnames(composite)) | grepl("sobs", colnames(composite))
+float_cols <- !int_cols
+float_cols[1:2] <- FALSE
+
+composite[,int_cols] <- format(composite[,int_cols], big.mark=',')
+composite[,float_cols] <- format(composite[,float_cols], big.mark=',', nsmall=1, digits=1)
 

### 2015-12-04 code/get_phylum_category_counts.R
Output counts instead of relative abundances

@@ -15,10 +15,10 @@ bact_phylum[bact_phylum=="Euryarchaeota"] <- NA #to be removed
 bact_phylum[bact_phylum=="aquifer1"] <- NA
 
 b_phylum_cat_count <- table(bact_phylum, bact$category)
-b_phylum_cat_ra <- prop.table(b_phylum_cat_count, margin=2)
-b_phylum_cat_ra <- cbind(b_phylum_cat_ra, total=table(bact_phylum)/length(bact_phylum))
-rownames(b_phylum_cat_ra) <- gsub("_?\\(.*\\)", "", rownames(b_phylum_cat_ra) )
-bacterial_data <- cbind(domain=rep("bacteria", nrow(b_phylum_cat_ra)), b_phylum_cat_ra)
+#b_phylum_cat_ra <- prop.table(b_phylum_cat_count, margin=2)
+#b_phylum_cat_ra <- cbind(b_phylum_cat_ra, total=table(bact_phylum)/length(bact_phylum))
+rownames(b_phylum_cat_count) <- gsub("_?\\(.*\\)", "", rownames(b_phylum_cat_count) )
+bacterial_data <- cbind(domain=rep("bacteria", nrow(b_phylum_cat_count)), b_phylum_cat_count)
 
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
@@ -27,10 +27,10 @@ arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD" #to remove
 arch_phylum <- get_phylum(arch$taxonomy)
 
 a_phylum_cat_count <- table(arch_phylum, factor(arch$category, levels=names(categories)))
-a_phylum_cat_ra <- prop.table(a_phylum_cat_count, margin=2)
-a_phylum_cat_ra <- cbind(a_phylum_cat_ra, total=table(arch_phylum)/length(arch_phylum))
-rownames(a_phylum_cat_ra) <- gsub("_?\\(.*\\)", "", rownames(a_phylum_cat_ra) )
-archaeal_data <- cbind(domain=rep("archaea", nrow(a_phylum_cat_ra)), a_phylum_cat_ra)
+#a_phylum_cat_ra <- prop.table(a_phylum_cat_count, margin=2)
+#a_phylum_cat_ra <- cbind(a_phylum_cat_ra, total=table(arch_phylum)/length(arch_phylum))
[…]

### 2015-12-04 code/get_phylum_category_counts.R
Add total column back in

@@ -15,8 +15,8 @@ bact_phylum[bact_phylum=="Euryarchaeota"] <- NA #to be removed
 bact_phylum[bact_phylum=="aquifer1"] <- NA
 
 b_phylum_cat_count <- table(bact_phylum, bact$category)
-#b_phylum_cat_ra <- prop.table(b_phylum_cat_count, margin=2)
-#b_phylum_cat_ra <- cbind(b_phylum_cat_ra, total=table(bact_phylum)/length(bact_phylum))
+b_phylum_cat_count <- cbind(b_phylum_cat_count, total=table(bact_phylum))
+
 rownames(b_phylum_cat_count) <- gsub("_?\\(.*\\)", "", rownames(b_phylum_cat_count) )
 bacterial_data <- cbind(domain=rep("bacteria", nrow(b_phylum_cat_count)), b_phylum_cat_count)
 
@@ -27,8 +27,8 @@ arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD" #to remove
 arch_phylum <- get_phylum(arch$taxonomy)
 
 a_phylum_cat_count <- table(arch_phylum, factor(arch$category, levels=names(categories)))
-#a_phylum_cat_ra <- prop.table(a_phylum_cat_count, margin=2)
-#a_phylum_cat_ra <- cbind(a_phylum_cat_ra, total=table(arch_phylum)/length(arch_phylum))
+a_phylum_cat_count <- cbind(a_phylum_cat_count, total=table(arch_phylum))
+
 rownames(a_phylum_cat_count) <- gsub("_?\\(.*\\)", "", rownames(a_phylum_cat_count) )
 archaeal_data <- cbind(domain=rep("archaea", nrow(a_phylum_cat_count)), a_phylum_cat_count)
 

### 2015-12-04 code/build_phylum_category_heatmap.R
Change to use counts as input

@@ -8,11 +8,13 @@ data <- read.table(file="data/process/phylum_category_counts.tsv", header=T)
 
 bacteria <- data[data$domain=="bacteria",-1]
 bacteria <- bacteria[order(bacteria$total, decreasing=T),]
-bacteria_top <- as.matrix(bacteria[1:10,rev_categories])
+bacteria_ra <- prop.table(as.matrix(bacteria), margin=2)
+bacteria_top <- as.matrix(bacteria_ra[1:10,rev_categories])
 
 archaea <- data[data$domain=="archaea",-1]
 archaea <- archaea[order(archaea$total, decreasing=T),]
-archaea_top <- as.matrix(archaea[1:10,rev_categories])
+archaea_ra <- prop.table(as.matrix(archaea), margin=2)
+archaea_top <- as.matrix(archaea_ra[1:10,rev_categories])
 
 rownames(archaea_top) <- gsub("Miscellaneous_Crenarchaeotic_Group", "Misc. Crenarchaeota", rownames(archaea_top))
 rownames(archaea_top) <- gsub("Hydrothermal", "Hydro.", rownames(archaea_top))

### 2015-12-04 Schloss_Census2_mBio_2015.Rmd
Add phylum by category supplemental tables

@@
 Progression of the microbial census since the first full-length 16S rRNA gene sequence was deposited into GenBank in 1983.*** The number of bacterial and archaeal 16S rRNA gene sequences deposited (A) and the new OTUs they represent (B) has increased exponentially until the last several years when the rate of change has plateaued.
-For both bacterial and archaeal sequences, the number of studies that are responsible for depositing more than 50% of the seuqences each year has been relatively small (C).
+For both bacterial and archaeal sequences, the number of studies that are responsible for depositing more than 50% of the sequences each year has been relatively small (C).
 
@@
 Heatmap depicting the relative abundance of the most common bacterial and archaeal phyla across different environments.** Each environmental category exhibited a phylum-level signature although the bacterial census was dominated by sequences from the Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes and the archaeal census was dominated by sequences from the Euryarchaeota and Thaumarchaeota.
+The ten most abundant phyla across all environmental categories are shown.
+The data for all bacterial and archaeal phyla are available in Supplemental Tables 2 and 3, respectively.
 
@@
 
-```{r figure4a, eval=FALSE}
-data <- read.table(file="data/mothur/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per.pdf")
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(40000)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
-points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
-points(data$X0.03.aerosol~data$numsampled, type="l", col="green", lwd=2)
-points(data$X0.03.aquatic~data$numsampled, type="l", col="blue", lwd=2)
-points(data$X0.03.built~data$numsampled, type="l", col="orange", lwd=2)
-points(data$X0.03.other~data$numsampled, type="l", col="black", lwd=2)
-points(data$X0.03.soil~data$numsampled, type="l", col="purple", lwd=2)
-points(data$X0.03.zoonotic~data$numsampled, type="l", col="deeppink", lwd=2)
-
-legend(x=5e5, y=20000, lwd=2, lty=1,
-       legend=c("Other", "Plant", "Aerosol", "Aquatic", "Built", "Soil", "Zoonotic"),
-       col=c("black", "red", "green", "blue", "orange", "purple", "deeppink"))
-
-axis(1, at=seq(0,700000,1e5), label=c("0", "100,000", "200,000", "300,000", "400,000", "500,000", "600,000", "700,000"), cex.axis=0.8)
-axis(2, at=seq(0,40000,5000), label=c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "35,000", "40,000"), las=2, cex.axis=0.8)
-box()
-mtext(side=2, "Number of OTUs", line=4)
-
-dev.off()
-
-```
-
-**Figure 4B. Archaeal sampling by environment**
-
-```{r figure4b, eval=FALSE}
-
-data <- read.table(file="data/mothur/all_archaea.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_3per.pdf")
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
-points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
-points(data$X0.03.aerosol~data$numsampled, type="l", col="green", lwd=2)
-points(data$X0.03.aquatic~data$numsampled, type="l", col="blue", lwd=2)
-points(data$X0.03.built~data$numsampled, type="l", col="orange", lwd=2)
-points(data$X0.03.other~data$numsampled, type="l", col="black", lwd=2)
-points(data$X0.03.soil~data$numsampled, type="l", col="purple", lwd=2)
-points(data$X0.03.zoonotic~data$numsampled, type="l", col="deeppink", lwd=2)
-
-legend(x=15000, y=1550, lwd=2, lty=1,
-       legend=c("Other", "Plant", "Aerosol", "Aquatic", "Built", "Soil", "Zoonotic"),
-       col=c("black", "red", "green", "blue", "orange", "purple", "deeppink"))
-
-axis(1, at=seq(0,35000,5000), label=c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "35,000"), cex.axis=0.8)
-axis(2, at=seq(0,5250,750), label=c("0", "750", "1500", "2250", "3,000", "3750", "4500", "5250"), las=2, cex.axis=0.8)
-box()
-mtext(side=2, "Number of OTUs", line=4)
-
-dev.off()
-```
-
 \newpage

### 2015-12-04 results/tables/build_archaeal_category_phylum_table.Rmd
Add phylum by category supplemental tables

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=1.0in
classoption: landscape
---

**Supplementary Table 3. Frequency that each archaeal phylum was found across each of the environmental categories.**

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)

categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other", total="Total")

data <- read.table(file="data/process/phylum_category_counts.tsv", header=T)

archaea <- data[data$domain=="archaea",-1]
archaea <- archaea[order(archaea$total, decreasing=T), names(categories)]
colnames(archaea) <- gsub("total", "Total", colnames(archaea))
rownames(archaea) <- gsub("Candidate_division_", "", rownames(archaea))
rownames(archaea) <- gsub("_", " ", rownames(archaea))

total <- apply(archaea, 2, sum)
archaea <- rbind(archaea, Total=total)
archaea <- cbind(rownames(archaea), archaea)

colnames(archaea) <- c("", categories)
composite <- xtable(archaea)

align(composite) <- paste0('ll', paste(rep('c', ncol(archaea)-1), collapse=""))

print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0,nrow(composite)-1, nrow(composite)), comment=FALSE, size="tiny", rotate.colnames=TRUE)

```

### 2015-12-04 results/tables/build_bacterial_category_phylum_table.Rmd
Add phylum by category supplemental tables

--- 2015-12-04 results/tables/build_archaeal_category_phylum_table.Rmd
@@
 tables: true
-geometry: margin=1.0in
+geometry: margin=0.1in
 classoption: landscape
@@
 
-**Supplementary Table 3. Frequency that each archaeal phylum was found across each of the environmental categories.**
+**Supplementary Table 2. Frequency that each bacterial phylum was found across each of the environmental categories.**
 
@@
 
-archaea <- data[data$domain=="archaea",-1]
-archaea <- archaea[order(archaea$total, decreasing=T), names(categories)]
-colnames(archaea) <- gsub("total", "Total", colnames(archaea))
-rownames(archaea) <- gsub("Candidate_division_", "", rownames(archaea))
-rownames(archaea) <- gsub("_", " ", rownames(archaea))
+bacteria <- data[data$domain=="bacteria",-1]
+bacteria <- bacteria[order(bacteria$total, decreasing=T), names(categories)]
+colnames(bacteria) <- gsub("total", "Total", colnames(bacteria))
+rownames(bacteria) <- gsub("Candidate_division_", "", rownames(bacteria))
+rownames(bacteria) <- gsub("_", " ", rownames(bacteria))
 
-total <- apply(archaea, 2, sum)
-archaea <- rbind(archaea, Total=total)
-archaea <- cbind(rownames(archaea), archaea)
+total <- apply(bacteria, 2, sum)
+bacteria <- rbind(bacteria, Total=total)
+bacteria <- cbind(rownames(bacteria), bacteria)
 
-colnames(archaea) <- c("", categories)
-composite <- xtable(archaea)
+colnames(bacteria) <- c("", categories)
+composite <- xtable(bacteria)
 
-align(composite) <- paste0('ll', paste(rep('c', ncol(archaea)-1), collapse=""))
+align(composite) <- paste0('ll', paste(rep('c', ncol(bacteria)-1), collapse=""))
 

### 2015-12-04 results/tables/build_archaeal_phylum_effort_table.Rmd
Add phylum sequencing effort supplementary table

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=1.0in
---

**Supplementary Table 5. Frequency that each archaeal phylum was sequenced before and after 2006.**

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)

data <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
data <- data[order(data$all, decreasing=T),]
data$ratio_of_ratios <- round(data$ratio_of_ratios, 2)
data$ratio_of_ratios[data$ratio_of_ratios=="Inf"] <- "NA"

rownames(data) <- gsub("\\(.*\\)", "", rownames(data))
rownames(data) <- gsub("_", " ", rownames(data))

total <- apply(data[, 1:3], 2, sum)
data <- rbind(data, Total=c(total, "1.00"))

data <- cbind(rownames(data), data)
colnames(data) <- c("Phylum",  "<2006 and Before", "After 2006", "Total", "Ratio of Ratios")

composite <- xtable(data)

align(composite) <- paste0('ll', paste(rep('c', ncol(data)-1), collapse=""))

print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0, nrow(composite)-1,nrow(composite)), comment=FALSE, size="normalsize")

```

### 2015-12-04 results/tables/build_bacterial_phylum_effort_table.Rmd
Add phylum sequencing effort supplementary table

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=0.5in
---

**Supplementary Table 4. Frequency that each bacterial phylum was sequenced before and after 2006.**

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)

data <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
data <- data[!rownames(data) %in% "aquifer1",]
data <- data[order(data$all, decreasing=T),]
data$ratio_of_ratios <- round(data$ratio_of_ratios, 2)
data$ratio_of_ratios[data$ratio_of_ratios=="Inf"] <- "NA"

rownames(data) <- gsub("Candidate_division_", "", rownames(data))
rownames(data) <- gsub("_", " ", rownames(data))

total <- apply(data[, 1:3], 2, sum)
data <- rbind(data, Total=c(total, "1.00"))

data <- cbind(rownames(data), data)
colnames(data) <- c("Phylum",  "<2006 and Before", "After 2006", "Total", "Ratio of Ratios")

composite <- xtable(data)

align(composite) <- paste0('ll', paste(rep('c', ncol(data)-1), collapse=""))

print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0, nrow(composite)-1, nrow(composite)), comment=FALSE, size="scriptsize")
```

### 2015-12-04 results/tables/build_environmental_categories_table.Rmd
Create the environmental category description table

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=1.0in
---

**Supplemental Table 1. Description of environmental categories and the criteria used to assign sequences to each category.**

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)

categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other")

text <- read.table("doc/notebook/ClassificationCode.md", sep=':', row.names=1, stringsAsFactors=FALSE)

text <- text[names(categories),]

text$V2 <- categories

text$V3 <- gsub("^ +", "", text$V3)
text$V3 <- gsub("^(.)", "\\U\\1", text$V3, perl=TRUE)

text$V3 <- gsub("AQM", "Aquatic Marine", text$V3, perl=TRUE)
text$V3 <- gsub("AQF", "Aquatic Freshwater", text$V3, perl=TRUE)
text$V3 <- gsub("AQB", "Aquatic Brackish", text$V3, perl=TRUE)

coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Zoological", "", "", "", "Other")

text <- cbind(coarse, text)

colnames(text) <- paste0("\\textbf{", c("Coarse", "Category", "Description"), "}")
composite <- xtable(text)
align(composite) <- "lllp{11cm}"
print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = seq(-1, nrow(composite)), comment=FALSE, size="footnotesize")

```

### 2015-12-05 results/tables/build_environmental_categories_table.Rmd
Add table for environmental categories

@@
       in_header: table_header.tex
+    template: notitle.tex
 tables: true
-geometry: margin=1.0in
+geometry: margin=1in
 ---
 
-**Supplemental Table 1. Description of environmental categories and the criteria used to assign sequences to each category.**
+\captionsetup{labelformat=empty}
 
@@
 
-categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other")
+categories <- c(AE="", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Other", PR="Plant root", PS="Plant surface", PO="Other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other", OT="")
 
@@
 
-coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Zoological", "", "", "", "Other")
+coarse <- c("\\midrule\\multirow{2}{*}{Aerosol}",
+			"\\midrule\\multirow{16}{*}{Aquatic}", "", "", "", "", "", "", "", "",
+			"\\midrule\\multirow{11}{*}{Built}", "", "", "", "",
+			"\\midrule\\multirow{6}{*}{Plant associated}", "", "",
+			"\\midrule\\multirow{6}{*}{Soil}", "", "", "",
+			"\\midrule\\multirow{6}{*}{Zoological}", "", "", "",
+			"\\midrule\\multirow{2}{*}{Other}")
 
@@
 
-colnames(text) <- paste0("\\textbf{", c("Coarse", "Category", "Description"), "}")
-composite <- xtable(text)
-align(composite) <- "lllp{11cm}"
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = seq(-1, nrow(composite)), comment=FALSE, size="footnotesize")
+colnames(text) <- paste0("\\textbf{", c("Coarse", "Fine", "Description"), "}")
 
+addtorow <- list(pos=as.list(1:nrow(text)),
+				command=c('',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'\\cmidrule(r){2-3}',
+						'',
+						'')
+
+						)
+
+caption <- "\\textbf{Supplemental Table 1. Description of environmental categories and the criteria used to assign sequences to each category.}"
+
+composite <- xtable(text, caption=caption, label="")
+align(composite) <- "lllp{10cm}"
+print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1, nrow(composite)), comment=FALSE, add.to.row=addtorow, caption.placement = "top", size="footnotesize")
 ```

### 2015-12-05 results/tables/build_environmental_categories_table.Rmd
Shrink size so we don't need notitle.tex

@@
       in_header: table_header.tex
-    template: notitle.tex
 tables: true
@@
 
-coarse <- c("\\midrule\\multirow{2}{*}{Aerosol}",
+coarse <- c("\\midrule\\multirow{1}{*}{Aerosol}",
 			"\\midrule\\multirow{16}{*}{Aquatic}", "", "", "", "", "", "", "", "",
-			"\\midrule\\multirow{11}{*}{Built}", "", "", "", "",
-			"\\midrule\\multirow{6}{*}{Plant associated}", "", "",
+			"\\midrule\\multirow{10}{*}{Built}", "", "", "", "",
+			"\\midrule\\multirow{5}{*}{Plant associated}", "", "",
 			"\\midrule\\multirow{6}{*}{Soil}", "", "", "",
 			"\\midrule\\multirow{6}{*}{Zoological}", "", "", "",
-			"\\midrule\\multirow{2}{*}{Other}")
+			"\\midrule\\multirow{1}{*}{Other}")
 
@@
 align(composite) <- "lllp{10cm}"
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1, nrow(composite)), comment=FALSE, add.to.row=addtorow, caption.placement = "top", size="footnotesize")
+print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1, nrow(composite)), comment=FALSE, add.to.row=addtorow, caption.placement = "top", size="scriptsize")
 ```

### 2015-12-05 results/tables/build_archaeal_culture_effort_table.Rmd
Add supplemental tables describing culturability

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=1.0in
---

\captionsetup{labelformat=empty}

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)

data <- read.table(file="data/process/archaea.cultured_by_time_counts.tsv", header=T, stringsAsFactors=F)

all <- data[,grepl("all_", colnames(data))]
all$all_uncult_nseqs <- all$all_total_nseqs - all$all_cult_nseqs
all$all_uncult_sobs <- all$all_total_sobs - all$all_cult_sobs

all <- all [,c("all_cult_nseqs", "all_uncult_nseqs", "all_total_nseqs",
				"all_cult_sobs", "all_uncult_sobs", "all_total_sobs" )]

all <- all[order(all$all_total_sobs, decreasing=T),]

rownames(all) <- gsub("Candidate_division_", "", rownames(all))
rownames(all) <- gsub("_", " ", rownames(all))

total <- apply(all, 2, sum)
all <- rbind(all, Total=total)

all <- cbind(rownames(all), all)

colnames(all) <- c("", "\\textbf{Cultured}", "\\textbf{Uncultured}", "\\textbf{Total}",  "\\textbf{Cultured}", "\\textbf{Uncultured}", "\\textbf{Total}")
#all <- rbind(header, all)

addtorow <- list(pos=list(-1),
				command=c('\\hline\\multirow{3}{*}{\\textbf{Phylum}} & \\multicolumn{3}{c}{\\textbf{Sequences}} & \\multicolumn{3}{c}{\\textbf{OTUs}} \\\\ \\cmidrule(r){2-4}	\\cmidrule(r){5-7}'))

caption <- "\\textbf{Supplementary Table 7. Frequency that each archaeal sequence or OTU was retrieved by cultivation or by culture-independent methods.}"

composite <- xtable(all, caption=caption, label="")

align(composite) <- paste0('llcccccc')

print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(0, nrow(composite)-1, nrow(composite)), comment=FALSE, add.to.row=addtorow, caption.placement = "top", size="scriptsize")
```

### 2015-12-05 results/tables/build_bacterial_culture_effort_table.Rmd
Add supplemental tables describing culturability

--- 2015-12-05 results/tables/build_archaeal_culture_effort_table.Rmd
@@
 tables: true
-geometry: margin=1.0in
+geometry: margin=0.2in
 ---
@@
 
-data <- read.table(file="data/process/archaea.cultured_by_time_counts.tsv", header=T, stringsAsFactors=F)
-
+data <- read.table(file="data/process/bacteria.cultured_by_time_counts.tsv", header=T, stringsAsFactors=F)
+data <- data[!rownames(data) %in% "aquifer1",]
 all <- data[,grepl("all_", colnames(data))]
@@
 
-caption <- "\\textbf{Supplementary Table 7. Frequency that each archaeal sequence or OTU was retrieved by cultivation or by culture-independent methods.}"
+caption <- "\\textbf{Supplementary Table 6. Frequency that each bacterial sequence or OTU was retrieved by cultivation or by culture-independent methods.}"
 

### 2015-12-07 results/tables/build_archaeal_category_phylum_table.Rmd
Fix caption and table placement

@@
 
-**Supplementary Table 3. Frequency that each archaeal phylum was found across each of the environmental categories.**
+\captionsetup{labelformat=empty}
 
@@
 colnames(archaea) <- c("", categories)
-composite <- xtable(archaea)
+
+colnames(archaea) <- paste0("\\textbf{", colnames(archaea), "}")
+
+caption <- "\\textbf{Supplementary Table 3. Frequency that each archaeal phylum was found across each of the environmental categories.}"
+
+composite <- xtable(archaea, caption=caption, label="")
 
@@
 
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0,nrow(composite)-1, nrow(composite)), comment=FALSE, size="tiny", rotate.colnames=TRUE)
+print(composite,
+		include.rownames=FALSE,
+		include.colnames=TRUE,
+		hline.after = c(-1,0,nrow(composite)-1, nrow(composite)),
+		comment=FALSE,
+		caption.placement = "top",
+		size="tiny",
+		table.placement="!ht",
+		rotate.colnames=TRUE
+		)
 

### 2015-12-07 results/tables/build_archaeal_phylum_effort_table.Rmd
Fix caption and table placement

@@
 
-**Supplementary Table 5. Frequency that each archaeal phylum was sequenced before and after 2006.**
+\captionsetup{labelformat=empty}
 
@@
 data <- cbind(rownames(data), data)
-colnames(data) <- c("Phylum",  "<2006 and Before", "After 2006", "Total", "Ratio of Ratios")
+colnames(data) <- c("\\textbf{Phylum}",  "\\textbf{<2006 and Before}", "\\textbf{After 2006}", "\\textbf{Total}", "\\textbf{Ratio of Ratios}")
 
-composite <- xtable(data)
+caption <- "\\textbf{Supplementary Table 5. Frequency that each archaeal phylum was sequenced before and after 2006.}"
+
+composite <- xtable(data, caption=caption, label="")
 
@@
 
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0, nrow(composite)-1,nrow(composite)), comment=FALSE, size="normalsize")
+print(composite,
+	include.rownames=FALSE,
+	include.colnames=TRUE,
+	hline.after = c(-1,0, nrow(composite)-1,nrow(composite)),
+	caption.placement = "top",
+	comment=FALSE,
+	table.placement="!ht",
+	size="normalsize")
 

### 2015-12-07 results/tables/build_bacterial_category_phylum_table.Rmd
Fix caption and table placement

@@
 
-**Supplementary Table 2. Frequency that each bacterial phylum was found across each of the environmental categories.**
+\captionsetup{labelformat=empty}
 
@@
 colnames(bacteria) <- c("", categories)
-composite <- xtable(bacteria)
+
+colnames(bacteria) <- paste0("\\textbf{", colnames(bacteria), "}")
+
+caption <- "\\textbf{Supplementary Table 2. Frequency that each bacterial phylum was found across each of the environmental categories.}"
+
+composite <- xtable(bacteria, caption=caption, label="")
 
@@
 
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0,nrow(composite)-1, nrow(composite)), comment=FALSE, size="tiny", rotate.colnames=TRUE)
+print(composite,
+	include.rownames=FALSE,
+	include.colnames=TRUE,
+	hline.after = c(-1,0,nrow(composite)-1, nrow(composite)),
+	comment=FALSE,
+	size="tiny",
+	caption.placement = "top",
+	table.placement="!ht",
+	rotate.colnames=TRUE)
 

### 2015-12-07 results/tables/build_bacterial_culture_effort_table.Rmd
Fix caption and table placement

@@
 tables: true
-geometry: margin=0.2in
+geometry: margin=1.0in
 ---
@@
 
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(0, nrow(composite)-1, nrow(composite)), comment=FALSE, add.to.row=addtorow, caption.placement = "top", size="scriptsize")
+print(composite,
+	include.rownames=FALSE,
+	include.colnames=TRUE,
+	hline.after = c(0, nrow(composite)-1, nrow(composite)),
+	comment=FALSE,
+	add.to.row=addtorow,
+	caption.placement = "top",
+	table.placement = "!ht",
+	size="scriptsize")
 ```

### 2015-12-07 results/tables/build_bacterial_phylum_effort_table.Rmd
Fix caption and table placement

@@
 tables: true
-geometry: margin=0.5in
+geometry: margin=1.0in
 ---
 
-**Supplementary Table 4. Frequency that each bacterial phylum was sequenced before and after 2006.**
+\captionsetup{labelformat=empty}
 
@@
 data <- cbind(rownames(data), data)
-colnames(data) <- c("Phylum",  "<2006 and Before", "After 2006", "Total", "Ratio of Ratios")
+colnames(data) <- c("\\textbf{Phylum}",  "\\textbf{<2006 and Before}", "\\textbf{After 2006}", "\\textbf{Total}", "\\textbf{Ratio of Ratios}")
 
-composite <- xtable(data)
+caption <- "\\textbf{Supplementary Table 4. Frequency that each bacterial phylum was sequenced before and after 2006.}"
+
+composite <- xtable(data, caption=caption, label="")
 
@@
 
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1,0, nrow(composite)-1, nrow(composite)), comment=FALSE, size="scriptsize")
+print(composite,
+		include.rownames=FALSE,
+		include.colnames=TRUE,
+		hline.after = c(-1,0, nrow(composite)-1, nrow(composite)),
+		comment=FALSE,
+		caption.placement = "top",
+		table.placement="!ht",
+		size="scriptsize")
 ```

### 2015-12-07 results/tables/build_coverage_by_category.Rmd
Fix caption and table placement

--- 2015-12-04 results/tables/build_table_1.Rmd
@@
     includes:
-      in_header: table_1_header.tex
+      in_header: table_header.tex
 tables: true
@@
 
-**Table 1.
-Status of microbial census by habitat classifications and domain.** The isolation_source field from the SILVA reference database was manually curated to assign bacterial and archaeal sequences coarse and fine scale habitat classifications.
-We calculated the number of sequences and OTUs observed and the percent coverage on a sequence or OTU basis for each classification and domain.
+\captionsetup{labelformat=empty}
 
@@
 
+header <- paste0("\\textbf{", header, "}")
+
+caption <- "\\textbf{Table 1. Status of microbial census by habitat classifications and domain.}"
+#The isolation_source field from the SILVA reference database was manually curated to assign bacterial and archaeal sequences coarse and fine scale habitat classifications.
+We calculated the number of sequences and OTUs observed and the percent coverage on a sequence or OTU basis for each classification and domain."
+
 composite <- rbind(header, composite)
 
-composite <- xtable(composite)
+composite <- xtable(composite, caption=caption, label="")
 
 addtorow <- list(pos=list(0, 1),
-				command=c('\\multirow{4}{*}{Coarse} & \\multirow{4}{*}{Fine} & \\multicolumn{4}{c}{Bacteria} & \\multicolumn{4}{c}{Archaea} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
+				command=c('\\multirow{4}{*}{\\textbf{Coarse}} & \\multirow{4}{*}{\\textbf{Fine}} & \\multicolumn{4}{c}{\\textbf{Bacteria}} & \\multicolumn{4}{c}{\\textbf{Archaea}} \\\\ \\cmidrule(r){3-6}	\\cmidrule(r){7-10}', '\\hline '))
 
 align(composite) <- "lllcccccccc"
-print(composite, add.to.row=addtorow, include.rownames=FALSE, include.colnames=FALSE, hline.after = c(-1,nrow(composite)-1, nrow(composite)), comment=FALSE, size="scriptsize")
+print(composite,
+	add.to.row=addtorow,
+	include.rownames=FALSE,
+	include.colnames=FALSE,
+	hline.after = c(-1,nrow(composite)-1, nrow(composite)),
+	comment=FALSE,
+	caption.placement = "top",
+	table.placement="!ht",
+	size="scriptsize")
 ```

### 2015-12-07 results/tables/build_environmental_categories_table.Rmd
Fix caption and table placement

@@
 align(composite) <- "lllp{10cm}"
-print(composite, include.rownames=FALSE, include.colnames=TRUE, hline.after = c(-1, nrow(composite)), comment=FALSE, add.to.row=addtorow, caption.placement = "top", size="scriptsize")
+print(composite,
+		include.rownames=FALSE,
+		include.colnames=TRUE,
+		hline.after = c(-1, nrow(composite)),
+		comment=FALSE,
+		add.to.row=addtorow,
+		caption.placement = "top",
+		table.placement="!ht",
+		size="footnotesize")
 ```

### 2015-12-07 results/tables/build_coverage_by_category.Rmd
Fix caption labeling

@@
 \captionsetup{labelformat=empty}
+
+**Table 1.
+Status of microbial census by habitat classifications and domain.** The isolation_source field from the SILVA reference database was manually curated to assign bacterial and archaeal sequences coarse and fine scale habitat classifications.
+We calculated the number of sequences and OTUs observed and the percent coverage on a sequence or OTU basis for each classification and domain.
 
@@
 
-caption <- "\\textbf{Table 1. Status of microbial census by habitat classifications and domain.}"
-#The isolation_source field from the SILVA reference database was manually curated to assign bacterial and archaeal sequences coarse and fine scale habitat classifications.
-We calculated the number of sequences and OTUs observed and the percent coverage on a sequence or OTU basis for each classification and domain."
-
 composite <- rbind(header, composite)
 
-composite <- xtable(composite, caption=caption, label="")
+composite <- xtable(composite)
 

### 2015-12-07 Schloss_Census2_mBio_2015.Rmd
Add phylum sequencing effort supplementary table
Incorporate Cameron's text edits

@@
 Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environemtnal categorizations are available at [@figshare].
+
+***Calculating coverage.***
+Sequencing coverage (C~Sequence~) was quantified by two methods. The first was to use Good's coverage according to
+
+$$C_{Sequence} = 1 - \frac{n_1}{N_t}$$
+
+where n~1~ is the number of OTUs represented by only one sequence and N~t~ is the total number of sequences [@ref].
+Although Good's coverage provides information about the success of the sequencing effort in sampling the most abundant organisms in a community, it does not directly provide information about the success of the sequencing effort in recovering previously unobserved OTUs.
+To quantify the ability of sequencing to identifying novel OTUs or, in other words, to quantify the "distance" in the peak of the rarefaction curves to their hypothetical asymptote, we defined "OTU coverage" (C~OTU~) as
+
+$$C_{OTU} = 1 - \frac{n_1}{S_t}$$
+
+where S~t~ is the total number of OTUs.
+Whereas Good’s coverage estimates the probability that a new sequence will have already been seen, OTU coverage estimates the probability that a new OTU will match an existing one.
+It is therefore an extension of Good’s coverage in that it quantifies the probability that, for any given set of sequences clustered into an OTU, that OTU will have already been seen.
+Thus, high Good’s coverage means that any new sequence is unlikely to be novel, and high OTU coverage means that any new OTU is unlikely to be novel.
 
@@
 We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure venn_otu_by_method.pdf).
-Considering that the percentage of sequences from bacterial cultures is greater than the percentage of OTUs from bacterial cultures, we suspect that bacterial cultivation efforts are largely resampling the same diversity relative to culture-independent approaches.
-In contrast, the relationship is reversed among the archaea indicating that archaeal cultivation efforts continue to sample greater biodiversity than culture-independent approaches.
+Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Fig.
+results/figures/phylum_cultured.pdf).
+The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
+This likely reflects the extremely high number of cultivated  Mycobacterium tuberculosis, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
+Conversely, clades such as the Actinobacteria, Fusobacteria, and Bacteroidetes had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapuslated within the 97% sequence identity cutoff.
 
@@
 Relative rate of sequence deposition for each bacterial and archaeal phylum before and after 2006 relative to the sequencing of all bacteria.** The figure shows the relative rates for those phyla with at least 1,000 sequences and the x-axis is on a log2 scale.
+The data for all bacterial and archaeal phyla are available in Supplemental Tables 4 and 5, respectively.
 
 **Figure results/figures/phylum_cultured.pdf.
-The rate that sequences and OTUs are generated from bacterial and archaeal cultures relative to all sequences and OTUs.** The phyla shown have at least 1,000 total sequences.
+The rate that sequences and OTUs are generated from bacterial and archaeal cultures relative to all sequences and OTUs by phlum.** Phyla with greater than 1,000 sequences are listed by domain.
+Open circles indicate the percentage of sequences in the database that match cultured organisms.
+Closed circles indicate the percentage of OTUs in this analysis that contain sequences belonging to a cultured organism.
 

### 2015-12-07 Schloss_Census2_mBio_2015.Rmd
Fix number formatting

@@
 In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
-Although this is still an impressive number of sequences, since `r start<-2007;start` the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(start:2014),'b_sobs'])` new OTUs per year (Figure 2B).
+Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure 2B).
 Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects.

### 2015-12-07 Schloss_Census2_mBio_2015.Rmd
Fix references and their handling
Update figure labels in text

--- 2015-12-07 Schloss_Census2_mBio_2015.Rmd
@@
 By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
-After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1).
+After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure results/figures/domain_rarefaction.pdf A, Table 1).
 Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once.
 In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once.
-Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1).
+Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure Figure results/figures/domain_rarefaction.pdf B, Table 1).
 Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once.
@@
 Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once it is likely that an even larger number of OTUs have yet to be sampled for both domains.
+
+Need to add text about categories
 
@@
 One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
-In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
-Although this is still an impressive number of sequences, since `r start<-2007;start` the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(start:2014),'b_sobs'])` new OTUs per year (Figure 2B).
+In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure results/figures/time_course_figure.pdf A).
+Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure results/figures/time_course_figure.pdf B).
 Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects.
-Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C).
+Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure results/figures/time_course_figure.pdf C).
 Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(dominant_pmids_2009_2012)` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences).
@@
 Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`).
-Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure phylum_effort.pdf).
+Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure results/figures/phylum_effort.pdf).
 Among the bacterial phyla with at least 1,000 sequences, `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios >= 4.0 & b_phylum$all >=1000,])))` were sequenced 4-fold more often while `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios < 0.5 & b_phylum$all >=1000,])))` were sequenced 2-fold less often than would have been expected since 2006.
@@
 The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
-Across all categories, the bacterial and archaeal sequencing data represented a limited number of phyla (Figure category_phylum_heatmap.pdf).
+Across all categories, the bacterial and archaeal sequencing data represented a limited number of phyla (Figure results/figures/category_phylum_heatmap.pdf).
 Among the bacterial data, the fine-scale categories were dominated by `r dominant_bacteria_list` and among the archaeal data, they were dominated by `r dominant_archaea_list`.
@@
 This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
-Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure phylum_effort.pdf).
+Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure results/figures/phylum_effort.pdf).
 So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
 Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
-We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure venn_otu_by_method.pdf).
+We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure results/figures/venn_otu_by_method.pdf.pdf).
 Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Fig.
@@
 Assembly of metagenomic and single cell shotgun sequence data offers the hope of identifying large fragments of genomic data from as yet uncultured organisms.
-To test the ability of single cell technologies to expand our knowledge of microbial diversity beyond that of the 16S rRNA gene and pure cultures, we compared the overlap of OTUs found by the three methods (Figure venn_otu_by_method.pdf).
+To test the ability of single cell technologies to expand our knowledge of microbial diversity beyond that of the 16S rRNA gene and pure cultures, we compared the overlap of OTUs found by the three methods (Figure results/figures/venn_otu_by_method.pdf).
 Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts['n_sc', 'bact_counts']` bacterial and `r overlap_counts['n_sc', 'arch_counts']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts["sc_all", "bact_counts"]` and `r overlap_counts["sc_all", "arch_counts"]` bacterial and archaeal OTUs, respectively.
@@
 \newpage
-\singlespacing
+\doublespacing
 
@@
 \newpage
-\doublespacing
 

### 2016-01-07 doc/notebook/Bacterial_data_acquisition.Rmd
Update processing status
Fix pre-cluster diffs

@@
 
-XXXXXXXXXXXXXXXX
-
 ## Pooling the data
@@
         unique.seqs();
-        pre.cluster(fasta=current, name=current, diffs=10);"
+        pre.cluster(fasta=current, name=current, diffs=12);"
 
@@
 mothur "#set.dir(input=data/mothur, output=data/mothur, seed=1);
-        cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
+	cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, splitmethod=classify, classic=T, taxlevel=2, cluster=F, processors=12)
+	cluster.split(file=all_bacteria.filter.unique.precluster.file, processors=8, classic=TRUE)"
 ```
+
+XXXXXXXXXX
 

### 2016-01-18 Schloss_Census2_mBio_2015.Rmd

@@
 ## Introduction
-In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was deposited into NCBI's GenBank making it the first of more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978].
-GenBank accessions represent nearly one-third of all sequences deposited in the database making it the best-represented gene.
-As Sanger sequencing has given way to to-called "next generation sequencing" technologies, hundreds of millions of 16S rRNA gene sequences have been deposited into the NCBI's Sequence Read Archive.
-The expansion in sequencing throughput and increased access to sequencing technology has allowed for more environments to be sequenced at a deeper coverage resulting in the identification of novel taxa.
-The ability to obtain sequence data from microorganisms without cultivation has radically altered our perspective of their role in nearly every environment from deep ocean sediment cores (e.g.
-accession AY436526) to the International Space Station (e.g.
-accession DQ497748).
 The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@reference].
@@
 
-The increase in sequencing volume has come at the cost of sequence length.
-The commonly used MiSeq-based sequencing platform from Illumina is widely used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
-Perhaps most disconcerting about this development is the sense that the increased read depth is being gained using short read platforms rather than the full-length sequences.
-Because these short reads are used for classification to existing taxa, we are missing the opportunity to propose novel candidate taxa and vastly underappreciating the biodiversity of microbial life.
-We likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
+Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
+In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was deposited into NCBI's GenBank making it the first of what is now more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978].
+16S rRNA gene accessions represent nearly one-third of all sequences deposited in GenBank, making it the best-represented gene.
+As Sanger sequencing has given way to so-called "next generation sequencing" technologies, hundreds of millions of 16S rRNA gene sequences have been deposited into the NCBI's Sequence Read Archive.
+The expansion in sequencing throughput and increased access to sequencing technology has allowed for more environments to be sequenced at a deeper coverage, resulting in the identification of novel taxa.
+The ability to obtain sequence data from microorganisms without cultivation has radically altered our perspective of their role in nearly every environment from deep ocean sediment cores (e.g.
+accession AY436526) to the International Space Station (e.g.
+accession DQ497748).
 
-Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998, @Curtis2002].
+Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial 16S rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998, @Curtis2002].
 That census called for a broader and deeper characterization of all environments.
@@
 There have been major investments in studying the Earth's microbiome using 16S rRNA gene sequencing through initiatives such as the Human Microbiome Project [@reference], the Earth Microbiome Project [@reference], and the International Census of Marine Microorganisms.
-But most important, the original census was performed on the cusp of radical developments in sequencing technologies.
-That advancement has largely moved the bulk of sequencing throughput from large sequencing centers to individual investigators and leveraged their diversity to expand the representation of organisms and environments represented in public databases.
+But most importantly, the original census was performed on the cusp of radical developments in sequencing technologies.
+That advancement has moved the generation of sequencing throughput from large sequencing centers to individual investigators and leveraged their diverse interests to expand the representation of organisms and environments represented in public databases.
+
+Disconcertingly, the increase in sequencing volume has come at the cost of sequence length.
+The commonly used MiSeq-based sequencing platform from Illumina is extensively leveraged to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
+Youssef et al.
+(2009) demonstrated high variability between the number of OTUs that different hypervariable sites estimated compared to full-length 16S rRNA gene sequences, and also that sample location influenced this behavior.
+Thus, it remains unclear to what degree richness estimates from short read technology over or underestimate the numbers from full-length sequences.
+We likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
 
@@
 As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.
-In this, continued technological advances, such as the reconstruction of nearly full-length sequences from metagenomics with EMIRGE (Miller 2011) and inovations in culturing (references needed: iChip, Epstein dormancy, Thrash 2015; Gifford 2014), will be vital.
+In this, continued technological advances, such as the reconstruction of nearly full-length sequences from metagenomics with EMIRGE (Miller 2011) and innovations in culturing (references needed: iChip, Epstein dormancy, Thrash 2015; Gifford 2014), will be vital.
 
@@
 "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata.
-The IMG sequences were aligned against the respective SILVA-based reference using mothur [@REF].
-The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur.
+Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included.
+These were obtained from XXXXXXXXXX.
+The IMG and EMIRGE sequences were aligned against the respective SILVA-based reference using mothur [@REF].
+The aligned bacterial and archaeal sequence sets were pooled and processed in parallel in mothur.
 Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates.
@@
 These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences (Table 1).
-For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
+For source definitions that were not identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
 There were `r category_summary["no_source", "bact_count"]` bacterial and `r category_summary["no_source", "arch_count"]` archaeal sequences where an "isolation_source" term was not collected.
 We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms.
-Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environemtnal categorizations are available at [@figshare].
+Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environmental categorizations are available at [@figshare].
 
@@
 By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
-After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure results/figures/domain_rarefaction.pdf A, Table 1).
-Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once.
-In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once.
-Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure Figure results/figures/domain_rarefaction.pdf B, Table 1).
-Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once.
-In contrast, only `r a_schloss`% of the OTUs that were observed had been seen more than once.
-The results for the bacterial and archaeal censuses indicate that regardless of the domain, if we continue sampling with the current strategies we will continue to sample OTUs that have already been observed even though a large fraction of OTUs have only been sampled once.
-Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once it is likely that an even larger number of OTUs have yet to be sampled for both domains.
-
-Need to add text about categories
+After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the respective thresholds (Figure results/figures/domain_rarefaction.pdf A, Table 1).
+Using only the OTUs generated using a 3% threshold, we calculated a `r b_goods`% Good's coverage (percent of sequences belonging to OTUs that have been observed more than once), but only `r b_schloss`% OTU coverage (percent of the OTUs that have been observed more than once).
+Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure results/figures/domain_rarefaction.pdf B, Table 1).
+Using only the OTUs generated using a 3% threshold, we calculated a `r a_goods`% Good's coverage, but only `r a_schloss`% OTU coverage.
+These results indicate that regardless of the domain, continued sampling with the current strategies for generating full-length sequences will largely reveal OTUs that have already been observed, even though a large fraction of OTUs have only been sampled once.
+Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once, it is likely that an even larger number of OTUs have yet to be sampled for both domains.
 
@@
 These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
+The heavy zoonotic focus is reflected in the rarefaction curve for this category (Figure results/figures/domain_rarefaction.pdf C).
 In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year.
@@
 With the exception of 2012 and 2014, the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has varied between `r min(time_data[as.character(c(1996:2011,2013)),'a_n50'])` and `r max(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions per year.
-The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea.
+The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea, which is clearly reflected in the relatively meager sampling effort across habitats, compared to bacteria (Figure results/figures/domain_rarefaction D),
 
@@
 Across these broad categories there was wide variation in the number of sequences that have been sampled.
-Among the bacterial sequences the three best represented groups were from `r oxford_comma(b_text[1:3])` sources and among the archaeal sequences the three best represented groups were from `r oxford_comma(a_text[1:3])` sources.
+Among bacterial sequences, the three best represented groups were from `r oxford_comma(b_text[1:3])` sources.
+Among the archaeal sequences the three best represented groups were the same, but ordered differently: `r oxford_comma(a_text[1:3])` (Figure 1C,D)).
 For both domains, soil samples were the fourth most represented category (bacteria: `r b_agg_count[b_agg_count$Group.1=="soil", "x"]`; archaea: `r a_agg_count[a_agg_count$Group.1=="soil", "x"]`).
 The orders of these categories was surprising considering soil and aquatic environments harbor the most microbial biomass and biodiversity [@Whitman1998].
-In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range across the fine-level categories for the bacterial OTU-based coverage only varied between `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.25))` to `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.75))` (median coverage=`r median(coverage_by_category$bact_schloss_coverage, na.rm=T)`%).
-The interquartile range in the OTU-based coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to`r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
+In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range across the fine-level categories for the bacterial OTU coverage only varied between `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.25))` to `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.75))` (median coverage=`r median(coverage_by_category$bact_schloss_coverage, na.rm=T)`%).
+The interquartile range in the OTU coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to`r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
 The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
@@
 The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
-This likely reflects the extremely high number of cultivated  Mycobacterium tuberculosis, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
+This likely reflects the extremely high number of cultivated  *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
 Conversely, clades such as the Actinobacteria, Fusobacteria, and Bacteroidetes had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
-Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapuslated within the 97% sequence identity cutoff.
+Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
+
+In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
+Overall, 18.6% of bacterial sequences and 6.8% of archaeal sequences have come from isolated organisms.
+Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by 2.5 and 2.4-fold for bacteria and archaea, respectively.
+Among the 67 bacterial phyla, 19 have no cultured representatives and 20 of the 10 archaeal phyla have cultured representatives.
+This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
+Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only 2.4% for the bacterial phyla and 1.7% for the archaeal phyla (Figure phylum_effort.pdf).
+So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
+Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
+We found that 13.0% of the 95,734 bacterial OTUs and 9.1% of the 4,205 archeael OTUs had at least one cultured representative (Figure venn_otu_by_method.pdf).
+Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Fig.
+results/figures/phylum_cultured.pdf).
+The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
+This likely reflects the extremely high number of cultivated *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
+Conversely, clades such as the Actinobacteria, Fusobacteria, and Bacteroidetes had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
 
@@
 
+* Would be nice if we could comment somewhere on whether or not the EMIRGE/SAG sequences are more or less “efficient” at uncovering new OTUs.
+I.e., does incorporating EMIRGE/SAG sequences increase the OTU coverage at a greater rate, per new sequence, than those created by other methods?
+
 ***New technologies to access novel biodiversity***
@@
 
+## Caveat Emptor
+Recent data suggests that a considerable short-read diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers (Furhman, 2015).
+Furthermore, Brown et al.
+(2015) have recently used metagenomic assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
+Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015).
+Thus, for many environments, the estimates for the census must be considered as lower bounds.
+
 ## Conclusions
 
-Future for PacBio in generating full-length sequences
+It is clear that considerable biodiversity has been discovered since the first census in 2004.
+However, much of it has been biased towards particular phyla and environments.
+Nevertheless, novel technologies such as single-cell genomics and algorithms to recover full-length sequences from shotgun metagenomic data have demonstrated promise in circumventing previous limitations in identifying new OTUs.
+Additional technologies coming on line that can provide full-length sequences, such as PacBio and potentially Nanopore, will likely provide considerable value.
+Additionally, as researchers attempt to answer questions about the physiology and biochemistry of the organisms identified in the massive deluge of cultivation-independent sequence data, a renewed emphasis on cultivation is beginning.
+Focused innovation to overcome the challenges of isolating new microorganisms will also contribute to improving out understanding of extant diversity.
 
@@
 
-Impact of EMIRGE
-- http://www.genomebiology.com/2011/12/5/R44
+What are the most significant improvements since 2004, and where are we still lacking the most data?
 
 Renewed call for cultivation
-
-What are the most significant improvements since 2004, and where are we still lacking the most data?
-
-Recent data suggests that a considerable diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers.
-This dataset does not include sequences from metagenomic assemblies but Brown et al.
-(2015) have used such assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
-Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015).
-Thus, for many environments, the estimates within must be considered as lower bounds.
-
-Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
-Indeed, continual community efforts to obtain 16S rRNA gene assessments of every environment possible have presented us with an ever-increasing estimate of total microbial diversity and the concomitant excitement of frontier science.
-While reliance on this gene subjects us to biases created by primer selection [REFS], differences in amplification strength [REFS] and fidelity [REFS], internal features which may disrupt traditional measurements [REFS], and potentially misleading classification due to infrequent horizontal gene transfer [REFS], the total data available from persistent collection of 16S rRNA gene sequences nevertheless dwarfs that of any other genetic marker.
-Thus, an attempt to quantify how much of the microbial world has been revealed inevitably starts there.
-
-Given the relatively low number of archaeal sequences that have been deposited for many of these categories, it is possible that the coverage for many categories may not be reliable (e.g.
-aerosol and plant surfaces).
-By decomposing the census by environmental categories, it is clear that even among the best sampled environments, our ability to claim anything but a basic characterization of microbial biodiversity is limited.
 
@@
 **Figure results/figures/domain_rarefaction.pdf.
-Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** The number of bacterial and archaeal OTUs observed among the longest sequences in the SILVA database continues to grow at a rate too slow to ever reach estimates of 10^6^ to 10^11^ bacterial species.
+Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B).
+Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).
+The number of bacterial and archaeal OTUs observed among the longest sequences in the SILVA database continues to grow at a rate too slow to ever reach estimates of 10^6^ to 10^11^ bacterial species.
 

### 2016-01-18 doc/notebook/Bacterial_data_acquisition.Rmd
Split at class level instead of phylum

@@
 mothur "#set.dir(input=data/mothur, output=data/mothur, seed=1);
-	cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, splitmethod=classify, classic=T, taxlevel=2, cluster=F, processors=12)
+	cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, splitmethod=classify, classic=T, taxlevel=3, cluster=F, processors=12)
 	cluster.split(file=all_bacteria.filter.unique.precluster.file, processors=8, classic=TRUE)"

### 2016-01-18 doc/notebook/Bacterial_data_acquisition.Rmd
Bring in EMIRGE metadata

@@
 
-XXXXXXXXXX
-
 ## Get metadata
@@
 
-#these sequences are missing from fasta file
-#metadata_genome_id[! metadata_genome_id %in% names(seq_numbers)]
-#probably not a horrible thing since not all genomes could give a 16S rRNA gene
-#sequence
+# these sequences are missing from fasta file
+# metadata_genome_id[! metadata_genome_id %in% names(seq_numbers)]
+# probably not a horrible thing since not all genomes could give a 16S rRNA gene
+# sequence
+#
+# [1] "SCGC AAA015-N04" "SCGC AAA536-N21" "SCGC AAA158-B04" "SCGC AAA036F23"
+# [5] "SCGC AAA001-B15" "SCGC AAA164-N20" "SCGC AAA536-J20" "SCGC AAA008-P15"
+# [9] "SCGC AAA536-G18" "SCGC AAA001-B15"
 
@@
 write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t", col.names=F, append=T)
+```
+
+We'd also like to get the metadata from the EMIRGE data.
+We have an metadata file was created above with `code/get_emirge.metadata.sh` that we'll read in and concatenate to the end of `bacteria.good.metadata`.
+
+```{r emirge_metadata}
+emirge_metadata <- read.table(file="data/raw/emirge.metadata", row.names=1)
+bacteria_sequences <- read.table(file="data/mothur/emirge.bacteria.taxonomy", stringsAsFactors=FALSE)[1]
+bacteria_emirge_metadata <- emirge_metadata[bacteria_sequences$V1,]
+
+write.table(bacteria_emirge_metadata, file="data/mothur/bacteria.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
 ```
@@
 
+table(metadata$category)
+
+#    AE    AQB   AQBS    AQF   AQFS    AQH    AQI    AQM   AQMS    AQO     BD
+#  3473   1094    390  59220  23979  10861   2073 134788  27811   8765  33152
+#    BF     BI     BO     BP     OT     PO     PR     PS     SA     SD     SO
+# 11813  16598   8556  38710  19417   9753  19695   4892  10051   3059  60335
+#    SP     ZA     ZN     ZO     ZV
+#  1922  13209   7476  10855 794492
+
 #need to change the name of this sequence to make it clear that it was not cultured

### 2016-01-18 doc/notebook/Archaeal_data_acquisition.Rmd
Output data to data/process

@@
 
-#... and we output a group file to use for splitting up the list file into the pooled categories...
-#   For archaea: AQH, AQM, AQMS, and ZV into one. the rest into the other
-#   per Rene 6/24/2015...
-pool1 <- c("AQH", "AQM", "AQMS", "ZV")
-pool_categories <- rep(NA, nrow(metadata))
-pool_categories[metadata$category %in% pool1] <- "pool1"
-pool_categories[!metadata$category %in% pool1] <- "pool2"
-pool_categories[is.na(metadata$category)] <- "NA"
-write.table(file="data/mothur/archaea.pool_category.groups", x=cbind(rownames(metadata), pool_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
-
 #... and we pool the categories by environment
@@
 mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.pool_category.groups, label=unique-0.03-0.05-0.10-0.20);
-        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.pool_category.shared);
-        rarefaction.single(shared=all_archaea.pool_category.shared, outputdir=data/process)"
-
-mothur "#set.dir(input=data/mothur, output=data/mothur);
         make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.env_category.groups, label=unique-0.03-0.05-0.10-0.20);

### 2016-01-18 doc/notebook/Bacterial_data_acquisition.Rmd

@@
 
-#need to change the name of this sequence to make it clear that it was not cultured
-rownames(metadata)[rownames(metadata) == 'AJ224540.MxeCultu'] <- 'AJ224540.UncMxeCultu'
+# need to change the name of this sequence to make it clear that it was not cultured
+# rownames(metadata)[rownames(metadata) == 'AJ224540.MxeCultu'] <- 'AJ224540.UncMxeCultu'
+# doing this here screws up the list file - need to do this in the actual
+# analysis
 
@@
 write.table(file="data/mothur/bacteria.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
-
-#... and we output a group file to use for splitting up the list file into the pooled categories...
-#   For bacteria: AQF, AQFS, AQM, AQMS, BD, BI, BP, OT, PR, SO, ZO, and ZV and the others into one category
-#   per Rene 6/24/2015...
-pool1 <- c("AQF", "AQFS", "AQM", "AQMS", "BD", "BI", "BP", "OT", "PR", "SO", "ZO", "ZV")
-pool_categories <- rep(NA, nrow(metadata))
-pool_categories[metadata$category %in% pool1] <- "pool1"
-pool_categories[!metadata$category %in% pool1] <- "pool2"
-pool_categories[is.na(metadata$category)] <- "NA"
-write.table(file="data/mothur/bacteria.pool_category.groups", x=cbind(rownames(metadata), pool_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
 
@@
         system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.all_categories.shared);
-        rarefaction.single(shared=all_bacteria.all_categories.shared);
-        summary.single(shared=all_bacteria.all_categories.shared, calc=nseqs-sobs-coverage)"
-
-mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.pool_category.groups, label=unique-0.03-0.05-0.10-0.20);
-        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.pool_category.shared);
-        rarefaction.single(shared=all_bacteria.pool_category.shared);
-        summary.single(shared=all_bacteria.pool_category.shared, calc=nseqs-sobs-coverage)"
+        rarefaction.single(shared=all_bacteria.all_categories.shared, outputdir=data/process)"
 
@@
         system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.env_category.shared);
-        rarefaction.single(shared=all_bacteria.env_category.shared);
-        summary.single(shared=all_bacteria.env_category.shared, calc=nseqs-sobs-coverage)"
+        rarefaction.single(shared=all_bacteria.env_category.shared, outputdir=data/process)"
 
-mothur "#rarefaction.single(list=data/mothur/all_bacteria.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20);"
+mothur "#rarefaction.single(list=data/mothur/all_bacteria.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20, outputdir=data/process);"
+
+rm data/mothur/*rabund data/mothur/mothur*logfile mothur*logfile
 ```

### 2016-01-18 code/build_culture_effort_plot.R
Modify to work with updated data

@@ -11,7 +11,6 @@ b_is_cultured <- b_cult_time$sobs_precent > 0
 b_cultured <- b_cult_time[b_is_cultured,colnames(b_cult_time) %in% c("nseqs_precent", "sobs_precent")]
 b_order <- order(b_cultured$sobs_precent, decreasing=T)
 b_cultured <- b_cultured[b_order,]
-b_cultured <- b_cultured[rownames(b_cultured) != "Euryarchaeota",] #needs to be removed
 b_n_taxa <- nrow(b_cultured)
 
 

### 2016-01-18 code/get_phylum_category_counts.R

@@ -8,21 +8,18 @@ get_phylum <- function(taxonomy){
 categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other")
 
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact$category[bact$category=="ZI" & !is.na(bact$category)] <- "ZA" #to remove
 
 bact_phylum <- get_phylum(bact$taxonomy)
-bact_phylum[bact_phylum=="Euryarchaeota"] <- NA #to be removed
 bact_phylum[bact_phylum=="aquifer1"] <- NA
 
 b_phylum_cat_count <- table(bact_phylum, bact$category)
 b_phylum_cat_count <- cbind(b_phylum_cat_count, total=table(bact_phylum))
 
 rownames(b_phylum_cat_count) <- gsub("_?\\(.*\\)", "", rownames(b_phylum_cat_count) )
-bacterial_data <- cbind(domain=rep("bacteria", nrow(b_phylum_cat_count)), b_phylum_cat_count)
+bacterial_data <- cbind(phylum=rownames(b_phylum_cat_count), domain=rep("bacteria", nrow(b_phylum_cat_count)), b_phylum_cat_count)
 
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch$category[arch$category=="BDBD" & !is.na(arch$category)] <- "BD" #to remove
 
 arch_phylum <- get_phylum(arch$taxonomy)
 
@@ -30,7 +27,7 @@ a_phylum_cat_count <- table(arch_phylum, factor(arch$category, levels=names(cate
 a_phylum_cat_count <- cbind(a_phylum_cat_count, total=table(arch_phylum))
 
[…]

### 2016-01-18 code/get_phylum_category_counts.R
Add aquifer2 data

@@ -10,7 +10,6 @@ categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Fre
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 
 bact_phylum <- get_phylum(bact$taxonomy)
-bact_phylum[bact_phylum=="aquifer1"] <- NA
 
 b_phylum_cat_count <- table(bact_phylum, bact$category)
 b_phylum_cat_count <- cbind(b_phylum_cat_count, total=table(bact_phylum))

### 2016-01-18 code/get_otu_overlap_by_method.R
Parse out EMIRGE data

@@ -1,13 +1,20 @@
 is_cultured <- function(db){
-	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db))
+	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db)) & !is_single_cell(db)
 }
 
 is_pcr <- function(db){
-	!((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db))
+
+	!((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db)) & !is_emirge(db)
 }
 
 is_single_cell <- function(db){
-	grepl("^[^.]*$", rownames(db)) | db$publication_doi == "10.1038/nature12352"
+	(grepl("^[^.]*$", rownames(db)) | (db$publication_doi == "10.1038/nature12352" & !is.na(db$publication_doi))) & !is_emirge(db)
+}
+
+is_emirge <- function(db){
+	emirge_dois <- c("10.1126/science.1224041", "10.1186/2049-2618-2-1", "10.1186/gb-2011-12-5-r44", "10.1371/journal.pone.0056018", "10.1371/journal.pone.0057819", "10.3389/fmicb.2015.00277")
+
+	db$publication_doi %in% emirge_dois
 }
 
 
@@ -18,6 +25,9 @@ get_data <- function(db){
 	pcrd <- is_pcr(db)
 	pcr_otus <- unique(db[pcrd, "otu"])
 
+	emirge <- is_emirge(db)
+	emirge_otus <- unique(db[emirge, "otu"])
+
 	sc <- is_single_cell(db)
 	sc_otus <- unique(db[sc, "otu"])
 
@@ -26,6 +36,7 @@ get_data <- function(db){
 
 	n_cult_otus <- length(cultured_otus) #Cultured OTUs	1
 	n_pcr_otus <- length(pcr_otus) #PCR OTUs				2
+	n_emirge_otus <- length(emirge_otus)
 	n_sc_otus <- length(sc_otus) #Single cell OTUs		3
 
 	n_cult_only <- sum(!cultured_otus %in% union(pcr_otus, sc_otus))

### 2016-01-19 code/get_coverage_by_category_and_time.R
Exclude single cell and EMIRGE data

@@ -114,17 +114,20 @@ summarize_category_data <- function(db){
 	cbind(db_count_pre, db_sobs_pre, db_good_coverage_pre, db_schloss_coverage_pre, db_count_post, db_sobs_post, db_good_coverage_post, db_schloss_coverage_post, db_count, db_sobs, db_good_coverage, db_schloss_coverage)
 }
 
+source("code/partition_data.R")
 
 abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQM", "AQMS", "AQH", "AQI", "AQO", "BD", "BF", "BI", "BP", "BO", "PR", "PS", "PO", "SA", "SD", "SP", "SO", "ZV", "ZA", "ZN",  "ZO", "OT", "no_source", "total")
 
 
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact_summary <- summarize_category_data(bact)
+
+bact_summary <- summarize_category_data(bact[is_pcr(bact) | is_cultured(bact),])
 colnames(bact_summary) <- gsub("db_", "bact_", colnames(bact_summary))
 
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch_summary <- summarize_category_data(arch)
+
+arch_summary <- summarize_category_data(arch[is_pcr(arch) | is_cultured(arch),])
 colnames(arch_summary) <- gsub("db_", "arch_", colnames(arch_summary))
 
 composite <- cbind(bact_summary, arch_summary)

### 2016-01-19 code/get_culture_by_category_counts.R

@@ -1,3 +1,5 @@
+source("code/partition_data.R")
+
 get_phylum <- function(taxonomy){
 	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
 	phylum[grepl(';', phylum)] <- "Unclassified"
@@ -41,6 +43,6 @@ run_domain <- function(domain){
 	output_file <- paste0('data/process/', domain, '.cultured_by_time_counts.tsv')
 
 	domain_data <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	domain_data <- get_domain_data(domain_data)
+	domain_data <- get_domain_data(domain_data[is_pcr(domain_data) | is_cultured(domain_data),])
 	write.table(domain_data, file=output_file, sep='\t', quote=FALSE)
 }

### 2016-01-19 code/get_culture_by_time_counts.R

@@ -1,3 +1,5 @@
+source("code/partition_data.R")
+
 get_phylum <- function(taxonomy){
 	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
 	phylum[grepl(';', phylum)] <- "Unclassified"
@@ -49,6 +51,6 @@ run_domain <- function(domain){
 	output_file <- paste0('data/process/', domain, '.cultured_by_time_counts.tsv')
 
 	domain_data <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	domain_data <- get_domain_data(domain_data)
+	domain_data <- get_domain_data(domain_data[is_pcr(domain_data) | is_cultured(domain_data),])
 	write.table(domain_data, file=output_file, sep='\t', quote=FALSE)
 }

### 2016-01-19 code/partition_data.R

@@ -1,0 +1,23 @@
+is_cultured <- function(db){
+	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db)) & !is_single_cell(db) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+}
+
+is_pcr <- function(db){
+	!((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db)) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+}
+
+is_single_cell <- function(db){
+	(grepl("^[^.]*$", rownames(db)) | (db$publication_doi == "10.1038/nature12352" & !is.na(db$publication_doi))) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+}
+
+is_emirge_pcr <- function(db){
+	emirge_dois <- c("10.1186/2049-2618-2-1", "10.1371/journal.pone.0056018", "10.1371/journal.pone.0057819")
+
+	db$publication_doi %in% emirge_dois
+}
+
+is_emirge_metag <- function(db){
+	emirge_dois <- c("10.1126/science.1224041", "10.1186/gb-2011-12-5-r44", "10.3389/fmicb.2015.00277")
+
+	db$publication_doi %in% emirge_dois
+}

### 2016-01-19 code/partition_data.R
Add comment indicating EMIRGE studies and their size

@@ -21,3 +21,10 @@ is_emirge_metag <- function(db){
 
 	db$publication_doi %in% emirge_dois
 }
+
+#	10.1126/science.1224041			99		genomes
+#	10.1186/gb-2011-12-5-r44		4		genomes
+#	10.3389/fmicb.2015.00277 		17		genomes
+#	10.1186/2049-2618-2-1 			21447	PCR
+#	10.1371/journal.pone.0056018	37386	PCR
+#	10.1371/journal.pone.0057819	23		PCR

### 2016-01-19 code/get_culture_by_category_counts.R
Fix partitioning

@@ -7,10 +7,6 @@ get_phylum <- function(taxonomy){
 	phylum
 }
 
-is_cultured <- function(db){
-	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db))
-}
-
 get_cultured_data <- function(cultured, phyla, otu){
 	nseqs <- table(phyla, cultured)
 	nseqs <- cbind(cult_nseqs=nseqs[,"TRUE"], total_nseqs=apply(nseqs, 1, sum))

### 2016-01-19 code/get_culture_by_time_counts.R

@@ -11,14 +11,6 @@ get_year <- function(date){
 	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date)
 }
 
-is_cultured <- function(db){
-	cultured <- (!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db))
-
-	cultured[rownames(db)=="AJ224540.MxeCultu"] <- FALSE
-	cultured[db$publication_doi=="10.1073/pnas.1219809110"] <- FALSE
-	cultured
-}
-
 get_cultured_data <- function(cultured, phyla, otu){
 	nseqs <- table(phyla, cultured)
 	nseqs <- cbind(cult_nseqs=nseqs[,"TRUE"], total_nseqs=apply(nseqs, 1, sum))

### 2016-01-19 code/partition_data.R

@@ -1,5 +1,8 @@
 is_cultured <- function(db){
-	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db)) & !is_single_cell(db) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+	cultured <- (!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db)) & !is_single_cell(db) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+
+	cultured[rownames(db)=="AJ224540.MxeCultu"] <- FALSE
+	cultured
 }
 
 is_pcr <- function(db){

### 2016-01-19 code/partition_data.R
Assign the weird sequence to PCR

@@ -6,7 +6,10 @@ is_cultured <- function(db){
 }
 
 is_pcr <- function(db){
-	!((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db)) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+	pcr <- !((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db)) & !(is_emirge_pcr(db) | is_emirge_metag(db))
+
+	pcr[rownames(db)=="AJ224540.MxeCultu"] <- TRUE
+	pcr
 }
 
 is_single_cell <- function(db){

### 2016-01-19 code/get_otu_overlap_by_method.R
Incorporate EMIRGE data

@@ -1,77 +1,68 @@
-is_cultured <- function(db){
-	(!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db)) & grepl("\\.", rownames(db)) & !is_single_cell(db)
+source("code/partition_data.R")
+
+run_analysis <- function(ref, otus){
+	ref_otus <- otus[[ref]]
+	n_otus <- length(ref_otus)
+
+	cultured <- sum(ref_otus %in% otus[["cultured"]])
+	pcr <- sum(ref_otus %in% otus[["pcr"]])
+	sc <- sum(ref_otus %in% otus[["sc"]])
+	em_pcr <- sum(ref_otus %in% otus[["em_pcr"]])
+	em_meta <- sum(ref_otus %in% otus[["em_meta"]])
+	freq <- c(cultured=cultured, pcr=pcr, sc=sc, em_pcr=em_pcr, em_meta=em_meta)
+	freq[ref] <- sum(!ref_otus %in% unique(unlist(otus[ !names(otus) %in% ref])))
+
+	if(n_otus != 0){
+		percentage <- 100 * freq / n_otus
+	} else {
+		percentage <- c(cultured=NA, pcr=NA, sc=NA, em_pcr=NA, em_meta=NA)
+	}
+
 }
 
-is_pcr <- function(db){
 
-	!((!is.na(db$strain) | !is.na(db$isolate)) & !grepl("\\.Unc", rownames(db))) & grepl("\\.", rownames(db)) & !is_emirge(db)
-}
-
-is_single_cell <- function(db){
-	(grepl("^[^.]*$", rownames(db)) | (db$publication_doi == "10.1038/nature12352" & !is.na(db$publication_doi))) & !is_emirge(db)
-}
-
-is_emirge <- function(db){
-	emirge_dois <- c("10.1126/science.1224041", "10.1186/2049-2618-2-1", "10.1186/gb-2011-12-5-r44", "10.1371/journal.pone.0056018", "10.1371/journal.pone.0057819", "10.3389/fmicb.2015.00277")
-
-	db$publication_doi %in% emirge_dois
-}
-
-
-get_data <- function(db){
+compare_methods <- function(db){
 	cultured <- is_cultured(db)
 	cultured_otus <- unique(db[cultured, "otu"])
 
-	pcrd <- is_pcr(db)
[…]

### 2016-01-20 Schloss_Census2_mBio_2015.Rmd

@@
 b_phylum <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
-b_phylum <- b_phylum[!rownames(b_phylum) %in% 'aquifer1',]
 rownames(b_phylum) <- gsub("Candidate_division_", "", rownames(b_phylum))
@@
 dominant_archaea_list <- oxford_comma(paste0(names(dominant_archaea), ' (N=', dominant_archaea, ')'))
-
 ```
@@
 This likely reflects the extremely high number of cultivated  *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
-Conversely, clades such as the Actinobacteria, Fusobacteria, and Bacteroidetes had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
-Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
-
-In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
-To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
-Overall, 18.6% of bacterial sequences and 6.8% of archaeal sequences have come from isolated organisms.
-Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by 2.5 and 2.4-fold for bacteria and archaea, respectively.
-Among the 67 bacterial phyla, 19 have no cultured representatives and 20 of the 10 archaeal phyla have cultured representatives.
-This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
-Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only 2.4% for the bacterial phyla and 1.7% for the archaeal phyla (Figure phylum_effort.pdf).
-So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
-Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
-We found that 13.0% of the 95,734 bacterial OTUs and 9.1% of the 4,205 archeael OTUs had at least one cultured representative (Figure venn_otu_by_method.pdf).
-Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Fig.
-results/figures/phylum_cultured.pdf).
-The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
-This likely reflects the extremely high number of cultivated *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
-Conversely, clades such as the Actinobacteria, Fusobacteria, and Bacteroidetes had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+Conversely, clades such as the *Actinobacteria*, *Fusobacteria*, and *Bacteroidetes* had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
 Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.

### 2016-01-20 results/tables/build_bacterial_culture_effort_table.Rmd

@@
 data <- read.table(file="data/process/bacteria.cultured_by_time_counts.tsv", header=T, stringsAsFactors=F)
-data <- data[!rownames(data) %in% "aquifer1",]
 all <- data[,grepl("all_", colnames(data))]

### 2016-01-20 results/tables/build_bacterial_phylum_effort_table.Rmd

@@
 data <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
-data <- data[!rownames(data) %in% "aquifer1",]
 data <- data[order(data$all, decreasing=T),]

### 2016-01-20 doc/notebook/Archaeal_data_acquisition.Rmd
Put all OTU and method information in metadata files. Will do rarefaction externally

@@
 
+spread_sheet_good$date <- "2013-07-25"
+
 write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
+
+#we'll output the sequence names so the sequences aren't included in the
+#rarefaction analysis
+write(rownames(spread_sheet_good), file="data/mothur/archaea.sc_emirge.accnos")
 ```
@@
 write.table(archaea_emirge_metadata, file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
+
+#we'll output the sequence names so the sequences aren't included in the
+#rarefaction analysis
+write(rownames(archaea_emirge_metadata), file="data/mothur/archaea.sc_emirge.accnos", append=T)
 ```
@@
 ```{r}
-#here we'll read in the list file and extract the OTU assignments for the 0.03
-#cutoff
-list_text <- scan(file="data/mothur/all_archaea.filter.unique.precluster.an.list", sep="\n", what="")[5]
+#here we'll read in the list file and extract the OTU assignments for the
+# unique, 0.03, 0.05, 0.10, 0.20 cutoffs
+list_text <- scan(file="data/mothur/all_archaea.filter.unique.precluster.an.list", sep="\n", what="")[c(2,5,7,12,22)]
+names(list_text) <- c("0.00", "0.03", "0.05", "0.10", "0.20")
 
-#let's remove the otu label and the number of OTUs at this cutoff
-list_otus <- unlist(strsplit(list_text, "\t"))[-c(1,2)]
+#let's remove the otu label and the number of OTUs at each cutoff
+parseList <- function(line){
+
+	list_otus <- unlist(strsplit(line, "\t"))[-c(1,2)]
 	n_otus <- length(list_otus)
@@
 	#now we get the OTU number for each sequence in the dataset
-otu_assignment <- unlist(lapply(1:n_otus, split_otu))
+	unlist(lapply(1:n_otus, split_otu))
+}
+otu_data <- sapply(list_text, parseList)
 
@@
 metadata <- read.table(file="data/mothur/archaea.good.metadata", header=T, stringsAsFactors=FALSE)
-metadata$otu <- otu_assignment[as.character(rownames(metadata))]
+metadata <- cbind(metadata, otu_data[as.character(rownames(metadata)),])
 metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]
@@
 
+#... and we pool the categories by environment
+metadata$environment <- rep(NA, nrow(metadata))
+metadata$environment[metadata$category == "AE"] <- "aerosol"
+metadata$environment[metadata$category == "OT"] <- "other"
+metadata$environment[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
+metadata$environment[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
+metadata$environment[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
+metadata$environment[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
+metadata$environment[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
+
+#let's flag each sequence to indicate how each was generated
+source("code/partition_data.R")
+metadata$pcr <- is_pcr(metadata)
+metadata$cultured <- is_cultured(metadata)
+metadata$single_cell <- is_single_cell(metadata)
+metadata$emirge_pcr <- is_emirge_pcr(metadata)
+metadata$emirge_metag <- is_emirge_metag(metadata)
+
 #... and we output the metadata file...
 write.table(metadata, file="data/process/archaea.v123.metadata", sep="\t")
-
-#... and we output a group file to use for splitting up the list file into the individual categories...
-write.table(file="data/mothur/archaea.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
-
-#... and we pool the categories by environment
-env_categories <- rep(NA, nrow(metadata))
-env_categories[metadata$category == "AE"] <- "aerosol"
-env_categories[metadata$category == "OT"] <- "other"
-env_categories[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
-env_categories[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
-env_categories[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
-env_categories[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
-env_categories[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
-write.table(file="data/mothur/archaea.env_category.groups", x=cbind(rownames(metadata), env_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
 ```
 
-One thing we'd like to look at are the rarefaction curves for each set of group files and the entire dataset. To do this we'll run the mothur command like so..
-
-```{r, engine="bash"}
-mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.all_category.groups, label=unique-0.03-0.05-0.10-0.20);
-        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.all_categories.shared);
-        rarefaction.single(shared=all_archaea.all_categories.shared, outputdir=data/process)"
-
-mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.env_category.groups, label=unique-0.03-0.05-0.10-0.20);
-        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.env_category.shared);
-        rarefaction.single(shared=all_archaea.env_category.shared, outputdir=data/process)"
-
-mothur "#rarefaction.single(list=data/mothur/all_archaea.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20, outputdir=data/process);"
-
-rm data/mothur/*rabund data/mothur/mothur*logfile mothur*logfile
-```
+Viola! We'll upload data/process/archaea.v123.metadata to FigShare and start analyzing the data for the paper.

### 2016-01-20 doc/notebook/Bacterial_data_acquisition.Rmd
Put all OTU and method information in metadata files. Will do rarefaction externally

@@
 Notes...
-
-* Looks like an archaeal sequence slipped through
 * Will want to get rid of sequences with large number of Ns in them - how large?
@@
 
+spread_sheet_good$date <- "2013-07-25"
+
 write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t", col.names=F, append=T)
+
+#we'll output the sequence names so the sequences aren't included in the
+#rarefaction analysis
+write(rownames(spread_sheet_good), file="data/mothur/bacteria.sc_emirge.accnos")
 ```
@@
 write.table(bacteria_emirge_metadata, file="data/mothur/bacteria.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
+
+#we'll output the sequence names so the sequences aren't included in the
+#rarefaction analysis
+write(rownames(bacteria_emirge_metadata), file="data/mothur/bacteria.sc_emirge.accnos", append=T)
 ```
@@
 ```{r}
-#here we'll read in the list file and extract the OTU assignments for the 0.03
-#cutoff (for some reason mothur was repeating the header line for each OTU threshold)
-list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")
+#here we'll read in the list file and extract the OTU assignments for the
+# unique, 0.03, 0.05, 0.10, 0.20 cutoffs
+list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")[c(2,5,7,12,22)]
+names(list_text) <- c("0.00", "0.03", "0.05", "0.10", "0.20")
 
-list_text_03 <- list_text[grepl("^0\\.03", list_text)]
+#let's remove the otu label and the number of OTUs at each cutoff
+parseList <- function(line){
 
-#let's remove the otu label and the number of OTUs at this cutoff
-list_otus <- unlist(strsplit(list_text_03, "\t"))[-c(1,2)]
+	list_otus <- unlist(strsplit(line, "\t"))[-c(1,2)]
 	n_otus <- length(list_otus)
@@
 metadata <- read.table(file="data/mothur/bacteria.good.metadata", header=T, stringsAsFactors=FALSE)
-metadata$otu <- otu_assignment[as.character(rownames(metadata))]
+metadata <- cbind(metadata, otu_data[as.character(rownames(metadata)),])
 metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]
@@
 #correct typos:
-metadata$category[metadata$category == 'AF'] <- 'AE'
-metadata$category[metadata$category == 'AQBF'] <- 'AQB'
 metadata$category[metadata$category == 'AQD'] <- 'AQO'
@@
 metadata$category[metadata$category == 'AQS'] <- 'AQMS'
-metadata$category[metadata$category == 'AWF'] <- 'AQF'
-metadata$category[metadata$category == 'DS'] <- 'BS'
-metadata$category[metadata$category == 'S'] <- 'SO'
-metadata$category[metadata$category == 'SI'] <- 'BI'
-metadata$category[metadata$category == 'V'] <- 'BO'
 metadata$category[metadata$category == 'VD'] <- 'BD'
-metadata$category[metadata$category == 'ZOZO'] <- 'ZO'
+metadata$category[metadata$category == 'BM'] <- 'BI'
 metadata$category[metadata$category == 'AH'] <- 'AQH'
-metadata$category[metadata$category == 'AQOS'] <- 'AQS'
-metadata$category[metadata$category == 'BS'] <- 'BD'
-metadata$category[metadata$category == 'AQS'] <- 'AQFS'
-metadata$category[metadata$category == 'ZL'] <- 'ZV'
-metadata$category[metadata$category == 'SR'] <- 'PR'
-metadata$category[metadata$category == 'M'] <- 'AQM'
-metadata$category[metadata$category == 'AQFM'] <- 'AQFS'
-metadata$category[metadata$category == 'SV'] <- 'ZV'
-metadata$category[metadata$category == 'ZI'] <- 'ZA'
+metadata$category[metadata$category == 'BDBD'] <- 'BD'
 
-table(metadata$category)
+#... and we pool the categories by environment
+metadata$environment <- rep(NA, nrow(metadata))
+metadata$environment[metadata$category == "AE"] <- "aerosol"
+metadata$environment[metadata$category == "OT"] <- "other"
+metadata$environment[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
+metadata$environment[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
+metadata$environment[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
+metadata$environment[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
+metadata$environment[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
 
-#    AE    AQB   AQBS    AQF   AQFS    AQH    AQI    AQM   AQMS    AQO     BD
-#  3473   1094    390  59220  23979  10861   2073 134788  27811   8765  33152
-#    BF     BI     BO     BP     OT     PO     PR     PS     SA     SD     SO
-# 11813  16598   8556  38710  19417   9753  19695   4892  10051   3059  60335
-#    SP     ZA     ZN     ZO     ZV
-#  1922  13209   7476  10855 794492
-
-# need to change the name of this sequence to make it clear that it was not cultured
-# rownames(metadata)[rownames(metadata) == 'AJ224540.MxeCultu'] <- 'AJ224540.UncMxeCultu'
-# doing this here screws up the list file - need to do this in the actual
-# analysis
+#let's flag each sequence to indicate how each was generated
+source("code/partition_data.R")
+metadata$pcr <- is_pcr(metadata)
+metadata$cultured <- is_cultured(metadata)
+metadata$single_cell <- is_single_cell(metadata)
+metadata$emirge_pcr <- is_emirge_pcr(metadata)
+metadata$emirge_metag <- is_emirge_metag(metadata)
 
@@
 write.table(metadata, file="data/process/bacteria.v123.metadata", sep="\t")
-
-#... and we output a group file to use for splitting up the list file into the individual categories...
-write.table(file="data/mothur/bacteria.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
-
-#... and we pool the categories by environment
-env_categories <- rep(NA, nrow(metadata))
-env_categories[metadata$category == "AE"] <- "aerosol"
-env_categories[metadata$category == "OT"] <- "other"
-env_categories[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
-env_categories[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
-env_categories[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
-env_categories[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
-env_categories[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
-write.table(file="data/mothur/bacteria.env_category.groups", x=cbind(rownames(metadata), env_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
 ```
 
-One thing we'd like to look at are the rarefaction curves for each set of group files and the entire dataset. To do this we'll run the mothur command like so..
-
-```{r, engine="bash"}
-mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.all_category.groups, label=unique-0.03-0.05-0.10-0.20);
-        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.all_categories.shared);
-        rarefaction.single(shared=all_bacteria.all_categories.shared, outputdir=data/process)"
-
-mothur "#set.dir(input=data/mothur, output=data/mothur);
-        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.env_category.groups, label=unique-0.03-0.05-0.10-0.20);
-        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.env_category.shared);
-        rarefaction.single(shared=all_bacteria.env_category.shared, outputdir=data/process)"
-
-mothur "#rarefaction.single(list=data/mothur/all_bacteria.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20, outputdir=data/process);"
-
-rm data/mothur/*rabund data/mothur/mothur*logfile mothur*logfile
-```
+Viola! We'll upload data/process/bacteria.v123.metadata to FigShare and start analyzing the data for the paper.

### 2016-01-20 code/get_coverage_by_category_and_time.R
Use method columns from metadata files instead of functions

@@ -114,20 +114,19 @@ summarize_category_data <- function(db){
 	cbind(db_count_pre, db_sobs_pre, db_good_coverage_pre, db_schloss_coverage_pre, db_count_post, db_sobs_post, db_good_coverage_post, db_schloss_coverage_post, db_count, db_sobs, db_good_coverage, db_schloss_coverage)
 }
 
-source("code/partition_data.R")
 
 abbreviation <- c("AE", "AQB", "AQBS", "AQF", "AQFS", "AQM", "AQMS", "AQH", "AQI", "AQO", "BD", "BF", "BI", "BP", "BO", "PR", "PS", "PO", "SA", "SD", "SP", "SO", "ZV", "ZA", "ZN",  "ZO", "OT", "no_source", "total")
 
 
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 
-bact_summary <- summarize_category_data(bact[is_pcr(bact) | is_cultured(bact),])
+bact_summary <- summarize_category_data(bact[bact$pcr | bact$cultured,])
 colnames(bact_summary) <- gsub("db_", "bact_", colnames(bact_summary))
 
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
 
-arch_summary <- summarize_category_data(arch[is_pcr(arch) | is_cultured(arch),])
+arch_summary <- summarize_category_data(arch[arch$pcr | arch$cultured,])
 colnames(arch_summary) <- gsub("db_", "arch_", colnames(arch_summary))
 
 composite <- cbind(bact_summary, arch_summary)

### 2016-01-20 code/get_culture_by_category_counts.R

@@ -1,5 +1,3 @@
-source("code/partition_data.R")
-
 get_phylum <- function(taxonomy){
 	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
 	phylum[grepl(';', phylum)] <- "Unclassified"
@@ -21,7 +19,7 @@ get_cultured_data <- function(cultured, phyla, otu){
 get_domain_data <- function(db, threshold=2006){
 
 	phyla <- factor(get_phylum(db$taxonomy))
-	cultured <- is_cultured(db)
+	cultured <- db$cultured
 
 	pre <- get_year(db$date) < threshold
 
@@ -39,6 +37,6 @@ run_domain <- function(domain){
 	output_file <- paste0('data/process/', domain, '.cultured_by_time_counts.tsv')
 
 	domain_data <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	domain_data <- get_domain_data(domain_data[is_pcr(domain_data) | is_cultured(domain_data),])
+	domain_data <- get_domain_data(domain_data[domain_data$pcr | domain_data$cultured,])
 	write.table(domain_data, file=output_file, sep='\t', quote=FALSE)
 }

### 2016-01-20 code/get_culture_by_time_counts.R

@@ -1,4 +1,3 @@
-source("code/partition_data.R")
 
 get_phylum <- function(taxonomy){
 	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
@@ -25,7 +24,7 @@ get_cultured_data <- function(cultured, phyla, otu){
 get_domain_data <- function(db, threshold=2006){
 
 	phyla <- factor(get_phylum(db$taxonomy))
-	cultured <- is_cultured(db)
+	cultured <- db$cultured
 
 	pre <- get_year(db$date) < threshold
 
@@ -43,6 +42,6 @@ run_domain <- function(domain){
 	output_file <- paste0('data/process/', domain, '.cultured_by_time_counts.tsv')
 
 	domain_data <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	domain_data <- get_domain_data(domain_data[is_pcr(domain_data) | is_cultured(domain_data),])
+	domain_data <- get_domain_data(domain_data[db$pcr | db$cultured,])
 	write.table(domain_data, file=output_file, sep='\t', quote=FALSE)
 }

### 2016-01-20 code/get_otu_overlap_by_method.R

@@ -1,5 +1,3 @@
-source("code/partition_data.R")
-
 run_analysis <- function(ref, otus){
 	ref_otus <- otus[[ref]]
 	n_otus <- length(ref_otus)
@@ -22,20 +20,12 @@ run_analysis <- function(ref, otus){
 
 
 compare_methods <- function(db){
-	cultured <- is_cultured(db)
-	cultured_otus <- unique(db[cultured, "otu"])
 
-	pcr <- (is_pcr(db))
-	pcr_otus <- unique(db[pcr, "otu"])
-
-	sc <- is_single_cell(db)
-	sc_otus <- unique(db[sc, "otu"])
-
-	em_pcr <- is_emirge_pcr(db)
-	em_pcr_otus <- unique(db[em_pcr, "otu"])
-
-	em_meta <- is_emirge_metag(db)
-	em_meta_otus <- unique(db[em_meta, "otu"])
+	cultured_otus <- unique(db[db$cultured, "otu"])
+	pcr_otus <- unique(db[db$pcr, "otu"])
+	sc_otus <- unique(db[db$single_cell, "otu"])
+	em_pcr_otus <- unique(db[db$emirge_pcr, "otu"])
+	em_meta_otus <- unique(db[emirge_meta, "otu"])
 
 	total_otus <- length(unique(db$otu))
 
@@ -49,8 +39,8 @@ compare_methods <- function(db){
 					em_pcr=em_pcr_otus, em_meta=em_meta_otus)
 
 	percentages <- t(sapply(names(otu_list), run_analysis, otus=otu_list))
-	n_seqs <- c(cultured=sum(cultured), pcr=sum(pcr), sc=sum(sc),
-					em_pcr=sum(em_pcr), em_meta=sum(em_meta))
+	n_seqs <- c(cultured=sum(db$cultured), pcr=sum(db$pcr), sc=sum(db$single_cell),
+					em_pcr=sum(db$emirge_pcr), em_meta=sum(db$emirge_meta))
 	n_otus <- c(cultured=length(cultured_otus), pcr=length(pcr_otus),
 				sc=length(sc_otus), em_pcr=length(em_pcr_otus),
  				em_meta=length(em_meta_otus))

### 2016-01-20 code/get_phylum_category_counts.R

@@ -1,5 +1,3 @@
-source("code/partition_data.R")
-
 get_phylum <- function(taxonomy){
 	phylum <- gsub("^[^;]*;([^;]*);.*", "\\1", taxonomy)
 	phylum[grepl(';', phylum)] <- "Unclassified"
@@ -10,7 +8,7 @@ get_phylum <- function(taxonomy){
 categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other")
 
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact <- bact[is_pcr(bact) | is_cultured(bact),]
+bact <- bact[bact$pcr | bact$cultured,]
 
 bact_phylum <- get_phylum(bact$taxonomy)
 
@@ -22,7 +20,7 @@ bacterial_data <- cbind(phylum=rownames(b_phylum_cat_count), domain=rep("bacteri
 
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch <- arch[is_pcr(arch) | is_cultured(arch),]
+arch <- arch[arch$pcr | arch$cultured,]
 arch_phylum <- get_phylum(arch$taxonomy)
 
 a_phylum_cat_count <- table(arch_phylum, factor(arch$category, levels=names(categories)))

### 2016-01-20 code/get_phylum_counts.R

@@ -1,5 +1,3 @@
-source("code/partition_data.R")
-
 get_year <- function(date){
 	gsub("^(\\d\\d\\d\\d)-.*", "\\1", date)
 }
@@ -38,5 +36,5 @@ generate_table <- function(domain, threshold=2006){
 
 
 	input <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	write.table(summarize_phylum_data(input[is_pcr(input) | is_cultured(input),], threshold), file=count_file, quote=FALSE, sep='\t')
+	write.table(summarize_phylum_data(input[input$pcr | input$cultured,], threshold), file=count_file, quote=FALSE, sep='\t')
 }

### 2016-01-20 code/get_time_course_submission_data.R

@@ -1,5 +1,3 @@
-source("code/partition_data.R")
-
 first_year <- 1983
 final_year  <- 2015
 
@@ -9,12 +7,12 @@ get_year <- function(date_vector){
 
 
 bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-bact <- bact[is_pcr(bact) | is_cultured(bact),]
+bact <- bact[bact$pcr | bact$cultured,]
 
 bact_year_deposited <- get_year(bact$date)
 
 arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
-arch <- arch[is_pcr(arch) | is_cultured(arch),]
+arch <- arch[arch$pcr | arch$cultured,]
 
 arch_year_deposited <- get_year(arch$date)
 

### 2016-01-20 code/get_coverage_by_category_and_time.R
Use 0.03 column from metadata files

@@ -34,11 +34,11 @@ get_sobs <- function(category, db){
 	sobs <- NA
 
 	if(category == "no_source"){
-		sobs <- length(table(db[is.na(db$category), "otu"]))
+		sobs <- length(table(db[is.na(db$category), "X0.03"]))
 	} else if(category == "total"){
-		sobs <- length(table(db[, "otu"]))
+		sobs <- length(table(db[, "X0.03"]))
 	} else {
-		sobs <- length(table(db[db$category==category, "otu"]))
+		sobs <- length(table(db[db$category==category, "X0.03"]))
 	}
 	return(sobs)
 }
@@ -48,11 +48,11 @@ get_sobs <- function(category, db){
 # specified dataset (db)
 get_goods_coverage <- function(category, db){
 	if(category=="no_source"){
-		category_table <- table(db[is.na(db$category), "otu"])
+		category_table <- table(db[is.na(db$category), "X0.03"])
 	} else if(category == "total") {
-		category_table <- table(db[, "otu"])
+		category_table <- table(db[, "X0.03"])
 	} else {
-		category_table <- table(db[db$category==category, "otu"])
+		category_table <- table(db[db$category==category, "X0.03"])
 	}
 
 	singletons <- sum(category_table == 1)
@@ -70,11 +70,11 @@ get_goods_coverage <- function(category, db){
 # specified dataset (db)
 get_schloss_coverage <- function(category, db){
 	if(category=="no_source"){
-		category_table <- table(db[is.na(db$category), "otu"])
+		category_table <- table(db[is.na(db$category), "X0.03"])
 	} else if(category == "total") {
-		category_table <- table(db[, "otu"])
+		category_table <- table(db[, "X0.03"])
 	} else {
-		category_table <- table(db[db$category==category, "otu"])
+		category_table <- table(db[db$category==category, "X0.03"])
 	}
 
 	singletons <- sum(category_table == 1)

### 2016-01-20 code/get_culture_by_category_counts.R

@@ -23,9 +23,9 @@ get_domain_data <- function(db, threshold=2006){
 
 	pre <- get_year(db$date) < threshold
 
-	all <- get_cultured_data(cultured, phyla, db$otu)
-	pre <- get_cultured_data(cultured[pre], phyla[pre], db$otu[pre])
-	post <- get_cultured_data(cultured[!pre], phyla[!pre], db$otu[!pre])
+	all <- get_cultured_data(cultured, phyla, db[,"X0.03"])
+	pre <- get_cultured_data(cultured[pre], phyla[pre], db[pre, "X0.03"])
+	post <- get_cultured_data(cultured[!pre], phyla[!pre], db[!pre, "X0.03"])
 	pool <- cbind(pre, post, all)
 	colnames(pool) <- paste0(c(rep("pre_", 3), rep("post_", 3), rep("all_", 3)), colnames(pool))
 	pool

### 2016-01-20 code/get_culture_by_time_counts.R

@@ -28,9 +28,9 @@ get_domain_data <- function(db, threshold=2006){
 
 	pre <- get_year(db$date) < threshold
 
-	all <- get_cultured_data(cultured, phyla, db$otu)
-	pre <- get_cultured_data(cultured[pre], phyla[pre], db$otu[pre])
-	post <- get_cultured_data(cultured[!pre], phyla[!pre], db$otu[!pre])
+	all <- get_cultured_data(cultured, phyla, db[, "X0.03"])
+	pre <- get_cultured_data(cultured[pre], phyla[pre], db[pre, "X0.03"])
+	post <- get_cultured_data(cultured[!pre], phyla[!pre], db[!pre, "X0.03"])
 	pool <- cbind(pre, post, all)
 	colnames(pool) <- paste0(c(rep("pre_", 4), rep("post_", 4), rep("all_", 4)), colnames(pool))
 	pool

### 2016-01-20 code/get_otu_overlap_by_method.R

@@ -21,13 +21,13 @@ run_analysis <- function(ref, otus){
 
 compare_methods <- function(db){
 
-	cultured_otus <- unique(db[db$cultured, "otu"])
-	pcr_otus <- unique(db[db$pcr, "otu"])
-	sc_otus <- unique(db[db$single_cell, "otu"])
-	em_pcr_otus <- unique(db[db$emirge_pcr, "otu"])
-	em_meta_otus <- unique(db[emirge_meta, "otu"])
+	cultured_otus <- unique(db[db$cultured, "X0.03"])
+	pcr_otus <- unique(db[db$pcr, "X0.03"])
+	sc_otus <- unique(db[db$single_cell, "X0.03"])
+	em_pcr_otus <- unique(db[db$emirge_pcr, "X0.03"])
+	em_meta_otus <- unique(db[emirge_meta, "X0.03"])
 
-	total_otus <- length(unique(db$otu))
+	total_otus <- length(unique(db[,"X0.03"]))
 
 	n_cult_otus <- length(cultured_otus)
 	n_pcr_otus <- length(pcr_otus)

### 2016-01-20 code/get_time_course_submission_data.R

@@ -55,10 +55,10 @@ get_annual_otu_deposits <- function(year_deposited, otus){
 	new_otus_by_year
 }
 
-bact_data <- get_annual_otu_deposits(bact_year_deposited, bact$otu)
+bact_data <- get_annual_otu_deposits(bact_year_deposited, bact[,"X0.03"])
 summary_table[names(bact_data),'b_sobs'] <- bact_data
 
-arch_data <- get_annual_otu_deposits(arch_year_deposited, arch$otu)
+arch_data <- get_annual_otu_deposits(arch_year_deposited, arch[,"X0.03"])
 summary_table[names(arch_data),'a_sobs'] <- arch_data
 
 

### 2016-01-20 code/rarefy_metadata.R

@@ -1,0 +1,62 @@
+rarefy <- function(data, breaks, n_partitions=20, n_iters=100){
+
+	n_seqs <- length(data)
+
+	sampling <- sample(data, n_seqs, replace=FALSE) #FALSE is the default
+
+	collectors_curve <- function(){
+		sapply(breaks, function(x){length(unique(sampling[1:x]))})
+	}
+
+	collectors <- replicate(n_iters, collectors_curve())
+	rarefied <- apply(collectors, 1, mean)
+	return(rarefied)
+}
+
+
+rarefy_by_coarse <- function(db_name, n_partitions=20, n_iters=100,
+						threshold="X0.03"){
+
+	db <- read.table(file=db_name, header=T, stringsAsFactors=FALSE)
+	db <- db[(db$cultured | db$pcr) & !is.na(db$environment), c("environment", threshold)]
+
+	env_counts <- table(db$environment)
+	breaks <- round(seq(1, max(env_counts), length.out=n_partitions))
+	breaks <- unique(unname(sort(c(breaks, env_counts))))
+
+	environments <- names(env_counts)
+
+	rarefied_data <- matrix(rep(NA, length(environments) * length(breaks)), nrow=length(breaks))
+	colnames(rarefied_data) <- environments
+	rownames(rarefied_data) <- breaks
+
+	for(e in environments){
+		env_otus <-db[db$environment == e,threshold]
+		env_breaks <- breaks[breaks <= env_counts[e]]
+
+		rarefied_data[as.character(env_breaks),e] <- rarefy(env_otus, env_breaks, n_iters)
+	}
+
+	output_name <- gsub("v123.metadata", "env_rarefaction.tsv", db_name)
+	write.table(file=output_name, rarefied_data, quote=F, sep="\t")
+}
+
+
+rarefy_all <- function(db_name, n_partitions=20, n_iters=100,
+						thresholds=c("X0.00", "X0.03", "X0.05", "X0.10", "X0.20")){
+
+	db <- read.table(file=db_name, header=T, stringsAsFactors=FALSE)
+	db <- db[db$cultured | db$pcr, thresholds]
+
+	breaks <- round(seq(1, nrow(db), length.out=n_partitions))
+	rarefied_data <- matrix(rep(0, n_partitions * length(thresholds)), nrow=n_partitions)
+	colnames(rarefied_data) <- thresholds
+	rownames(rarefied_data) <- breaks
+
[…]

### 2016-01-20 Schloss_Census2_mBio_2015.Rmd
Replace typo corrections that were overwritten

@@
 
-***The ability to sample microbial life is taxonomically skewed (meh.)***
+***The ability to sample microbial life is taxonomically skewed.***
 The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences.
@@
 
-***Focusing the census by environment***
+***Focusing the census by environment.***
 We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database.
@@
 
-***The cultured census***
+***The cultured census.***
 In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
@@
 
-***New technologies to access novel biodiversity***
-Assembly of metagenomic and single cell shotgun sequence data offers the hope of identifying large fragments of genomic data from as yet uncultured organisms.
+***New technologies to access novel biodiversity.***
+Given the shift away from Sanger sequencing to platforms that offer higher throughput but with shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled.
+Several culture-idnependent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
+One option is the use of single cell genomics where uncultured bacteria are isolated by dilution and then their genomes are amplified and sequenced.
 To test the ability of single cell technologies to expand our knowledge of microbial diversity beyond that of the 16S rRNA gene and pure cultures, we compared the overlap of OTUs found by the three methods (Figure results/figures/venn_otu_by_method.pdf).
@@
 Although the majority of single-cell genome projects have originated from a single study [@REF], `r b_expand_genome` and `r a_expand_genome` of the bacterial and archaeal OTUs, respectively, were from previously uncultured organisms.
-This represents an encouraging avenue to expanding our knowledge of bacterial diversity beyond the 16S rRNA gene.
+Another approach is to assemble sequence reads from metagenomic sequencing projects that correspond to the 16S rRNA gene.
+This has been difficult because of the risk of generating assembly artifacts as a result of the high level of similarity among conserved domains within 16S rRNA genes.
+The development of the EMIRGE algorithm has facilitated the accurate assembly of 16S rRNA gene fragments [@REF].
+We obtained XXX sequences from 3 studies that assembled 16S rRNA genes from metagenomes, which belonged to XX OTUs.
+Comparing the OTUs found from the metagenomes to the other methods we found no overlap with the OTUs generated by single cell genome sequencing
 

### 2016-01-20 doc/notebook/Bacterial_data_acquisition.Rmd
Replace typo corrections that were overwritten

@@
 #correct typos:
+metadata$category[metadata$category == 'AF'] <- 'AE'
+metadata$category[metadata$category == 'AQBF'] <- 'AQB'
 metadata$category[metadata$category == 'AQD'] <- 'AQO'
@@
 metadata$category[metadata$category == 'AQS'] <- 'AQMS'
+metadata$category[metadata$category == 'AWF'] <- 'AQF'
+metadata$category[metadata$category == 'DS'] <- 'BS'
+metadata$category[metadata$category == 'S'] <- 'SO'
+metadata$category[metadata$category == 'SI'] <- 'BI'
+metadata$category[metadata$category == 'V'] <- 'BO'
 metadata$category[metadata$category == 'VD'] <- 'BD'
-metadata$category[metadata$category == 'BM'] <- 'BI'
+metadata$category[metadata$category == 'ZOZO'] <- 'ZO'
 metadata$category[metadata$category == 'AH'] <- 'AQH'
-metadata$category[metadata$category == 'BDBD'] <- 'BD'
+metadata$category[metadata$category == 'AQOS'] <- 'AQS'
+metadata$category[metadata$category == 'BS'] <- 'BD'
+metadata$category[metadata$category == 'AQS'] <- 'AQFS'
+metadata$category[metadata$category == 'ZL'] <- 'ZV'
+metadata$category[metadata$category == 'SR'] <- 'PR'
+metadata$category[metadata$category == 'M'] <- 'AQM'
+metadata$category[metadata$category == 'AQFM'] <- 'AQFS'
+metadata$category[metadata$category == 'SV'] <- 'ZV'
+metadata$category[metadata$category == 'ZI'] <- 'ZA'
 

### 2016-01-20 code/get_otu_overlap_by_method.R
Fix typo

@@ -25,7 +25,7 @@ compare_methods <- function(db){
 	pcr_otus <- unique(db[db$pcr, "X0.03"])
 	sc_otus <- unique(db[db$single_cell, "X0.03"])
 	em_pcr_otus <- unique(db[db$emirge_pcr, "X0.03"])
-	em_meta_otus <- unique(db[emirge_meta, "X0.03"])
+	em_meta_otus <- unique(db[db$emirge_meta, "X0.03"])
 
 	total_otus <- length(unique(db[,"X0.03"]))
 

### 2016-01-20 code/build_domain_rarefaction_plot.R
Update rarefaction plot code for new format

@@ -1,23 +1,21 @@
-my_domain_col <- c(unique=rainbow(5)[1], X0.03='orange', X0.05=rainbow(5)[3], X0.10=rainbow(5)[4], X0.20=rainbow(5)[5] )
-my_domain_pch <- c(unique=16, X0.03=16, X0.05=17, X0.10=21, X0.20=22)
+my_domain_col <- c( X0.00=rainbow(5)[1], X0.03='orange', X0.05=rainbow(5)[3], X0.10=rainbow(5)[4], X0.20=rainbow(5)[5] )
+my_domain_pch <- c(X0.00=16, X0.03=16, X0.05=17, X0.10=21, X0.20=22)
 
 plot_domain_curves <- function(rarefaction){
-	nseqs <- rarefaction$numsampled
-	rarefaction <- rarefaction[,-1]
-	rarefaction <- rarefaction[,!grepl("ci", colnames(rarefaction))]
+	nseqs <- as.numeric(rownames(rarefaction))
 
-	xlim <- c(1,max(nseqs))
-	ylim <- c(1, 1.05* max(rarefaction$unique))
+	points <- round(seq(1, length(nseqs), length.out=5))
+
+	xlim <- c(1, max(nseqs))
+	ylim <- c(1, 1.15*max(rarefaction))
 
 	plot(NA, xlim=xlim, ylim=ylim, type='n', xlab="", ylab="", main="", axes=FALSE)
 	for(cutoff in colnames(rarefaction)){
 		points(rarefaction[,cutoff]~nseqs, type="l", lwd=2, col=my_domain_col[cutoff])
 
-		spacing <- floor(seq(1,length(nseqs),length.out=10))[-1]
-		points(rarefaction[spacing,cutoff]~nseqs[spacing], col=my_domain_col[cutoff], pch=my_domain_pch[cutoff], bg="white", cex=1.25)
+		points(rarefaction[points,cutoff]~nseqs[points], col=my_domain_col[cutoff], pch=my_domain_pch[cutoff], bg="white", cex=1.25)
 	}
 	box()
-
 }
 
 my_categories <- c(aerosol="Aerosol", aquatic="Aquatic", built="Built", plant="Plant", soil="Soil", zoonotic="Zoonotic", other="Other")
@@ -27,39 +25,35 @@ names(my_categories_col) <- names(my_categories)
 
[…]

### 2016-01-20 code/get_culture_by_time_counts.R
Fix typo in code

@@ -42,6 +42,6 @@ run_domain <- function(domain){
 	output_file <- paste0('data/process/', domain, '.cultured_by_time_counts.tsv')
 
 	domain_data <- read.table(file=metadata_file, header=T, row.names=1, stringsAsFactors=FALSE)
-	domain_data <- get_domain_data(domain_data[db$pcr | db$cultured,])
+	domain_data <- get_domain_data(domain_data[domain_data$pcr | domain_data$cultured,])
 	write.table(domain_data, file=output_file, sep='\t', quote=FALSE)
 }

### 2016-01-20 code/build_culture_effort_plot.R
Stretch x-axis and move legend down

@@ -32,7 +32,7 @@ total_lines <- b_n_taxa + a_n_taxa + 2
 
 pdf("results/figures/phylum_cultured.pdf",  width=3.75, height=4.75)
 par(mar=c(5,10,1.0,0.5), cex=0.7)
-plot(NA, xlim=c(0,40), ylim=c(1,total_lines), axes=F, xlab="", ylab="")
+plot(NA, xlim=c(0,50), ylim=c(1,total_lines), axes=F, xlab="", ylab="")
 abline(h=c(1:a_n_taxa, (a_n_taxa+3):total_lines), lty=3, col="gray")
 
 points(x=a_cultured$nseqs_precent, y=a_n_taxa:1, pch=1)
@@ -47,7 +47,7 @@ axis(2, at=a_n_taxa:1, label=rownames(a_cultured), las=1, tick=F)
 box()
 axis(1, at=seq(0,50,10), label=seq(0,50,10))
 
-legend(x=20, y=20, legend=c("Cultured sequences", "Cultured OTUs"), pch=c(1, 19), bg="white", cex=0.8)
+legend(x=20, y=15, legend=c("Cultured sequences", "Cultured OTUs"), pch=c(1, 19), bg="white", cex=0.8)
 mtext("Bacteria", side=2, line=5, at=total_lines+1.2, las=1, font=2, cex=0.8, xpd=T)
 mtext("Archaea", side=2, line=5, at=a_n_taxa+1.2, las=1, font=2, cex=0.8)
 

### 2016-01-21 code/build_otu_overlap_by_method_figure.R
First version of new overlap plot

@@ -1,80 +1,46 @@
-require("plotrix")
 
-pds_venn <- function(A, B, C, AB, AC, BC, ABC, categories,
-	cir.lwd = 2
-){
+overlap <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
 
-	par(mar=c(1,2,3,2), cex=0.7)
-	plot(NA, xlim=c(0,1), ylim=c(0,1), axes=F, xlab="", ylab="")
 
-	radius <- 0.325
-	draw.circle(radius, 1-radius, radius, lwd=cir.lwd)	#A
-	draw.circle(1-radius, 1-radius, radius, lwd=cir.lwd)	#B
-	draw.circle(0.5, radius, radius, lwd=cir.lwd)	#C
 
-	n_ABC <- ABC
-	n_A <- A - AB - AC + n_ABC
-	n_B <- B - AB - BC + n_ABC
-	n_C <- C - AC - BC + n_ABC
+plot_overlap <- function(domain){
 
-	n_AB <- AB - n_ABC
-	n_AC <- AC - n_ABC
-	n_BC <- BC - n_ABC
+	par(mar=c(0.5, 3,0.5,0.5))
+	d_overlap <- overlap[overlap$domain==domain,-1]
+	rownames(d_overlap) <- d_overlap$method
+	d_overlap <- d_overlap[,-1]
 
-	text(x=0.2, y=0.7, format(n_A, big.mark=','))
-	text(x=0.8, y=0.7, format(n_B, big.mark=','))
-	text(x=0.5, y=0.2, format(n_C, big.mark=','))
 
-	text(x=0.3, y=0.43, format(n_AC, big.mark=','))
-	text(x=0.7, y=0.43, format(n_BC, big.mark=','))
-	text(x=0.5, y=0.75, format(n_AB, big.mark=','))
+	d_subset <- d_overlap[c("sc", "em_pcr", "em_meta"), c("cultured", "pcr", "sc", "em_pcr", "em_meta")]
 
-	text(x=0.5, y=0.56, format(n_ABC, big.mark=','), pos=1)
-
-	text(x=-0.025, y=1.025, categories[1], font=2, adj=0, xpd=T)
-	text(x=1.025, y=1.025, categories[2], font=2, adj=1, xpd=T)
-	text(x=0.5, y=0, categories[3], font=2, pos=1, xpd=T)
[…]

### 2016-01-21 code/build_otu_overlap_by_method_figure.R
Add legend, title, and color

@@ -1,7 +1,7 @@
+#devtools::install_github("karthik/wesanderson")
+library("wesanderson")
 
-overlap <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
-
-
+palette <- "Darjeeling2"
 
 plot_overlap <- function(domain){
 
@@ -13,13 +13,14 @@ plot_overlap <- function(domain){
 
 	d_subset <- d_overlap[c("sc", "em_pcr", "em_meta"), c("cultured", "pcr", "sc", "em_pcr", "em_meta")]
 
-	barplot(t(as.matrix(d_subset)), beside=T, ylim=c(0,105), axes=F, ylab="", names.arg=c("","",""))
+	barplot(t(as.matrix(d_subset)), beside=T, ylim=c(0,105), axes=F, ylab="", names.arg=c("","",""), col=wes_palette(palette))
 	axis(2, las=2)
 	box()
-
+	text(x=17, y=100, gsub("^(.)", "\\U\\1", domain, perl=TRUE), cex=1.5, font=2)
 }
 
 
+overlap <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
 
 
 pdf(file="results/figures/otu_overlap_by_method.pdf", width=6.0, height=5.0)
@@ -27,6 +28,8 @@ pdf(file="results/figures/otu_overlap_by_method.pdf", width=6.0, height=5.0)
 layout(matrix(c(3,1,3,2,0,4), nrow=3, byrow=T), widths=c(0.1,1), heights=c(1, 1, 0.18))
 plot_overlap("bacteria")
 plot_overlap("archaea")
+
+legend(x=5, y=80, legend=c("Cultured", "PCR", "Single Cell", "EMIRGE (PCR)", "EMIRGE (Metagenome)"), fill=wes_palette(palette))
 
 plot.new()
 par(mar=c(0,0,0,0))
@@ -39,8 +42,3 @@ text(x=10.0,y=0.6, "EMIRGE\n(PCR)", cex=1.4, xpd=T)
 text(x=16.3,y=0.6, "EMIRGE\n(Metagenome)", cex=1.4, xpd=T)
 
 dev.off()
-
-
-# Domain headings
-# Legend
-# Color scheme

### 2016-01-21 code/build_domain_rarefaction_plot.R
Alter the colors to use wes anderson palette

@@ -1,4 +1,8 @@
-my_domain_col <- c( X0.00=rainbow(5)[1], X0.03='orange', X0.05=rainbow(5)[3], X0.10=rainbow(5)[4], X0.20=rainbow(5)[5] )
+#devtools::install_github("karthik/wesanderson")
+library("wesanderson")
+clrs <- c(wes_palette("Darjeeling"))
+
+my_domain_col <- c( X0.00=clrs[1], X0.03=clrs[2], X0.05=clrs[3], X0.10=clrs[4], X0.20=clrs[5] )
 my_domain_pch <- c(X0.00=16, X0.03=16, X0.05=17, X0.10=21, X0.20=22)
 
 plot_domain_curves <- function(rarefaction){
@@ -20,7 +24,8 @@ plot_domain_curves <- function(rarefaction){
 
 my_categories <- c(aerosol="Aerosol", aquatic="Aquatic", built="Built", plant="Plant", soil="Soil", zoonotic="Zoonotic", other="Other")
 
-my_categories_col <- rainbow(7)
+set.seed(1)
+my_categories_col <- sample(wes_palette("Zissou", 7, type = "continuous"), 7)#rainbow(7)
 names(my_categories_col) <- names(my_categories)
 
 my_categories_pch <- c(aerosol=19, aquatic=22, built=23, plant=24, soil=15, zoonotic=17, other=21)

### 2016-01-21 code/build_time_course_plots.R

@@ -1,3 +1,9 @@
+#devtools::install_github("karthik/wesanderson")
+library("wesanderson")
+clrs <- wes_palette("Zissou")[c(2,5)]
+
+
+
 time_data <- read.table(file="data/process/by_year_analysis.tsv", header=T)
 year <- as.numeric(rownames(time_data))
 
@@ -15,17 +21,17 @@ pdf(file="results/figures/time_course_figure.pdf", width=3.75, height=6.0)
 	time_data[time_data$a_nseqs == 0, "a_nseqs"] <- 1
 	time_data[time_data$b_nseqs == 0, "b_nseqs"] <- 1
 
-	plot(time_data$b_nseqs~year, log='y', type="l", col="blue", lwd=2, xlab="", ylab="Number of new\nsequences per year", ylim=c(1,1e6), xlim=c(1980, 2015), axes=F)
+	plot(time_data$b_nseqs~year, log='y', type="l", col=clrs[1], lwd=2, xlab="", ylab="Number of new\nsequences per year", ylim=c(1,1e6), xlim=c(1980, 2015), axes=F)
 
-	points(time_data$a_nseqs~year, type="l", col="red", lwd=2)
+	points(time_data$a_nseqs~year, type="l", col=clrs[2], lwd=2)
 
-	points(time_data$b_nseqs[year %% 5==0]~year[year %% 5==0], pch=17, col="blue", cex=1.5)
-	points(time_data$a_nseqs[year %% 5==0]~year[year %% 5==0], pch=19, col="red", cex=1.5)
+	points(time_data$b_nseqs[year %% 5==0]~year[year %% 5==0], pch=17, col=clrs[1], cex=1.5)
+	points(time_data$a_nseqs[year %% 5==0]~year[year %% 5==0], pch=19, col=clrs[2], cex=1.5)
 
 	axis(1, labels=FALSE)
 	axis(2, at=c(1,10,1e2,1e3,1e4,1e5,1e6), labels=parse(text=paste("10^", 0:6)), las=1)
 	box()
-	legend(x=2003, y=100, legend=c("Bacteria", "Archaea"), col=c("blue", "red"), pch=c(17,19), lty=1, lwd=2, pt.cex=1.5)
+	legend(x=2003, y=100, legend=c("Bacteria", "Archaea"), col=c(clrs[1], clrs[2]), pch=c(17,19), lty=1, lwd=2, pt.cex=1.5)
 
[…]

### 2016-01-21 Schloss_Census2_mBio_2015.Rmd
Finish alternative methods section

@@
 
-b_per_unique_sc <- 100*overlap_counts["sc_only", "bact_counts"]/overlap_counts["sc_all", "bact_counts"]
-
-a_per_unique_sc <- 100*overlap_counts["sc_only", "arch_counts"]/overlap_counts["sc_all", "arch_counts"]
-
-b_per_shared <- 100*overlap_counts["cult-pcr-sc", "bact_counts"]/overlap_counts["sc_all","bact_counts"]
-
-b_expand_genome <- 100*(overlap_counts["sc_only", "bact_counts"]+overlap_counts["pcr-sc","bact_counts"]-overlap_counts["cult-pcr-sc","bact_counts"])/overlap_counts["sc_all", "bact_counts"]
-
-a_expand_genome <- 100*(overlap_counts["sc_only", "arch_counts"]+overlap_counts["pcr-sc","arch_counts"]-overlap_counts["cult-pcr-sc","arch_counts"])/overlap_counts["sc_all", "arch_counts"]
-
 ```
 
-* Would be nice if we could comment somewhere on whether or not the EMIRGE/SAG sequences are more or less “efficient” at uncovering new OTUs.
-I.e., does incorporating EMIRGE/SAG sequences increase the OTU coverage at a greater rate, per new sequence, than those created by other methods?
-
 ***New technologies to access novel biodiversity.***
-Given the shift away from Sanger sequencing to platforms that offer higher throughput but with shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled.
-Several culture-idnependent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
-One option is the use of single cell genomics where uncultured bacteria are isolated by dilution and then their genomes are amplified and sequenced.
-To test the ability of single cell technologies to expand our knowledge of microbial diversity beyond that of the 16S rRNA gene and pure cultures, we compared the overlap of OTUs found by the three methods (Figure results/figures/venn_otu_by_method.pdf).
-Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts['n_sc', 'bact_counts']` bacterial and `r overlap_counts['n_sc', 'arch_counts']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts["sc_all", "bact_counts"]` and `r overlap_counts["sc_all", "arch_counts"]` bacterial and archaeal OTUs, respectively.
-Interestingly, only `r b_per_unique_sc` and `r a_per_unique_sc`% of the bacterial and archaeal OTUs, respectively, that the single-cell 16S rRNA gene sequences belonged to had not been observed by cultivation or PCR-based efforts.
-For both domains, the fractions of the single-cell 16S rRNA gene sequences that were recovered by cultivation and PCR were similar.
-Furthermore, among the bacteria `r b_per_shared`% of the single-cell OTUs were previously observed by both cultivation and PCR-based methods.
-Although the majority of single-cell genome projects have originated from a single study [@REF], `r b_expand_genome` and `r a_expand_genome` of the bacterial and archaeal OTUs, respectively, were from previously uncultured organisms.
-Another approach is to assemble sequence reads from metagenomic sequencing projects that correspond to the 16S rRNA gene.
-This has been difficult because of the risk of generating assembly artifacts as a result of the high level of similarity among conserved domains within 16S rRNA genes.
-The development of the EMIRGE algorithm has facilitated the accurate assembly of 16S rRNA gene fragments [@REF].
-We obtained XXX sequences from 3 studies that assembled 16S rRNA genes from metagenomes, which belonged to XX OTUs.
-Comparing the OTUs found from the metagenomes to the other methods we found no overlap with the OTUs generated by single cell genome sequencing
+Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled.
+Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
+These have included single cell genomics and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data using the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@refs].
+To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional culutre-independent amplification and cultivation-based approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure results/figures/venn_otu_by_method.pdf).
+Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
+Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by cultivation or PCR-based efforts.
+Next, we identified three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences.
+Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively.
+Only
+`r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'pcr']`% of the bacterial OTUs generated by this approach were previously identified by this traditional cultivation and PCR-based approaches, respectively.
+Although the application of this approach to Archaea has been limited, it was still suprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
+Finally, we
+pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@REF].
+These sequences were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_otus']` OTUs.
+We were surprised that only  `r 100-overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'em_pcr']`% of these OTUs were previously found by a more traditional approach.
+Although these PCR-based EMIRGE results may be valid, the high degree of novelty that was observed suggests that the error of the assembled reads may be too high for generating reference sequences.
+Each of these methods represent promising opportunities to continue the bacterial census using full-length sequences as well as genomic information.
 

### 2016-01-21 Schloss_Census2_mBio_2015.Rmd
Get 'make write.paper' to compile

@@
 ```{r rarefaction}
-# take these data out of the rarefaciton files so that we can have the unique,
+# take these data out of the rarefaction files so that we can have the unique,
 # 0.03, 0.05, 0.10, and 0.20 OTU definitions for both the bacterial and archaeal
@@
 
-get_sobs <- function(data){
-	rownames(data) <- data[,1]
-	data <- data[,-1]
-	data <- data[,!grepl('ci', colnames(data))]
-	unlist(data[nrow(data),])
-}
+b_sobs <- read.table(file="data/process/bacteria.all_rarefaction.tsv", header=T)
+b_sobs <- b_sobs[nrow(b_sobs),]
+b_nseqs <- as.numeric(rownames(b_sobs))
 
-b_sobs <- get_sobs(read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T))
-a_sobs <- get_sobs(read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T))
+a_sobs <- read.table(file="data/process/archaea.all_rarefaction.tsv", header=T)
+a_sobs <- a_sobs[nrow(a_sobs),]
+a_nseqs <- as.numeric(rownames(a_sobs))
 ```
@@
 By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
-After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the respective thresholds (Figure results/figures/domain_rarefaction.pdf A, Table 1).
+After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['X0.00']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the respective thresholds (Figure results/figures/domain_rarefaction.pdf A, Table 1).
 Using only the OTUs generated using a 3% threshold, we calculated a `r b_goods`% Good's coverage (percent of sequences belonging to OTUs that have been observed more than once), but only `r b_schloss`% OTU coverage (percent of the OTUs that have been observed more than once).
-Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure results/figures/domain_rarefaction.pdf B, Table 1).
+Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['X0.00']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure results/figures/domain_rarefaction.pdf B, Table 1).
 Using only the OTUs generated using a 3% threshold, we calculated a `r a_goods`% Good's coverage, but only `r a_schloss`% OTU coverage.

### 2016-01-22 Schloss_Census2_mBio_2015.Rmd
Complete remainig text sections

@@
 ```
+\vspace{35mm}
+
+Running title: The microbial census
 
@@
 
-Abstract goes here.
+A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity.
+Although a census is typically carried out for people at a national level, microbial ecologists have carried out a molecular census of bacteria and archaea by sequencing their 16S rRNA genes.
+The goal of this study was to update an earlier effort from 2004 by assessing how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies.
+We found that nearly all full-length sequences belong to OTUs that have been previously observed, but that most bacterial and archaeal operational taxonomic units (OTUs) have only been observed once.
+This suggests that there is still considerable microbial diversity to be explored.
+We discuss the role that traditional culture-dependent and -independent approaches have had in advancing the census as well as the possible role that new technologies such as single cell genomics and short read assembly might have in allowing researchers to improve our ability to sample rare OTUs.
+We also show that environments that are known to be rich in bacterial diversity (e.g.
+soil and aquatic environments) have been poorly sampled.
+The ability to contextualize the biodiversity that is sampled using short read sequencing technologies depends on the ability to obtain similar sequence depths of full-length sequences from diverse environments.
+Furthermore, these sequences and their environmental metadata must be made available in public databases.
 
-\newpage
+### Importance
 
-## Importance
-
-Importance goes here.
+The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical.
+Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes.
+By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census.
+We actually found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms.
+Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms.
+Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of microbial diversity.
 
@@
 
+## Results and Discussion
+
+```{r rarefaction}
+# take these data out of the rarefaction files so that we can have the unique,
+# 0.03, 0.05, 0.10, and 0.20 OTU definitions for both the bacterial and archaeal
+# data
+
+b_sobs <- read.table(file="data/process/bacteria.all_rarefaction.tsv", header=T)
+b_sobs <- b_sobs[nrow(b_sobs),]
+b_nseqs <- as.numeric(rownames(b_sobs))
+
+a_sobs <- read.table(file="data/process/archaea.all_rarefaction.tsv", header=T)
+a_sobs <- a_sobs[nrow(a_sobs),]
+a_nseqs <- as.numeric(rownames(a_sobs))
+```
+
+***The status of the bacterial and archaeal census.***
+To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%.
+Although it is not possible to link a specific taxonomic level (e.g.
+species, genus, family, etc.) to a specific distance threshold, we selected distances of 0, 3, 5, 10, and 20% because they are widely regarded as representing the range of genetic diversity of the 16S rRNA gene within each domain.
+By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
+After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['X0.00']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the respective thresholds (Figure 1A, Table 1).
+Using only the OTUs generated using a 3% threshold, we calculated a `r b_goods`% Good's coverage (percent of sequences belonging to OTUs that have been observed more than once), but only `r b_schloss`% OTU coverage (percent of the OTUs that have been observed more than once).
+Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['X0.00']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1).
+Using only the OTUs generated using a 3% threshold, we calculated a `r a_goods`% Good's coverage, but only `r a_schloss`% OTU coverage.
+These results indicate that regardless of the domain, continued sampling with the current strategies for generating full-length sequences will largely reveal OTUs that have already been observed, even though a large fraction of OTUs have only been sampled once.
+Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once, it is likely that an even larger number of OTUs have yet to be sampled for both domains.
+
+```{r time_course}
+time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE, stringsAsFactors=FALSE)
+
+all_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_nseqs'])
+dominant_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_n50_nseqs'])
+
+# what were the papers that these sequences came from?
+dominant_pmids_2009_2012 <- unique(unlist(strsplit(time_data[as.character(2009:2012), "b_n50_doi"], split=',')))
+
+#require(knitcitations, quiet=TRUE)
+#citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
+
+stopifnot(dominant_pmids_2009_2012 == c("19478181", "22310478", "20668241", "22719818", "22832344"))
+#if this stops then will need to update stuff below
+```
+
+***The status of the bacterial census.***
+One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
+In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
+Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure 2B).
+Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects.
+Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C).
+Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(dominant_pmids_2009_2012)` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences).
+These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
+The heavy zoonotic focus is reflected in the rarefaction curve for this category (Figure 1C).
+In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year.
+Although these deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.
+
+***The status of the archaeal census.***
+The depth of sequencing being done to advance the archaeal census has been `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census (Table 1).
+The annual number of sequences submitted has largely paralleled that of the bacterial census with a plateau starting in 2009 and an average of `r mean(time_data[as.character(2009:2014),'a_nseqs'])` sequences each year since then.
+The number of new archaeal OTUs represented by these sequences began to slow in 2005 with an average of `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year.
+With the exception of 2012 and 2014, the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has varied between `r min(time_data[as.character(c(1996:2011,2013)),'a_n50'])` and `r max(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions per year.
+The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea, which is clearly reflected in the relatively meager sampling effort across habitats, compared to bacteria (Figure 1D),
+
+```{r taxon_census}
+oxford_comma <- function(ids){
+	if(length(ids)==0){
+		"ERROR"
+	} else if(length(ids) == 1){
+		ids
+	} else if(length(ids) == 2){
+		paste(ids, collapse=" and ")
+	} else {
+		paste(paste(ids[1:(length(ids)-1)], collapse=", "), ids[length(ids)], sep=", and ")
+	}
+}
+
+b_phylum <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
+rownames(b_phylum) <- gsub("Candidate_division_", "", rownames(b_phylum))
+rownames(b_phylum) <- gsub("_", " ", rownames(b_phylum))
+
+b_new_phyla <- rownames(b_phylum)[b_phylum$pre == 0]
+b_new_phyla_names <- oxford_comma(b_new_phyla)
+
+a_phylum <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
+a_new_phyla <- rownames(a_phylum)[a_phylum$pre == 0]
+
+a_new_phyla_names <- oxford_comma(a_new_phyla)
+a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
+a_new_phyla_names <- gsub("_", " ", a_new_phyla_names)
+```
+
+***The ability to sample microbial life is taxonomically skewed.***
+The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences.
+We sought to understand how the representation of individual phyla has changed relative to the state of the census in 2006.
+We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published and ushered in a radical change in how microbial communities are studied [@Sogin2006].
+In 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
+Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`).
+Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure 3).
+Among the bacterial phyla with at least 1,000 sequences, `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios >= 4.0 & b_phylum$all >=1000,])))` were sequenced 4-fold more often while `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios < 0.5 & b_phylum$all >=1000,])))` were sequenced 2-fold less often than would have been expected since 2006.
+Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r a_phylum["Thaumarchaeota", "ratio_of_ratios"]`-fold more often and the Crenarchaeota were sequenced `r 1/a_phylum["Crenarchaeota", "ratio_of_ratios"]`-fold less often than expected.
+Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006.
+
+```{r environment}
+strip <- function(x){	as.numeric(gsub(',', '', x))	}
+
+coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
+
+time <- grepl('_pre', colnames(coverage_by_category_and_time)) | grepl('_post', colnames(coverage_by_category_and_time))
+coverage_by_category <- coverage_by_category_and_time[,!time]
+
+totals <- unlist(coverage_by_category[nrow(coverage_by_category),])
+coverage_by_category <- coverage_by_category[-nrow(coverage_by_category),]
+
+b_class_percent <- 100*sum(coverage_by_category[-nrow(coverage_by_category), "bact_count"])/sum(totals["bact_count"])
+
+a_class_percent <- 100*sum(coverage_by_category[-nrow(coverage_by_category), "arch_count"])/sum(totals["arch_count"])
+
+coarse_map <- c("A"=="aerosol", "AQ" = "aquatic", "B" = "built environment", "O" = "other", "P" = "plant", "S" = "soil", "Z" = "zoonotic")
+
+coarse <- gsub("^(.)[A-Z]$", "\\1", rownames(coverage_by_category))
+coarse <- gsub("^(..)[A-Z]+$", "\\1", coarse)
+
+b_agg_count <- aggregate(coverage_by_category$bact_count, by=list(coarse_map[coarse]), sum)
+b_text <- paste0(b_agg_count$Group.1, ' (N=', format(b_agg_count$x, big.mark=',', trim=TRUE), ')')[order(b_agg_count$x, decreasing=T)]
+
+a_agg_count <- aggregate(coverage_by_category$arch_count, by=list(coarse_map[coarse]), sum)
+a_text <- paste0(a_agg_count$Group.1, ' (N=', format(a_agg_count$x, big.mark=',', trim=TRUE), ')')[order(a_agg_count$x, decreasing=T)]
+
+phylum_category <- read.table(file="data/process/phylum_category_counts.tsv")
+bacteria <- phylum_category[phylum_category$domain == "bacteria", -1]
+
+dominant_bacteria <- sort(table(apply(bacteria, 2, which.max)), decreasing=T)
+names(dominant_bacteria) <- rownames(bacteria)[as.numeric(names(dominant_bacteria))]
+dominant_bacteria_list <- oxford_comma(paste0(names(dominant_bacteria), ' (N=', dominant_bacteria, ')'))
+
+archaea <- phylum_category[phylum_category$domain == "archaea", -1]
+dominant_archaea <- sort(table(apply(archaea, 2, function(x){z<-which.max(x);if(length(z)==0){z<-NA};z})), decreasing=T)
+names(dominant_archaea) <- rownames(archaea)[as.numeric(names(dominant_archaea))]
+dominant_archaea_list <- oxford_comma(paste0(names(dominant_archaea), ' (N=', dominant_archaea, ')'))
+```
+
+***Focusing the census by environment.***
+We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database.
+Across these broad categories there was wide variation in the number of sequences that have been sampled.
+Among bacterial sequences, the three best represented groups were from `r oxford_comma(b_text[1:3])` sources.
+Among the archaeal sequences the three best represented groups were the same, but ordered differently: `r oxford_comma(a_text[1:3])` (Figure 1C,D)).
+For both domains, soil samples were the fourth most represented category (bacteria: `r b_agg_count[b_agg_count$Group.1=="soil", "x"]`; archaea: `r a_agg_count[a_agg_count$Group.1=="soil", "x"]`).
+The orders of these categories was surprising considering soil and aquatic environments harbor the most microbial biomass and biodiversity [@Whitman1998].
+In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range across the fine-level categories for the bacterial OTU coverage only varied between `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.25))` to `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.75))` (median coverage=`r median(coverage_by_category$bact_schloss_coverage, na.rm=T)`%).
+The interquartile range in the OTU coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to`r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
+The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
+Across all categories, the bacterial and archaeal sequencing data represented a limited number of phyla (Figure 4).
+Among the bacterial data, the fine-scale categories were dominated by `r dominant_bacteria_list` and among the archaeal data, they were dominated by `r dominant_archaea_list`.
+Regardless, there were clear phylum-level signatures that differentiated the various categories.
+Within each of the bacterial and archaeal phyla, there was considerable variation in the relative abundance of each across the categories confirming that taxonomic signatures exist to differentiate different environments even at a broad taxonomic level.
+
+```{r culture}
+b_cult_time <- read.table(file='data/process/bacteria.cultured_by_time_counts.tsv', header=T)
+a_cult_time <- read.table(file='data/process/archaea.cultured_by_time_counts.tsv', header=T)
+
+b_percent_cult_nseqs_all <- 100*sum(b_cult_time[,"all_cult_nseqs"])/sum(b_cult_time[,"all_total_nseqs"])
+a_percent_cult_nseqs_all <- 100*sum(a_cult_time[,"all_cult_nseqs"])/sum(a_cult_time[,"all_total_nseqs"])
+
+b_shift_cult_nseqs_all <- 1/((sum(b_cult_time[,"post_cult_nseqs"])/sum(b_cult_time[,"pre_cult_nseqs"])) / (sum(b_cult_time[,"post_total_nseqs"])/sum(b_cult_time[,"pre_total_nseqs"])))
+
+a_shift_cult_nseqs_all <- 1/((sum(a_cult_time[,"post_cult_nseqs"])/sum(a_cult_time[,"pre_cult_nseqs"])) / (sum(a_cult_time[,"post_total_nseqs"])/sum(a_cult_time[,"pre_total_nseqs"])))
+
+b_rate_cult_nseqs_phyulm <- b_cult_time[, 'all_cult_nseqs'] / b_cult_time[, 'all_total_nseqs']
+a_rate_cult_nseqs_phyulm <- a_cult_time[, 'all_cult_nseqs'] / a_cult_time[, 'all_total_nseqs']
+
+b_percent_cult_otu_all <- 100*sum(b_cult_time[,'all_cult_sobs'])/sum(b_cult_time[,'all_total_sobs'])
+a_percent_cult_otu_all <- 100*sum(a_cult_time[,'all_cult_sobs'])/sum(a_cult_time[,'all_total_sobs'])
+```
+
+***The cultured census.***
+In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
+Overall, `r b_percent_cult_nseqs_all`% of bacterial sequences and `r a_percent_cult_nseqs_all`% of archaeal sequences have come from isolated organisms.
+Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_shift_cult_nseqs_all` and `r a_shift_cult_nseqs_all`-fold for bacteria and archaea, respectively.
+Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r nrow(a_cult_time)` of the `r sum(a_cult_time$all_cult_nseqs == 0)` archaeal phyla have cultured representatives.
+This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
+Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure 3).
+So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
+Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
+We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure 5).
+Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Figure 5).
+The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
+This likely reflects the extremely high number of cultivated  *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
+Conversely, clades such as the *Actinobacteria*, *Fusobacteria*, and *Bacteroidetes* had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
+
+```{r single_cell}
+overlap_counts <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
+```
+
+***New technologies to access novel biodiversity.***
+Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled.
+Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
+These have included single cell genomics and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data using the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@refs].
+To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional culutre-independent amplification and cultivation-based approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure 6).
+Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
+Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by cultivation or PCR-based efforts.
+Next, we identified three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences.
+Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively.
+Only
+`r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'pcr']`% of the bacterial OTUs generated by this approach were previously identified by this traditional cultivation and PCR-based approaches, respectively.
+Although the application of this approach to Archaea has been limited, it was still suprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
+Finally, we
+pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@REF].
+These sequences were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_otus']` OTUs.
+We were surprised that only  `r 100-overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'em_pcr']`% of these OTUs were previously found by a more traditional approach.
+Although these PCR-based EMIRGE results may be valid, the high degree of novelty that was observed suggests that the error of the assembled reads may be too high for generating reference sequences.
+Each of these methods represent promising opportunities to continue the bacterial census using full-length sequences as well as genomic information.
+
+## Conclusions
+
+In spite of estimates suggesting the global bacterial species richness ranges between 10^XX^ and 10^XX^ species, the current census based on full-length 16S rRNA gene sequences suggests that our current methods and sampling schemes would never reach these estimates.
+As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations.
+Given this low level of OTU coverage, it is likely that there are many more bacterial and archaeal populations yet to be sampled.
+The recent advances in sequencing technologies have done little to expand the deposition of full-length sequences representing microbial biodiversity.
+We found that most sequences deposited into public databases are being made by a small number of projects that have deeply sampled similar environments.
+Furthermore, the number of full-lenth reads deposited into the databases has stalled.
+We fear that these trend will worsen unless we can leverage new sequencing and cultivations technologies to generate large numbers of full-length sequences.
+
+During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to to generate dozens or hundreds of sequences per sample.
+The existing databases that are used for classifying sequences are based on these sequences.
+Thus, the sequences from those studies represent organisms that are generally abundant.
+We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms is because our databases do not contain full-length references for those sequences.
+As stated in the 2004 census, it remains true that we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage.
+Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.
+
+It is clear that considerable biodiversity has been discovered since the first census in 2004.
+However, much of it has been biased towards particular phyla and environments.
+Our analysis suggests that `r b_goods`% of new full-length bacterial and archaeal sequences are likely to have already been seen.
+Meanwhile, `r b_schloss`% of bacterial and `r a_schloss`% of archaeal OTUs have only been observed once.
+It appears that although brute force deep sequencing of communities may increase our coverage of OTUs, new methods are needed to more directly sample rare organisms.
+Nevertheless, novel technologies such as single-cell genomics, metagenomics, and algorithms to recover full-length sequences from new sequencing platforms have demonstrated promise in circumventing previous limitations in identifying new OTUs.
+Using EMIRGE to assemble fragmented 16S rRNA gene amplicons may allow us to obtain deep coverage of communities; however, it is still unclear how faithful the assembled sequence is to that of the original organism.
+Additional sequencing technologies also offer the ability to directly generate full-length sequences, such as PacBio and potentially Oxford Nanopore.
+Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate.
+To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques.
+The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing.
+One strength of cultivation, single cell genomics, and metagenomics approaches is that not only do the methods yield a full 16S rRNA gene sequence, but they also allow researchers to obtain the complete genome of the organism.
+These approaches are still in active development, we hope that through continuous refinement, they may allow us to significantly improve the coverage of OTUs in public databases.
+
+Our ability to assess the microbial census is limited by the quality of the data in the databases.
+First, it is widely acknowledged that the primers used to amplify 16S rRNA genes are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads (Furhman, 2015).
+Furthermore, assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g.
+Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard PCR-based approaches [@Brown et al.
+2015].
+Thus, for many environments, the estimates for the census must be considered as lower bounds.
+Although the primers used to amplify the entire gene are also biased, obtaining a better characterization of microbial diversity will help improve the design of primers targeting subregions of the gene.
+Second, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases.
+Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983.
+Similarly, most of the studies that used EMIRGE provided their sequences in supplements, not GenBank making them effectively invisible to 16S rRNA gene-specific databases.
+As assembly and long read technologies advance, a mechanism is needed to assess the quality of the consensus sequences and to make them easily accessible to the 16S rRNA gene-specific databases.
+
+The advance in sequencing technologies has significantly expanded our ability to peer into the depths of microbial communities.
+We now have unprecedented abilities to sequence large numbers of individual communities deeply for the same cost of generating hundreds of full length sequences.
+Yet, unless we invest in improving our full-length databases, it will continue to be difficult to contextualize this newly discovered biodiversity within the tree of life.
+
 ## Materials and Methods
@@
 ```{r overview}
-category_summary <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
+b_nseqs <- coverage_by_category_and_time["total", "bact_count"]
+b_goods <- coverage_by_category_and_time["total", "bact_good_coverage"]
+b_schloss <- coverage_by_category_and_time["total", "bact_schloss_coverage"]
 
-b_nseqs <- category_summary["total", "bact_count"]
-b_goods <- category_summary["total", "bact_good_coverage"]
-b_schloss <- category_summary["total", "bact_schloss_coverage"]
-
-a_nseqs <- category_summary["total", "arch_count"]
-a_goods <- category_summary["total", "arch_good_coverage"]
-a_schloss <- category_summary["total", "arch_schloss_coverage"]
+a_nseqs <- coverage_by_category_and_time["total", "arch_count"]
+a_goods <- coverage_by_category_and_time["total", "arch_good_coverage"]
+a_schloss <- coverage_by_category_and_time["total", "arch_schloss_coverage"]
 ```
@@
 `r packageDescription("openxlsx")$Version`) [@Walker2015] packages.
-A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_mBio_2015.
-
-## Results and Discussion
-
-```{r rarefaction}
-# take these data out of the rarefaction files so that we can have the unique,
-# 0.03, 0.05, 0.10, and 0.20 OTU definitions for both the bacterial and archaeal
-# data
-
-b_sobs <- read.table(file="data/process/bacteria.all_rarefaction.tsv", header=T)
-b_sobs <- b_sobs[nrow(b_sobs),]
-b_nseqs <- as.numeric(rownames(b_sobs))
-
-a_sobs <- read.table(file="data/process/archaea.all_rarefaction.tsv", header=T)
-a_sobs <- a_sobs[nrow(a_sobs),]
-a_nseqs <- as.numeric(rownames(a_sobs))
-```
-
-***The status of the bacterial and archaeal census.***
-To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%.
-Although it is not possible to link a specific taxonomic level (e.g.
-species, genus, family, etc.) to a specific distance threshold, we selected distances of 0, 3, 5, 10, and 20% because they are widely regarded as representing the range of genetic diversity of the 16S rRNA gene within each domain.
-By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs.
-After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['X0.00']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the respective thresholds (Figure results/figures/domain_rarefaction.pdf A, Table 1).
-Using only the OTUs generated using a 3% threshold, we calculated a `r b_goods`% Good's coverage (percent of sequences belonging to OTUs that have been observed more than once), but only `r b_schloss`% OTU coverage (percent of the OTUs that have been observed more than once).
-Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['X0.00']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure results/figures/domain_rarefaction.pdf B, Table 1).
-Using only the OTUs generated using a 3% threshold, we calculated a `r a_goods`% Good's coverage, but only `r a_schloss`% OTU coverage.
-These results indicate that regardless of the domain, continued sampling with the current strategies for generating full-length sequences will largely reveal OTUs that have already been observed, even though a large fraction of OTUs have only been sampled once.
-Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once, it is likely that an even larger number of OTUs have yet to be sampled for both domains.
-
-```{r time_course}
-time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE, stringsAsFactors=FALSE)
-
-all_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_nseqs'])
-dominant_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_n50_nseqs'])
-
-# what were the papers that these sequences came from?
-dominant_pmids_2009_2012 <- unique(unlist(strsplit(time_data[as.character(2009:2012), "b_n50_doi"], split=',')))
-
-#require(knitcitations, quiet=TRUE)
-#citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))
-
-stopifnot(dominant_pmids_2009_2012 == c("19478181", "22310478", "20668241", "22719818", "22832344"))
-#if this stops then will need to update stuff below
-```
-
-***The status of the bacterial census.***
-One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
-In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure results/figures/time_course_figure.pdf A).
-Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure results/figures/time_course_figure.pdf B).
-Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects.
-Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure results/figures/time_course_figure.pdf C).
-Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(dominant_pmids_2009_2012)` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences).
-These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
-The heavy zoonotic focus is reflected in the rarefaction curve for this category (Figure results/figures/domain_rarefaction.pdf C).
-In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year.
-Although these deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.
-
-***The status of the archaeal census.***
-The depth of sequencing being done to advance the archaeal census has been `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census (Table 1).
-The annual number of sequences submitted has largely paralleled that of the bacterial census with a plateau starting in 2009 and an average of `r mean(time_data[as.character(2009:2014),'a_nseqs'])` sequences each year since then.
-The number of new archaeal OTUs represented by these sequences began to slow in 2005 with an average of `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year.
-With the exception of 2012 and 2014, the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has varied between `r min(time_data[as.character(c(1996:2011,2013)),'a_n50'])` and `r max(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions per year.
-The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea, which is clearly reflected in the relatively meager sampling effort across habitats, compared to bacteria (Figure results/figures/domain_rarefaction D),
-
-```{r taxon_census}
-oxford_comma <- function(ids){
-	if(length(ids)==0){
-		"ERROR"
-	} else if(length(ids) == 1){
-		ids
-	} else if(length(ids) == 2){
-		paste(ids, collapse=" and ")
-	} else {
-		paste(paste(ids[1:(length(ids)-1)], collapse=", "), ids[length(ids)], sep=", and ")
-	}
-}
-
-b_phylum <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
-rownames(b_phylum) <- gsub("Candidate_division_", "", rownames(b_phylum))
-rownames(b_phylum) <- gsub("_", " ", rownames(b_phylum))
-
-b_new_phyla <- rownames(b_phylum)[b_phylum$pre == 0]
-b_new_phyla_names <- oxford_comma(b_new_phyla)
-
-a_phylum <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
-a_new_phyla <- rownames(a_phylum)[a_phylum$pre == 0]
-
-a_new_phyla_names <- oxford_comma(a_new_phyla)
-a_new_phyla_names <- gsub("\\(\\S*\\)", "", a_new_phyla_names)
-a_new_phyla_names <- gsub("_", " ", a_new_phyla_names)
-```
-
-***The ability to sample microbial life is taxonomically skewed.***
-The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences.
-We sought to understand how the representation of individual phyla has changed relative to the state of the census in 2006.
-We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published and ushered in a radical change in how microbial communities are studied [@Sogin2006].
-In 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
-Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`).
-Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure results/figures/phylum_effort.pdf).
-Among the bacterial phyla with at least 1,000 sequences, `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios >= 4.0 & b_phylum$all >=1000,])))` were sequenced 4-fold more often while `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios < 0.5 & b_phylum$all >=1000,])))` were sequenced 2-fold less often than would have been expected since 2006.
-Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r a_phylum["Thaumarchaeota", "ratio_of_ratios"]`-fold more often and the Crenarchaeota were sequenced `r 1/a_phylum["Crenarchaeota", "ratio_of_ratios"]`-fold less often than expected.
-Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006.
-
-```{r environment}
-strip <- function(x){	as.numeric(gsub(',', '', x))	}
-
-coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
-
-time <- grepl('_pre', colnames(coverage_by_category_and_time)) | grepl('_post', colnames(coverage_by_category_and_time))
-coverage_by_category <- coverage_by_category_and_time[,!time]
-
-totals <- unlist(coverage_by_category[nrow(coverage_by_category),])
-coverage_by_category <- coverage_by_category[-nrow(coverage_by_category),]
-
-b_class_percent <- 100*sum(coverage_by_category[-nrow(coverage_by_category), "bact_count"])/sum(totals["bact_count"])
-
-a_class_percent <- 100*sum(coverage_by_category[-nrow(coverage_by_category), "arch_count"])/sum(totals["arch_count"])
-
-coarse_map <- c("A"=="aerosol", "AQ" = "aquatic", "B" = "built environment", "O" = "other", "P" = "plant", "S" = "soil", "Z" = "zoonotic")
-
-coarse <- gsub("^(.)[A-Z]$", "\\1", rownames(coverage_by_category))
-coarse <- gsub("^(..)[A-Z]+$", "\\1", coarse)
-
-b_agg_count <- aggregate(coverage_by_category$bact_count, by=list(coarse_map[coarse]), sum)
-b_text <- paste0(b_agg_count$Group.1, ' (N=', format(b_agg_count$x, big.mark=',', trim=TRUE), ')')[order(b_agg_count$x, decreasing=T)]
-
-a_agg_count <- aggregate(coverage_by_category$arch_count, by=list(coarse_map[coarse]), sum)
-a_text <- paste0(a_agg_count$Group.1, ' (N=', format(a_agg_count$x, big.mark=',', trim=TRUE), ')')[order(a_agg_count$x, decreasing=T)]
-
-phylum_category <- read.table(file="data/process/phylum_category_counts.tsv")
-bacteria <- phylum_category[phylum_category$domain == "bacteria", -1]
-
-dominant_bacteria <- sort(table(apply(bacteria, 2, which.max)), decreasing=T)
-names(dominant_bacteria) <- rownames(bacteria)[as.numeric(names(dominant_bacteria))]
-dominant_bacteria_list <- oxford_comma(paste0(names(dominant_bacteria), ' (N=', dominant_bacteria, ')'))
-
-archaea <- phylum_category[phylum_category$domain == "archaea", -1]
-dominant_archaea <- sort(table(apply(archaea, 2, function(x){z<-which.max(x);if(length(z)==0){z<-NA};z})), decreasing=T)
-names(dominant_archaea) <- rownames(archaea)[as.numeric(names(dominant_archaea))]
-dominant_archaea_list <- oxford_comma(paste0(names(dominant_archaea), ' (N=', dominant_archaea, ')'))
-```
-
-***Focusing the census by environment.***
-We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database.
-Across these broad categories there was wide variation in the number of sequences that have been sampled.
-Among bacterial sequences, the three best represented groups were from `r oxford_comma(b_text[1:3])` sources.
-Among the archaeal sequences the three best represented groups were the same, but ordered differently: `r oxford_comma(a_text[1:3])` (Figure 1C,D)).
-For both domains, soil samples were the fourth most represented category (bacteria: `r b_agg_count[b_agg_count$Group.1=="soil", "x"]`; archaea: `r a_agg_count[a_agg_count$Group.1=="soil", "x"]`).
-The orders of these categories was surprising considering soil and aquatic environments harbor the most microbial biomass and biodiversity [@Whitman1998].
-In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range across the fine-level categories for the bacterial OTU coverage only varied between `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.25))` to `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.75))` (median coverage=`r median(coverage_by_category$bact_schloss_coverage, na.rm=T)`%).
-The interquartile range in the OTU coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to`r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
-The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
-Across all categories, the bacterial and archaeal sequencing data represented a limited number of phyla (Figure results/figures/category_phylum_heatmap.pdf).
-Among the bacterial data, the fine-scale categories were dominated by `r dominant_bacteria_list` and among the archaeal data, they were dominated by `r dominant_archaea_list`.
-Regardless, there were clear phylum-level signatures that differentiated the various categories.
-Within each of the bacterial and archaeal phyla, there was considerable variation in the relative abundance of each across the categories confirming that taxonomic signatures exist to differentiate different environments even at a broad taxonomic level.
-
-```{r culture}
-b_cult_time <- read.table(file='data/process/bacteria.cultured_by_time_counts.tsv', header=T)
-a_cult_time <- read.table(file='data/process/archaea.cultured_by_time_counts.tsv', header=T)
-
-b_percent_cult_nseqs_all <- 100*sum(b_cult_time[,"all_cult_nseqs"])/sum(b_cult_time[,"all_total_nseqs"])
-a_percent_cult_nseqs_all <- 100*sum(a_cult_time[,"all_cult_nseqs"])/sum(a_cult_time[,"all_total_nseqs"])
-
-b_shift_cult_nseqs_all <- 1/((sum(b_cult_time[,"post_cult_nseqs"])/sum(b_cult_time[,"pre_cult_nseqs"])) / (sum(b_cult_time[,"post_total_nseqs"])/sum(b_cult_time[,"pre_total_nseqs"])))
-
-a_shift_cult_nseqs_all <- 1/((sum(a_cult_time[,"post_cult_nseqs"])/sum(a_cult_time[,"pre_cult_nseqs"])) / (sum(a_cult_time[,"post_total_nseqs"])/sum(a_cult_time[,"pre_total_nseqs"])))
-
-b_rate_cult_nseqs_phyulm <- b_cult_time[, 'all_cult_nseqs'] / b_cult_time[, 'all_total_nseqs']
-a_rate_cult_nseqs_phyulm <- a_cult_time[, 'all_cult_nseqs'] / a_cult_time[, 'all_total_nseqs']
-
-b_percent_cult_otu_all <- 100*sum(b_cult_time[,'all_cult_sobs'])/sum(b_cult_time[,'all_total_sobs'])
-a_percent_cult_otu_all <- 100*sum(a_cult_time[,'all_cult_sobs'])/sum(a_cult_time[,'all_total_sobs'])
-```
-
-***The cultured census.***
-In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
-To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
-Overall, `r b_percent_cult_nseqs_all`% of bacterial sequences and `r a_percent_cult_nseqs_all`% of archaeal sequences have come from isolated organisms.
-Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_shift_cult_nseqs_all` and `r a_shift_cult_nseqs_all`-fold for bacteria and archaea, respectively.
-Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r nrow(a_cult_time)` of the `r sum(a_cult_time$all_cult_nseqs == 0)` archaeal phyla have cultured representatives.
-This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
-Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure results/figures/phylum_effort.pdf).
-So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
-Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
-We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure results/figures/venn_otu_by_method.pdf.pdf).
-Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Fig.
-results/figures/phylum_cultured.pdf).
-The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
-This likely reflects the extremely high number of cultivated  *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
-Conversely, clades such as the *Actinobacteria*, *Fusobacteria*, and *Bacteroidetes* had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
-Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
-
-```{r single_cell}
-overlap_counts <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
-
-```
-
-***New technologies to access novel biodiversity.***
-Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled.
-Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
-These have included single cell genomics and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data using the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@refs].
-To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional culutre-independent amplification and cultivation-based approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure results/figures/venn_otu_by_method.pdf).
-Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
-Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by cultivation or PCR-based efforts.
-Next, we identified three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences.
-Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively.
-Only
-`r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'pcr']`% of the bacterial OTUs generated by this approach were previously identified by this traditional cultivation and PCR-based approaches, respectively.
-Although the application of this approach to Archaea has been limited, it was still suprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
-Finally, we
-pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@REF].
-These sequences were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_otus']` OTUs.
-We were surprised that only  `r 100-overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'em_pcr']`% of these OTUs were previously found by a more traditional approach.
-Although these PCR-based EMIRGE results may be valid, the high degree of novelty that was observed suggests that the error of the assembled reads may be too high for generating reference sequences.
-Each of these methods represent promising opportunities to continue the bacterial census using full-length sequences as well as genomic information.
-
-## Caveat Emptor
-Recent data suggests that a considerable short-read diversity of microorganisms may be missing based on biases in existing 16S rRNA gene primers (Furhman, 2015).
-Furthermore, Brown et al.
-(2015) have recently used metagenomic assemblies to show evidence for introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (CPR- Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard cultivation-independent microbial surveys.
-Furthermore, many of these CPR organisms are very small and frequently pass through 0.2 µm filters (Luef, 2015).
-Thus, for many environments, the estimates for the census must be considered as lower bounds.
-
-## Conclusions
-
-It is clear that considerable biodiversity has been discovered since the first census in 2004.
-However, much of it has been biased towards particular phyla and environments.
-Nevertheless, novel technologies such as single-cell genomics and algorithms to recover full-length sequences from shotgun metagenomic data have demonstrated promise in circumventing previous limitations in identifying new OTUs.
-Additional technologies coming on line that can provide full-length sequences, such as PacBio and potentially Nanopore, will likely provide considerable value.
-Additionally, as researchers attempt to answer questions about the physiology and biochemistry of the organisms identified in the massive deluge of cultivation-independent sequence data, a renewed emphasis on cultivation is beginning.
-Focused innovation to overcome the challenges of isolating new microorganisms will also contribute to improving out understanding of extant diversity.
-
-The first 16S sequence was published in 1978, not deposited until 1983. A bit of an allegory for our time.
-
-What are the most significant improvements since 2004, and where are we still lacking the most data?
-
-Renewed call for cultivation
+A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_mBio_2016.
 
@@
 
-**Figure results/figures/domain_rarefaction.pdf.
+**Figure 1.
 Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B).
@@
 
-**Figure results/figures/time_course_figure.pdf.
+**Figure 2.
 Progression of the microbial census since the first full-length 16S rRNA gene sequence was deposited into GenBank in 1983.*** The number of bacterial and archaeal 16S rRNA gene sequences deposited (A) and the new OTUs they represent (B) has increased exponentially until the last several years when the rate of change has plateaued.
@@
 
-**Figure results/figures/category_phylum_heatmap.pdf.
+**Figure 3.
+Relative rate of sequence deposition for each bacterial and archaeal phylum before and after 2006 relative to the sequencing of all bacteria.** The figure shows the relative rates for those phyla with at least 1,000 sequences and the x-axis is on a log2 scale.
+The data for all bacterial and archaeal phyla are available in Supplemental Tables 2 and 3, respectively.
+
+**Figure 4.
 Heatmap depicting the relative abundance of the most common bacterial and archaeal phyla across different environments.** Each environmental category exhibited a phylum-level signature although the bacterial census was dominated by sequences from the Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes and the archaeal census was dominated by sequences from the Euryarchaeota and Thaumarchaeota.
 The ten most abundant phyla across all environmental categories are shown.
-The data for all bacterial and archaeal phyla are available in Supplemental Tables 2 and 3, respectively.
-
-**Figure results/figures/phylum_effort.pdf.
-Relative rate of sequence deposition for each bacterial and archaeal phylum before and after 2006 relative to the sequencing of all bacteria.** The figure shows the relative rates for those phyla with at least 1,000 sequences and the x-axis is on a log2 scale.
 The data for all bacterial and archaeal phyla are available in Supplemental Tables 4 and 5, respectively.
 
-**Figure results/figures/phylum_cultured.pdf.
+**Figure 5.
 The rate that sequences and OTUs are generated from bacterial and archaeal cultures relative to all sequences and OTUs by phlum.** Phyla with greater than 1,000 sequences are listed by domain.
@@
 Closed circles indicate the percentage of OTUs in this analysis that contain sequences belonging to a cultured organism.
+The data for all bacterial and archaeal phyla are available in Supplemental Tables 6 and 7, respectively.
 
-**Figure results/figures/venn_otu_by_method.pdf. The number of OTUs that were found by either cultivation, PCR, single-cell genomics or multiple methods for bacterial and archaeal sequences.**
+**Figure 6.
+The percentage of bacterial and archaeal OTUs found by single cell genomics and EMIRGE using PCR or metagenomics that were also detected by other.** The bars comparing a method to itself indicate the percentage of OTUs that were only detected by that method.
 

### 2016-01-22 Schloss_Census2_mBio_2015.Rmd
Get Rmd file to compile with 'make write.paper'

@@
 a_nseqs <- as.numeric(rownames(a_sobs))
+
+coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
+
+b_nseqs <- coverage_by_category_and_time["total", "bact_count"]
+b_goods <- coverage_by_category_and_time["total", "bact_good_coverage"]
+b_schloss <- coverage_by_category_and_time["total", "bact_schloss_coverage"]
+
+a_nseqs <- coverage_by_category_and_time["total", "arch_count"]
+a_goods <- coverage_by_category_and_time["total", "arch_good_coverage"]
+a_schloss <- coverage_by_category_and_time["total", "arch_schloss_coverage"]
 ```
@@
 
-coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
-
 time <- grepl('_pre', colnames(coverage_by_category_and_time)) | grepl('_post', colnames(coverage_by_category_and_time))
@@
 ***Focusing the census by environment.***
-We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database.
+We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database (Table S1).
 Across these broad categories there was wide variation in the number of sequences that have been sampled.
@@
 
-```{r overview}
-b_nseqs <- coverage_by_category_and_time["total", "bact_count"]
-b_goods <- coverage_by_category_and_time["total", "bact_good_coverage"]
-b_schloss <- coverage_by_category_and_time["total", "bact_schloss_coverage"]
-
-a_nseqs <- coverage_by_category_and_time["total", "arch_count"]
-a_goods <- coverage_by_category_and_time["total", "arch_good_coverage"]
-a_schloss <- coverage_by_category_and_time["total", "arch_schloss_coverage"]
-```
-
 ***Sequence data curation.***
@@
 The metadata that was contained within the SSU Ref database was used to expand our analysis beyond a basic count of sequences and the number of OTUs in each domain.
-The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table 1).
+The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table S1).
 These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences (Table 1).
 For source definitions that were not identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
-There were `r category_summary["no_source", "bact_count"]` bacterial and `r category_summary["no_source", "arch_count"]` archaeal sequences where an "isolation_source" term was not collected.
+There were `r coverage_by_category_and_time["no_source", "bact_count"]` bacterial and `r coverage_by_category_and_time["no_source", "arch_count"]` archaeal sequences where an "isolation_source" term was not collected.
 We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms.

### 2016-01-22 Schloss_Census2_mBio_2016.Rmd
Update year from 2015 to 2016

--- 2016-01-22 Schloss_Census2_mBio_2015.Rmd
@@
 
-inline_hook <- function(x){
-	print(x)
+custom_print <- function(x){
+
 	if(is.numeric(x)){
@@
 	}
+
+}
+
+inline_hook <- function(x){
+	print(x)
+	custom_print(x)
 }

### 2016-01-22 Schloss_Census2_mBio_2016.Rmd
Editing

--- 2016-01-22 Schloss_Census2_mBio_2015.Rmd
@@
 	print(x)
+
+	if(is.list(x)){
+		x <- unlist(x)
+	}
+
 	if(is.numeric(x)){
@@
 		}
-
 	} else {
@@
 The goal of this study was to update an earlier effort from 2004 by assessing how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies.
-We found that nearly all full-length sequences belong to OTUs that have been previously observed, but that most bacterial and archaeal operational taxonomic units (OTUs) have only been observed once.
+We found that nearly all full-length sequences belong to operational taxonomic units (OTUs) that have been previously observed, but that most bacterial and archaeal OTUs have only been observed once.
 This suggests that there is still considerable microbial diversity to be explored.
@@
 The commonly used MiSeq-based sequencing platform from Illumina is extensively leveraged to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
-Youssef et al.
-(2009) demonstrated high variability between the number of OTUs that different hypervariable sites estimated compared to full-length 16S rRNA gene sequences, and also that sample location influenced this behavior.
-Thus, it remains unclear to what degree richness estimates from short read technology over or underestimate the numbers from full-length sequences.
+The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef et al.
+2009; Schloss 200X].
+Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences.
 We likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
 
-Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences.
-In the 13 years since the collection of data for Schloss and Handelsman’s analysis, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies.
-This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study.
-Although there has been a robust growth in the number of full-length sequences deposited to GenBank since its creation in 1983, the rate of growth has stalled over the past 5 years and the deposits have been dominated by a handful of research groups studying a limited number of environments.
+Here we update the status of the microbial census with full-length 16S rRNA gene sequences.
+In the 13 years since the collection of data for Schloss and Handelsman’s initial census, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies (Figure 1A).
+This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study.
 As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.
-In this, continued technological advances, such as the reconstruction of nearly full-length sequences from metagenomics with EMIRGE (Miller 2011) and innovations in culturing (references needed: iChip, Epstein dormancy, Thrash 2015; Gifford 2014), will be vital.
 
@@
 ***The status of the bacterial and archaeal census.***
-To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%.
+To assess the field's progress in characterizing the biodiversity of bacteria and archaea, we assigned each 16S rRNA gene sequence to OTUs using distance thresholds that varied between 0 and 20%.
 Although it is not possible to link a specific taxonomic level (e.g.
@@
 Paralleling the bacterial results, after sampling `r a_nseqs` archaeal 16S rRNA gene sequences we have identified `r a_sobs['X0.00']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1).
-Using only the OTUs generated using a 3% threshold, we calculated a `r a_goods`% Good's coverage, but only `r a_schloss`% OTU coverage.
+Using only the OTUs generated with a 3% threshold, we calculated a `r a_goods`% Good's coverage, but only `r a_schloss`% OTU coverage.
 These results indicate that regardless of the domain, continued sampling with the current strategies for generating full-length sequences will largely reveal OTUs that have already been observed, even though a large fraction of OTUs have only been sampled once.
@@
 ***The status of the bacterial census.***
-One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
-In fact, since 2009 the number of new bacterial sequences generated has slowed two an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
+One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
+In fact, since 2009 the number of new bacterial sequences generated has slowed to an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A).
 Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure 2B).
@@
 We sought to understand how the representation of individual phyla has changed relative to the state of the census in 2006.
-We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published and ushered in a radical change in how microbial communities are studied [@Sogin2006].
+We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published [@Sogin2006].
 In 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
@@
 ***Focusing the census by environment.***
-We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database (Table S1).
+We were able to assign `r b_class_percent` and `r a_class_percent`% of the sequences to one of seven broad environmental categories based on the metadata that accompanied the SILVA database (Tables 1).
 Across these broad categories there was wide variation in the number of sequences that have been sampled.
@@
 ***The cultured census.***
-In the 2004 bacterial census there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+In the 2004 bacterial census, there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
 To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
@@
 
-In spite of estimates suggesting the global bacterial species richness ranges between 10^XX^ and 10^XX^ species, the current census based on full-length 16S rRNA gene sequences suggests that our current methods and sampling schemes would never reach these estimates.
+In spite of current estimates suggesting the global bacterial species richness may be as high as 10^11^ species [@reference], the current census based on full-length 16S rRNA gene sequences suggests that our current methods and sampling schemes would never reach these estimates.
 As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations.
@@
 Furthermore, the number of full-lenth reads deposited into the databases has stalled.
-We fear that these trend will worsen unless we can leverage new sequencing and cultivations technologies to generate large numbers of full-length sequences.
+We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.
 
@@
 Thus, the sequences from those studies represent organisms that are generally abundant.
-We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms is because our databases do not contain full-length references for those sequences.
-As stated in the 2004 census, it remains true that we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage.
+We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms is because our databases do not contain full-length references for those sequences [@reference].
+As stated in the 2004 census, we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage.
 Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.
@@
 Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate.
-To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques.
+To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@iChip, Epstein dormancy, Thrash 2015; Gifford 2014].
 The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing.
@@
 Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983.
-Similarly, most of the studies that used EMIRGE provided their sequences in supplements, not GenBank making them effectively invisible to 16S rRNA gene-specific databases.
+Similarly, only XX of the 6 studies that used the EMIRGE algorithm deposited their sequences in GenBank.
+This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences.
 As assembly and long read technologies advance, a mechanism is needed to assess the quality of the consensus sequences and to make them easily accessible to the 16S rRNA gene-specific databases.

### 2016-01-22 doc/notebook/Archaeal_data_acquisition.Rmd
Harvest EMIRGE data that was deposited into GenBank

@@
 write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
-
-#we'll output the sequence names so the sequences aren't included in the
-#rarefaction analysis
-write(rownames(spread_sheet_good), file="data/mothur/archaea.sc_emirge.accnos")
 ```
@@
 write.table(archaea_emirge_metadata, file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
-
-#we'll output the sequence names so the sequences aren't included in the
-#rarefaction analysis
-write(rownames(archaea_emirge_metadata), file="data/mothur/archaea.sc_emirge.accnos", append=T)
 ```

### 2016-01-22 doc/notebook/Bacterial_data_acquisition.Rmd
Harvest EMIRGE data that was deposited into GenBank

@@
 write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t", col.names=F, append=T)
-
-#we'll output the sequence names so the sequences aren't included in the
-#rarefaction analysis
-write(rownames(spread_sheet_good), file="data/mothur/bacteria.sc_emirge.accnos")
 ```
@@
 write.table(bacteria_emirge_metadata, file="data/mothur/bacteria.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
-
-#we'll output the sequence names so the sequences aren't included in the
-#rarefaction analysis
-write(rownames(bacteria_emirge_metadata), file="data/mothur/bacteria.sc_emirge.accnos", append=T)
 ```

### 2016-01-22 code/partition_data.R

@@ -18,19 +18,36 @@ is_single_cell <- function(db){
 
 is_emirge_pcr <- function(db){
 	emirge_dois <- c("10.1186/2049-2618-2-1", "10.1371/journal.pone.0056018", "10.1371/journal.pone.0057819")
+	is_emirge_doi <- db$publication_doi %in% emirge_dois
 
-	db$publication_doi %in% emirge_dois
+
+	gb <- gsub("^(.*)\\..*", "\\1", rownames(metdata))
+	search <- c(paste0("KC", 716084:731398), paste0("JX", 221725:226064))
+	is_emirge_gb <- gb %in% search
+
+	is_emirge_doi | is_emirge_gb
 }
 
 is_emirge_metag <- function(db){
-	emirge_dois <- c("10.1126/science.1224041", "10.1186/gb-2011-12-5-r44", "10.3389/fmicb.2015.00277")
+	emirge_dois <- c("10.1126/science.1224041", "10.1186/gb-2011-12-5-r44", "10.3389/fmicb.2015.00277, 10.1128/AEM.00032-11")
+	is_emirge_doi <- db$publication_doi %in% emirge_dois
 
-	db$publication_doi %in% emirge_dois
+	gb <- gsub("^(.*)\\..*", "\\1", rownames(db))
+	search <- c(paste0("KM", 410305:410928), paste0("JX", 125436:125454))
+	is_emirge_gb <- gb %in% search
+
+	is_emirge_doi | is_emirge_gb
 }
 
 #	10.1126/science.1224041			99		genomes
 #	10.1186/gb-2011-12-5-r44		4		genomes
 #	10.3389/fmicb.2015.00277 		17		genomes
+#	10.1128/AEM.00032-11			21		genomes
+#	10.3389/fmicb.2014.00756		479		genomes	need to search: KM410305-KM410928
+#	10.1038/ismej.2012.148			~19	 	genomes need to search: JX125436–JX125454
+
 #	10.1186/2049-2618-2-1 			21447	PCR
 #	10.1371/journal.pone.0056018	37386	PCR
 #	10.1371/journal.pone.0057819	23		PCR
+#	10.1111/1462-2920.12467			15315	PCR		need to search: KC716084-KC731398
+#	10.1021/es502701u				1909	PCR		need to search: JX221725-JX226064

### 2016-01-22 code/partition_data.R
Fix typo

@@ -21,7 +21,7 @@ is_emirge_pcr <- function(db){
 	is_emirge_doi <- db$publication_doi %in% emirge_dois
 
 
-	gb <- gsub("^(.*)\\..*", "\\1", rownames(metdata))
+	gb <- gsub("^(.*)\\..*", "\\1", rownames(metadata))
 	search <- c(paste0("KC", 716084:731398), paste0("JX", 221725:226064))
 	is_emirge_gb <- gb %in% search
 

### 2016-01-22 Schloss_Census2_mBio_2016.Rmd
Rerun entire makefile with updated metadata files

@@
 The SILVA curators then screened the 7,168,241 resulting sequences based on a minimum length criteria (<300 nt), number of ambiguous base calls (>2%), length of sequence homopolymers (>2%), presence of vector contamination (>2%), low alignment quality value (<75), and likelihood of being chimeric (Pintail value < 50).
-Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nt and the archaeal reference set retained those archaeal sequences longer than 900 nt.
+Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nucleotides and the archaeal reference set retained those archaeal sequences longer than 900 nucleotides.
 The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata.
@@
 "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata.
-Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included.
-These were obtained from XXXXXXXXXX.
+Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included [@REF].
 The IMG and EMIRGE sequences were aligned against the respective SILVA-based reference using mothur [@REF].
-The aligned bacterial and archaeal sequence sets were pooled and processed in parallel in mothur.
+The aligned bacterial and archaeal sequence sets were pooled and processed in parallel.
 Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates.
-The sequences in the resulting bacterial dataset overlapped bases 113 through 1344 of an *E.
-coli* reference sequence (V00348) and had a median length of 1,227 nt.
-The sequences in the resulting archaeal dataset overlapped positions 362 to 937 of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of 580 nt.
+The sequences in the resulting bacterial dataset overlapped bases X113X through X1344X of an *E.
+coli* reference sequence (V00348) and had a median length of X1,227X nt.
+The sequences in the resulting archaeal dataset overlapped positions X362X to X937X of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of X580X nt.
 The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences.
@@
 The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table S1).
-These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences (Table 1).
+These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences.
 For source definitions that were not identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category.
@@
 Our analysis made use of ARB (OS X v.6.0) [@Pruesse2007], mothur (v.1.37.0) [@Schloss2009], and R (v.`r paste(version$major, version$minor, sep='.')`) [@language2015].
-Within R we utilized knitr (v.`r packageDescription("knitr")$Version`) [@xie2013dynamic] and openxlsx (v.
+Within R we utilized the knitr (v.`r packageDescription("knitr")$Version`) [@xie2013dynamic], wesanderson (v.`r packageDescription("wesanderson")$Version`)[@Ram2016], and openxlsx (v.
 `r packageDescription("openxlsx")$Version`) [@Walker2015] packages.

### 2016-01-22 code/build_culture_effort_plot.R
Fix x-axis scale and labels

@@ -32,7 +32,7 @@ total_lines <- b_n_taxa + a_n_taxa + 2
 
 pdf("results/figures/phylum_cultured.pdf",  width=3.75, height=4.75)
 par(mar=c(5,10,1.0,0.5), cex=0.7)
-plot(NA, xlim=c(0,50), ylim=c(1,total_lines), axes=F, xlab="", ylab="")
+plot(NA, xlim=c(0,70), ylim=c(1,total_lines), axes=F, xlab="", ylab="")
 abline(h=c(1:a_n_taxa, (a_n_taxa+3):total_lines), lty=3, col="gray")
 
 points(x=a_cultured$nseqs_precent, y=a_n_taxa:1, pch=1)
@@ -45,9 +45,9 @@ axis(2, at=total_lines:(a_n_taxa+3), label=rownames(b_cultured), las=1, tick=F)
 axis(2, at=a_n_taxa:1, label=rownames(a_cultured), las=1, tick=F)
 
 box()
-axis(1, at=seq(0,50,10), label=seq(0,50,10))
+axis(1, at=seq(0,100,10), label=seq(0,100,10))
 
-legend(x=20, y=15, legend=c("Cultured sequences", "Cultured OTUs"), pch=c(1, 19), bg="white", cex=0.8)
+legend(x=30, y=15, legend=c("Cultured sequences", "Cultured OTUs"), pch=c(1, 19), bg="white", cex=0.8)
 mtext("Bacteria", side=2, line=5, at=total_lines+1.2, las=1, font=2, cex=0.8, xpd=T)
 mtext("Archaea", side=2, line=5, at=a_n_taxa+1.2, las=1, font=2, cex=0.8)
 

### 2016-01-22 Schloss_Census2_mBio_2016.Rmd
Update number of EMIRGE studies

@@
 Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by cultivation or PCR-based efforts.
-Next, we identified three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences.
+Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences.
 Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively.
@@
 `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'pcr']`% of the bacterial OTUs generated by this approach were previously identified by this traditional cultivation and PCR-based approaches, respectively.
-Although the application of this approach to Archaea has been limited, it was still suprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
-Finally, we
-pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from three studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@REF].
+Although the application of this approach to Archaea has been limited, it was still surprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
+Finally, we pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from five studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@REF].
 These sequences were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_otus']` OTUs.

### 2016-01-25 code/renes_code/Arc_3_percent_Uniq.R
Remove extra code that wasn't used

@@ -1,45 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_archaea.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_3per.pdf")
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
[…]

### 2016-01-25 code/renes_code/Arc_5_percent_fixedY.R

@@ -1,46 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_archaea.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_5per_FixedYx.pdf")
-
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
[…]

### 2016-01-25 code/renes_code/Arc_Uniq_10per_adjustedY.R

@@ -1,46 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_archaea.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_10per.pdf")
-
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
[…]

### 2016-01-25 code/renes_code/Bac_3_percent_FixedY.R

@@ -1,45 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per_fixedY.pdf")
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
[…]

### 2016-01-25 code/renes_code/Bac_3_percent_Uniq.R

@@ -1,45 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per.pdf")
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(40000)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
-points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
[…]

### 2016-01-25 code/renes_code/Bac_5_percent_fixedY.R

@@ -1,46 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_5per_FixedYx.pdf")
-
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
[…]

### 2016-01-25 code/renes_code/Bac_5_percent_uniq.R

@@ -1,46 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_5per.pdf")
-
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(30000)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
-points(data$X0.05.plant~data$numsampled, type="l", col="red", lwd=2)
[…]

### 2016-01-25 code/renes_code/Bac_Uniq_10per_adjustedY.R

@@ -1,46 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_10per.pdf")
-
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(10000)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
-points(data$X0.10.plant~data$numsampled, type="l", col="red", lwd=2)
[…]

### 2016-01-25 code/renes_code/Bac_Uniq_10per_fixedY.R

@@ -1,46 +1,0 @@
-data <- read.table(file="/Users/renegirard7gfy/Desktop/Thrash_Lab/Mix/all_bacteria.env_category.groups.rarefaction", header=T)
-category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
-plant_columns <- grep(category, colnames(data))
-category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
-aerosol_columns <- grep(category, colnames(data))
-category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
-aquatic_columns <- grep(category, colnames(data))
-category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
-built_columns <- grep(category, colnames(data))
-category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
-other_columns <- grep(category, colnames(data))
-category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
-soil_columns <- grep(category, colnames(data))
-category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
-zoonotic_columns <- grep(category, colnames(data))
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
-nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
-plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
-pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_10per_FixedY.pdf")
-
-
-par(mar=c(5,5,0.5,0.5))
-plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
-     xlim=c(0,max(data$numsampled)),
-     xlab="Number of sequences sampled",
-     ylab="",
-     axes=FALSE)
-
[…]

### 2016-01-25 Schloss_Census2_mBio_2016.Rmd
Add references

@@
 ## Introduction
-The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@reference].
+The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@McGill2007, @Hubbell2001].
 At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process.
-Furthermore, creating standards for delimiting what makes one microbe "different” from another has been contentious [@reference].
+Furthermore, creating standards for delimiting what makes one microbe "different" from another has been contentious [@Konstantinidis2006, @Oren2013].
 In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, single cell technologies, and metagenomics.
@@
 Refreshingly, this challenge was largely met.
-There have been major investments in studying the Earth's microbiome using 16S rRNA gene sequencing through initiatives such as the Human Microbiome Project [@reference], the Earth Microbiome Project [@reference], and the International Census of Marine Microorganisms.
+There have been major investments in studying the Earth's microbiome using 16S rRNA gene sequencing through initiatives such as the Human Microbiome Project [@HMP2012], the Earth Microbiome Project [@Gilbert2014], and the International Census of Marine Microorganisms [@AmaralZettler2010].
 But most importantly, the original census was performed on the cusp of radical developments in sequencing technologies.
@@
 
-Disconcertingly, the increase in sequencing volume has come at the cost of sequence length.
+It has been disconcerting that the increase in sequencing volume has come at the cost of sequence length.
 The commonly used MiSeq-based sequencing platform from Illumina is extensively leveraged to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
-The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef et al.
-2009; Schloss 200X].
+The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef2009, @Schloss2010].
 Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences.
@@
 Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
-These have included single cell genomics and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data using the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@refs].
+These have included single cell genomics [@Rinke2013] and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data using the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@Miller2011, @Miller2013].
 To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional culutre-independent amplification and cultivation-based approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure 6).
-Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
+Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
 Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by cultivation or PCR-based efforts.
-Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences.
+Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences [@Wrighton2012, @Miller2011, @NunesdaRocha2015, @Hamilton2015, @Handley2012, @Gladden2011].
 Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively.
@@
 Although the application of this approach to Archaea has been limited, it was still surprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
-Finally, we pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from five studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@REF].
+Finally, we pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from five studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@Brooks2014, @Miller2013, @Wilkins2013, @Handley2014, @Alessi2014].
 These sequences were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_otus']` OTUs.
@@
 
-In spite of current estimates suggesting the global bacterial species richness may be as high as 10^11^ species [@reference], the current census based on full-length 16S rRNA gene sequences suggests that our current methods and sampling schemes would never reach these estimates.
+In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that our current methods and sampling schemes would never reach these estimates.
 As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations.
@@
 Thus, the sequences from those studies represent organisms that are generally abundant.
-We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms is because our databases do not contain full-length references for those sequences [@reference].
+We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms is because our databases do not contain full-length references for those sequences.
 As stated in the 2004 census, we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage.
@@
 Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate.
-To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@iChip, Epstein dormancy, Thrash 2015; Gifford 2014].
+To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010, @Buerger2012, @Thrash2014, @Das2015].
 The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing.
@@
 Furthermore, assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g.
-Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard PCR-based approaches [@Brown et al.
-2015].
+Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard PCR-based approaches [@Brown2015].
 Thus, for many environments, the estimates for the census must be considered as lower bounds.
@@
 Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983.
-Similarly, only XX of the 6 studies that used the EMIRGE algorithm deposited their sequences in GenBank.
+Similarly, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank.
 This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences.
@@
 ***Sequence data curation.***
-The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@REF].
+The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@Pruesse2007].
 This release is based on the EMBL-EBI/ENA Release 123, which was released in March 2015.
@@
 The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata.
-Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], many of which were recently obtained via the GEBA-MDM effort in Rinke et al.
-[@REF].
+Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], many of which were recently obtained via the GEBA-MDM effort in Rinke et al.
+[-@Rinke2013].
 "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata.
-Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included [@REF].
-The IMG and EMIRGE sequences were aligned against the respective SILVA-based reference using mothur [@REF].
+Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included [@Miller2011, @Miller2013].
+The IMG and EMIRGE sequences were aligned against the respective SILVA-based reference using mothur [@Schloss2009].
 The aligned bacterial and archaeal sequence sets were pooled and processed in parallel.
 Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates.
+
 The sequences in the resulting bacterial dataset overlapped bases X113X through X1344X of an *E.
@@
 The sequences in the resulting archaeal dataset overlapped positions X362X to X937X of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of X580X nt.
+
 The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences.
 The final datasets contained `r b_nseqs` bacterial and `r a_nseqs` archaeal 16S rRNA gene sequences.
-Sequences were assigned to OTUs using the average neighbor clustering algorithm [@REF].
+Sequences were assigned to OTUs using the average neighbor clustering algorithm [@Westcott2015].
 
@@
 We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms.
-Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environmental categorizations are available at [@figshare].
+Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environmental categorizations are available at FigShare for the bacterial (XXXXX) and archaeal (XXXXX) data.
 
@@
 
-where n~1~ is the number of OTUs represented by only one sequence and N~t~ is the total number of sequences [@ref].
+where n~1~ is the number of OTUs represented by only one sequence and N~t~ is the total number of sequences [@Good1953].
 Although Good's coverage provides information about the success of the sequencing effort in sampling the most abundant organisms in a community, it does not directly provide information about the success of the sequencing effort in recovering previously unobserved OTUs.
@@
 Our analysis made use of ARB (OS X v.6.0) [@Pruesse2007], mothur (v.1.37.0) [@Schloss2009], and R (v.`r paste(version$major, version$minor, sep='.')`) [@language2015].
-Within R we utilized the knitr (v.`r packageDescription("knitr")$Version`) [@xie2013dynamic], wesanderson (v.`r packageDescription("wesanderson")$Version`)[@Ram2016], and openxlsx (v.
-`r packageDescription("openxlsx")$Version`) [@Walker2015] packages.
+Within R we utilized the knitr (v.`r packageDescription("knitr")$Version`), wesanderson (v.`r packageDescription("wesanderson")$Version`), and openxlsx (v.
+`r packageDescription("openxlsx")$Version`) packages.
 A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_mBio_2016.

### 2016-01-25 Schloss_Census2_mBio_2016.Rmd
Create submission folder
Add FigShare information

@@
 We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms.
-Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environmental categorizations are available at FigShare for the bacterial (XXXXX) and archaeal (XXXXX) data.
+Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environmental categorizations are available at FigShare for the bacterial (https://dx.doi.org/10.6084/m9.figshare.2064927) and archaeal (https://dx.doi.org/10.6084/m9.figshare.2064942) data.
 

### 2016-01-28 Schloss_Census2_mBio_2016.Rmd
Incorporate Thrash comments

@@
 
+b_sobs <- read.table(file="data/process/bacteria.all_rarefaction.tsv", header=T)
+a_sobs <- read.table(file="data/process/archaea.all_rarefaction.tsv", header=T)
+
+coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
+
+coarse_count <- aggregate(coverage_by_category_and_time$bact_count, by=list(gsub("^(.).*", "\\1", rownames(coverage_by_category_and_time))), sum)
+
+get_coarse_count <- function(letter){
+	rows <- grepl(letter, rownames(coverage_by_category_and_time))
+	sum(coverage_by_category_and_time[rows, "bact_count"])
+}
+
+coarse_counts <- sapply(c("A", "B", "P", "S", "Z", "O"), get_coarse_count)
+coarse_percentages <- 100 * coarse_counts / sum(coarse_counts)
+
+time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE, stringsAsFactors=FALSE)
+
+b_phylum <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
+a_phylum <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
+
+phylum_category <- read.table(file="data/process/phylum_category_counts.tsv", header=T)
+
+b_cult_time <- read.table(file='data/process/bacteria.cultured_by_time_counts.tsv', header=T)
+a_cult_time <- read.table(file='data/process/archaea.cultured_by_time_counts.tsv', header=T)
+
+per_cult_b_otus <- 100 * sum(b_cult_time$all_cult_sobs) / sum(b_cult_time$all_total_sobs)
+
+overlap_counts <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
 ```
@@
 
-A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity.
-Although a census is typically carried out for people at a national level, microbial ecologists have carried out a molecular census of bacteria and archaea by sequencing their 16S rRNA genes.
-The goal of this study was to update an earlier effort from 2004 by assessing how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies.
-We found that nearly all full-length sequences belong to operational taxonomic units (OTUs) that have been previously observed, but that most bacterial and archaeal OTUs have only been observed once.
-This suggests that there is still considerable microbial diversity to be explored.
-We discuss the role that traditional culture-dependent and -independent approaches have had in advancing the census as well as the possible role that new technologies such as single cell genomics and short read assembly might have in allowing researchers to improve our ability to sample rare OTUs.
-We also show that environments that are known to be rich in bacterial diversity (e.g.
-soil and aquatic environments) have been poorly sampled.
-The ability to contextualize the biodiversity that is sampled using short read sequencing technologies depends on the ability to obtain similar sequence depths of full-length sequences from diverse environments.
-Furthermore, these sequences and their environmental metadata must be made available in public databases.
+A census is typically carried out for people at a national level; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes.
+We assessed how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies.
+Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once.
+Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_good_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the archaeal OTUs have been observed more than once.
+Thus, there is still considerable microbial diversity to be explored.
+Unfortunately, the number of new full-length sequences has been declining and these sequences are primarily deposited by a small number of studies.
+Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the sequences while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the sequences.
+Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling biases.
+The success of ongoing efforts to use short read sequencing to characterize microbial communities requires that researchers strive to expand the depth and breadth of the microbial census.
 
 ### Importance
+
+A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity.
 
@@
 ## Introduction
-The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@McGill2007, @Hubbell2001].
+The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@McGill2007; @Hubbell2001].
 At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process.
-Furthermore, creating standards for delimiting what makes one microbe "different" from another has been contentious [@Konstantinidis2006, @Oren2013].
+Furthermore, creating standards for delimiting what makes one microbe "different" from another has been contentious [@Konstantinidis2006; @Oren2013].
 In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, single cell technologies, and metagenomics.
@@
 
-Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial 16S rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998, @Curtis2002].
+Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial 16S rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998; @Curtis2002].
 That census called for a broader and deeper characterization of all environments.
@@
 The commonly used MiSeq-based sequencing platform from Illumina is extensively leveraged to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
-The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef2009, @Schloss2010].
+The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef2009; @Schloss2010].
 Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences.
@@
 Here we update the status of the microbial census with full-length 16S rRNA gene sequences.
-In the 13 years since the collection of data for Schloss and Handelsman’s initial census, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies (Figure 1A).
+In the 13 years since the collection of data for Schloss and Handelsman’s initial census, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies.
 This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study.
@@
 
-b_sobs <- read.table(file="data/process/bacteria.all_rarefaction.tsv", header=T)
 b_sobs <- b_sobs[nrow(b_sobs),]
@@
 
-a_sobs <- read.table(file="data/process/archaea.all_rarefaction.tsv", header=T)
 a_sobs <- a_sobs[nrow(a_sobs),]
 a_nseqs <- as.numeric(rownames(a_sobs))
-
-coverage_by_category_and_time <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T, stringsAsFactors=FALSE)
 
@@
 ```{r time_course}
-time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE, stringsAsFactors=FALSE)
-
 all_nseqs_2009_2012 <- sum(time_data[as.character(2009:2012), 'b_nseqs'])
@@
 
-***The status of the bacterial census.***
+***Sequencing efforts are a source of bias in the census.***
 One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined.
@@
 
-***The status of the archaeal census.***
 The depth of sequencing being done to advance the archaeal census has been `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census (Table 1).
@@
 
-b_phylum <- read.table(file="data/process/bacteria.phyla.counts.tsv", header=T)
 rownames(b_phylum) <- gsub("Candidate_division_", "", rownames(b_phylum))
@@
 
-a_phylum <- read.table(file="data/process/archaea.phyla.counts.tsv", header=T)
 a_new_phyla <- rownames(a_phylum)[a_phylum$pre == 0]
@@
 We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published [@Sogin2006].
-In 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
+Based on the representation of sequences within the SILVA database, in 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
 Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`).
@@
 Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r a_phylum["Thaumarchaeota", "ratio_of_ratios"]`-fold more often and the Crenarchaeota were sequenced `r 1/a_phylum["Crenarchaeota", "ratio_of_ratios"]`-fold less often than expected.
-Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006.
+Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006 and encouragingly, show that some underrepresented phyla are becoming better sampled.
 
@@
 
-phylum_category <- read.table(file="data/process/phylum_category_counts.tsv")
-bacteria <- phylum_category[phylum_category$domain == "bacteria", -1]
+bacteria <- phylum_category[phylum_category$domain == "bacteria", ]
+rownames(bacteria) <- bacteria$phylum
+bacteria <- bacteria[,-c(1,2, ncol(bacteria))]
 
@@
 names(dominant_bacteria) <- rownames(bacteria)[as.numeric(names(dominant_bacteria))]
+
 dominant_bacteria_list <- oxford_comma(paste0(names(dominant_bacteria), ' (N=', dominant_bacteria, ')'))
 
-archaea <- phylum_category[phylum_category$domain == "archaea", -1]
-dominant_archaea <- sort(table(apply(archaea, 2, function(x){z<-which.max(x);if(length(z)==0){z<-NA};z})), decreasing=T)
+archaea <- phylum_category[phylum_category$domain == "archaea", ]
+rownames(archaea) <- archaea$phylum
+archaea <- archaea[,-c(1,2, ncol(archaea))]
+
+dominant_archaea <- sort(table(apply(archaea, 2, which.max)), decreasing=T)
 names(dominant_archaea) <- rownames(archaea)[as.numeric(names(dominant_archaea))]
+
 dominant_archaea_list <- oxford_comma(paste0(names(dominant_archaea), ' (N=', dominant_archaea, ')'))
@@
 In spite of wide variation in sequencing depth and coverage (Table 1), the interquartile range across the fine-level categories for the bacterial OTU coverage only varied between `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.25))` to `r quantile(coverage_by_category$bact_schloss_coverage, prob=c(0.75))` (median coverage=`r median(coverage_by_category$bact_schloss_coverage, na.rm=T)`%).
-The interquartile range in the OTU coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to`r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
+The interquartile range in the OTU coverage by environment for the archaeal data was `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.25), na.rm=T)` to `r quantile(coverage_by_category$arch_schloss_coverage, prob=c(0.75), na.rm=T)` (median coverage=`r median(coverage_by_category$arch_schloss_coverage, na.rm=T)`%).
 The archaeal coverage was higher than that of the bacterial OTU coverage for all categories except the food-associated, plant surface, and other invertebrate categories.
@@
 ```{r culture}
-b_cult_time <- read.table(file='data/process/bacteria.cultured_by_time_counts.tsv', header=T)
-a_cult_time <- read.table(file='data/process/archaea.cultured_by_time_counts.tsv', header=T)
-
 b_percent_cult_nseqs_all <- 100*sum(b_cult_time[,"all_cult_nseqs"])/sum(b_cult_time[,"all_total_nseqs"])
@@
 This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
-Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure 3).
+Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure 5).
 So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
+
 Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative.
@@
 Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Figure 5).
-The Tenericutes, Proteobacteria, Firmicutes, and Spirochaeatae all had considerably higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
-This likely reflects the extremely high number of cultivated  *Mycobacterium tuberculosis*, *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, *Borrelia*, and others.
-Conversely, clades such as the *Actinobacteria*, *Fusobacteria*, and *Bacteroidetes* had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
-Regardless of these observations, the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
-
-```{r single_cell}
-overlap_counts <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
-```
+The Firmicutes and Proteobacteria both had higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
+This likely reflects the extremely high number of cultivated *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others.
+Conversely, most phyla had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+It is clear that the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
 
@@
 Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
-These have included single cell genomics [@Rinke2013] and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data using the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@Miller2011, @Miller2013].
-To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional culutre-independent amplification and cultivation-based approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure 6).
-Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences that met our criteria, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
-Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by cultivation or PCR-based efforts.
-Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences [@Wrighton2012, @Miller2011, @NunesdaRocha2015, @Hamilton2015, @Handley2012, @Gladden2011].
+These have included single cell genomics [@Rinke2013] and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data with the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@Miller2011; @Miller2013].
+To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure 6).
+Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
+Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by previous efforts.
+Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences [@Wrighton2012; @Miller2011; @NunesdaRocha2015; @Hamilton2015; @Handley2012; @Gladden2011].
 Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively.
@@
 Although the application of this approach to Archaea has been limited, it was still surprising that `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'cultured']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'pcr']`% of the archaeal OTUs had been previously recovered by traditional cultivation and PCR-based approaches, respectively.
-Finally, we pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from five studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@Brooks2014, @Miller2013, @Wilkins2013, @Handley2014, @Alessi2014].
+Finally, we pooled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_seqs']` bacterial sequences from five studies that utilized EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons [@Brooks2014; @Miller2013; @Wilkins2013; @Handley2014; @Alessi2014].
 These sequences were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_pcr", 'n_otus']` OTUs.
@@
 
-In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that our current methods and sampling schemes would never reach these estimates.
+In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that our current sampling methods will prevent us from acquiring full-length sequences for that level of diversity.
 As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations.
@@
 We found that most sequences deposited into public databases are being made by a small number of projects that have deeply sampled similar environments.
-Furthermore, the number of full-lenth reads deposited into the databases has stalled.
+Furthermore, the number of full-length reads deposited into the databases has stalled.
 We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.
@@
 Thus, the sequences from those studies represent organisms that are generally abundant.
-We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms is because our databases do not contain full-length references for those sequences.
+We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms are because our databases do not contain full-length references for those sequences.
 As stated in the 2004 census, we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage.
@@
 Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate.
-To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010, @Buerger2012, @Thrash2014, @Das2015].
+To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010; @Buerger2012; @Thrash2014; @Das2015].
 The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing.
@@
 "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata.
-Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included [@Miller2011, @Miller2013].
+Further, sequences generated from amplicon and shotgun metagenomic data using the EMIRGE program were also included [@Miller2011; @Miller2013].
 The IMG and EMIRGE sequences were aligned against the respective SILVA-based reference using mothur [@Schloss2009].

### 2016-01-28 Schloss_Census2_mBio_2016.Rmd
Update alignment coordiantes (didn't change)

@@
 Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates.
-
-The sequences in the resulting bacterial dataset overlapped bases X113X through X1344X of an *E.
-coli* reference sequence (V00348) and had a median length of X1,227X nt.
-The sequences in the resulting archaeal dataset overlapped positions X362X to X937X of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of X580X nt.
-
+The sequences in the resulting bacterial dataset overlapped bases 113 through 1350 of an *E.
+coli* reference sequence (V00348) and had a median length of 1,233 nt.
+The sequences in the resulting archaeal dataset overlapped positions 362 to 937 of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of 580 nt.
 The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences.
-The final datasets contained `r b_nseqs` bacterial and `r a_nseqs` archaeal 16S rRNA gene sequences.
+The final datasets contained 1,411,234 bacterial and 53,546 archaeal 16S rRNA gene sequences.
 Sequences were assigned to OTUs using the average neighbor clustering algorithm [@Westcott2015].

### 2016-02-02 Schloss_Census2_mBio_2016.Rmd
Integrate Cameron's Thrashing of the Discussion :)

@@
 Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once.
-Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_good_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the archaeal OTUs have been observed more than once.
+Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_schloss_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_schloss_coverage"]`% of the archaeal OTUs have been observed more than once.
 Thus, there is still considerable microbial diversity to be explored.
@@
 
-### Importance
+### Importance - too long
 
 A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity.
-
 The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical.
@@
 We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure 5).
-Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Figure 5).
-The Firmicutes and Proteobacteria both had higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs.
-This likely reflects the extremely high number of cultivated *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others.
-Conversely, most phyla had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
-It is clear that the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
+Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed a strong cultivation bias within the Firmicutes, which had a higher percentage of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs (Figure 5).
+This likely reflects the extremely high number of cultivated biomedically relevant cultivars such as *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others.
+Conversely, many phyla, including Cyanobacteria, Actinobacteria, Bacteroidetes, and Nitrospirae, had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+Nevertheless, it is clear that the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
 
@@
 
-In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that our current sampling methods will prevent us from acquiring full-length sequences for that level of diversity.
-As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations.
-Given this low level of OTU coverage, it is likely that there are many more bacterial and archaeal populations yet to be sampled.
-The recent advances in sequencing technologies have done little to expand the deposition of full-length sequences representing microbial biodiversity.
-We found that most sequences deposited into public databases are being made by a small number of projects that have deeply sampled similar environments.
-Furthermore, the number of full-length reads deposited into the databases has stalled.
-We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.
-
-During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to to generate dozens or hundreds of sequences per sample.
-The existing databases that are used for classifying sequences are based on these sequences.
-Thus, the sequences from those studies represent organisms that are generally abundant.
-We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms are because our databases do not contain full-length references for those sequences.
-As stated in the 2004 census, we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage.
-Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.
-
 It is clear that considerable biodiversity has been discovered since the first census in 2004.
@@
 Meanwhile, `r b_schloss`% of bacterial and `r a_schloss`% of archaeal OTUs have only been observed once.
-It appears that although brute force deep sequencing of communities may increase our coverage of OTUs, new methods are needed to more directly sample rare organisms.
-Nevertheless, novel technologies such as single-cell genomics, metagenomics, and algorithms to recover full-length sequences from new sequencing platforms have demonstrated promise in circumventing previous limitations in identifying new OTUs.
-Using EMIRGE to assemble fragmented 16S rRNA gene amplicons may allow us to obtain deep coverage of communities; however, it is still unclear how faithful the assembled sequence is to that of the original organism.
-Additional sequencing technologies also offer the ability to directly generate full-length sequences, such as PacBio and potentially Oxford Nanopore.
-Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate.
-To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010; @Buerger2012; @Thrash2014; @Das2015].
-The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing.
-One strength of cultivation, single cell genomics, and metagenomics approaches is that not only do the methods yield a full 16S rRNA gene sequence, but they also allow researchers to obtain the complete genome of the organism.
-These approaches are still in active development, we hope that through continuous refinement, they may allow us to significantly improve the coverage of OTUs in public databases.
+In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that existing sampling methods will prevent us from acquiring full-length sequences for that level of diversity.
+As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations.
+Given this low level of OTU coverage, it is likely that there are many more bacterial and archaeal populations yet to be sampled.
 
-Our ability to assess the microbial census is limited by the quality of the data in the databases.
-First, it is widely acknowledged that the primers used to amplify 16S rRNA genes are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads (Furhman, 2015).
-Furthermore, assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g.
-Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard PCR-based approaches [@Brown2015].
-Thus, for many environments, the estimates for the census must be considered as lower bounds.
-Although the primers used to amplify the entire gene are also biased, obtaining a better characterization of microbial diversity will help improve the design of primers targeting subregions of the gene.
-Second, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases.
+There are several additional reasons to suspect that the current census should be considered conservative.
+First, we found that most sequences recently deposited into public databases are being made by a small number of projects that have deeply sampled similar environments, and the number of full-length reads deposited into the databases has stalled.
+Second, it is widely acknowledged that 16S rRNA gene primers are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads [@Parada2015].
+Assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g.
+Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.) that would preclude detection with standard PCR-based approaches [@Brown2015; @EloeFadrosh2016].
+Third, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases.
 Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983.
@@
 
-The advance in sequencing technologies has significantly expanded our ability to peer into the depths of microbial communities.
-We now have unprecedented abilities to sequence large numbers of individual communities deeply for the same cost of generating hundreds of full length sequences.
-Yet, unless we invest in improving our full-length databases, it will continue to be difficult to contextualize this newly discovered biodiversity within the tree of life.
+Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.
+During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to generate dozens or hundreds of sequences per sample.
+The existing databases that are used for classifying sequences are based on those sequences, which represent organisms that are generally abundant.
+We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms are because our databases do not contain full-length references for those sequences.
+We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.
+
+Novel technologies such as single-cell genomics, metagenomics, and algorithms to recover full-length sequences from new sequencing platforms have demonstrated promise in circumventing previous limitations in identifying new OTUs.
+Using EMIRGE to assemble fragmented 16S rRNA gene amplicons may allow us to obtain deep coverage of communities; however, it is still unclear how faithful the assembled sequence is to that of the original organism.
+Additional sequencing technologies also offer the ability to directly generate full-length sequences, such as PacBio and potentially Oxford Nanopore.
+Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate [@Schloss2015].
+To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010; @Buerger2012; @Thrash2015; @Das2015].
+The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing.
+These approaches are still in active development, and we hope that through continuous refinement, they may allow us to significantly improve the coverage of OTUs in public databases.
 

### 2016-02-02 Schloss_Census2_mBio_2016.Rmd
Pat's final edits

@@
 Thus, there is still considerable microbial diversity to be explored.
-Unfortunately, the number of new full-length sequences has been declining and these sequences are primarily deposited by a small number of studies.
-Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the sequences while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the sequences.
-Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling biases.
+Unfortunately, the rate of new full-length sequences has been declining and new sequences are primarily being deposited by a small number of studies.
+Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the census while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the census.
+Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling bias.
 The success of ongoing efforts to use short read sequencing to characterize microbial communities requires that researchers strive to expand the depth and breadth of the microbial census.
 
-### Importance - too long
+### Importance
 
-A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity.
 The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical.
@@
 
-Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial 16S rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998; @Curtis2002].
+Previously, Schloss and Handelsman [-@Schloss2004] assigned the 56,215 partial 16S rRNA gene sequences that were available in the Ribosomal Database Project to operational taxonomic units (OTUs) and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998; @Curtis2002].
 That census called for a broader and deeper characterization of all environments.
@@
 
-It has been disconcerting that the increase in sequencing volume has come at the cost of sequence length.
-The commonly used MiSeq-based sequencing platform from Illumina is extensively leveraged to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
+It is disconcerting that the increase in sequencing volume has come at the cost of sequence length.
+The commonly used MiSeq-based sequencing platform from Illumina is extensively used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
 The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef2009; @Schloss2010].
 Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences.
-We likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
+Furthermore, we likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
 
@@
 These results indicate that regardless of the domain, continued sampling with the current strategies for generating full-length sequences will largely reveal OTUs that have already been observed, even though a large fraction of OTUs have only been sampled once.
-Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once, it is likely that an even larger number of OTUs have yet to be sampled for both domains.
+Considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once, it is likely that an even larger number of OTUs have yet to be sampled for both domains.
 
@@
 In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year.
-Although these deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.
+Although the recent deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.
 
@@
 Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_shift_cult_nseqs_all` and `r a_shift_cult_nseqs_all`-fold for bacteria and archaea, respectively.
-Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r nrow(a_cult_time)` of the `r sum(a_cult_time$all_cult_nseqs == 0)` archaeal phyla have cultured representatives.
+Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r sum(a_cult_time$all_cult_nseqs == 0)` of the `r nrow(a_cult_time)`  archaeal phyla have no cultured representatives.
 This lag is likely due to the differences in throughput of culture-dependent and -independent approaches.
 Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure 5).
-So, even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
+Even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.
 
@@
 Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed a strong cultivation bias within the Firmicutes, which had a higher percentage of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs (Figure 5).
-This likely reflects the extremely high number of cultivated biomedically relevant cultivars such as *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others.
-Conversely, many phyla, including Cyanobacteria, Actinobacteria, Bacteroidetes, and Nitrospirae, had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity.
+This likely reflects the extremely high number of cultivated biomedically relevant cultivars from genera such as *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others.
+Conversely, many phyla, including Cyanobacteria, Actinobacteria, Bacteroidetes, and Nitrospirae, had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively inefficient with regards to available diversity.
 Nevertheless, it is clear that the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.
@@
 Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively.
-Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively had not been observed by previous efforts.
+Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively, had not been observed by previous efforts.
 Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences [@Wrighton2012; @Miller2011; @NunesdaRocha2015; @Hamilton2015; @Handley2012; @Gladden2011].

### 2016-02-02 Schloss_Census2_mBio_2016.Rmd
Fresh 'make write.paper'

@@
 By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census.
-We actually found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms.
+We found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms.
 Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms.

### 2016-02-02 results/tables/build_archaeal_category_phylum_table.Rmd
Rerun entire makefile with updated metadata files
Fix row labels in supplemental figures

@@
 data <- read.table(file="data/process/phylum_category_counts.tsv", header=T)
+archaea <- data[data$domain=="archaea",]
+rownames(archaea) <- archaea$phylum
+archaea <- archaea[,-c(1,2)]
 
-archaea <- data[data$domain=="archaea",-1]
 archaea <- archaea[order(archaea$total, decreasing=T), names(categories)]
@@
 
-caption <- "\\textbf{Supplementary Table 3. Frequency that each archaeal phylum was found across each of the environmental categories.}"
+caption <- "\\textbf{Supplementary Table 5. Frequency that each archaeal phylum was found across each of the environmental categories.}"
 

### 2016-02-02 results/tables/build_bacterial_category_phylum_table.Rmd
Rerun entire makefile with updated metadata files
Fix row labels in supplemental figures

@@
 data <- read.table(file="data/process/phylum_category_counts.tsv", header=T)
+bacteria <- data[data$domain=="bacteria",]
+rownames(bacteria) <- bacteria$phylum
+bacteria <- bacteria[,-c(1,2)]
 
-bacteria <- data[data$domain=="bacteria",-1]
 bacteria <- bacteria[order(bacteria$total, decreasing=T), names(categories)]
@@
 
-caption <- "\\textbf{Supplementary Table 2. Frequency that each bacterial phylum was found across each of the environmental categories.}"
+caption <- "\\textbf{Supplementary Table 4. Frequency that each bacterial phylum was found across each of the environmental categories.}"
 

### 2016-02-03 Schloss_Census2_mBio_2016.Rmd
Put supplemental table captions into paper

@@
 
+**Supplemental Table 1. Description of environmental categories and the criteria used to assign sequences to each category.**
+
+**Supplementary Table 2. Frequency that each bacterial phylum was sequenced before and after 2006.**
+
+**Supplementary Table 3. Frequency that each archaeal phylum was sequenced before and after 2006.**
+
+**Supplementary Table 4. Frequency that each bacterial phylum was found across each of the environmental categories.**
+
+**Supplementary Table 5. Frequency that each archaeal phylum was found across each of the environmental categories.**
+
+**Supplementary Table 6. Frequency that each bacterial sequence or OTU was retrieved by cultivation or by culture-independent methods.**
+
+**Supplementary Table 7. Frequency that each archaeal sequence or OTU was retrieved by cultivation or by culture-independent methods.**
+
 \newpage

### 2016-02-03 code/build_culture_effort_plot.R
Change figure format from PDF to EPS

@@ -30,7 +30,7 @@ a_n_taxa <- nrow(a_cultured)
 
 total_lines <- b_n_taxa + a_n_taxa + 2
 
-pdf("results/figures/phylum_cultured.pdf",  width=3.75, height=4.75)
+cairo_ps("results/figures/phylum_cultured.eps",  width=3.75, height=4.75)
 par(mar=c(5,10,1.0,0.5), cex=0.7)
 plot(NA, xlim=c(0,70), ylim=c(1,total_lines), axes=F, xlab="", ylab="")
 abline(h=c(1:a_n_taxa, (a_n_taxa+3):total_lines), lty=3, col="gray")

### 2016-02-03 code/build_domain_rarefaction_plot.R

@@ -53,7 +53,7 @@ arch_domain <- read.table(file='data/process/archaea.all_rarefaction.tsv', heade
 bact_category <- read.table(file='data/process/bacteria.env_rarefaction.tsv', header=T)
 arch_category <- read.table(file='data/process/archaea.env_rarefaction.tsv', header=T)
 
-pdf('results/figures/domain_rarefaction.pdf', width=7.5, height=7.5)
+cairo_ps('results/figures/domain_rarefaction.eps', width=7.5, height=7.5)
 par(mar=c(2.5,4,0.5,0.5))
 
 layout(matrix(c(5,1,2,5,3,4,0,6,6), nrow=3, byrow=T), widths=c(0.1,1,1), heights=c(1, 1, 0.10))

### 2016-02-03 code/build_otu_overlap_by_method_figure.R

@@ -23,7 +23,7 @@ plot_overlap <- function(domain){
 overlap <- read.table(file="data/process/otu_overlap_by_method.tsv", header=T)
 
 
-pdf(file="results/figures/otu_overlap_by_method.pdf", width=6.0, height=5.0)
+cairo_ps(file="results/figures/otu_overlap_by_method.eps", width=6.0, height=5.0)
 
 layout(matrix(c(3,1,3,2,0,4), nrow=3, byrow=T), widths=c(0.1,1), heights=c(1, 1, 0.18))
 plot_overlap("bacteria")

### 2016-02-03 code/build_phylum_category_heatmap.R

@@ -27,7 +27,7 @@ rownames(archaea_top) <- gsub("_", " ", rownames(archaea_top))
 
 
 
-pdf(file="results/figures/category_phylum_heatmap.pdf", width=7.5, height=5)
+cairo_ps(file="results/figures/category_phylum_heatmap.eps", width=7.5, height=5)
 
 layout(matrix(c(3,1,2), nrow=1), widths=c(0.5,1,1))
 par(mar=c(11, 0.5, 0.5, 0.5))

### 2016-02-03 code/build_phylum_effort_plot.R

@@ -18,7 +18,7 @@ names(arch_ratios) <- gsub("Miscellaneous_Crenarchaeotic_Group", "Misc. Crenarch
 arch_ratios <- sort(arch_ratios)
 n_arch_ratios <- length(arch_ratios)
 
-pdf(file="results/figures/phylum_effort.pdf", width=3.75, height=5)
+cairo_ps(file="results/figures/phylum_effort.eps", width=3.75, height=5)
 	par(mar=c(5,10, 1.0, 0.5), cex=0.7)
 
 	plot("", ylim=c(1,n_bact_ratios+n_arch_ratios+2), xlim=c(-3,4), type="n", xlab="", ylab="", axes=F)

### 2016-02-03 code/build_time_course_plots.R

@@ -12,7 +12,7 @@ exclude_year <- 2015
 time_data <- time_data[year != exclude_year,]
 year <- year[year != exclude_year]
 
-pdf(file="results/figures/time_course_figure.pdf", width=3.75, height=6.0)
+cairo_ps(file="results/figures/time_course_figure.eps", width=3.75, height=6.0)
 	layout(matrix(c(1,2,3,4), nrow=4), height=c(1,1,1,0.25))
 
 	#A

### 2016-03-14 rebuttal.md
Incorporate reviewers' comments

**Reviewer #1 (Comments for the Author):**

**The manuscript by Patrick Schloss et al provides an updated on the status of the microbial census. By microbial census the authors refer to the census of bacteria and archaea assessed through the full length 16S rRNA gene sequences.**

**The manuscript is well written and the authors have done a good job in summarizing the current status of the census and the bottlenecks in scaling up.**

**A few minor issues to consider below**
**Page 1. While the manuscript describes a census of bacteria and archaea only, it is referred to as microbial census which however includes microbial eukaryotes. As such the word microbial in the context of this paper and this census is misleading. I assume that the authors would prefer not to use the word prokaryotic census, but microbial is not accurate either.**

**Lines 22-24: The meaning of the sentence "Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes", is a bit confusing. There are several databases that keep good track of all the available 16S sequences so the authors should clarify here what they mean.**

**Lines 25-27: The authors claim that "We found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms". However, recent literature is not very much supportive of that. In a recent publication (Eloe-Fadrosh et al 2016) it is shown that there is a large number of unaffiliated 16S sequences by looking at the assembled metagenomics sequences. Since typically, most of the 16S sequences that get assembled from metagenomic datasets are not of very low abundance organisms, perhaps the statement above doesn't seem to hold up? Furthermore, the authors agree in the discussion that the primers uses for PCR have biases, which contradicts the statement that ongoing efforts have done an excellent job in sampling the most abundant organisms.**

**Lines 142-143: the sentence "Based on the representation of sequences within the SILVA database, in 2006 there were 61 bacterial and 18 phyla" seems to be incomplete. Do they mean 18 "archaeal" phyla?**

**Lines 253-256: The authors discuss the contribution of assembled metagenomics data in showing the presence of introns in 16S rRNA genes. They don't discuss at all the 16S primer mismatches though as another source of significant diversity missed even from abundant organisms as has been recently shown.**

**The authors are also using throughout the text the term "zoonotic" as one of the broad environmental categories. The use of this term is confusing. Do the authors mean host-associated? If yes, I suggest to adopt this term, since this is the term used in the SILVA database and the term recommended from the Genomics Standards Consortium and the MIxS standards. Zoonotic also refers to pathogenic/disease states and unless the authors refer specifically to this type of environments rather than host-associated environments, I would recommend changing the term here.**

**Also, while Table 1 refers to Zoological samples, one of the curves in Figure 1C refers to Zoonotic. I assume that the zoonotic in the figure corresponds to the zoological in the Table, but keeping two terms is confusing, I suggest the authors will also use zoological in the figure to match the Table.**

**It would be also very interesting if the authors would discuss how many of the totally number of OTUs, they generate from the SILVA data remain unaffiliated with any of the known phyla, if any.**

**Reviewer #2 (Comments for the Author):**

**Overall I believe this manuscript is an important and useful extension of the previous microbial census paper. I think the results will be important for the entire community of microbial ecologists and also many others.**

**I have some minor concerns with various statements in the manuscript. They are below**

**Line 2: A census is typically carried out for people at a national level.**
**This is simply not true - censuses are carried out for states, counties, cities, towns, and more**

**Line 6: it would be good in the abstract to mention why only full length sequences are being considered for this census**

**Line 21: I think it would be worth addressing the recent issues in regard to whether the archaea are in fact a separate domain from eukaryotes**

**Line 24: again I think it would be important to explain here why the focus on full length sequences?**

**Line 33: I would argue this is fundamental to more than just ecology**

**Line 42. Regarding the deposition of this sequence in 1983 in Genbank at the NCBI.. I don't think this is possible. I think NCBI did not take over Genbank until later. Plus the paper was in 1978 and the precursor toe Genbank existed before 1983 so it would be good to expain the discrepancies.**

**Line 45: "making it the best-represented gene." details and or/ reference needed for this sentence**

**Line 68-69. "The number of OTUs that are sampled when using different regions". Not completely clear what is meant here - do you mean the # that are inferred from the data or the # that are actually sampled by the PCR for the different regions?**

**Line 74: it would be helpful to clarify what is meant here by "lack the references necessary"**

**Line 92. I think the only thing that can be said from this is saturation for the types of sites being sampled**

**Line 104-105. "it is likely that an even larger number of OTUs have yet to be sampled for both domains." Maybe this is a wording thing but it is unclear to me how we can have "started to saturate" yet have an even larger number of OTUs to be sampled**

**Line 143-145. Since then there have been 4 new bacterial (CKC4, OC31, S2R-29, and SBYG-2791) and 2 new archaeal candidate phyla (Ancient Archaeal Group and TVG8AR30). Citations for this? and is this classification simply pulled fro Silva or based upon other analysis?**

**Line 177. "Great concern" seems excessive. I think "concern expressed" would be more appropriate**

**Line 208. "There is concern". Can you clarify - or, perhaps it would be better to say "we are concerned"**

**Line 233. Problems with EMIRGE. I think this is possibly true. But not enough evidence is presented in this paper to make this type of statement. What evidence do you have that any errors are coming from EMIRGE?**

**Line 292 - 322. Much of the text here discusses data from Silva which appears to have mostly come from EMBL. And other text discusses sequences from other sources. This is all fine. But the text as it is written in the main parts of the paper does not provide enough of this for context. I think it would be very beneficial to add a bit more to the main text on the sources of the sequences.**

**Line 337-338. Figshare. Really good to see what is being released in Figshare and Github for this paper. I would further recommend that sequeneces used in this paper also be included in Figshare so that people do not have to rerun the code in order to get access to the sequences (this is what seems to be required)**

**Line 363. " grow at a rate too slow to ever reach estimates of 10^6 to 10^11 bacterial species." This needs further details in my opinion. What would it take, for example to get to 10^6? How many years at current rates? What if the rate went up a bit? Then how many years?**

**Line 388 "that were also detected by other." I assume some word is missing here.**

### 2016-03-14 submission/rebuttal.md
Pat's first whack at revisions

--- 2016-03-14 rebuttal.md
@@
 
+We appreciate the reviewer's kind words regarding our manuscript as well as their suggestions that we have done our best to incorporate into the new version of the manuscript.
+
 **A few minor issues to consider below**
@@
 
+Where appropriate, we have made the text more specific to describing the "archaeal and  bacterial census" rather than the microbial census.
+
 **Lines 22-24: The meaning of the sentence "Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes", is a bit confusing.
 There are several databases that keep good track of all the available 16S sequences so the authors should clarify here what they mean.**
+
+We have edited this sentence to hopefully make it more clear that we're referring to the fact that the ongoing collection and deposition of sequences in public databases is somewhat haphazard.
+In contrast, our analysis is more formal and allows us to take stock of the quality of the biodiversity within those databases.
 
@@
 
+This observation fits with the overall story of our analysis.
+In the Importance section, we are referring to the large number of singleton OTUs in the database.
+It is not hard to conceive the problem described by Eloe-Fardrosh when one combines the limited environmental sampling that has been performed.
+
 **Lines 142-143: the sentence "Based on the representation of sequences within the SILVA database, in 2006 there were 61 bacterial and 18 phyla" seems to be incomplete.
@@
 
+This has been corrected.
+
 **Lines 253-256: The authors discuss the contribution of assembled metagenomics data in showing the presence of introns in 16S rRNA genes.
 They don't discuss at all the 16S primer mismatches though as another source of significant diversity missed even from abundant organisms as has been recently shown.**
+
+This is actually described in the preceding sentence.
 
@@
 
+We have changed "zoonotic" to "host-associated."
+
 **Also, while Table 1 refers to Zoological samples, one of the curves in Figure 1C refers to Zoonotic.
@@
 
+We have changed "zoonotic"/"zoological" to "host-associated."
+
 **It would be also very interesting if the authors would discuss how many of the totally number of OTUs, they generate from the SILVA data remain unaffiliated with any of the known phyla, if any.**
+
+All of the bacterial sequences we obtained from the SILVA database affiliated with a phylum (Supplementary Table 6).
+Among the archaeal sequences two OTUs did not affiliate with a phylum (Supplementary Table 7).
 
@@
 
+We appreciate the reviewer's support for our manuscript. We have done our best to incorporate their comments into the new version of the manuscript.
+
 **Line 2: A census is typically carried out for people at a national level.**
@@
 
+Agreed. We have changed this to point out that a census is typically done "across a range of geographic levels."
+
 **Line 6: it would be good in the abstract to mention why only full length sequences are being considered for this census**
+
+We have added a sentence to the abstract.
 
@@
 
+This is certainly an interesting question; however, given our clarification that we are only concerned with archaeal and bacterial 16S rRNA gene sequences, we feel that this debate is not entirely pertinent to the overall message of the current manuscript.
+
 **Line 24: again I think it would be important to explain here why the focus on full length sequences?**
 
+Given the lack of space allowed to us for the Abstract and Importance sections, we have only included this explanation in the Abstract.
+We also justify this more completely in the main body of the manuscript.
+
 **Line 33: I would argue this is fundamental to more than just ecology**
+
+?????
 
@@
 
+We have edited the sentence here and in the Conclusion section to remove emphasis on the dates. Here we point out that it was released with the rest of GenBank in 1983.
+
 **Line 45: "making it the best-represented gene." details and or/ reference needed for this sentence**
+
+?????
 
@@
 
+We have added a clause to point out the number of OTUs that are inferred from the data. It's also likely true that the PCR sampling is also biased between regions.
+
 **Line 74: it would be helpful to clarify what is meant here by "lack the references necessary"**
+
+We have edited this sentence to clarify that the necessary references are "full-length reference sequences"
 
@@
 
+We hope this is implied from the analysis and discussion we have provided in the manuscript.
+
 **Line 104-105.
 "it is likely that an even larger number of OTUs have yet to be sampled for both domains." Maybe this is a wording thing but it is unclear to me how we can have "started to saturate" yet have an even larger number of OTUs to be sampled**
+
+?????
 
@@
 
+These names were simply pulled from SILVA.
+
 **Line 177. "Great concern" seems excessive. I think "concern expressed" would be more appropriate**
 
+This has been edited as suggested.
+
 **Line 208. "There is concern". Can you clarify - or, perhaps it would be better to say "we are concerned"**
+
+This has been edited as suggested.
 
@@
 What evidence do you have that any errors are coming from EMIRGE?**
+
+It was the high degree of novelty that was observed among the EMIRGE sequences that makes us concerned.
+Furthermore, considering EMIRGE hasn't been used to sequence a mock community, it is unclear what the error rates are within contigs generated by the approach.
 
@@
 
+We feel that this section of text adequately describes the sources of data used in the paper: SILVA, IMG, and the individual EMIRGE datatsets.
+These are rolled out in the Results section as we describe the main analysis (SILVA) and the newer methods (IMG and EMIRGE).
+
 **Line 337-338.
@@
 I would further recommend that sequeneces used in this paper also be included in Figshare so that people do not have to rerun the code in order to get access to the sequences (this is what seems to be required)**
+
+We are reluctant to do this since the sequences were not generated by us.
+In addition, the SILVA sequences are subject to a somewhat odd licensing agreement.
+The code posted in GitHub describes in detail how to obtain and process the sequences.
 
@@
 
+We have removed this sentence since it is not directly discussed in the main body of the manuscript.
+
 **Line 388 "that were also detected by other." I assume some word is missing here.**
+
+We have corrected the text by adding "methods."

### 2016-03-14 submission/Schloss_Census2_mBio_2016.Rmd
Pat's first whack at revisions

--- 2016-02-03 Schloss_Census2_mBio_2016.Rmd
@@
 ---
-title: "**The status of the microbial census: an update**"
+title: "**The status of the archaeal and bacterial census: an update**"
 bibliography: references.bib
@@
   pdf_document:
+    keep_tex: true
     includes:
@@
 
-Running title: The microbial census
+Running title: The archaeal and bacterial census
 
@@
 
-A census is typically carried out for people at a national level; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes.
-We assessed how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies.
+A census is typically carried out for people across a range of geographical levels; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes.
+We assessed how well the census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies because full-length sequences are typically used as references for classification of the short sequences generated by newer technologies.
 Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once.
 Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_schloss_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_schloss_coverage"]`% of the archaeal OTUs have been observed more than once.
-Thus, there is still considerable microbial diversity to be explored.
+Thus, there is still considerable diversity to be explored.
 Unfortunately, the rate of new full-length sequences has been declining and new sequences are primarily being deposited by a small number of studies.
-Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the census while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the census.
+Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the census while sequences associated with host-associated environments represent `r coarse_percentages["Z"]`% of the census.
 Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling bias.
-The success of ongoing efforts to use short read sequencing to characterize microbial communities requires that researchers strive to expand the depth and breadth of the microbial census.
+The success of ongoing efforts to use short read sequencing to characterize archaeal and bacterial communities requires that researchers strive to expand the depth and breadth of this census.
 
@@
 The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical.
-Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes.
+Surprisingly, we have done a relatively poor job of formally tracking  the quality of the biodiversity as represented in full-length 16S rRNA genes.
 By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census.
@@
 Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms.
-Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of microbial diversity.
+Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of archaeal and bacterial diversity.
 
@@
 Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene.
-In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was deposited into NCBI's GenBank making it the first of what is now more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978].
+In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was released as part of NCBI's GenBank making it the first of what is now more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978].
 16S rRNA gene accessions represent nearly one-third of all sequences deposited in GenBank, making it the best-represented gene.
@@
 The commonly used MiSeq-based sequencing platform from Illumina is extensively used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp.
-The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef2009; @Schloss2010].
+The number of OTUs that are inferred from the sequencing data when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation with the genetic diversity of the full-length sequence [@Youssef2009; @Schloss2010].
 Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences.
-Furthermore, we likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
+Furthermore, we likely lack the full-length reference sequences necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.
 
-Here we update the status of the microbial census with full-length 16S rRNA gene sequences.
+Here we update the status of the archaeal and bacterial census with full-length 16S rRNA gene sequences.
 In the 13 years since the collection of data for Schloss and Handelsman’s initial census, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies.
@@
 These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012].
-The heavy zoonotic focus is reflected in the rarefaction curve for this category (Figure 1C).
+The heavy focus on host-associated communities is reflected in the rarefaction curve for this category (Figure 1C).
 In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year.
@@
 
-***The ability to sample microbial life is taxonomically skewed.***
+***The ability to sample archaea and bacteria is taxonomically skewed.***
 The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences.
@@
 We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published [@Sogin2006].
-Based on the representation of sequences within the SILVA database, in 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla.
+Based on the representation of sequences within the SILVA database, in 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` archaeal phyla.
 Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`).
@@
 ***The cultured census.***
-In the 2004 bacterial census, there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
+In the 2004 bacterial census, there was concern expressed that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives.
 To update this assessment, we identified those sequences that came from cultured and uncultured organisms.
@@
 ***New technologies to access novel biodiversity.***
-Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled.
+Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, we are concerned that our ability to harvest full-length sequences from communities will remain stalled.
 Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome.
@@
 Third, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases.
-Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983.
-Similarly, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank.
+As an illustration of this problem, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank.
 This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences.
@@
 
-Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.
+Efforts to census archaea and bacteria using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of archaeal and bacterial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.
 During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to generate dozens or hundreds of sequences per sample.
@@
 Additional sequencing technologies also offer the ability to directly generate full-length sequences, such as PacBio and potentially Oxford Nanopore.
-Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate [@Schloss2015].
+Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffered from a high error rate [@Schloss2015].
 To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010; @Buerger2012; @Thrash2015; @Das2015].
@@
 Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).
-The number of bacterial and archaeal OTUs observed among the longest sequences in the SILVA database continues to grow at a rate too slow to ever reach estimates of 10^6^ to 10^11^ bacterial species.
 
 **Figure 2.
-Progression of the microbial census since the first full-length 16S rRNA gene sequence was deposited into GenBank in 1983.*** The number of bacterial and archaeal 16S rRNA gene sequences deposited (A) and the new OTUs they represent (B) has increased exponentially until the last several years when the rate of change has plateaued.
+Progression of the archaeal and bacterial census since the first full-length 16S rRNA gene sequence was deposited into GenBank in 1983.*** The number of bacterial and archaeal 16S rRNA gene sequences deposited (A) and the new OTUs they represent (B) has increased exponentially until the last several years when the rate of change has plateaued.
 For both bacterial and archaeal sequences, the number of studies that are responsible for depositing more than 50% of the sequences each year has been relatively small (C).
@@
 **Figure 6.
-The percentage of bacterial and archaeal OTUs found by single cell genomics and EMIRGE using PCR or metagenomics that were also detected by other.** The bars comparing a method to itself indicate the percentage of OTUs that were only detected by that method.
+The percentage of bacterial and archaeal OTUs found by single cell genomics and EMIRGE using PCR or metagenomics that were also detected by other methods.** The bars comparing a method to itself indicate the percentage of OTUs that were only detected by that method.
+
+\newpage
 

### 2016-03-14 submission/Schloss_Census2_mBio_2016.tex
Pat's first whack at revisions

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={The status of the archaeal and bacterial census: an update},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{The status of the archaeal and bacterial census: an update}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\vspace{35mm}

Running title: The archaeal and bacterial census

\vspace{35mm}

Patrick D. Schloss\textsuperscript{1\(\dagger\)}, Rene
Girard\textsuperscript{2}, Thomas Martin\textsuperscript{2}, Joshua
Edwards\textsuperscript{2}, and J. Cameron
Thrash\textsuperscript{2\(\dagger\)}

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}} and
\href{mailto:thrashc@lsu.edu}{\nolinkurl{thrashc@lsu.edu}}

1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

2. Department of Biological Sciences, Louisiana State University, Baton
Rouge, LA 70803

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

A census is typically carried out for people across a range of
geographical levels; however, microbial ecologists have implemented a
molecular census of bacteria and archaea by sequencing their 16S rRNA
genes. We assessed how well the census of full-length 16S rRNA gene
sequences is proceeding in the context of recent advances in high
throughput sequencing technologies because full-length sequences are
typically used as references for classification of the short sequences
generated by newer technologies. Among the 1,411,234 and 53,546
full-length bacterial and archaeal sequences sequences, 94.5\% and
95.1\% of the bacterial and archeaeal sequences, respectively, belonged
to operational taxonomic units (OTUs) that have been observed more than
once. Although these metrics suggest that the census is approaching
completion, 29.2\% of the bacterial and 38.5\% of the archaeal OTUs have
been observed more than once. Thus, there is still considerable
diversity to be explored. Unfortunately, the rate of new full-length
sequences has been declining and new sequences are primarily being
deposited by a small number of studies. Furthermore, sequences from soil
and aquatic environments, which are known to be rich in bacterial
diversity, only represent 7.8 and 16.5\% of the census while sequences
associated with host-associated environments represent 55.0\% of the
census. Continued use of traditional approaches and new technologies
such as single cell genomics and short read assembly are likely to
improve our ability to sample rare OTUs if it is possible to overcome
this sampling bias. The success of ongoing efforts to use short read
sequencing to characterize archaeal and bacterial communities requires
that researchers strive to expand the depth and breadth of this census.

\subsubsection{Importance}\label{importance}

The biodiversity contained within the bacterial and archaeal domains
dwarfs that of the eukaryotes and the services these organisms provide
to the biosphere are critical. Surprisingly, we have done a relatively
poor job of formally tracking the quality of the biodiversity as
represented in full-length 16S rRNA genes. By understanding how this
census is proceeding, it is possible to suggest the best allocation of
resources for advancing the census. We found that the ongoing effort has
done an excellent job of sampling the most abundant organisms, but
struggles to sample the more rare organisms. Through the use of new
sequencing technologies we should be able to obtain full-length
sequences from these rare organisms. Furthermore, we suggest that by
allocating more resources to sampling environments known to have the
greatest biodiversity we will be able to make significant advances in
our characterization of archaeal and bacterial diversity.

\newpage

\subsection{Introduction}\label{introduction}

The effort to quantify the number of different organisms in a system
remains fundamental to understanding ecology (1, 2). At the scale of
microorganisms, small physical sizes, morphological ambiguity, and
highly variable population sizes complicate this process. Furthermore,
creating standards for delimiting what makes one microbe ``different''
from another has been contentious (3, 4). In spite of these challenges,
we continue to peel back the curtain on the microbial world with the aid
of more and more informative, if still limited, technologies like
cultivation, 16S rRNA gene surveys, single cell technologies, and
metagenomics.

Generating a comprehensive understanding of any system with a single
gene may seem a fool's errand, yet we have learned a considerable amount
regarding the diversity, dynamics, and natural history of microorganisms
using the venerable 16S rRNA gene. In 1983, the full-length 16S rRNA
gene sequence of \emph{Escherichia coli} (accession J01695) was released
as part of NCBI's GenBank making it the first of what is now more than
10 million 16S rRNA gene sequences to be deposited into the database
(5). 16S rRNA gene accessions represent nearly one-third of all
sequences deposited in GenBank, making it the best-represented gene. As
Sanger sequencing has given way to so-called ``next generation
sequencing'' technologies, hundreds of millions of 16S rRNA gene
sequences have been deposited into the NCBI's Sequence Read Archive. The
expansion in sequencing throughput and increased access to sequencing
technology has allowed for more environments to be sequenced at a deeper
coverage, resulting in the identification of novel taxa. The ability to
obtain sequence data from microorganisms without cultivation has
radically altered our perspective of their role in nearly every
environment from deep ocean sediment cores (e.g.~accession AY436526) to
the International Space Station (e.g.~accession DQ497748).

Previously, Schloss and Handelsman (6) assigned the 56,215 partial 16S
rRNA gene sequences that were available in the Ribosomal Database
Project to operational taxonomic units (OTUs) and concluded that the
sampling methods of the time were insufficient to identify the
previously estimated 10\textsuperscript{7} to 10\textsuperscript{9}
different species (7, 8). That census called for a broader and deeper
characterization of all environments. Refreshingly, this challenge was
largely met. There have been major investments in studying the Earth's
microbiome using 16S rRNA gene sequencing through initiatives such as
the Human Microbiome Project (9), the Earth Microbiome Project (10), and
the International Census of Marine Microorganisms (11). But most
importantly, the original census was performed on the cusp of radical
developments in sequencing technologies. That advancement has moved the
generation of sequencing throughput from large sequencing centers to
individual investigators and leveraged their diverse interests to expand
the representation of organisms and environments represented in public
databases.

It is disconcerting that the increase in sequencing volume has come at
the cost of sequence length. The commonly used MiSeq-based sequencing
platform from Illumina is extensively used to sequence the approximately
250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have
used different parts of the gene that are generally shorter than 500 bp.
The number of OTUs that are inferred from the sequencing data when using
different regions within the 16S rRNA gene can vary considerably and the
genetic diversity within these regions typically has only a modest
correlation with the genetic diversity of the full-length sequence (12,
13). Thus, it remains unclear to what degree richness estimates from
short read technologies over or underestimate the numbers from
full-length sequences. Furthermore, we likely lack the full-length
reference sequences necessary to adequately classify the novel
biodiversity we are sampling when we generate 100-times the sequence
data from a community than we did using full-length sequencing.

Here we update the status of the archaeal and bacterial census with
full-length 16S rRNA gene sequences. In the 13 years since the
collection of data for Schloss and Handelsman's initial census, the
number of full-length sequences has grown exponentially, despite the
overwhelming contemporary focus by most researchers on short-read
technologies. This update to the census allows us to evaluate the
relative sampling thoroughness for different environments and clades and
make an argument for the continued need to collection full-length
sequence data from many systems that have a long history of study. As
researchers consider coalescing into a Unified Microbiome Initiative
(14), it will be important to balance the need for mechanism-based
studies with the need to generate full-length reference sequences from a
diversity of environments.

\subsection{Results and Discussion}\label{results-and-discussion}

\textbf{\emph{The status of the bacterial and archaeal census.}} To
assess the field's progress in characterizing the biodiversity of
bacteria and archaea, we assigned each 16S rRNA gene sequence to OTUs
using distance thresholds that varied between 0 and 20\%. Although it is
not possible to link a specific taxonomic level (e.g.~species, genus,
family, etc.) to a specific distance threshold, we selected distances of
0, 3, 5, 10, and 20\% because they are widely regarded as representing
the range of genetic diversity of the 16S rRNA gene within each domain.
By rarefaction, it was clear that the ongoing sampling efforts have
started to saturate the number of current OTUs. After sampling 1,411,234
near full-length bacterial 16S rRNA gene sequences we have identified
217,645, 108,950, 66,819, 15,743, and 3,731 OTUs at the respective
thresholds (Figure 1A, Table 1). Using only the OTUs generated using a
3\% threshold, we calculated a 94.5\% Good's coverage (percent of
sequences belonging to OTUs that have been observed more than once), but
only 29.2\% OTU coverage (percent of the OTUs that have been observed
more than once). Paralleling the bacterial results, after sampling
53,546 archaeal 16S rRNA gene sequences we have identified 11,040,
4,252, 2,364, 812, and 110 OTUs (Figure 1B, Table 1). Using only the
OTUs generated with a 3\% threshold, we calculated a 95.1\% Good's
coverage, but only 38.5\% OTU coverage. These results indicate that
regardless of the domain, continued sampling with the current strategies
for generating full-length sequences will largely reveal OTUs that have
already been observed, even though a large fraction of OTUs have only
been sampled once. Considering more than 70.8\% of the OTUs have only
been observed once, it is likely that an even larger number of OTUs have
yet to be sampled for both domains.

\textbf{\emph{Sequencing efforts are a source of bias in the census.}}
One explanation for the large number of OTUs that have only been
observed once is that with the the broad adoption of sequencing
platforms that generate short sequence reads, the rate of full-length
sequence generation has declined. In fact, since 2009 the number of new
bacterial sequences generated has slowed to an average of 189,960
sequences per year (Figure 2A). Although this is still an impressive
number of sequences, since 2007 the number of new bacterial OTUs has
plateaued at an average of 11,184 new OTUs per year (Figure 2B). Given
the expense of generating full-length sequences using the Sanger
sequencing technology and the transition to other platforms at that
time, we expected that the large number of sequences were being
deposited by a handful of large projects. Indeed, when we counted the
number of submissions responsible for depositing 50\% of the sequences,
we found that with the exception of 2006 and 2013, eight or fewer
studies were responsible for depositing the majority of the full-length
sequences each year since 2005 (Figure 2C). Between 2009 and 2012,
908,190 total sequences were submitted and 6 submissions from 5 studies
were responsible for depositing 550,274 (60.6\% of all sequences). These
studies generated sequences from the human gastrointestinal tract (15),
human skin (16, 17), murine skin (18), and hypersaline microbial mats
(19). The heavy focus on host-associated communities is reflected in the
rarefaction curve for this category (Figure 1C). In contrast to recent
years, between 1995 and 2006, an average of 39.3 studies were
responsible for submitting more than half of the sequences each year.
Although the recent deep surveys represent significant contributions to
our knowledge of bacterial biogeography, their small number and lack of
environmental diversity is indicative of the broader problems in
advancing the bacterial census.

The depth of sequencing being done to advance the archaeal census has
been 26-times less than that of the bacterial census (Table 1). The
annual number of sequences submitted has largely paralleled that of the
bacterial census with a plateau starting in 2009 and an average of 7,075
sequences each year since then. The number of new archaeal OTUs
represented by these sequences began to slow in 2005 with an average of
355.5 new OTUs per year. With the exception of 2012 and 2014, the number
of submissions responsible for more than 50\% of the archaeal sequences
submitted per year has varied between 2 and 11 submissions per year. The
clear bias towards sequencing bacterial 16S rRNA genes has limited the
ability to more fully characterize the biodiversity of the archaea,
which is clearly reflected in the relatively meager sampling effort
across habitats, compared to bacteria (Figure 1D),

\textbf{\emph{The ability to sample archaea and bacteria is
taxonomically skewed.}} The Firmicutes, Proteobacteria, Actinobacteria,
and Bacteroidetes represent 89.2\% of the bacterial sequences and the
Euryarchaeota and Thaumarchaeota 86.5\% of the archaeal sequences. We
sought to understand how the representation of individual phyla has
changed relative to the state of the census in 2006. We used 2006 as a
reference point for calibrating the dynamics of the bacterial and
archaeal censuses since that was the year that the first highly
parallelized 16S rRNA gene sequence dataset was published (20). Based on
the representation of sequences within the SILVA database, in 2006 there
were 61 bacterial and 18 archaeal phyla. Since then there have been 4
new bacterial (CKC4, OC31, S2R-29, and SBYG-2791) and 2 new archaeal
candidate phyla (Ancient Archaeal Group and TVG8AR30). Relative to the
overall sequencing trends before and after 2006, several phyla stand out
for being over and underrepresented in sequence submissions (Figure 3).
Among the bacterial phyla with at least 1,000 sequences, Atribacteria
and Kazan-3B-09 were sequenced 4-fold more often while
Deinococcus-Thermus and Tenericutes were sequenced 2-fold less often
than would have been expected since 2006. Among the archaeal phyla with
at least 1,000 sequences, the Thaumarchaeota were sequenced 2.0-fold
more often and the Crenarchaeota were sequenced 6.7-fold less often than
expected. Together, these results demonstrate a change in the
phylum-level lineages represented in the census from before and after
2006 and encouragingly, show that some underrepresented phyla are
becoming better sampled.

\textbf{\emph{Focusing the census by environment.}} We were able to
assign 89.3 and 95.1\% of the sequences to one of seven broad
environmental categories based on the metadata that accompanied the
SILVA database (Tables 1). Across these broad categories there was wide
variation in the number of sequences that have been sampled. Among
bacterial sequences, the three best represented groups were from
zoonotic (N=804,585), aquatic (N=214,085), and built environment
(N=108,799) sources. Among the archaeal sequences the three best
represented groups were the same, but ordered differently: aquatic
(N=34,400), built environment (N=7,286), and zoonotic (N=5,597) (Figure
1C,D)). For both domains, soil samples were the fourth most represented
category (bacteria: 74,870; archaea: 2,517). The orders of these
categories was surprising considering soil and aquatic environments
harbor the most microbial biomass and biodiversity (21). In spite of
wide variation in sequencing depth and coverage (Table 1), the
interquartile range across the fine-level categories for the bacterial
OTU coverage only varied between 34.5 to 40.0 (median coverage=36.7\%).
The interquartile range in the OTU coverage by environment for the
archaeal data was 41.5 to 53.1 (median coverage=44.9\%). The archaeal
coverage was higher than that of the bacterial OTU coverage for all
categories except the food-associated, plant surface, and other
invertebrate categories. Across all categories, the bacterial and
archaeal sequencing data represented a limited number of phyla (Figure
4). Among the bacterial data, the fine-scale categories were dominated
by Proteobacteria (N=24), Firmicutes (N=2), and Actinobacteria (N=1) and
among the archaeal data, they were dominated by Euryarchaeota (N=16),
Thaumarchaeota (N=10), and Aenigmarchaeota (N=1). Regardless, there were
clear phylum-level signatures that differentiated the various
categories. Within each of the bacterial and archaeal phyla, there was
considerable variation in the relative abundance of each across the
categories confirming that taxonomic signatures exist to differentiate
different environments even at a broad taxonomic level.

\textbf{\emph{The cultured census.}} In the 2004 bacterial census, there
was concern expressed that although culture-independent methods were
significantly enhancing our knowledge of microbial life, there were
numerous bacterial phyla with no or only a few cultured representatives.
To update this assessment, we identified those sequences that came from
cultured and uncultured organisms. Overall, 18.9\% of bacterial
sequences and 6.8\% of archaeal sequences have come from isolated
organisms. Comparing the fraction of sequences deposited during and
before 2006 from isolates to those collected after 2006, we found that
culturing rates lag by 2.4 and 2.5-fold for bacteria and archaea,
respectively. Among the 65 bacterial phyla, 24 have no cultured
representatives and 14 of the 20 archaeal phyla have no cultured
representatives. This lag is likely due to the differences in throughput
of culture-dependent and -independent approaches. Of the phyla with at
least one cultured representative, the median percentage of sequences
coming from a culture was only 2.8\% for the bacterial phyla and 1.7\%
for the archaeal phyla (Figure 5). Even though many phyla have cultured
representatives, there is still a skew in the representation of most
phyla found in cultivation efforts.

Considering the possibility that large culture-independent sequencing
efforts may only be re-sequencing organisms that already exist in
culture, we asked what percentage of OTUs had at least one cultured
representative. We found that 16.9\% of the 117,385 bacterial OTUs and
13.1\% of the 4,574 archeael OTUs had at least one cultured
representative (Figure 5). Comparing the percentage of sequences with
cultured representatives to the percentage of OTUs containing a sequence
from a cultured representative revealed a strong cultivation bias within
the Firmicutes, which had a higher percentage of sequences generated by
cultivated representatives than would be expected based on the number of
cultured organisms represented by OTUs (Figure 5). This likely reflects
the extremely high number of cultivated biomedically relevant cultivars
from genera such as \emph{Bacillus}, \emph{Streptococcus},
\emph{Lactobacillus}, \emph{Staphylococcus}, and others. Conversely,
many phyla, including Cyanobacteria, Actinobacteria, Bacteroidetes, and
Nitrospirae, had a lower percentage of sequences belonging to cultivated
representatives than would be expected based on the percentage of OTUs
that have sequences from cultured organisms, indicating that the
cultivation efforts in these clades are relatively inefficient with
regards to available diversity. Nevertheless, it is clear that the
majority of OTUs from any phylum remain uncultivated, to say nothing of
the diversity of organisms that may be encapsulated within the 97\%
sequence identity cutoff.

\textbf{\emph{New technologies to access novel biodiversity.}} Given the
shift from Sanger sequencing to platforms that offer higher throughput
but shorter reads, we are concerned that our ability to harvest
full-length sequences from communities will remain stalled. Several
culture-independent methods have been developed that offer the ability
to obtain full-length sequences of the 16S rRNA gene and even the
complete genome. These have included single cell genomics (22) and
assembly of short 16S rRNA gene fragments using data generated from PCR
amplicons or metagenomic shotgun sequence data with the
Expectation-Maximization Iterative Reconstruction of Genes from the
Environment (EMIRGE) algorithm (23, 24). To test the ability of these
technologies to expand our knowledge of microbial diversity beyond that
of traditional approaches, we compared the overlap of OTUs found using
each of the new methods with the traditional approaches (Figure 6).
Utilizing the 16S rRNA gene sequences extracted from the single-cell
genomes available on the Integrated Microbial Genomes (IMG) system (25),
we identified 311 bacterial and 70 archaeal sequences, which were
assigned to 115 and 27 bacterial and archaeal OTUs, respectively.
Interestingly, only 8.7 and 3.7\% of the bacterial and archaeal single
celled OTUs, respectively, had not been observed by previous efforts.
Next, we identified six studies that utilized EMIRGE to assemble 16S
rRNA gene sequences from metagenomic sequences (23, 26--30). Together
these studies assembled 599 bacterial and 9 archaeal full-length
sequences, which were assigned to 335 and 7 bacterial and archaeal OTUs,
respectively. Only 40.6 and 60.3\% of the bacterial OTUs generated by
this approach were previously identified by this traditional cultivation
and PCR-based approaches, respectively. Although the application of this
approach to Archaea has been limited, it was still surprising that 85.7
and 85.7\% of the archaeal OTUs had been previously recovered by
traditional cultivation and PCR-based approaches, respectively. Finally,
we pooled 76,080 bacterial sequences from five studies that utilized
EMIRGE to assemble 16S rRNA gene sequences from fragmented amplicons
(24, 31--34). These sequences were assigned to 40,213 OTUs. We were
surprised that only 7.6\% of these OTUs were previously found by a more
traditional approach. Although these PCR-based EMIRGE results may be
valid, the high degree of novelty that was observed suggests that the
error of the assembled reads may be too high for generating reference
sequences. Each of these methods represent promising opportunities to
continue the bacterial census using full-length sequences as well as
genomic information.

\subsection{Conclusions}\label{conclusions}

It is clear that considerable biodiversity has been discovered since the
first census in 2004. However, much of it has been biased towards
particular phyla and environments. Our analysis suggests that 94.5\% of
new full-length bacterial and archaeal sequences are likely to have
already been seen. Meanwhile, 29.2\% of bacterial and 38.5\% of archaeal
OTUs have only been observed once. In spite of current estimates
suggesting the global bacterial species richness may be as high as
10\textsuperscript{12} species (35), the current census based on
full-length 16S rRNA gene sequences suggests that existing sampling
methods will prevent us from acquiring full-length sequences for that
level of diversity. As we have shown, current strategies repeatedly
sample the same OTUs and do a poor job of resampling rarer populations.
Given this low level of OTU coverage, it is likely that there are many
more bacterial and archaeal populations yet to be sampled.

There are several additional reasons to suspect that the current census
should be considered conservative. First, we found that most sequences
recently deposited into public databases are being made by a small
number of projects that have deeply sampled similar environments, and
the number of full-length reads deposited into the databases has
stalled. Second, it is widely acknowledged that 16S rRNA gene primers
are biased; these biases are amplified when designing primers to amplify
subregions used in sequencing short reads (36). Assembly of metagenomic
data has shown the presence of introns in the 16S rRNA genes of
organisms within the so-called ``Candidate Phyla Radiation''
(e.g.~Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58),
WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.) that would
preclude detection with standard PCR-based approaches (37, 38). Third,
the willingness of researchers to contribute their sequences and the
metadata describing the environment that the sequences were sampled from
is critical for assessing the progress of the census and to accrue the
benefits from having full-length sequences in the databases. As an
illustration of this problem, only 5 of the 11 studies that used the
EMIRGE algorithm deposited their sequences in GenBank. This makes the
sequences from the other studies effectively invisible to the search
algorithms used by 16S rRNA gene-specific databases to harvest
sequences. As assembly and long read technologies advance, a mechanism
is needed to assess the quality of the consensus sequences and to make
them easily accessible to the 16S rRNA gene-specific databases.

Efforts to census archaea and bacteria using short read technology such
as the International Census of Marine Microbes, the Earth Microbiome
Project, and the Human Microbiome Project have significantly advanced
our knowledge of archaeal and bacterial biogeography; however, these
analyses have demonstrated the limitations of databases and taxonomies
that are based on sequences from common and abundant organisms. During
the period prior to the introduction of massively parallelized high
throughput sequencing, it was common for a study to generate dozens or
hundreds of sequences per sample. The existing databases that are used
for classifying sequences are based on those sequences, which represent
organisms that are generally abundant. We hypothesize that recent
difficulties obtaining adequate classification for short sequences
captured from more rare organisms are because our databases do not
contain full-length references for those sequences. We fear that these
trends will worsen unless researchers can leverage new sequencing and
cultivation technologies to generate large numbers of full-length
sequences from a large number of diverse samples.

Novel technologies such as single-cell genomics, metagenomics, and
algorithms to recover full-length sequences from new sequencing
platforms have demonstrated promise in circumventing previous
limitations in identifying new OTUs. Using EMIRGE to assemble fragmented
16S rRNA gene amplicons may allow us to obtain deep coverage of
communities; however, it is still unclear how faithful the assembled
sequence is to that of the original organism. Additional sequencing
technologies also offer the ability to directly generate full-length
sequences, such as PacBio and potentially Oxford Nanopore. Initial
application of PacBio to sequencing full-length fragments suggests that
the sequences suffered from a high error rate (39). To obtain a more
direct investigation of rare organisms, microbiologists are developing
novel cultivation and single cell genomics techniques (\textbf{???},
40--42). The ability to enrich or select for specific populations using
these approaches could limit the need for redundant brute force
sequencing. These approaches are still in active development, and we
hope that through continuous refinement, they may allow us to
significantly improve the coverage of OTUs in public databases.

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{\emph{Sequence data curation.}} The July 19, 2015 release of the
ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref
v.123) was downloaded from
\url{http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz}
(43). This release is based on the EMBL-EBI/ENA Release 123, which was
released in March 2015. The SILVA curators identify potential SSU
sequences using keyword searches and sequence-based search using RNAmmer
(\url{http://www.arb-silva.de/documentation/release-123/}). The SILVA
curators then screened the 7,168,241 resulting sequences based on a
minimum length criteria (\textless{}300 nt), number of ambiguous base
calls (\textgreater{}2\%), length of sequence homopolymers
(\textgreater{}2\%), presence of vector contamination
(\textgreater{}2\%), low alignment quality value (\textless{}75), and
likelihood of being chimeric (Pintail value \textless{} 50). Of the
remaining sequences, the bacterial reference set retained those
bacterial sequences longer than 1,200 nucleotides and the archaeal
reference set retained those archaeal sequences longer than 900
nucleotides. The aligned 1,515,024 bacterial and 59,240 archaeal
sequences were exported from the database using ARB along with the
complete set of metadata. Additional sequence data was included from
single-cell genomes available on the Integrated Microbial Genomes (IMG)
system (25), many of which were recently obtained via the GEBA-MDM
effort in Rinke et al. (22). ``SCGC'' was searched on the IMG database
March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S
rRNA gene sequences and their associated metadata. Further, sequences
generated from amplicon and shotgun metagenomic data using the EMIRGE
program were also included (23, 24). The IMG and EMIRGE sequences were
aligned against the respective SILVA-based reference using mothur (44).
The aligned bacterial and archaeal sequence sets were pooled and
processed in parallel. Using mothur, sequences were further screened to
remove any sequence with more than 2 ambiguous base calls and trimmed to
overlap the same alignment coordinates. The sequences in the resulting
bacterial dataset overlapped bases 113 through 1350 of an \emph{E. coli}
reference sequence (V00348) and had a median length of 1,233 nt. The
sequences in the resulting archaeal dataset overlapped positions 362 to
937 of a \emph{Sulfolobus solfataricus} reference sequence (X03235) and
had a median length of 580 nt. The archaeal sequences were considerably
shorter than their initial length because it was necessary to find a
common overlapping region across the sequences. The final datasets
contained 1,411,234 bacterial and 53,546 archaeal 16S rRNA gene
sequences. Sequences were assigned to OTUs using the average neighbor
clustering algorithm (45).

\textbf{\emph{Metadata curation.}} The metadata that was contained
within the SSU Ref database was used to expand our analysis beyond a
basic count of sequences and the number of OTUs in each domain. The
environmental origins of the 16S rRNA gene sequences were manually
classified using seven broad ``coarse'' categories, and further refined
to facilitate additional analyses with twenty-six more specific ``fine''
categories (Table S1). These were assigned based on manual curation of
the ``isolation\_source'' category within the ARB database associated
with each of the sequences. For source definitions that were not
identifiable by online searches, educated guesses were made or they were
placed into the coarse ``Other'' category. There were 151,669 bacterial
and 2,565 archaeal sequences where an ``isolation\_source'' term was not
collected. We ascertained whether a sequence came from a cultured
organism by including those sequences that had data in their ``strain''
or ``isolate'' fields within the database and excluded any sequences
that had ``Unc'' as part of their database name as this is a convention
in the database that represents sequences from uncultured organisms.
Complete tables containing the ARB-provided metadata, taxonomic
information, OTU assignment, and our environmental categorizations are
available at FigShare for the bacterial
(\url{https://dx.doi.org/10.6084/m9.figshare.2064927}) and archaeal
(\url{https://dx.doi.org/10.6084/m9.figshare.2064942}) data.

\textbf{\emph{Calculating coverage.}} Sequencing coverage
(C\textsubscript{Sequence}) was quantified by two methods. The first was
to use Good's coverage according to

\[C_{Sequence} = 1 - \frac{n_1}{N_t}\]

where n\textsubscript{1} is the number of OTUs represented by only one
sequence and N\textsubscript{t} is the total number of sequences (46).
Although Good's coverage provides information about the success of the
sequencing effort in sampling the most abundant organisms in a
community, it does not directly provide information about the success of
the sequencing effort in recovering previously unobserved OTUs. To
quantify the ability of sequencing to identifying novel OTUs or, in
other words, to quantify the ``distance'' in the peak of the rarefaction
curves to their hypothetical asymptote, we defined ``OTU coverage''
(C\textsubscript{OTU}) as

\[C_{OTU} = 1 - \frac{n_1}{S_t}\]

where S\textsubscript{t} is the total number of OTUs. Whereas Good's
coverage estimates the probability that a new sequence will have already
been seen, OTU coverage estimates the probability that a new OTU will
match an existing one. It is therefore an extension of Good's coverage
in that it quantifies the probability that, for any given set of
sequences clustered into an OTU, that OTU will have already been seen.
Thus, high Good's coverage means that any new sequence is unlikely to be
novel, and high OTU coverage means that any new OTU is unlikely to be
novel.

\textbf{\emph{Data analysis.}} Our analysis made use of ARB (OS X v.6.0)
(43), mothur (v.1.37.0) (44), and R (v.3.2.3) (47). Within R we utilized
the knitr (v.1.12.3), wesanderson (v.0.3.3.99), and openxlsx (v. 2.4.0)
packages. A reproducible version of this manuscript including data
extraction and processing is available at
\url{https://www.github.com/SchlossLab/Schloss_Census2_mBio_2016}.

\newpage

\doublespacing

\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
16S rRNA gene sequences for different OTU definitions and level of
sequencing effort.} Rarefaction curves for different OTU definitions of
Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
environments in Table 1 for Bacteria (C) and Archaea (D).

\textbf{Figure 2. Progression of the archaeal and bacterial census since
the first full-length 16S rRNA gene sequence was deposited into GenBank
in 1983.}* The number of bacterial and archaeal 16S rRNA gene sequences
deposited (A) and the new OTUs they represent (B) has increased
exponentially until the last several years when the rate of change has
plateaued. For both bacterial and archaeal sequences, the number of
studies that are responsible for depositing more than 50\% of the
sequences each year has been relatively small (C).

\textbf{Figure 3. Relative rate of sequence deposition for each
bacterial and archaeal phylum before and after 2006 relative to the
sequencing of all bacteria.} The figure shows the relative rates for
those phyla with at least 1,000 sequences and the x-axis is on a log2
scale. The data for all bacterial and archaeal phyla are available in
Supplemental Tables 2 and 3, respectively.

\textbf{Figure 4. Heatmap depicting the relative abundance of the most
common bacterial and archaeal phyla across different environments.} Each
environmental category exhibited a phylum-level signature although the
bacterial census was dominated by sequences from the Firmicutes,
Proteobacteria, Actinobacteria, and Bacteroidetes and the archaeal
census was dominated by sequences from the Euryarchaeota and
Thaumarchaeota. The ten most abundant phyla across all environmental
categories are shown. The data for all bacterial and archaeal phyla are
available in Supplemental Tables 4 and 5, respectively.

\textbf{Figure 5. The rate that sequences and OTUs are generated from
bacterial and archaeal cultures relative to all sequences and OTUs by
phlum.} Phyla with greater than 1,000 sequences are listed by domain.
Open circles indicate the percentage of sequences in the database that
match cultured organisms. Closed circles indicate the percentage of OTUs
in this analysis that contain sequences belonging to a cultured
organism. The data for all bacterial and archaeal phyla are available in
Supplemental Tables 6 and 7, respectively.

\textbf{Figure 6. The percentage of bacterial and archaeal OTUs found by
single cell genomics and EMIRGE using PCR or metagenomics that were also
detected by other methods.} The bars comparing a method to itself
indicate the percentage of OTUs that were only detected by that method.

\newpage

\textbf{Supplemental Table 1. Description of environmental categories
and the criteria used to assign sequences to each category.}

\textbf{Supplementary Table 2. Frequency that each bacterial phylum was
sequenced before and after 2006.}

\textbf{Supplementary Table 3. Frequency that each archaeal phylum was
sequenced before and after 2006.}

\textbf{Supplementary Table 4. Frequency that each bacterial phylum was
found across each of the environmental categories.}

\textbf{Supplementary Table 5. Frequency that each archaeal phylum was
found across each of the environmental categories.}

\textbf{Supplementary Table 6. Frequency that each bacterial sequence or
OTU was retrieved by cultivation or by culture-independent methods.}

\textbf{Supplementary Table 7. Frequency that each archaeal sequence or
OTU was retrieved by cultivation or by culture-independent methods.}

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-McGill2007}{}
1. \textbf{McGill BJ}, \textbf{Etienne RS}, \textbf{Gray JS},
\textbf{Alonso D}, \textbf{Anderson MJ}, \textbf{Benecha HK},
\textbf{Dornelas M}, \textbf{Enquist BJ}, \textbf{Green JL}, \textbf{He
F}, \textbf{Hurlbert AH}, \textbf{Magurran AE}, \textbf{Marquet PA},
\textbf{Maurer BA}, \textbf{Ostling A}, \textbf{Soykan CU},
\textbf{Ugland KI}, \textbf{White EP}. 2007. Species abundance
distributions: Moving beyond single prediction theories to integration
within an ecological framework. Ecology Letters \textbf{10}:995--1015.
doi:\href{https://doi.org/10.1111/j.1461-0248.2007.01094.x}{10.1111/j.1461-0248.2007.01094.x}.

\hypertarget{ref-Hubbell2001}{}
2. \textbf{Hubbell SP}. 2001. A Unified Theory of Biodiversity and
Biogeography. Princeton University Press, Princeton.

\hypertarget{ref-Konstantinidis2006}{}
3. \textbf{Konstantinidis KT}, \textbf{Ramette A}, \textbf{Tiedje JM}.
2006. The bacterial species definition in the genomic era. Philosophical
Transactions of the Royal Society B: Biological Sciences
\textbf{361}:1929--1940.
doi:\href{https://doi.org/10.1098/rstb.2006.1920}{10.1098/rstb.2006.1920}.

\hypertarget{ref-Oren2013}{}
4. \textbf{Oren A}, \textbf{Garrity GM}. 2013. Then and now: A
systematic review of the systematics of prokaryotes in the last
80~years. Antonie van Leeuwenhoek \textbf{106}:43--56.
doi:\href{https://doi.org/10.1007/s10482-013-0084-1}{10.1007/s10482-013-0084-1}.

\hypertarget{ref-Brosius1978}{}
5. \textbf{Brosius J}, \textbf{Palmer ML}, \textbf{Kennedy PJ},
\textbf{Noller HF}. 1978. Complete nucleotide sequence of a 16S
ribosomal RNA gene from \emph{Escherichia coli.} Proceedings of the
National Academy of Sciences \textbf{75}:4801--4805.
doi:\href{https://doi.org/10.1073/pnas.75.10.4801}{10.1073/pnas.75.10.4801}.

\hypertarget{ref-Schloss2004}{}
6. \textbf{Schloss PD}, \textbf{Handelsman J}. 2004. Status of the
microbial census. Microbiology and Molecular Biology Reviews
\textbf{68}:686--691.
doi:\href{https://doi.org/10.1128/mmbr.68.4.686-691.2004}{10.1128/mmbr.68.4.686-691.2004}.

\hypertarget{ref-Dykhuizen1998}{}
7. \textbf{Dykhuizen DE}. 1998. Santa Rosalia revisited: Why are there
so many species of bacteria? Antonie van Leeuwenhoek \textbf{73}:25--33.
doi:\href{https://doi.org/10.1023/a:1000665216662}{10.1023/a:1000665216662}.

\hypertarget{ref-Curtis2002}{}
8. \textbf{Curtis TP}, \textbf{Sloan WT}, \textbf{Scannell JW}. 2002.
Estimating prokaryotic diversity and its limits. Proceedings of the
National Academy of Sciences \textbf{99}:10494--10499.
doi:\href{https://doi.org/10.1073/pnas.142680199}{10.1073/pnas.142680199}.

\hypertarget{ref-HMP2012}{}
9. \textbf{The Human Microbiome Consortium}. 2012. Structure, function
and diversity of the healthy human microbiome. Nature
\textbf{486}:207--214.
doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\hypertarget{ref-Gilbert2014}{}
10. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2014.
The Earth Microbiome Project: Successes and aspirations. BMC Biology
\textbf{12}:69.
doi:\href{https://doi.org/10.1186/s12915-014-0069-1}{10.1186/s12915-014-0069-1}.

\hypertarget{ref-AmaralZettler2010}{}
11. \textbf{Amaral-Zettler L}, \textbf{Artigas LF}, \textbf{Baross J},
\textbf{P.A. LB}, \textbf{Boetius A}, \textbf{Chandramohan D},
\textbf{Herndl G}, \textbf{Kogure K}, \textbf{Neal P},
\textbf{Pedrós-Alió C}, \textbf{Ramette A}, \textbf{Schouten S},
\textbf{Stal L}, \textbf{Thessen A}, \textbf{Leeuw J de}, \textbf{Sogin
M}. 2010. A global census of marine microbes, pp. 221--245. \emph{In}
Life in the worlds oceans. Wiley-Blackwell.

\hypertarget{ref-Youssef2009}{}
12. \textbf{Youssef N}, \textbf{Sheik CS}, \textbf{Krumholz LR},
\textbf{Najar FZ}, \textbf{Roe BA}, \textbf{Elshahed MS}. 2009.
Comparison of species richness estimates obtained using nearly complete
fragments and simulated pyrosequencing-generated fragments in 16S rRNA
gene-based environmental surveys. Applied and Environmental Microbiology
\textbf{75}:5227--5236.
doi:\href{https://doi.org/10.1128/aem.00592-09}{10.1128/aem.00592-09}.

\hypertarget{ref-Schloss2010}{}
13. \textbf{Schloss PD}. 2010. The effects of alignment quality,
distance calculation method, sequence filtering, and region on the
analysis of 16S rRNA gene-based studies. PLoS Comput Biol
\textbf{6}:e1000844.
doi:\href{https://doi.org/10.1371/journal.pcbi.1000844}{10.1371/journal.pcbi.1000844}.

\hypertarget{ref-Alivisatos2015}{}
14. \textbf{Alivisatos AP}, \textbf{Blaser MJ}, \textbf{Brodie EL},
\textbf{Chun M}, \textbf{Dangl JL}, \textbf{Donohue TJ},
\textbf{Dorrestein PC}, \textbf{Gilbert JA}, \textbf{Green JL},
\textbf{Jansson JK}, \textbf{Knight R}, \textbf{Maxon ME},
\textbf{McFall-Ngai MJ}, \textbf{Miller JF}, \textbf{Pollard KS},
\textbf{Ruby EG}, \textbf{Taha SA}. 2015. A unified initiative to
harness Earth's microbiomes. Science \textbf{350}:507--508.
doi:\href{https://doi.org/10.1126/science.aac8480}{10.1126/science.aac8480}.

\hypertarget{ref-Liux5f2012}{}
15. \textbf{Li E}, \textbf{Hamm CM}, \textbf{Gulati AS}, \textbf{Sartor
RB}, \textbf{Chen H}, \textbf{Wu X}, \textbf{Zhang T}, \textbf{Rohlf
FJ}, \textbf{Zhu W}, \textbf{Gu C}, \textbf{Robertson CE}, \textbf{Pace
NR}, \textbf{Boedeker EC}, \textbf{Harpaz N}, \textbf{Yuan J},
\textbf{Weinstock GM}, \textbf{Sodergren E}, \textbf{Frank DN}. 2012.
Inflammatory bowel diseases phenotype, textitC. difficile and NOD2
genotype are associated with shifts in human ileum associated microbial
composition. PLoS ONE \textbf{7}:e26284.
doi:\href{https://doi.org/10.1371/journal.pone.0026284}{10.1371/journal.pone.0026284}.

\hypertarget{ref-Kongux5f2012}{}
16. \textbf{Kong HH}, \textbf{Oh J}, \textbf{Deming C}, \textbf{Conlan
S}, \textbf{Grice EA}, \textbf{Beatson MA}, \textbf{Nomicos E},
\textbf{Polley EC}, \textbf{Komarow HD}, \textbf{Murray PR},
\textbf{Turner ML}, \textbf{Segre JA}. 2012. Temporal shifts in the skin
microbiome associated with disease flares and treatment in children with
atopic dermatitis. Genome Research \textbf{22}:850--859.
doi:\href{https://doi.org/10.1101/gr.131029.111}{10.1101/gr.131029.111}.

\hypertarget{ref-Griceux5f2009}{}
17. \textbf{Grice EA}, \textbf{Kong HH}, \textbf{Conlan S},
\textbf{Deming CB}, \textbf{Davis J}, \textbf{Young AC},
\textbf{Bouffard GG}, \textbf{Blakesley RW}, \textbf{Murray PR},
\textbf{Green ED}, \textbf{Turner ML}, \textbf{Segre JA}. 2009.
Topographical and temporal diversity of the human skin microbiome.
Science \textbf{324}:1190--1192.
doi:\href{https://doi.org/10.1126/science.1171700}{10.1126/science.1171700}.

\hypertarget{ref-Griceux5f2010}{}
18. \textbf{Grice EA}, \textbf{Snitkin ES}, \textbf{Yockey LJ},
\textbf{Bermudez DM}, \textbf{Liechty KW}, \textbf{Segre JA},
\textbf{Mullikin J}, \textbf{Blakesley R}, \textbf{Young A}, \textbf{Chu
G}, \textbf{Ramsahoye C}, \textbf{Lovett S}, \textbf{Han J},
\textbf{Legaspi R}, \textbf{Fuksenko T}, \textbf{Reddix-Dugue N},
\textbf{Sison C}, \textbf{Gregory M}, \textbf{Montemayor C},
\textbf{Gestole M}, \textbf{Hargrove A}, \textbf{Johnson T},
\textbf{Myrick J}, \textbf{Riebow N}, \textbf{Schmidt B},
\textbf{Novotny B}, \textbf{Gupti J}, \textbf{Benjamin B},
\textbf{Brooks S}, \textbf{Coleman H}, \textbf{Ho S-l},
\textbf{Schandler K}, \textbf{Smith L}, \textbf{Stantripop M},
\textbf{Maduro Q}, \textbf{Bouffard G}, \textbf{Dekhtyar M},
\textbf{Guan X}, \textbf{Masiello C}, \textbf{Maskeri B},
\textbf{McDowell J}, \textbf{Park M}, \textbf{Thomas PJ}. 2010.
Longitudinal shift in diabetic wound microbiota correlates with
prolonged skin defense response. Proceedings of the National Academy of
Sciences \textbf{107}:14799--14804.
doi:\href{https://doi.org/10.1073/pnas.1004204107}{10.1073/pnas.1004204107}.

\hypertarget{ref-Kirkux5fHarrisux5f2012}{}
19. \textbf{Harris JK}, \textbf{Caporaso JG}, \textbf{Walker JJ},
\textbf{Spear JR}, \textbf{Gold NJ}, \textbf{Robertson CE},
\textbf{Hugenholtz P}, \textbf{Goodrich J}, \textbf{McDonald D},
\textbf{Knights D}, \textbf{Marshall P}, \textbf{Tufo H}, \textbf{Knight
R}, \textbf{Pace NR}. 2012. Phylogenetic stratigraphy in the guerrero
negro hypersaline microbial mat. The ISME Journal \textbf{7}:50--60.
doi:\href{https://doi.org/10.1038/ismej.2012.79}{10.1038/ismej.2012.79}.

\hypertarget{ref-Sogin2006}{}
20. \textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber JA},
\textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR}, \textbf{Arrieta
JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in the deep sea and
the underexplored ``rare biosphere''. Proceedings of the National
Academy of Sciences \textbf{103}:12115--12120.
doi:\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.

\hypertarget{ref-Whitman1998}{}
21. \textbf{Whitman WB}, \textbf{Coleman DC}, \textbf{Wiebe WJ}. 1998.
Prokaryotes: The unseen majority. Proceedings of the National Academy of
Sciences \textbf{95}:6578--6583.
doi:\href{https://doi.org/10.1073/pnas.95.12.6578}{10.1073/pnas.95.12.6578}.

\hypertarget{ref-Rinke2013}{}
22. \textbf{Rinke C}, \textbf{Schwientek P}, \textbf{Sczyrba A},
\textbf{Ivanova NN}, \textbf{Anderson IJ}, \textbf{Cheng J-F},
\textbf{Darling A}, \textbf{Malfatti S}, \textbf{Swan BK}, \textbf{Gies
EA}, \textbf{Dodsworth JA}, \textbf{Hedlund BP}, \textbf{Tsiamis G},
\textbf{Sievert SM}, \textbf{Liu W-T}, \textbf{Eisen JA}, \textbf{Hallam
SJ}, \textbf{Kyrpides NC}, \textbf{Stepanauskas R}, \textbf{Rubin EM},
\textbf{Hugenholtz P}, \textbf{Woyke T}. 2013. Insights into the
phylogeny and coding potential of microbial dark matter. Nature
\textbf{499}:431--437.
doi:\href{https://doi.org/10.1038/nature12352}{10.1038/nature12352}.

\hypertarget{ref-Miller2011}{}
23. \textbf{Miller CS}, \textbf{Baker BJ}, \textbf{Thomas BC},
\textbf{Singer SW}, \textbf{Banfield JF}. 2011. EMIRGE: Reconstruction
of full-length ribosomal genes from microbial community short read
sequencing data. Genome Biol \textbf{12}:R44.
doi:\href{https://doi.org/10.1186/gb-2011-12-5-r44}{10.1186/gb-2011-12-5-r44}.

\hypertarget{ref-Miller2013}{}
24. \textbf{Miller CS}, \textbf{Handley KM}, \textbf{Wrighton KC},
\textbf{Frischkorn KR}, \textbf{Thomas BC}, \textbf{Banfield JF}. 2013.
Short-read assembly of full-length 16S amplicons reveals bacterial
diversity in subsurface sediments. PLoS ONE \textbf{8}:e56018.
doi:\href{https://doi.org/10.1371/journal.pone.0056018}{10.1371/journal.pone.0056018}.

\hypertarget{ref-Markowitz2013}{}
25. \textbf{Markowitz VM}, \textbf{Chen I-MA}, \textbf{Palaniappan K},
\textbf{Chu K}, \textbf{Szeto E}, \textbf{Pillay M}, \textbf{Ratner A},
\textbf{Huang J}, \textbf{Woyke T}, \textbf{Huntemann M},
\textbf{Anderson I}, \textbf{Billis K}, \textbf{Varghese N},
\textbf{Mavromatis K}, \textbf{Pati A}, \textbf{Ivanova NN},
\textbf{Kyrpides NC}. 2013. IMG 4 version of the integrated microbial
genomes comparative analysis system. Nucleic Acids Research
\textbf{42}:D560--D567.
doi:\href{https://doi.org/10.1093/nar/gkt963}{10.1093/nar/gkt963}.

\hypertarget{ref-Wrighton2012}{}
26. \textbf{Wrighton KC}, \textbf{Thomas BC}, \textbf{Sharon I},
\textbf{Miller CS}, \textbf{Castelle CJ}, \textbf{VerBerkmoes NC},
\textbf{Wilkins MJ}, \textbf{Hettich RL}, \textbf{Lipton MS},
\textbf{Williams KH}, \textbf{Long PE}, \textbf{Banfield JF}. 2012.
Fermentation, hydrogen, and sulfur metabolism in multiple uncultivated
bacterial phyla. Science \textbf{337}:1661--1665.
doi:\href{https://doi.org/10.1126/science.1224041}{10.1126/science.1224041}.

\hypertarget{ref-NunesdaRocha2015}{}
27. \textbf{Rocha UN da}, \textbf{Cadillo-Quiroz H}, \textbf{Karaoz U},
\textbf{Rajeev L}, \textbf{Klitgord N}, \textbf{Dunn S}, \textbf{Truong
V}, \textbf{Buenrostro M}, \textbf{Bowen BP}, \textbf{Garcia-Pichel F},
\textbf{Mukhopadhyay A}, \textbf{Northen TR}, \textbf{Brodie EL}. 2015.
Isolation of a significant fraction of non-phototroph diversity from a
desert biological soil crust. Front Microbiol \textbf{6}:277.
doi:\href{https://doi.org/10.3389/fmicb.2015.00277}{10.3389/fmicb.2015.00277}.

\hypertarget{ref-Hamilton2015}{}
28. \textbf{Hamilton TL}, \textbf{Jones DS}, \textbf{Schaperdoth I},
\textbf{Macalady JL}. 2015. Metagenomic insights into S(0) precipitation
in a terrestrial subsurface lithoautotrophic ecosystem. Front Microbiol
\textbf{5}:756.
doi:\href{https://doi.org/10.3389/fmicb.2014.00756}{10.3389/fmicb.2014.00756}.

\hypertarget{ref-Handley2012}{}
29. \textbf{Handley KM}, \textbf{VerBerkmoes NC}, \textbf{Steefel CI},
\textbf{Williams KH}, \textbf{Sharon I}, \textbf{Miller CS},
\textbf{Frischkorn KR}, \textbf{Chourey K}, \textbf{Thomas BC},
\textbf{Shah MB}, \textbf{Long PE}, \textbf{Hettich RL},
\textbf{Banfield JF}. 2012. Biostimulation induces syntrophic
interactions that impact c, s and n cycling in a sediment microbial
community. The ISME Journal \textbf{7}:800--816.
doi:\href{https://doi.org/10.1038/ismej.2012.148}{10.1038/ismej.2012.148}.

\hypertarget{ref-Gladden2011}{}
30. \textbf{Gladden JM}, \textbf{Allgaier M}, \textbf{Miller CS},
\textbf{Hazen TC}, \textbf{VanderGheynst JS}, \textbf{Hugenholtz P},
\textbf{Simmons BA}, \textbf{Singer SW}. 2011. Glycoside hydrolase
activities of thermophilic bacterial consortia adapted to switchgrass.
Applied and Environmental Microbiology \textbf{77}:5804--5812.
doi:\href{https://doi.org/10.1128/aem.00032-11}{10.1128/aem.00032-11}.

\hypertarget{ref-Brooks2014}{}
31. \textbf{Brooks B}, \textbf{Firek BA}, \textbf{Miller CS},
\textbf{Sharon I}, \textbf{Thomas BC}, \textbf{Baker R},
\textbf{Morowitz MJ}, \textbf{Banfield JF}. 2014. Microbes in the
neonatal intensive care unit resemble those found in the gut of
premature infants. Microbiome \textbf{2}:1.
doi:\href{https://doi.org/10.1186/2049-2618-2-1}{10.1186/2049-2618-2-1}.

\hypertarget{ref-Wilkins2013}{}
32. \textbf{Wilkins MJ}, \textbf{Wrighton KC}, \textbf{Nicora CD},
\textbf{Williams KH}, \textbf{McCue LA}, \textbf{Handley KM},
\textbf{Miller CS}, \textbf{Giloteaux L}, \textbf{Montgomery AP},
\textbf{Lovley DR}, \textbf{Banfield JF}, \textbf{Long PE},
\textbf{Lipton MS}. 2013. Fluctuations in species-level protein
expression occur during element and nutrient cycling in the subsurface.
PLoS ONE \textbf{8}:e57819.
doi:\href{https://doi.org/10.1371/journal.pone.0057819}{10.1371/journal.pone.0057819}.

\hypertarget{ref-Handley2014}{}
33. \textbf{Handley KM}, \textbf{Wrighton KC}, \textbf{Miller CS},
\textbf{Wilkins MJ}, \textbf{Kantor RS}, \textbf{Thomas BC},
\textbf{Williams KH}, \textbf{Gilbert JA}, \textbf{Long PE},
\textbf{Banfield JF}. 2014. Disturbed subsurface microbial communities
follow equivalent trajectories despite different structural starting
points. Environ Microbiol \textbf{17}:622--636.
doi:\href{https://doi.org/10.1111/1462-2920.12467}{10.1111/1462-2920.12467}.

\hypertarget{ref-Alessi2014}{}
34. \textbf{Alessi DS}, \textbf{Lezama-Pacheco JS}, \textbf{Janot N},
\textbf{Suvorova EI}, \textbf{Cerrato JM}, \textbf{Giammar DE},
\textbf{Davis JA}, \textbf{Fox PM}, \textbf{Williams KH}, \textbf{Long
PE}, \textbf{Handley KM}, \textbf{Bernier-Latmani R}, \textbf{Bargar
JR}. 2014. Speciation and reactivity of uranium products formed during
in situ bioremediation in a shallow alluvial aquifer. Environmental
Science \& Technology \textbf{48}:12842--12850.
doi:\href{https://doi.org/10.1021/es502701u}{10.1021/es502701u}.

\hypertarget{ref-Locey2015}{}
35. \textbf{Locey KJ}, \textbf{Lennon JT}. 2015. Scaling laws predict
global microbial diversity. PeerJ PrePrints.
doi:\href{https://doi.org/10.7287/peerj.preprints.1451v1}{10.7287/peerj.preprints.1451v1}.

\hypertarget{ref-Parada2015}{}
36. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
Every base matters: Assessing small subunit rRNA primers for marine
microbiomes with mock communities, time series and global field samples.
Environ Microbiol n/a--n/a.
doi:\href{https://doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.

\hypertarget{ref-Brown2015}{}
37. \textbf{Brown CT}, \textbf{Hug LA}, \textbf{Thomas BC},
\textbf{Sharon I}, \textbf{Castelle CJ}, \textbf{Singh A},
\textbf{Wilkins MJ}, \textbf{Wrighton KC}, \textbf{Williams KH},
\textbf{Banfield JF}. 2015. Unusual biology across a group comprising
more than 15\% of domain bacteria. Nature \textbf{523}:208--211.
doi:\href{https://doi.org/10.1038/nature14486}{10.1038/nature14486}.

\hypertarget{ref-EloeFadrosh2016}{}
38. \textbf{Eloe-Fadrosh EA}, \textbf{Ivanova NN}, \textbf{Woyke T},
\textbf{Kyrpides NC}. 2016. Metagenomics uncovers gaps in amplicon-based
detection of microbial diversity. Nat Microbiol 15032.
doi:\href{https://doi.org/10.1038/nmicrobiol.2015.32}{10.1038/nmicrobiol.2015.32}.

\hypertarget{ref-Schloss2015}{}
39. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Jenior ML},
\textbf{Highlander SK}. 2015. Sequencing 16S rRNA gene fragments using
the pacBio sMRT dNA sequencing system. PeerJ PrePrints.
doi:\href{https://doi.org/10.7287/peerj.preprints.778v1}{10.7287/peerj.preprints.778v1}.

\hypertarget{ref-Nichols2010}{}
40. \textbf{Nichols D}, \textbf{Cahoon N}, \textbf{Trakhtenberg EM},
\textbf{Pham L}, \textbf{Mehta A}, \textbf{Belanger A}, \textbf{Kanigan
T}, \textbf{Lewis K}, \textbf{Epstein SS}. 2010. Use of iChip for
high-throughput in situ cultivation of ``uncultivable'' microbial
species. Applied and Environmental Microbiology \textbf{76}:2445--2450.
doi:\href{https://doi.org/10.1128/aem.01754-09}{10.1128/aem.01754-09}.

\hypertarget{ref-Buerger2012}{}
41. \textbf{Buerger S}, \textbf{Spoering A}, \textbf{Gavrish E},
\textbf{Leslin C}, \textbf{Ling L}, \textbf{Epstein SS}. 2012. Microbial
scout hypothesis, stochastic exit from dormancy, and the nature of slow
growers. Applied and Environmental Microbiology \textbf{78}:3221--3228.
doi:\href{https://doi.org/10.1128/aem.07307-11}{10.1128/aem.07307-11}.

\hypertarget{ref-Das2015}{}
42. \textbf{Das N}, \textbf{Tripathi N}, \textbf{Basu S}, \textbf{Bose
C}, \textbf{Maitra S}, \textbf{Khurana S}. 2015. Progress in the
development of gelling agents for improved culturability of
microorganisms. Front Microbiol \textbf{6}:698.
doi:\href{https://doi.org/10.3389/fmicb.2015.00698}{10.3389/fmicb.2015.00698}.

\hypertarget{ref-Pruesse2007}{}
43. \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Knittel K},
\textbf{Fuchs BM}, \textbf{Ludwig W}, \textbf{Peplies J},
\textbf{Glockner FO}. 2007. SILVA: A comprehensive online resource for
quality checked and aligned ribosomal RNA sequence data compatible with
ARB. Nucleic Acids Research \textbf{35}:7188--7196.
doi:\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.

\hypertarget{ref-Schloss2009}{}
44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Westcott2015}{}
45. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.

\hypertarget{ref-Good1953}{}
46. \textbf{Good IJ}. 1953. The population frequencies of species and
the estimation of population parameters. Biometrika
\textbf{40}:237--264.
doi:\href{https://doi.org/10.1093/biomet/40.3-4.237}{10.1093/biomet/40.3-4.237}.

\hypertarget{ref-language2015}{}
47. \textbf{R Core Team}. 2015. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\end{document}

### 2016-03-14 code/build_domain_rarefaction_plot.R

@@ -22,7 +22,7 @@ plot_domain_curves <- function(rarefaction){
 	box()
 }
 
-my_categories <- c(aerosol="Aerosol", aquatic="Aquatic", built="Built", plant="Plant", soil="Soil", zoonotic="Zoonotic", other="Other")
+my_categories <- c(aerosol="Aerosol", aquatic="Aquatic", built="Built", plant="Plant", soil="Soil", zoonotic="Host-Associated", other="Other")
 
 set.seed(1)
 my_categories_col <- sample(wes_palette("Zissou", 7, type = "continuous"), 7)#rainbow(7)

### 2016-03-14 results/tables/build_bacterial_category_phylum_table.Rmd
Fix residual zoological labels

@@
 
-categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other zoological", OT="Other", total="Total")
+categories <- c(AE="Aerosol", AQB="Brackish", AQBS="Brackish sediment", AQF="Freshwater", AQFS="Freshwater sediment", AQM="Marine", AQMS="Marine sediment", AQH="Hydrothermal vent", AQI="Ice", AQO="Aquatic other", BD="Digesters", BF="Food-associated", BI="Industrial/mining", BP="Pollution associated", BO="Built other", PR="Plant root", PS="Plant surface", PO="Plant other", SA="Agricultural soil", SD="Desert soil", SP="Permafrost", SO="Other soils", ZV="Vertebrate", ZA="Arthropod", ZN="Other invertebrate", ZO="Other host-associated", OT="Other", total="Total")
 

### 2016-03-14 results/tables/build_coverage_by_category.Rmd
Editing
Fix residual zoological labels

@@
 We calculated the number of sequences and OTUs observed and the percent coverage on a sequence or OTU basis for each classification and domain.
+Descriptions of each category are provided in Table S1.
 
@@
 
-coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Zoological", "", "", "", "Other", "No source data", "Total")
+coarse <- c("Aerosol", "Aquatic", "", "", "", "", "", "", "", "", "Built", "", "", "", "", "Plant associated", "", "", "Soil", "", "", "", "Host-associated", "", "", "", "Other", "No source data", "Total")
 

### 2016-03-14 results/tables/build_environmental_categories_table.Rmd
Fix residual zoological labels

@@
 			"\\midrule\\multirow{6}{*}{Soil}", "", "", "",
-			"\\midrule\\multirow{6}{*}{Zoological}", "", "", "",
+			"\\midrule\\multirow{6}{*}{Host-associated}", "", "", "",
 			"\\midrule\\multirow{1}{*}{Other}")

### 2016-03-14 submission/Schloss_Census2_mBio_2016.tex
Fresh remake on axiom

@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \setlength{\parindent}{0pt}
@@
 bacterial sequences, the three best represented groups were from
-zoonotic (N=804,585), aquatic (N=214,085), and built environment
+host-associated (N=804,585), aquatic (N=214,085), and built environment
 (N=108,799) sources. Among the archaeal sequences the three best
 represented groups were the same, but ordered differently: aquatic
-(N=34,400), built environment (N=7,286), and zoonotic (N=5,597) (Figure
-1C,D)). For both domains, soil samples were the fourth most represented
-category (bacteria: 74,870; archaea: 2,517). The orders of these
-categories was surprising considering soil and aquatic environments
-harbor the most microbial biomass and biodiversity (21). In spite of
-wide variation in sequencing depth and coverage (Table 1), the
+(N=34,400), built environment (N=7,286), and host-associated (N=5,597)
+(Figure 1C,D)). For both domains, soil samples were the fourth most
+represented category (bacteria: 74,870; archaea: 2,517). The orders of
+these categories was surprising considering soil and aquatic
+environments harbor the most microbial biomass and biodiversity (21). In
+spite of wide variation in sequencing depth and coverage (Table 1), the
 interquartile range across the fine-level categories for the bacterial
@@
 direct investigation of rare organisms, microbiologists are developing
-novel cultivation and single cell genomics techniques (\textbf{???},
+novel cultivation and single cell genomics techniques ({\textbf{???}},
 40--42). The ability to enrich or select for specific populations using
@@
 \textbf{\emph{Data analysis.}} Our analysis made use of ARB (OS X v.6.0)
-(43), mothur (v.1.37.0) (44), and R (v.3.2.3) (47). Within R we utilized
-the knitr (v.1.12.3), wesanderson (v.0.3.3.99), and openxlsx (v. 2.4.0)
+(43), mothur (v.1.37.0) (44), and R (v.3.2.0) (47). Within R we utilized
+the knitr (v.1.10.5), wesanderson (v.0.3.2), and openxlsx (v. 2.4.0)
 packages. A reproducible version of this manuscript including data
@@
 
+\hyperdef{}{references}{\label{references}}
 \subsection*{References}\label{references}
@@
 
-\hypertarget{refs}{}
-\hypertarget{ref-McGill2007}{}
+\hyperdef{}{ref-McGill2007}{\label{ref-McGill2007}}
 1. \textbf{McGill BJ}, \textbf{Etienne RS}, \textbf{Gray JS},
@@
 within an ecological framework. Ecology Letters \textbf{10}:995--1015.
-doi:\href{https://doi.org/10.1111/j.1461-0248.2007.01094.x}{10.1111/j.1461-0248.2007.01094.x}.
-
-\hypertarget{ref-Hubbell2001}{}
+doi:\url{http://doi.org/10.1111/j.1461-0248.2007.01094.x}.
+
+\hyperdef{}{ref-Hubbell2001}{\label{ref-Hubbell2001}}
 2. \textbf{Hubbell SP}. 2001. A Unified Theory of Biodiversity and
@@
 
-\hypertarget{ref-Konstantinidis2006}{}
+\hyperdef{}{ref-Konstantinidis2006}{\label{ref-Konstantinidis2006}}
 3. \textbf{Konstantinidis KT}, \textbf{Ramette A}, \textbf{Tiedje JM}.
@@
 \textbf{361}:1929--1940.
-doi:\href{https://doi.org/10.1098/rstb.2006.1920}{10.1098/rstb.2006.1920}.
-
-\hypertarget{ref-Oren2013}{}
+doi:\url{http://doi.org/10.1098/rstb.2006.1920}.
+
+\hyperdef{}{ref-Oren2013}{\label{ref-Oren2013}}
 4. \textbf{Oren A}, \textbf{Garrity GM}. 2013. Then and now: A
@@
 80~years. Antonie van Leeuwenhoek \textbf{106}:43--56.
-doi:\href{https://doi.org/10.1007/s10482-013-0084-1}{10.1007/s10482-013-0084-1}.
-
-\hypertarget{ref-Brosius1978}{}
+doi:\url{http://doi.org/10.1007/s10482-013-0084-1}.
+
+\hyperdef{}{ref-Brosius1978}{\label{ref-Brosius1978}}
 5. \textbf{Brosius J}, \textbf{Palmer ML}, \textbf{Kennedy PJ},
@@
 National Academy of Sciences \textbf{75}:4801--4805.
-doi:\href{https://doi.org/10.1073/pnas.75.10.4801}{10.1073/pnas.75.10.4801}.
-
-\hypertarget{ref-Schloss2004}{}
+doi:\url{http://doi.org/10.1073/pnas.75.10.4801}.
+
+\hyperdef{}{ref-Schloss2004}{\label{ref-Schloss2004}}
 6. \textbf{Schloss PD}, \textbf{Handelsman J}. 2004. Status of the
@@
 \textbf{68}:686--691.
-doi:\href{https://doi.org/10.1128/mmbr.68.4.686-691.2004}{10.1128/mmbr.68.4.686-691.2004}.
-
-\hypertarget{ref-Dykhuizen1998}{}
+doi:\url{http://doi.org/10.1128/mmbr.68.4.686-691.2004}.
+
+\hyperdef{}{ref-Dykhuizen1998}{\label{ref-Dykhuizen1998}}
 7. \textbf{Dykhuizen DE}. 1998. Santa Rosalia revisited: Why are there
 so many species of bacteria? Antonie van Leeuwenhoek \textbf{73}:25--33.
-doi:\href{https://doi.org/10.1023/a:1000665216662}{10.1023/a:1000665216662}.
-
-\hypertarget{ref-Curtis2002}{}
+doi:\url{http://doi.org/10.1023/a:1000665216662}.
+
+\hyperdef{}{ref-Curtis2002}{\label{ref-Curtis2002}}
 8. \textbf{Curtis TP}, \textbf{Sloan WT}, \textbf{Scannell JW}. 2002.
@@
 National Academy of Sciences \textbf{99}:10494--10499.
-doi:\href{https://doi.org/10.1073/pnas.142680199}{10.1073/pnas.142680199}.
-
-\hypertarget{ref-HMP2012}{}
+doi:\url{http://doi.org/10.1073/pnas.142680199}.
+
+\hyperdef{}{ref-HMP2012}{\label{ref-HMP2012}}
 9. \textbf{The Human Microbiome Consortium}. 2012. Structure, function
 and diversity of the healthy human microbiome. Nature
-\textbf{486}:207--214.
-doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
-
-\hypertarget{ref-Gilbert2014}{}
+\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
+
+\hyperdef{}{ref-Gilbert2014}{\label{ref-Gilbert2014}}
 10. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2014.
 The Earth Microbiome Project: Successes and aspirations. BMC Biology
-\textbf{12}:69.
-doi:\href{https://doi.org/10.1186/s12915-014-0069-1}{10.1186/s12915-014-0069-1}.
-
-\hypertarget{ref-AmaralZettler2010}{}
+\textbf{12}:69. doi:\url{http://doi.org/10.1186/s12915-014-0069-1}.
+
+\hyperdef{}{ref-AmaralZettler2010}{\label{ref-AmaralZettler2010}}
 11. \textbf{Amaral-Zettler L}, \textbf{Artigas LF}, \textbf{Baross J},
@@
 
-\hypertarget{ref-Youssef2009}{}
+\hyperdef{}{ref-Youssef2009}{\label{ref-Youssef2009}}
 12. \textbf{Youssef N}, \textbf{Sheik CS}, \textbf{Krumholz LR},
@@
 gene-based environmental surveys. Applied and Environmental Microbiology
-\textbf{75}:5227--5236.
-doi:\href{https://doi.org/10.1128/aem.00592-09}{10.1128/aem.00592-09}.
-
-\hypertarget{ref-Schloss2010}{}
+\textbf{75}:5227--5236. doi:\url{http://doi.org/10.1128/aem.00592-09}.
+
+\hyperdef{}{ref-Schloss2010}{\label{ref-Schloss2010}}
 13. \textbf{Schloss PD}. 2010. The effects of alignment quality,
@@
 \textbf{6}:e1000844.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1000844}{10.1371/journal.pcbi.1000844}.
-
-\hypertarget{ref-Alivisatos2015}{}
+doi:\url{http://doi.org/10.1371/journal.pcbi.1000844}.
+
+\hyperdef{}{ref-Alivisatos2015}{\label{ref-Alivisatos2015}}
 14. \textbf{Alivisatos AP}, \textbf{Blaser MJ}, \textbf{Brodie EL},
@@
 harness Earth's microbiomes. Science \textbf{350}:507--508.
-doi:\href{https://doi.org/10.1126/science.aac8480}{10.1126/science.aac8480}.
-
-\hypertarget{ref-Liux5f2012}{}
+doi:\url{http://doi.org/10.1126/science.aac8480}.
+
+\hyperdef{}{ref-Liux5f2012}{\label{ref-Liux5f2012}}
 15. \textbf{Li E}, \textbf{Hamm CM}, \textbf{Gulati AS}, \textbf{Sartor
@@
 composition. PLoS ONE \textbf{7}:e26284.
-doi:\href{https://doi.org/10.1371/journal.pone.0026284}{10.1371/journal.pone.0026284}.
-
-\hypertarget{ref-Kongux5f2012}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0026284}.
+
+\hyperdef{}{ref-Kongux5f2012}{\label{ref-Kongux5f2012}}
 16. \textbf{Kong HH}, \textbf{Oh J}, \textbf{Deming C}, \textbf{Conlan
@@
 atopic dermatitis. Genome Research \textbf{22}:850--859.
-doi:\href{https://doi.org/10.1101/gr.131029.111}{10.1101/gr.131029.111}.
-
-\hypertarget{ref-Griceux5f2009}{}
+doi:\url{http://doi.org/10.1101/gr.131029.111}.
+
+\hyperdef{}{ref-Griceux5f2009}{\label{ref-Griceux5f2009}}
 17. \textbf{Grice EA}, \textbf{Kong HH}, \textbf{Conlan S},
@@
 Science \textbf{324}:1190--1192.
-doi:\href{https://doi.org/10.1126/science.1171700}{10.1126/science.1171700}.
-
-\hypertarget{ref-Griceux5f2010}{}
+doi:\url{http://doi.org/10.1126/science.1171700}.
+
+\hyperdef{}{ref-Griceux5f2010}{\label{ref-Griceux5f2010}}
 18. \textbf{Grice EA}, \textbf{Snitkin ES}, \textbf{Yockey LJ},
@@
 Sciences \textbf{107}:14799--14804.
-doi:\href{https://doi.org/10.1073/pnas.1004204107}{10.1073/pnas.1004204107}.
-
-\hypertarget{ref-Kirkux5fHarrisux5f2012}{}
+doi:\url{http://doi.org/10.1073/pnas.1004204107}.
+
+\hyperdef{}{ref-Kirkux5fHarrisux5f2012}{\label{ref-Kirkux5fHarrisux5f2012}}
 19. \textbf{Harris JK}, \textbf{Caporaso JG}, \textbf{Walker JJ},
@@
 negro hypersaline microbial mat. The ISME Journal \textbf{7}:50--60.
-doi:\href{https://doi.org/10.1038/ismej.2012.79}{10.1038/ismej.2012.79}.
-
-\hypertarget{ref-Sogin2006}{}
+doi:\url{http://doi.org/10.1038/ismej.2012.79}.
+
+\hyperdef{}{ref-Sogin2006}{\label{ref-Sogin2006}}
 20. \textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber JA},
@@
 Academy of Sciences \textbf{103}:12115--12120.
-doi:\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.
-
-\hypertarget{ref-Whitman1998}{}
+doi:\url{http://doi.org/10.1073/pnas.0605127103}.
+
+\hyperdef{}{ref-Whitman1998}{\label{ref-Whitman1998}}
 21. \textbf{Whitman WB}, \textbf{Coleman DC}, \textbf{Wiebe WJ}. 1998.
@@
 Sciences \textbf{95}:6578--6583.
-doi:\href{https://doi.org/10.1073/pnas.95.12.6578}{10.1073/pnas.95.12.6578}.
-
-\hypertarget{ref-Rinke2013}{}
+doi:\url{http://doi.org/10.1073/pnas.95.12.6578}.
+
+\hyperdef{}{ref-Rinke2013}{\label{ref-Rinke2013}}
 22. \textbf{Rinke C}, \textbf{Schwientek P}, \textbf{Sczyrba A},
@@
 phylogeny and coding potential of microbial dark matter. Nature
-\textbf{499}:431--437.
-doi:\href{https://doi.org/10.1038/nature12352}{10.1038/nature12352}.
-
-\hypertarget{ref-Miller2011}{}
+\textbf{499}:431--437. doi:\url{http://doi.org/10.1038/nature12352}.
+
+\hyperdef{}{ref-Miller2011}{\label{ref-Miller2011}}
 23. \textbf{Miller CS}, \textbf{Baker BJ}, \textbf{Thomas BC},
@@
 sequencing data. Genome Biol \textbf{12}:R44.
-doi:\href{https://doi.org/10.1186/gb-2011-12-5-r44}{10.1186/gb-2011-12-5-r44}.
-
-\hypertarget{ref-Miller2013}{}
+doi:\url{http://doi.org/10.1186/gb-2011-12-5-r44}.
+
+\hyperdef{}{ref-Miller2013}{\label{ref-Miller2013}}
 24. \textbf{Miller CS}, \textbf{Handley KM}, \textbf{Wrighton KC},
@@
 diversity in subsurface sediments. PLoS ONE \textbf{8}:e56018.
-doi:\href{https://doi.org/10.1371/journal.pone.0056018}{10.1371/journal.pone.0056018}.
-
-\hypertarget{ref-Markowitz2013}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0056018}.
+
+\hyperdef{}{ref-Markowitz2013}{\label{ref-Markowitz2013}}
 25. \textbf{Markowitz VM}, \textbf{Chen I-MA}, \textbf{Palaniappan K},
@@
 genomes comparative analysis system. Nucleic Acids Research
-\textbf{42}:D560--D567.
-doi:\href{https://doi.org/10.1093/nar/gkt963}{10.1093/nar/gkt963}.
-
-\hypertarget{ref-Wrighton2012}{}
+\textbf{42}:D560--D567. doi:\url{http://doi.org/10.1093/nar/gkt963}.
+
+\hyperdef{}{ref-Wrighton2012}{\label{ref-Wrighton2012}}
 26. \textbf{Wrighton KC}, \textbf{Thomas BC}, \textbf{Sharon I},
@@
 bacterial phyla. Science \textbf{337}:1661--1665.
-doi:\href{https://doi.org/10.1126/science.1224041}{10.1126/science.1224041}.
-
-\hypertarget{ref-NunesdaRocha2015}{}
+doi:\url{http://doi.org/10.1126/science.1224041}.
+
+\hyperdef{}{ref-NunesdaRocha2015}{\label{ref-NunesdaRocha2015}}
 27. \textbf{Rocha UN da}, \textbf{Cadillo-Quiroz H}, \textbf{Karaoz U},
@@
 desert biological soil crust. Front Microbiol \textbf{6}:277.
-doi:\href{https://doi.org/10.3389/fmicb.2015.00277}{10.3389/fmicb.2015.00277}.
-
-\hypertarget{ref-Hamilton2015}{}
+doi:\url{http://doi.org/10.3389/fmicb.2015.00277}.
+
+\hyperdef{}{ref-Hamilton2015}{\label{ref-Hamilton2015}}
 28. \textbf{Hamilton TL}, \textbf{Jones DS}, \textbf{Schaperdoth I},
@@
 in a terrestrial subsurface lithoautotrophic ecosystem. Front Microbiol
-\textbf{5}:756.
-doi:\href{https://doi.org/10.3389/fmicb.2014.00756}{10.3389/fmicb.2014.00756}.
-
-\hypertarget{ref-Handley2012}{}
+\textbf{5}:756. doi:\url{http://doi.org/10.3389/fmicb.2014.00756}.
+
+\hyperdef{}{ref-Handley2012}{\label{ref-Handley2012}}
 29. \textbf{Handley KM}, \textbf{VerBerkmoes NC}, \textbf{Steefel CI},
@@
 community. The ISME Journal \textbf{7}:800--816.
-doi:\href{https://doi.org/10.1038/ismej.2012.148}{10.1038/ismej.2012.148}.
-
-\hypertarget{ref-Gladden2011}{}
+doi:\url{http://doi.org/10.1038/ismej.2012.148}.
+
+\hyperdef{}{ref-Gladden2011}{\label{ref-Gladden2011}}
 30. \textbf{Gladden JM}, \textbf{Allgaier M}, \textbf{Miller CS},
@@
 Applied and Environmental Microbiology \textbf{77}:5804--5812.
-doi:\href{https://doi.org/10.1128/aem.00032-11}{10.1128/aem.00032-11}.
-
-\hypertarget{ref-Brooks2014}{}
+doi:\url{http://doi.org/10.1128/aem.00032-11}.
+
+\hyperdef{}{ref-Brooks2014}{\label{ref-Brooks2014}}
 31. \textbf{Brooks B}, \textbf{Firek BA}, \textbf{Miller CS},
@@
 premature infants. Microbiome \textbf{2}:1.
-doi:\href{https://doi.org/10.1186/2049-2618-2-1}{10.1186/2049-2618-2-1}.
-
-\hypertarget{ref-Wilkins2013}{}
+doi:\url{http://doi.org/10.1186/2049-2618-2-1}.
+
+\hyperdef{}{ref-Wilkins2013}{\label{ref-Wilkins2013}}
 32. \textbf{Wilkins MJ}, \textbf{Wrighton KC}, \textbf{Nicora CD},
@@
 PLoS ONE \textbf{8}:e57819.
-doi:\href{https://doi.org/10.1371/journal.pone.0057819}{10.1371/journal.pone.0057819}.
-
-\hypertarget{ref-Handley2014}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0057819}.
+
+\hyperdef{}{ref-Handley2014}{\label{ref-Handley2014}}
 33. \textbf{Handley KM}, \textbf{Wrighton KC}, \textbf{Miller CS},
@@
 points. Environ Microbiol \textbf{17}:622--636.
-doi:\href{https://doi.org/10.1111/1462-2920.12467}{10.1111/1462-2920.12467}.
-
-\hypertarget{ref-Alessi2014}{}
+doi:\url{http://doi.org/10.1111/1462-2920.12467}.
+
+\hyperdef{}{ref-Alessi2014}{\label{ref-Alessi2014}}
 34. \textbf{Alessi DS}, \textbf{Lezama-Pacheco JS}, \textbf{Janot N},
@@
 Science \& Technology \textbf{48}:12842--12850.
-doi:\href{https://doi.org/10.1021/es502701u}{10.1021/es502701u}.
-
-\hypertarget{ref-Locey2015}{}
+doi:\url{http://doi.org/10.1021/es502701u}.
+
+\hyperdef{}{ref-Locey2015}{\label{ref-Locey2015}}
 35. \textbf{Locey KJ}, \textbf{Lennon JT}. 2015. Scaling laws predict
 global microbial diversity. PeerJ PrePrints.
-doi:\href{https://doi.org/10.7287/peerj.preprints.1451v1}{10.7287/peerj.preprints.1451v1}.
-
-\hypertarget{ref-Parada2015}{}
+doi:\url{http://doi.org/10.7287/peerj.preprints.1451v1}.
+
+\hyperdef{}{ref-Parada2015}{\label{ref-Parada2015}}
 36. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
@@
 Environ Microbiol n/a--n/a.
-doi:\href{https://doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.
-
-\hypertarget{ref-Brown2015}{}
+doi:\url{http://doi.org/10.1111/1462-2920.13023}.
+
+\hyperdef{}{ref-Brown2015}{\label{ref-Brown2015}}
 37. \textbf{Brown CT}, \textbf{Hug LA}, \textbf{Thomas BC},
@@
 more than 15\% of domain bacteria. Nature \textbf{523}:208--211.
-doi:\href{https://doi.org/10.1038/nature14486}{10.1038/nature14486}.
-
-\hypertarget{ref-EloeFadrosh2016}{}
+doi:\url{http://doi.org/10.1038/nature14486}.
+
+\hyperdef{}{ref-EloeFadrosh2016}{\label{ref-EloeFadrosh2016}}
 38. \textbf{Eloe-Fadrosh EA}, \textbf{Ivanova NN}, \textbf{Woyke T},
@@
 detection of microbial diversity. Nat Microbiol 15032.
-doi:\href{https://doi.org/10.1038/nmicrobiol.2015.32}{10.1038/nmicrobiol.2015.32}.
-
-\hypertarget{ref-Schloss2015}{}
+doi:\url{http://doi.org/10.1038/nmicrobiol.2015.32}.
+
+\hyperdef{}{ref-Schloss2015}{\label{ref-Schloss2015}}
 39. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Jenior ML},
@@
 the pacBio sMRT dNA sequencing system. PeerJ PrePrints.
-doi:\href{https://doi.org/10.7287/peerj.preprints.778v1}{10.7287/peerj.preprints.778v1}.
-
-\hypertarget{ref-Nichols2010}{}
+doi:\url{http://doi.org/10.7287/peerj.preprints.778v1}.
+
+\hyperdef{}{ref-Nichols2010}{\label{ref-Nichols2010}}
 40. \textbf{Nichols D}, \textbf{Cahoon N}, \textbf{Trakhtenberg EM},
@@
 species. Applied and Environmental Microbiology \textbf{76}:2445--2450.
-doi:\href{https://doi.org/10.1128/aem.01754-09}{10.1128/aem.01754-09}.
-
-\hypertarget{ref-Buerger2012}{}
+doi:\url{http://doi.org/10.1128/aem.01754-09}.
+
+\hyperdef{}{ref-Buerger2012}{\label{ref-Buerger2012}}
 41. \textbf{Buerger S}, \textbf{Spoering A}, \textbf{Gavrish E},
@@
 growers. Applied and Environmental Microbiology \textbf{78}:3221--3228.
-doi:\href{https://doi.org/10.1128/aem.07307-11}{10.1128/aem.07307-11}.
-
-\hypertarget{ref-Das2015}{}
+doi:\url{http://doi.org/10.1128/aem.07307-11}.
+
+\hyperdef{}{ref-Das2015}{\label{ref-Das2015}}
 42. \textbf{Das N}, \textbf{Tripathi N}, \textbf{Basu S}, \textbf{Bose
@@
 microorganisms. Front Microbiol \textbf{6}:698.
-doi:\href{https://doi.org/10.3389/fmicb.2015.00698}{10.3389/fmicb.2015.00698}.
-
-\hypertarget{ref-Pruesse2007}{}
+doi:\url{http://doi.org/10.3389/fmicb.2015.00698}.
+
+\hyperdef{}{ref-Pruesse2007}{\label{ref-Pruesse2007}}
 43. \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Knittel K},
@@
 ARB. Nucleic Acids Research \textbf{35}:7188--7196.
-doi:\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.
-
-\hypertarget{ref-Schloss2009}{}
+doi:\url{http://doi.org/10.1093/nar/gkm864}.
+
+\hyperdef{}{ref-Schloss2009}{\label{ref-Schloss2009}}
 44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
-
-\hypertarget{ref-Westcott2015}{}
+\textbf{75}:7537--7541. doi:\url{http://doi.org/10.1128/aem.01541-09}.
+
+\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
 45. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
@@
 sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
-
-\hypertarget{ref-Good1953}{}
+doi:\url{http://doi.org/10.7717/peerj.1487}.
+
+\hyperdef{}{ref-Good1953}{\label{ref-Good1953}}
 46. \textbf{Good IJ}. 1953. The population frequencies of species and
@@
 \textbf{40}:237--264.
-doi:\href{https://doi.org/10.1093/biomet/40.3-4.237}{10.1093/biomet/40.3-4.237}.
-
-\hypertarget{ref-language2015}{}
+doi:\url{http://doi.org/10.1093/biomet/40.3-4.237}.
+
+\hyperdef{}{ref-language2015}{\label{ref-language2015}}
 47. \textbf{R Core Team}. 2015. R: A language and environment for

### 2016-03-14 code/build_domain_rarefaction_plot.R
Move figure 1C legend

@@ -76,7 +76,7 @@ axis(2, las=1, at=seq(0,50000,10000), label=format(seq(0,50000,10000), big.mark=
 
 text(par()$usr[2]*0.95, par()$usr[4]*0.95, "C", cex=2, font=2)
 
-legend(x=6e5, y=20000, legend=my_categories, lwd=2, pch=my_categories_pch, col=my_categories_col, pt.bg="white", bty="n", pt.cex=1.25)
+legend(x=5e5, y=20000, legend=my_categories, lwd=2, pch=my_categories_pch, col=my_categories_col, pt.bg="white", bty="n", pt.cex=1.25)
 
 
 

### 2016-03-16 submission/rebuttal.md
Incorporate Cameron's changes

@@
 
-?????
+We would be happy to entertain this argument and possibly add text if the reviewer would provide additional details.
 
@@
 
-?????
+We have edited this line to now read "A search of GenBank's nucleotide database reveals that 16S rRNA gene accessions represent nearly one-third of the database..."
 
@@
 
-?????
+Since the rarefaction curves are not actually fully saturated at the OTU level (3% threshold), this means there are still OTUs that we expect to observe with new sampling.
+Consistent with this, the sampled group itself is made up primarily of singletons, meaning that we are only barely detecting most of the OTUs in the first place.
+Thus, we believe these two sentences do not contradict each other.
 
@@
 Furthermore, considering EMIRGE hasn't been used to sequence a mock community, it is unclear what the error rates are within contigs generated by the approach.
+We have tried to be conservative here by using “suggests”, and considering this is a combined Results/Discussion, we felt it was appropriate to add some amount of interpretation to the results.
 
@@
 
-We are reluctant to do this since the sequences were not generated by us.
+We are reluctant to do this since the sequences were not generated by us, and we would like to avoid duplicating an existing public database.
 In addition, the SILVA sequences are subject to a somewhat odd licensing agreement.
